% !TeX root = RJwrapper.tex
\title{spinifex: An R Package for Creating a Manual Tour of
Low-dimensional Projections of Multivariate Data}
\author{by Nicholas Spyrison and Dianne Cook}

\maketitle

\abstract{%
Dynamic low-dimensional linear projections of multivariate data
collectively known as tours provide an important tool for exploring
multivariate data and models. The R package tourr provides functions for
several types of tours: grand, guided, little, local and frozen. Each of
these can be viewed dynamically, or saved into a data object for
animation. This paper describes a new package, spinifex, which provides
a manual tour of multivariate data where the projection coefficient of a
single variable is controlled. The variable is rotated fully into the
projection, or completely out of the projection. The resulting sequence
of projections can be displayed as an animation, with functions from
either the plotly or gganimate packages. By varying the coefficient of a
single variable, it is possible to explore the sensitivity of structure
in the projection to that variable. This is particularly useful when
used with a projection pursuit guided tour to simplify and understand
the solution. The use of the manual tour is applied particle physics
data to illustrate the sensitivity of structure in a projection to
specific variable contributions.
}

@CRANpkg(spinifex)

\begin{verbatim}
\end{verbatim}


\address{%
Nicholas Spyrison\\
Monash University\\%
\\
%
%
%
%
}

\address{%
Dianne Cook\\
Monash University\\%
\\
%
%
%
%
}
