---
title: "Appendix: Practical Implementation and Extended Models - A Comprehensive Guide to the exametrika Package"
author:  
  - "Koji Kosugi^[Senshu University, School of Human Sciences, kosugi@psy.senshu-u.ac.jp]"
  - "Kojiro Shojima^[University Entrance Examination Center, Research and Development Department, shojima@rd.dnc.ac.jp]"
output:
  pdf_document:
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
set.seed(17)
```

# Introduction

The `exametrika` package implements statistical models for test data engineering introduced by Shojima (2022), providing a comprehensive framework for analyzing educational test data. This document focuses on two key approaches: Latent Rank Analysis (LRA) for ordinal ability assessment and Biclustering for simultaneous clustering of examinees and items. LRA extends traditional Latent Class Analysis by incorporating ordered structures, while Biclustering enables the exploration of complex response patterns through two-mode clustering. The package implements efficient EM algorithms and rich visualization tools.

```{r packages, message=FALSE}
# Load the package
library(exametrika)
```

# Data Preparation and Conversion

The most important step in using the `exametrika` package is converting your data into a format that can be processed by the package (exametrika class). This conversion is done using the `dataFormat` function.

## The exametrika Class

The exametrika class is a data structure that includes the following information:

- `U`: Response data matrix (binary 0/1 data by default)
- `Z`: Missing value indicator matrix (1 for observed values, 0 for missing values)
- `ItemLabel`: List of item names
- `ID`: List of examinee IDs
- `response.type`: Data type ("binary", "nominal", "ordinal", or "rated")

## Using Sample Datasets

The `exametrika` package includes several sample datasets that have already been converted to the exametrika class format.

```{r sample_data}
# List of sample datasets included in the package
data(package = "exametrika")

# Example: Loading J15S500 (15 items, 500 samples)
data(J15S500)

# Checking the data structure
head(J15S500$U, 3) # First 3 rows of the response data matrix
head(J15S500$Z, 3) # First 3 rows of the missing value indicator matrix
J15S500$response.type # Data type
```

These sample datasets are useful for testing the package's functionality before analyzing your own data.

## Creating and Converting Data

### Converting from a Data Frame

```{r data_frame_conversion}
# Numeric ID column
data1 <- data.frame(
  id = 1:5,
  item1 = c(0, 1, 1, 0, 1),
  item2 = c(1, 0, 1, 0, 0)
)
result1 <- dataFormat(data1)
print(result1)
```

In this code, the first column is automatically recognized as the ID column, and the remaining columns are treated as items.

### Specifying the ID Column

If the ID column is not the first column, you can specify its position using the `id` parameter.

```{r id_specification}
# String ID column in the third position
data2 <- data.frame(
  item1 = c(0, 1, 1, 0, 1),
  item2 = c(1, 0, 1, 0, 99),
  id = paste0("student", 1:5)
)
result2 <- dataFormat(data2, id = 3, na = 99)
print(result2)
```

### Handling Missing Values

Missing values are specified with the `na` parameter. By default, R's standard missing value (`NA`) is used.

```{r missing_values}
# Data with NA values
data3 <- data.frame(
  id = c(1, 2, 3, 4, 5),
  item1 = c(1, NA, 2, 4, 1),
  item2 = c(1, 3, 2, 4, 5)
)
result3 <- dataFormat(data3)
print(result3)
```
```{r missing_values_ex2}
# Data with custom missing value (-999)
data4 <- data.frame(
  id = c(1, 2, 3, 4, 5),
  item1 = c(1, -999, 2, 4, 1),
  item2 = c(1, 3, 2, 4, -999)
)
result4 <- dataFormat(data4, na = -999)
print(result4)
```

### Loading from CSV Files

When loading data from a CSV file, first use the `read.csv` function to load the data, then convert it using the `dataFormat` function.

```{r csv_loading, eval=FALSE}
# Loading from a CSV file
test_data <- read.csv("test_data.csv")
result <- dataFormat(test_data, na = -99) # Specifying -99 as missing value
```

### A Complete Example with Real Data

Let's examine a practical example where we create test data for 5 examinees, save it as a CSV file, then load and analyze the data.

```{r csv_loading example}
# Create sample test data
sample_test_data <- data.frame(
  student_id = c("S001", "S002", "S003", "S004", "S005"),
  Q1 = c(1, 0, 1, 1, 0),
  Q2 = c(0, 1, 1, 0, 0),
  Q3 = c(1, 1, 0, 1, -99), # -99 represents missing value
  Q4 = c(0, 0, 1, 1, 1),
  Q5 = c(1, 0, 0, -99, 1) # -99 represents missing value
)

# View the raw data
head(sample_test_data)

# Save data to a CSV file
write.csv(sample_test_data, "sample_test.csv", row.names = FALSE)

# Now read it back as if it was an external file
read_test_data <- read.csv("sample_test.csv")

# Convert to exametrika format
test_data_exam <- dataFormat(read_test_data, id = 1, na = -99)

# Display the converted data
print(test_data_exam)

# Remove created csv file
file.remove("sample_test.csv")
```

```{r simulated_csv}
# Simulated example (creating a virtual CSV-like dataset)
test_data <- data.frame(
  ID = paste0("S", 1:10),
  Q1 = sample(c(0, 1, -99), 10, replace = TRUE),
  Q2 = sample(c(0, 1, -99), 10, replace = TRUE),
  Q3 = sample(c(0, 1, -99), 10, replace = TRUE)
)
result <- dataFormat(test_data, na = -99)
print(result)
```

## Specifying Data Types

The `exametrika` package can handle various types of data. The data type is specified using the `response.type` parameter.

- `binary`: Binary data (0/1) - default
- `nominal`: Nominal scale data (categorical)
- `ordinal`: Ordinal scale data (Likert scale, etc.)
- `rated`: Nominal scale data with correct answers (multiple-choice items, etc.)

```{r data_types_ordinal}
# Ordinal scale data example
ordinal_data <- data.frame(
  id = 1:5,
  Q1 = sample(1:5, 5, replace = TRUE),
  Q2 = sample(1:5, 5, replace = TRUE)
)
ordinal_result <- dataFormat(ordinal_data, response.type = "ordinal")
print(ordinal_result)
```

```{r data_types_nominal}
# Multiple-choice data with correct answers
rated_data <- data.frame(
  id = 1:5,
  Q1 = sample(1:4, 5, replace = TRUE),
  Q2 = sample(1:4, 5, replace = TRUE)
)
# Correct answers: Q1 = option 4, Q2 = option 2
rated_result <- dataFormat(rated_data, response.type = "rated", CA = c(4, 2))
print(rated_result)
```

## Using the Weight (w) Parameter

The `w` parameter is used to specify item weights. It is primarily used in Classical Test Theory (CTT) analysis to adjust the contribution of each item to the total score. However, in modern test theories such as Latent Rank Analysis (LRA) and Biclustering, item characteristics (such as difficulty) are estimated separately, so this parameter is typically not used.

```{r weights}
# Example of specifying weights (used in CTT)
weighted_data <- data.frame(
  id = 1:5,
  Q1 = c(1, 0, 1, 0, 1),
  Q2 = c(0, 1, 1, 0, 0)
)
# Weight for Q1 = 2, weight for Q2 = 1
weighted_result <- dataFormat(weighted_data, w = c(3, 2))
weighted_result
TestStatistics(weighted_result)
```

# Checking Data

To check the contents of the converted data, simply enter the variable name or use the `summary` function.

```{r check_data}
# Basic overview
summary(result1)

# Checking individual components
head(result1$U, 3) # First 3 rows of the response data matrix
head(result1$Z, 3) # First 3 rows of the missing value indicator matrix
result1$ItemLabel # Item names
result1$ID # Examinee IDs
result1$response.type # Data type
```

# Analysis Examples

Once you have converted your data to the exametrika class format, you can perform various analyses. In this section, we'll demonstrate the major analytical methods using sample datasets included in the `exametrika` package.

## Latent Rank Analysis (LRA)

Latent Rank Analysis (LRA) is a statistical method that classifies examinees into ordered ranks based on their response patterns. It extends traditional Latent Class Analysis by incorporating an ordinal structure.

```{r lra_example}
# Load sample data
data(J15S500) # 15 items, 500 samples binary dataset

# Run LRA with 6 ranks
lra_result <- LRA(J15S500, nrank = 6)

# Examine the Item Reference Profile (IRP)
# IRP shows the probability of correct responses for each item at each rank
head(lra_result$IRP)
```

### Model Fit Indices

Model fit indices help evaluate the appropriateness of the specified number of ranks:

```{r lra_fit_indices}
# Display test fit indices
lra_result$TestFitIndices
# Display item fit indices
lra_result$ItemFitIndices
```

These indices compare the current model to saturated and null models. Important metrics include:
- NFI, RFI, IFI, TLI, CFI: Values close to 1 indicate good fit
- RMSEA: Values less than 0.08 suggest reasonable fit
- AIC, BIC, CAIC: Lower values indicate better models

### Examining Individual Results

We can also examine rank membership profiles for individual examinees:

```{r lra_students}
# Display rank membership for the first few examinees
head(lra_result$Students)
```

The results include:
- Membership probabilities for each rank
- Estimated rank (the rank with highest probability)
- Rank-Up and Rank-Down Odds (indicating likelihood of transitioning to adjacent ranks)

### Visualizing LRA Results

```{r lra_plots, fig.cap='Item Reference Profile (IRP) plots'}
# Plot Item Reference Profiles (IRP) for the first 4 items
plot(lra_result, type = "IRP", items = 1:4, nc = 2, nr = 2)
```

```{r lra_plots2, fig.cap='Rank Membership Profile (RMP) plots'}
# Plot Rank Membership Profiles (RMP) for the first 4 examinees
plot(lra_result, type = "RMP", students = 1:4, nc = 2, nr = 2)
```

The IRP plots show how the probability of correct response increases with rank for each item. Steeper slopes indicate items with better discrimination between ranks.

## Biclustering

Biclustering simultaneously clusters examinees into classes and items into fields, creating a two-mode clustering solution. This allows for a more detailed analysis of response patterns and item groups.

```{r biclustering_example}
# Load sample data
data(J35S515) # 35 items, 515 examinees

# Run Biclustering with 5 fields and 6 classes
bi_result <- Biclustering(J35S515, nfld = 5, ncls = 6, method = "B")

# Examine the Field Reference Profile (FRP)
# This matrix shows the probability of correct responses for each field at each class
bi_result$FRP
```

### Biclustering Model Fit

```{r biclustering_fit}
# Display test fit indices
bi_result$TestFitIndices
```

### Field Membership

Item membership in fields shows how items are clustered:

```{r biclustering_field}
# Display field membership for the first few items
head(bi_result$FieldMembership, 10)
```

Each row represents an item, with probabilities of belonging to each field. The highest probability indicates the most likely field for that item.

### Class Membership

Examinee membership in classes:

```{r biclustering_students}
# Display class membership for the first few examinees
head(bi_result$Students)
```

### Visualizing Biclustering Results

```{r biclustering_plots,fig.cap='Array plot'}
# Array plot: visualize the raw data and the clustered result
plot(bi_result, type = "Array")
```

The Array plot shows the original data pattern on the left and the rearranged (clustered) pattern on the right. Black cells represent correct responses (1), and white cells represent incorrect responses (0). This visualization helps identify structure in the response patterns.

```{r biclustering_plot2, ,fig.cap='Field Reference Profile(FRP) plots'}
# Plot of Field Reference Profile
plot(bi_result, type = "FRP", nc = 2, nr = 3)
```


## Ranklustering

Ranklustering is an extension of Biclustering that incorporates ordered ranks for examinees instead of unordered classes, making it particularly useful for educational assessment.

```{r ranklustering_example}
# Run Ranklustering with 5 fields and 6 ranks
rank_result <- Biclustering(J35S515, nfld = 5, ncls = 6, method = "R")

# Examine the Field-Rank Profile (FRP)
# Shows the probability of correct responses for each field at each rank
rank_result$FRP
```

### Ranklustering Model Fit

```{r ranklustering_fit}
# Display test fit indices
rank_result$TestFitIndices
```

### Field Membership in Ranklustering

```{r ranklustering_field}
# Display field membership for the first few items
head(rank_result$FieldMembership, 10)
```

### Rank Membership for Examinees

```{r ranklustering_students}
# Display rank membership for the first few examinees
head(rank_result$Students)
```

The results include Rank-Up and Rank-Down Odds, which are unique to Ranklustering and indicate the likelihood of moving to a higher or lower rank.

### Visualizing Ranklustering Results

```{r ranklustering_plots,fig.cap='Array plot and Field Reference Profile(FRP) plots' }
# Array plot
plot(rank_result, type = "Array")

# Field Reference Profile
plot(rank_result, type = "FRP", nr = 2, nc = 3)
```

The key difference in visualization between Biclustering and Ranklustering is that the latter maintains an ordinal structure among examinee clusters, which is evident in the Array plot's pattern.

## Selecting the Optimal Number of Ranks and Fields

Finding the optimal number of ranks and fields is crucial for meaningful interpretation. While model fit indices provide guidance, domain knowledge about the test content should also inform this decision.

```{r model_comparison}
fit_indices <- data.frame()
for (r in 3:6) {
  result <- LRA(J15S500, nrank = r)
  fit_indices <- rbind(
    fit_indices,
    data.frame(
      Rank = r,
      AIC = result$TestFitIndices$AIC,
      BIC = result$TestFitIndices$BIC
    )
  )
}
print(fit_indices)
```

In educational contexts, the number of ranks should reflect meaningful developmental stages rather than being determined solely by statistical criteria.

## Bayesian network model

The Bayesian network model provided by the `exametrika` package represents the conditional probabilities between items in a network format based on the pass rates of the items. This approach allows for modeling the hierarchical relationships and dependencies between test items, providing insights into knowledge structures and potential learning pathways.

### Theoretical Background

Bayesian networks in educational assessment model the probabilistic relationships between test items, where edges in the network represent prerequisite relationships or cognitive dependencies. The model assumes that mastery of certain concepts (represented by correctly answering specific items) influences the probability of mastering dependent concepts.


### Implementation in exametrika

By providing a Directed Acyclic Graph (DAG) between items externally, the package calculates the conditional probabilities based on the specified graph. The igraph package is used for the analysis and representation of the network.
There are three ways to specify the graph structure:

+ Pass a matrix-type DAG to the argument adj_matrix
+ Pass a DAG described in a CSV file to the argument adj_file
+ Pass a graph-type object g used in the igraph package to the argument g

### Interpretation and Educational Applications

The BNM results provide several valuable insights:

+ Conditional Correct Response Rates (CCRR): These indicate how the probability of answering an item correctly changes based on responses to prerequisite items
+ Parameter Learning: The model estimates the strength of dependencies between items
+ Model Fit Indices: These help evaluate whether the specified network structure adequately represents the relationships in the data

It is natural to assume directed paths from items with higher pass rates to those with lower pass rates. The model calculates conditional pass rates from parent nodes (higher pass rates) to child nodes (lower pass rates).

### Practical Applications

The Bayesian network approach offers several practical benefits for educational assessment:

+ Learning Pathways: By identifying prerequisite relationships, educators can design more effective learning sequences
+ Diagnostic Assessment: The model can identify specific knowledge gaps based on response patterns
+ Adaptive Testing: Results can inform which items to present next based on current performance
Curriculum Design: Understanding item relationships helps in structuring educational content

The model fit can be evaluated by comparing it with both the complete independence model and the saturated model. Furthermore, it can provide a learning roadmap by identifying which items are easier to answer correctly and which concepts should be mastered first.
For more advanced applications, the package also offers features to infer graph structures from data using genetic algorithms, as well as integration with biclustering to identify item groups and student classes simultaneously.

### Example

There are three ways to specify the graph. You can either pass a matrix-type DAG to the argument adj_matrix, pass a DAG described in a CSV file to the argument adj_file, or pass a graph-type object g used in the igraph package to the argument g.

The methods to create the matrix-type adj_matrix and the graph object g are as follows:

```{r setup-igraph, message=FALSE, warning=FALSE}
library(igraph)
DAG <-
  matrix(
    c(
      "Item01", "Item02",
      "Item02", "Item03",
      "Item02", "Item04",
      "Item03", "Item05",
      "Item04", "Item05"
    ),
    ncol = 2, byrow = T
  )
## graph object
g <- igraph::graph_from_data_frame(DAG)
g
## Adjacency matrix
adj_mat <- as.matrix(igraph::get.adjacency(g))
print(adj_mat)
```

A CSV file with the same information as the graph above in the following format. The first line contains column names (headers) and will not be read as data.

```{r print-dag, echo=FALSE, message=FALSE, warning=FALSE}
cat("From,To\n")
for (i in 1:nrow(DAG)) {
  cat(sprintf("%s,%s\n", DAG[i, 1], DAG[i, 2]))
}
```

While only one specification is sufficient, if multiple specifications are provided, they will be prioritized in the order of file, matrix, and graph object.

An example of executing BNM by providing a graph structure (DAG) is as follows:

```{r model-bnm, message=FALSE, warning=FALSE}
result.BNM <- BNM(J5S10, adj_matrix = adj_mat)
result.BNM
```


The package also offers features to infer graph structures from data using genetic algorithms, as well as a model that integrates with biclustering. For more details, please refer to the package manual or website(Kosugi,2023).

# Conclusion
This appendix has provided an overview of the basic usage of the exametrika package and detailed the Bayesian Network Model (BNM), which could not be elaborated in the main text of the paper. The overall workflow can be summarized as:

1. Prepare your data (create a data frame or load from a CSV file)
2. Convert it to the exametrika class format using the dataFormat function
3. Select and run an appropriate analysis method (LRA, Biclustering, Ranklustering, BNM, etc.)
4. Check the results and visualize them as needed

Within this workflow, data conversion using the dataFormat function represents the most critical step. Properly specifying the data type, ID column, and missing values in this function ensures smooth analysis in subsequent steps.

The `exametrika` package, based on the theoretical framework proposed by Shojima (2022) in "Test Data Engineering," provides an integrated approach for educational test data analysis. While the main paper focused primarily on LRA and Biclustering/Ranklustering, the package implements several other models. In particular, the BNM introduced in this appendix is valuable for discovering prerequisite relationships between items and presenting learning pathways.

The visualization capabilities of the package also play a crucial role in providing educational feedback. Through IRP and FRP plots, array plots, and network diagrams, statistical results can be transformed into practical insights for educators and test-takers.
Additionally, this appendix has addressed the package's application range, including handling ordinal data and accommodating polytomous data. These features enable appropriate analysis for a wider variety of test data.

The `exametrika` package goes beyond simple numerical evaluation, enabling a deeper understanding of examinee abilities and providing clear learning goals. By feeding back the results of rank analysis and clustering to educational settings, effective educational support becomes possible. For further details on the exametrika package, readers are encouraged to refer to the package website (Kosugi, 2023) and Shojima's (2022) book. These resources contain abundant information on other models and applications not covered in this appendix.
