---
title: 'metafrontier: Unified Metafrontier Analysis for Efficiency and Productivity
  in R'
date: '2026-09-30'
abstract: |
  Comparing the technical efficiency of groups of firms that face different restrictions of a common production technology requires a metafrontier framework that envelops the group-specific frontiers. Existing R packages cover parts of this workflow, but none combines SFA- and DEA-based metafrontier estimation with inference for technology gap ratios and the metafrontier Malmquist decomposition. We introduce metafrontier, a package that provides a unified interface for three complementary approaches: the deterministic metafrontier of Battese, Rao, and O'Donnell (2004) under both of its identification criteria, the stochastic metafrontier of Huang, Huang, and Liu (2014), and DEA-based metafrontier models with radial, directional, hyperbolic, and FDH technologies. The package estimates group-specific frontiers, constructs the metafrontier envelope, and decomposes efficiency into group technical efficiency and the technology gap ratio. Additional features include the metafrontier Malmquist productivity index with three-way decomposition, bootstrap confidence intervals for technology gap ratios, Murphy--Topel corrected standard errors, poolability tests for both SFA and DEA fits, convergence diagnostics, latent class metafrontier estimation via the EM algorithm, panel stochastic frontier models with time-varying inefficiency on balanced or unbalanced panels, and ggplot2 visualisation methods. Pre-fitted group frontiers from sfaR and frontier are accepted directly, and estimation can be delegated to sfaR, frontier, or Benchmarking as backends. We illustrate the package with simulated and real data and a Monte Carlo study covering parameter and efficiency recovery, small samples, and unbalanced group sizes. All examples and simulations are fully reproducible and complete in under ten minutes on a standard laptop.
draft: no
author:
- name: Erik Enstad
  affiliation: NHH Norwegian School of Economics
  address:
  - Department of Business and Management Science
  - Bergen, Norway
  orcid: 0009-0006-1053-3849
  email: erik.enstad@nhh.no
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
header-includes:
- \providecommand{\proglang}[1]{\textsf{#1}}
- \makeatletter\setlength{\@fptop}{0pt}\setlength{\@fpsep}{2.5em plus 2pt}\setlength{\@fpbot}{0pt
  plus 1fil}\makeatother
- \renewcommand{\topfraction}{0.85}
- \renewcommand{\bottomfraction}{0.7}
- \renewcommand{\textfraction}{0.1}
- \renewcommand{\floatpagefraction}{0.8}
- \setcounter{topnumber}{3}
- \setcounter{bottomnumber}{2}
- \setcounter{totalnumber}{5}
- \makeatletter\renewcommand*{\fps@figure}{htbp}\renewcommand*{\fps@table}{htbp}\makeatother
bibliography: RJreferences.bib
date_received: '2026-04-22'
volume: 18
issue: 3
slug: RJ-2026-055

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  cache = TRUE,
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  out.width = "100%"
)
options(width = 72)
library(metafrontier)
library(ggplot2)

# Helper: apply R Journal booktabs styling for PDF (avoids double-caption in HTML)
ktbl <- function(x) {
  if (knitr::is_latex_output()) {
    kableExtra::kable_styling(x,
      latex_options = "scale_down",
      position      = "center")
  } else {
    x
  }
}
```

```{r css-fix, echo=FALSE, results='asis'}
if (knitr::is_html_output()) {
  cat('<style>
  d-article pre { overflow-x: auto; max-width: 100%; }
  </style>')
}
```

# Introduction {#sec:intro}

Firms within the same industry often face different restrictions on the set
of production plans available to them. In microeconomic terms the industry
technology set contains all technically feasible input-output combinations,
but regulation, environmental conditions, resource endowments, and the cost
of switching production techniques can prevent particular groups of firms
from accessing every point in it. Steel producers, for example, operate
either the capital-intensive blast furnace--basic oxygen furnace (BF-BOF)
route or the scrap-based electric arc furnace (EAF) route; both belong to
the same industry technology set, but capital vintage and scrap availability
restrict which points an individual producer can reach. Similarly, dairy
farms in different countries face distinct climatic conditions and input
markets that restrict their feasible production plans [@moreira2010]. When
groups of producers face different restricted technology sets, pooling all
observations into a single stochastic frontier or data envelopment analysis
model conflates within-group inefficiency with between-group differences in
the accessible technology, potentially yielding misleading efficiency
rankings.

The metafrontier framework, introduced by @battese2004 and extended by
@odonnell2008 and @huang2014, addresses this problem by defining an
overarching production frontier, the *metafrontier*, that envelops the
group-specific frontiers. This framework decomposes a firm's distance from
the metafrontier into two components:

\begin{equation}
\text{TE}^*_i = \text{TE}_i \times \text{TGR}_i
(\#eq:decomposition)
\end{equation}

where $\text{TE}_i$ is the firm's *technical efficiency* relative to its own
group frontier, measuring managerial performance within the available
technology, and $\text{TGR}_i$ is the *technology gap ratio*, measuring how
close the group's frontier lies to the best-practice metafrontier. The product
$\text{TE}^*_i$ is the firm's *metafrontier technical efficiency*, that is,
its performance measured against the industry-wide best practice.
Figure \@ref(fig:conceptual) in Section `r if(knitr::is_latex_output()) "\\@ref(sec:methodology)" else "[2](#sec:methodology)"` illustrates this decomposition for
the single-input case.

The package is designed for applied economists, regulators, and
operations researchers who need to compare technical efficiency across
groups operating under different production technologies.

Several \proglang{R} [@R] packages address parts of the efficiency analysis
workflow. The \CRANpkg{frontier} package [@frontier], the \CRANpkg{sfaR}
package [@sfaR], and the newer \CRANpkg{sfa} package [@sfa] provide
stochastic frontier estimation across a wide set of specifications but
include no metafrontier layer. The \CRANpkg{Benchmarking} package [@bogetoft2011] implements a wide
range of DEA models; its `XREF`/`YREF` arguments score observations against
an externally supplied reference technology, so pooled-frontier DEA scores
can be assembled manually, but the user must orchestrate the per-group runs,
form and summarise the technology gap ratios, and forgo inference on them:
its bootstrap targets Farrell scores with respect to a single technology,
and its Malmquist functions compute the standard index rather than the
metafrontier decomposition. The \CRANpkg{deaR} package [@deaR] offers a
general DEA toolkit, including directional distance functions via
`model_dir()`, but no metafrontier workflow. The \CRANpkg{smfa} package
[@smfa], the CRAN release of the earlier GitHub package \pkg{metafrontieR}
[@metafrontieR], wraps \CRANpkg{sfaR} to estimate deterministic (LP and QP),
stochastic, latent class, and sample selection [@greene2010] metafrontiers,
but it is SFA-only: it provides no DEA-based metafrontier, no productivity
decomposition, no bootstrap or Murphy--Topel inference for technology gap
ratios, and no panel or visualisation support. No package, on CRAN or
elsewhere, combines SFA- and DEA-based metafrontier estimation with
inference for technology gap ratios and the metafrontier Malmquist
decomposition of @odonnell2008.

The \CRANpkg{metafrontier} package fills this gap. Its principal contributions
are:

- **Unified estimation framework.** A single function, `metafrontier()`,
  dispatches across SFA and DEA methods with deterministic or stochastic
  metafrontier formulations through the `method` and `meta_type` arguments.
- **Malmquist productivity decomposition.** The `malmquist_meta()` function
  computes the metafrontier Malmquist total factor productivity index with
  the three-way decomposition of @odonnell2008.
- **Bootstrap confidence intervals.** The `boot_tgr()` function provides
  parametric and nonparametric bootstrap inference for technology gap ratios,
  with parallel computation support.
- **Murphy--Topel corrected standard errors.** The stochastic metafrontier
  accounts for the generated-regressor problem using the variance correction
  of @murphy1985.
- **Latent class metafrontier.** The `latent_class_metafrontier()` function
  estimates group membership endogenously when technology classes are
  unobserved.
- **Interoperability.** Pre-fitted group frontiers from \CRANpkg{sfaR} and
  \CRANpkg{frontier} are accepted directly via the `models` argument, and
  the `engine` argument delegates group-frontier estimation to
  \CRANpkg{sfaR}, \CRANpkg{frontier}, or (for DEA) \CRANpkg{Benchmarking}.
- **Diagnostics and user safety.** Every estimation stage records its
  convergence status, inspectable via `check_convergence()` and
  `summary()`; poolability of the group frontiers is testable for both SFA
  (likelihood ratio) and DEA (permutation) fits.

The remainder of this article is organised as follows. Section `r if(knitr::is_latex_output()) "\\@ref(sec:methodology)" else "[2](#sec:methodology)"` reviews the
metafrontier methodology. Section `r if(knitr::is_latex_output()) "\\@ref(sec:package-design)" else "[3](#sec:package-design)"` describes the package design. Section `r if(knitr::is_latex_output()) "\\@ref(sec:basic-usage)" else "[4](#sec:basic-usage)"`
demonstrates basic usage with simulated data. Section `r if(knitr::is_latex_output()) "\\@ref(sec:advanced-features)" else "[5](#sec:advanced-features)"` covers advanced
features. Section `r if(knitr::is_latex_output()) "\\@ref(sec:monte-carlo)" else "[6](#sec:monte-carlo)"` presents Monte Carlo evidence on estimator performance.
Section `r if(knitr::is_latex_output()) "\\@ref(sec:comparison)" else "[7](#sec:comparison)"` compares \CRANpkg{metafrontier} with related packages, and Section `r if(knitr::is_latex_output()) "\\@ref(sec:summary)" else "[8](#sec:summary)"`
concludes.

# Metafrontier methodology {#sec:methodology}

Readers already familiar with stochastic frontier analysis and
metafrontier models may wish to proceed directly to Section `r if(knitr::is_latex_output()) "\\@ref(sec:package-design)" else "[3](#sec:package-design)"`.

Throughout, we use the term *technology* in the sense of the metafrontier
literature. The industry *metatechnology* $T^*$ is the technology set of
microeconomic production theory: the set of all input-output combinations
that are technically feasible given the current state of knowledge in the
industry [@odonnell2008]. Group-specific *technologies* $T_j \subset T^*$
are restricted subsets of the metatechnology, arising because regulation,
environmental conditions, resource endowments, or the cost of switching
production techniques prevent the firms in group $j$ from choosing every
point of $T^*$. Following the convention of @battese2004 and @odonnell2008,
we refer to these restricted sets simply as group technologies; the
metafrontier is the boundary of $T^*$, and the group frontiers are the
boundaries of the sets $T_j$.

## Group-specific stochastic frontiers {#sec:group-frontiers}

Consider $J$ groups of firms, where group $j$ ($j = 1, \ldots, J$) has
access to a potentially restricted technology set $T_j$. The
stochastic frontier production function for group $j$ is

\begin{equation}
y_{ij} = \mathbf{x}_{ij}'\boldsymbol{\beta}_j + v_{ij} - u_{ij},
\quad i = 1, \ldots, N_j
(\#eq:sfa)
\end{equation}

where $y_{ij}$ is the log output of firm $i$ in group $j$,
$\mathbf{x}_{ij}$ is a vector of log inputs (including a constant),
$\boldsymbol{\beta}_j$ is the group-specific parameter vector,
$v_{ij} \sim N(0, \sigma^2_{v,j})$ is a symmetric noise term, and
$u_{ij} \geq 0$ is a one-sided inefficiency term [@aigner1977]. The composed
error is $\varepsilon_{ij} = v_{ij} - u_{ij}$.

The \CRANpkg{metafrontier} package supports three distributional assumptions
for $u_{ij}$: the half-normal
($u_{ij} \sim |N(0, \sigma^2_{u,j})|$), the truncated normal
($u_{ij} \sim N^+(\mu_j, \sigma^2_{u,j})$), and the exponential
($u_{ij} \sim \text{Exp}(\lambda_j)$). Two estimators of technical
efficiency from the composed error are in common use, and the package
computes both. The default is the conditional expectation of @battese1988
(BC88),

\begin{equation}
\text{TE}_{ij} = E\!\left[\exp(-u_{ij}) \mid \varepsilon_{ij}\right]
= \frac{\Phi\!\left(\mu^*_{ij}/\sigma^*_j - \sigma^*_j\right)}
       {\Phi\!\left(\mu^*_{ij}/\sigma^*_j\right)}
  \exp\!\left(-\mu^*_{ij} + \tfrac{1}{2}\sigma^{*2}_j\right)
(\#eq:te)
\end{equation}

where $\mu^*_{ij}$ and $\sigma^*_j$ are the mean and standard deviation of
the conditional distribution of $u_{ij}$ given $\varepsilon_{ij}$, and
$\Phi(\cdot)$ is the standard normal distribution function. BC88 provides
consistent estimates of technical efficiency scores. The alternative
estimator $\exp(-\widehat{u}_{ij})$ with
$\widehat{u}_{ij} = E[u_{ij} \mid \varepsilon_{ij}]$ [@jondrow1982, JLMS]
does not account for Jensen's inequality; the difference is negligible when
$\sigma_u$ is small relative to $\sigma_v$ but grows for highly inefficient
samples. The `estimator` argument of `metafrontier()` selects between the
two (`"bc88"`, the default, or `"jlms"`); both are stored on the fitted
object, and `efficiencies()` can switch between them without refitting. For
a book-length treatment of stochastic frontier analysis, see
@kumbhakar2000.

## Deterministic metafrontier {#sec:det-metafrontier}

@battese2004 [building on @battese2002] define the metafrontier as the deterministic envelope of the
group-specific frontiers. The metafrontier production function satisfies
$f^*(\mathbf{x}) \geq f_j(\mathbf{x})$ for all $\mathbf{x}$ and all $j$,
meaning that it represents the output that would be attainable if all firms
had access to the most productive technology available in the industry.

Let $\widehat{\boldsymbol{\beta}}_j$ denote the maximum likelihood estimate from
group $j$, and let $\widehat{f}_{ij} = \mathbf{x}_{ij}'\widehat{\boldsymbol{\beta}}_j$
denote the fitted group frontier value for observation $i$ in group $j$.
The \CRANpkg{metafrontier} package obtains $\boldsymbol{\beta}^*$ by solving
the linear program

\begin{equation}
\min_{\boldsymbol{\beta}^*} \sum_{i=1}^{N}
\mathbf{x}_{i}'\boldsymbol{\beta}^*
\quad \text{s.t.} \quad
\mathbf{x}_{ij}'\boldsymbol{\beta}^* \geq \widehat{f}_{ij}
\quad \forall\, i, j
(\#eq:deterministic)
\end{equation}

which minimises the total metafrontier output subject to the envelope
constraint, producing the tightest feasible metafrontier. @battese2004
propose two criteria for identifying $\boldsymbol{\beta}^*$ under the same
envelope constraints: minimising the sum of absolute deviations, which
reduces to the linear program in Equation \@ref(eq:deterministic) because the
constraints force every deviation to be non-negative [@odonnell2008,
Eqs. 23--25], and minimising the sum of squared deviations, a convex
quadratic program. The two share the same constraint set but differ in
objective and can yield different $\boldsymbol{\beta}^*$ when not all
constraints are binding; Section `r if(knitr::is_latex_output()) "\\@ref(sec:monte-carlo)" else "[6](#sec:monte-carlo)"` reports their maximum absolute TGR
difference on the simulation designs. The package implements both:
`objective = "lp"` (the default) is solved via \CRANpkg{lpSolveAPI}
[@lpSolveAPI], and `objective = "qp"` is solved exactly via
\CRANpkg{quadprog} [@quadprog], with an adaptive-barrier fallback via
`constrOptim()` when \CRANpkg{quadprog} is unavailable. The technology gap
ratio under the deterministic metafrontier is

\begin{equation}
\text{TGR}_{ij} =
\frac{\exp(\mathbf{x}_{ij}'\widehat{\boldsymbol{\beta}}_j)}
     {\exp(\mathbf{x}_{ij}'\widehat{\boldsymbol{\beta}}^*)}
(\#eq:tgr-det)
\end{equation}

which lies in $[0, 1]$ by construction. Because the metafrontier is
deterministic, no standard errors are available for $\boldsymbol{\beta}^*$
without resampling methods.

## Stochastic metafrontier {#sec:stoch-metafrontier}

@huang2014 propose an alternative in which the metafrontier itself is
stochastic. Define the fitted group frontier value
$\widehat{y}_{ij} = \mathbf{x}_{ij}'\widehat{\boldsymbol{\beta}}_j$. The stochastic
metafrontier is estimated as a second-stage SFA:

\begin{equation}
\widehat{y}_{ij} = \mathbf{x}_{ij}'\boldsymbol{\beta}^* + v^*_{ij} - u^*_{ij}
(\#eq:stochastic)
\end{equation}

where $v^*_{ij}$ captures approximation noise and $u^*_{ij} \geq 0$ represents
the technology gap. This formulation provides standard errors and hypothesis
tests for $\boldsymbol{\beta}^*$, overcoming a key limitation of the
deterministic approach. Note that because the stochastic metafrontier need
not envelop the group frontiers at every point (due to the noise term
$v^*_{ij}$), the resulting TGR can exceed 1 for some observations: at those points
the group frontier locally surpasses the estimated metafrontier.
How often this occurs is sample-dependent: in the Monte Carlo design used
below no observation reaches $\text{TGR} > 1$, whereas on the
`sfaR::utility` example used later around 40\% of observations do,
with values up to about 1.25. Such excess values reflect noise rather
than superior-to-best-practice technology. For group-level means or
ranking applications we recommend leaving them untruncated so the noise
averages out; for firm-level interpretation we recommend truncating at
one (`pmin(tgr, 1)`) and reporting the truncated share as a
diagnostic.

However, because the dependent variable $\widehat{y}_{ij}$ is itself estimated
from the first stage, the standard errors of $\boldsymbol{\beta}^*$ are
understated unless corrected for the generated-regressor problem. The
\CRANpkg{metafrontier} package addresses this with the variance correction of
@murphy1985, as described in Section `r if(knitr::is_latex_output()) "\\@ref(sec:advanced-features)" else "[5](#sec:advanced-features)"`.

## DEA-based metafrontier {#sec:dea-metafrontier}

The metafrontier framework extends naturally to nonparametric methods
[@odonnell2008; @charnes1978]. For each group $j$, a data envelopment analysis
(DEA) frontier is estimated using only the observations in that group, yielding
group-specific efficiency scores $\text{TE}^{\text{group}}_{ij}$. A second DEA
problem pools all observations across groups, producing metafrontier scores
$\text{TE}^{\text{pool}}_{ij}$. In the output orientation (the default), the
score of observation $(\mathbf{x}_i, \mathbf{y}_i)$ against a reference set
$R$ is $\text{TE}_i = 1/\phi^*_i$, where $\phi^*_i$ solves the envelopment
linear program

\begin{equation}
\max_{\phi_i, \boldsymbol{\lambda}} \phi_i
\quad \text{s.t.} \quad
\sum_{j \in R} \lambda_j \mathbf{x}_j \leq \mathbf{x}_i, \quad
\sum_{j \in R} \lambda_j \mathbf{y}_j \geq \phi_i \mathbf{y}_i, \quad
\boldsymbol{\lambda} \in \Lambda,
(\#eq:dea-lp)
\end{equation}

and in the input orientation the score is $\theta^*_i$ from
$\min \theta_i$ subject to
$\sum_j \lambda_j \mathbf{x}_j \leq \theta_i \mathbf{x}_i$,
$\sum_j \lambda_j \mathbf{y}_j \geq \mathbf{y}_i$,
$\boldsymbol{\lambda} \in \Lambda$. The set $\Lambda$ encodes the returns to
scale: $\lambda_j \geq 0$ unrestricted (CRS), adding
$\sum_j \lambda_j = 1$ (VRS), $\sum_j \lambda_j \leq 1$ (DRS),
$\sum_j \lambda_j \geq 1$ (IRS), or $\lambda_j \in \{0, 1\}$ with
$\sum_j \lambda_j = 1$ (FDH, dropping convexity). The group-frontier
problem restricts the reference set $R$ to the observations of group $j$;
the metafrontier problem pools all $N$ observations into $R$. This
distinction is the entire metafrontier construction: the same LP is solved
twice per observation against nested reference sets. The technology gap
ratio is then

\begin{equation}
\text{TGR}_{ij} =
\frac{\text{TE}^{\text{pool}}_{ij}}{\text{TE}^{\text{group}}_{ij}}
(\#eq:tgr-dea)
\end{equation}

which lies in $(0, 1]$ because the pooled reference set contains the group
reference set. Beyond radial measures, the DEA path exposes directional
distance functions (DDF) via `type = "directional"`, with `direction`
either a preset (`"proportional"`, `"output"`, `"input"`), a common numeric
direction vector, or a matrix of firm-specific directions, and hyperbolic
(graph) efficiency via `type = "hyperbolic"`, which contracts inputs and
expands outputs simultaneously and remains feasible in cross-period
comparisons where radial cross-period programs may not be. Second-stage
slack maximisation is available via `slack = TRUE`, flagging observations
that are radially efficient but not Koopmans-efficient. The resulting
object carries the same `metafrontier` class and supports the standard
extractor and plotting methods. Because DEA is nonparametric, it avoids
functional form assumptions, but it does not account for statistical noise:
positive noise realisations shift the estimated frontier upwards, while
negative realisations are attributed to inefficiency. SFA, by contrast,
accounts for noise through the composed error, although neither approach
can separate the unobserved noise and inefficiency realisations for an
individual observation; SFA efficiency scores are conditional expectations
(Section `r if(knitr::is_latex_output()) "\\@ref(sec:group-frontiers)" else "[2.1](#sec:group-frontiers)"`).

## Efficiency decomposition {#sec:efficiency-decomp}

Regardless of the estimation method, the fundamental identity holds:

\begin{equation}
\text{TE}^*_{ij} = \text{TE}_{ij} \times \text{TGR}_{ij}
(\#eq:identity)
\end{equation}

where $\text{TE}_{ij}$ measures how efficiently firm $i$ uses its own group's
technology, $\text{TGR}_{ij}$ measures how close that group's technology is to
the best-practice metafrontier, and $\text{TE}^*_{ij}$ measures overall
performance relative to the industry-wide frontier.

Figure \@ref(fig:conceptual) illustrates this decomposition for a single-input
production function. The vertical distance from a firm's observed output to its
group frontier measures technical inefficiency, while the distance from the
group frontier to the metafrontier measures the technology gap. A firm can be
technically efficient within its group ($\text{TE}_i = 1$) yet still fall short
of the metafrontier if its group's technology is inferior
($\text{TGR}_i < 1$).

```{r conceptual, echo=FALSE, fig.width=7, fig.height=4.2, fig.cap="Metafrontier decomposition for a single-input production function (diagram drawn with ggplot2; code in the supplementary materials). For Firm A, technical efficiency (TE) is the distance from observed output to the group frontier, the technology gap ratio (TGR) the distance from the group frontier to the metafrontier (solid black), and TE* = TE x TGR.", fig.alt="A line plot showing two group-specific production frontiers as upward-sloping dashed lines and a metafrontier as a solid black line above both. A point labelled Firm A sits below the Group 1 frontier. Two vertical segments from the firm to the group frontier (labelled TE) and from the group frontier to the metafrontier (labelled TGR) illustrate the efficiency decomposition."}
# Conceptual diagram of metafrontier decomposition
f_g1 <- function(x) 0.5 + 0.7 * x
f_g2 <- function(x) 1.3 + 0.45 * x
f_meta <- function(x) 1.4 + 0.58 * x

x_firm <- 2.5
y_firm <- 2.0
y_g1 <- f_g1(x_firm)
y_meta <- f_meta(x_firm)

ggplot(data.frame(x = c(0.5, 4.5)), aes(x = x)) +
  stat_function(fun = f_g1, aes(colour = "Group 1 frontier"),
                linetype = "dashed", linewidth = 0.8) +
  stat_function(fun = f_g2, aes(colour = "Group 2 frontier"),
                linetype = "dashed", linewidth = 0.8) +
  stat_function(fun = f_meta, aes(colour = "Metafrontier"),
                linetype = "solid", linewidth = 1) +
  geom_point(aes(x = x_firm, y = y_firm), size = 3, shape = 16) +
  geom_segment(aes(x = x_firm + 0.02, y = y_firm,
                   xend = x_firm + 0.02, yend = y_g1),
               colour = "#2166AC", linewidth = 0.7,
               arrow = arrow(length = unit(0.12, "cm"), ends = "both")) +
  geom_segment(aes(x = x_firm + 0.02, y = y_g1,
                   xend = x_firm + 0.02, yend = y_meta),
               colour = "#B2182B", linewidth = 0.7,
               arrow = arrow(length = unit(0.12, "cm"), ends = "both")) +
  annotate("text", x = x_firm + 0.18, y = (y_firm + y_g1) / 2,
           label = "TE", fontface = "italic", size = 4, hjust = 0) +
  annotate("text", x = x_firm + 0.18, y = (y_g1 + y_meta) / 2,
           label = "TGR", fontface = "italic", size = 4, hjust = 0) +
  annotate("text", x = x_firm - 0.15, y = y_firm - 0.08,
           label = "Firm A", size = 3.5, hjust = 1) +
  scale_colour_manual(
    values = c("Group 1 frontier" = "#2166AC",
               "Group 2 frontier" = "#D6604D",
               "Metafrontier" = "black"),
    guide = guide_legend(override.aes = list(
      linetype = c("dashed", "dashed", "solid")))
  ) +
  labs(x = "Input (log)", y = "Output (log)", colour = NULL) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

Table \@ref(tab:approaches) summarises the key properties of the three
metafrontier approaches.

```{r approaches, echo=FALSE}
approaches <- data.frame(
  Property = c("Functional form", "Statistical noise",
               "Inference for metafrontier",
               "TGR bounds", "Key reference"),
  `Deterministic MF` = c("Parametric", "First stage only",
                          "None (resample)",
                          "[0, 1]", "BRO (2004)"),
  `Stochastic MF` = c("Parametric", "Both stages",
                       "Yes (with M--T correction)",
                       "Can exceed 1", "HHL (2014)"),
  DEA = c("Nonparametric", "None",
          "None (bootstrap)",
          "[0, 1]", "ORB (2008)"),
  check.names = FALSE
)
knitr::kable(approaches, caption = "Comparison of metafrontier approaches.",
             align = "llll", booktabs = TRUE, linesep = "") |>
  ktbl()
```

## Malmquist productivity decomposition {#sec:malmquist-methodology}

For panel data, the metafrontier Malmquist total factor productivity (TFP)
index extends the standard Malmquist index to incorporate technology gap
dynamics [@odonnell2008]. The index between periods $t$ and $t+1$ decomposes
as

\begin{equation}
M^*_{i,t,t+1} = \text{TEC}_{i,t,t+1} \times \text{TGC}_{i,t,t+1}
\times \text{TC}^*_{i,t,t+1}
(\#eq:malmquist)
\end{equation}

where $\text{TEC}$ is the *technical efficiency change* (catching up to the
group frontier), $\text{TGC}$ is the *technology gap change* (movement of the
group frontier relative to the metafrontier), and $\text{TC}^*$ is the
*metafrontier technical change* (shift in the metafrontier itself). This
three-way decomposition allows researchers to distinguish productivity growth
driven by improved managerial practice from gains attributable to technology
adoption or frontier innovation.

One interpretive caveat applies. The Malmquist index is not a *proper*
productivity index in the sense of @odonnell2012: except under restrictive
conditions on the technology, it is not multiplicatively complete and fails
the transitivity axiom of index theory, so chained multi-period or
cross-firm comparisons of index levels are not generally valid
[@odonnell2018]. Interpretation should be confined to binary
period-to-period comparisons, which is how the package reports the index.
Multiplicatively complete alternatives such as the Färe-Primont and Lowe
indexes are planned extensions (Section `r if(knitr::is_latex_output()) "\\@ref(sec:summary)" else "[8](#sec:summary)"`).

# Package design and architecture {#sec:package-design}

## Design principles and S3 interface {#sec:design-principles}

The \CRANpkg{metafrontier} package is designed around two principles: a unified
interface and a consistent S3 method suite. The package exports 13 standalone
functions and registers 38 S3 methods across five object classes:
`metafrontier` (the primary estimation result), `malmquist_meta` (productivity
decomposition), `boot_tgr` (bootstrap inference), `lc_metafrontier`
(latent class models), and `metafrontier_convergence` (diagnostics).
Table \@ref(tab:functions) lists all 16 user-facing
functions grouped by category, including 3 `autoplot` methods that are
registered as S3 methods of the \CRANpkg{ggplot2} generic rather than
standalone exports.

```{r functions, echo=FALSE}
funs <- data.frame(
  Category = c(rep("Estimation", 3),
               rep("Analysis", 5),
               rep("Inference", 3),
               rep("Visualisation", 3),
               rep("Simulation", 2)),
  Function = c("`metafrontier()`", "`malmquist_meta()`",
               "`latent_class_metafrontier()`",
               "`efficiencies()`", "`technology_gap_ratio()`",
               "`tgr_summary()`", "`poolability_test()`",
               "`check_convergence()`",
               "`boot_tgr()`", "`select_n_classes()`",
               "`as_metafrontier_model()`",
               "`autoplot.metafrontier()`",
               "`autoplot.malmquist_meta()`",
               "`autoplot.boot_tgr()`",
               "`simulate_metafrontier()`",
               "`simulate_panel_metafrontier()`"),
  Description = c("Main metafrontier estimation (SFA or DEA)",
                   "Metafrontier Malmquist TFP index",
                   "EM algorithm for latent class metafrontier",
                   "Extract TE, TGR, or TE* scores",
                   "Technology gap ratios by group",
                   "Summary statistics for TGR",
                   "Poolability test (LR for SFA, permutation for DEA)",
                   "Convergence diagnostics for all estimation stages",
                   "Bootstrap CIs for technology gap ratios",
                   "BIC-based selection of latent classes",
                   "Convert external models to metafrontier format",
                   "TGR density, efficiency, decomposition, frontier plots",
                   "Malmquist decomposition, TGR evolution, MPI trend plots",
                   "Bootstrap distribution and CI plots",
                   "Cross-sectional simulation with known DGP",
                   "Panel simulation with time-varying inefficiency")
)
knitr::kable(funs, caption = "User-facing functions in the metafrontier package.",
             align = "lll", booktabs = TRUE, linesep = "") |>
  ktbl()
```

A single `metafrontier()` entry point was chosen over separate functions
for each method--meta-type combination because users frequently compare
estimators on the same data; a unified interface guarantees that the
returned objects share the same S3 class and methods regardless of the
estimation path. S3 was preferred over S4 and R6 because it provides
the lightest-weight dispatch mechanism, requires no additional
dependencies, and matches the class system used by the upstream packages
(\CRANpkg{frontier}, \CRANpkg{Benchmarking}) with which \CRANpkg{metafrontier}
interoperates.

All primary estimation is accessed through `metafrontier()`, which dispatches
via its `method` argument (`"sfa"` or `"dea"`) and `meta_type` argument
(`"deterministic"` or `"stochastic"`). The returned `metafrontier` object
supports the standard \proglang{R} model interface: `summary()`, `coef()`,
`vcov()`, `confint()`, `fitted()`, `residuals()`, `logLik()`, `nobs()`,
`predict()`, and `plot()`, so fitted objects work with the standard
model-handling idiom without adaptation.

## Estimation workflow {#sec:estimation-workflow}

The package supports three paths to metafrontier estimation. In the
**formula path**, the user supplies raw data, a formula, and a group
identifier; the package estimates all group-specific frontiers internally
before constructing the metafrontier. In the **pre-fitted model path**, the
user first estimates group frontiers using external packages (\CRANpkg{sfaR}
or \CRANpkg{frontier}) and passes the fitted objects directly to
`metafrontier()` via the `models` argument; conversion to the internal
format happens automatically. In the **engine path**, the user stays in the
formula interface but delegates the group-frontier estimation to an
external backend via the `engine` argument: `engine = "sfaR"` or
`engine = "frontier"` for SFA group frontiers, and
`engine = "Benchmarking"` for the DEA path, where both the group frontiers
and the pooled metafrontier are computed by `Benchmarking::dea()` using its
`XREF`/`YREF` external-reference facility.

The formula path is the simplest workflow. A single call to `metafrontier()`
handles all estimation steps: parsing the formula, splitting data by group,
estimating group-specific frontiers, constructing the metafrontier envelope,
and computing the TGR decomposition. The pre-fitted model and engine paths
are useful when the user requires fine-grained control over individual group
frontier specifications or needs features of the specialised packages
that the internal estimators lack. The metafrontier stage of the SFA methods is always
estimated internally, because the Murphy--Topel variance correction
(Section `r if(knitr::is_latex_output()) "\\@ref(sec:advanced-features)" else "[5](#sec:advanced-features)"`) differentiates the internal stage-2 likelihood.

## Interoperability {#sec:interoperability}

Fitted model objects from \CRANpkg{sfaR} (class `sfacross`) and
\CRANpkg{frontier} (class `frontier`) can be passed directly to the
`models` argument of `metafrontier()`; the conversion to the internal
format is performed automatically by the `as_metafrontier_model()` generic,
which users only need to call explicitly for hand-built list models (the
conversion is idempotent, so pre-converting is harmless). Farrell objects
from \CRANpkg{Benchmarking} store neither coefficients nor the underlying
data, so they cannot be used with `models`; the `engine = "Benchmarking"`
path covers that use case instead. The following example illustrates the
pre-fitted model path (not evaluated):

```{r interop-example, eval=FALSE}
library(sfaR)
# Estimate group frontiers externally
mod1 <- sfacross(log_y ~ log_x1 + log_x2,
                 data = subset(dat, group == "G1"))
mod2 <- sfacross(log_y ~ log_x1 + log_x2,
                 data = subset(dat, group == "G2"))

# Pass the fitted objects directly; conversion is automatic
fit <- metafrontier(models = list(G1 = mod1, G2 = mod2))
```

## Computational details {#sec:computational-details}

The package implements its own maximum likelihood estimator for group-specific
stochastic frontiers, using BFGS optimisation with Nelder--Mead fallback for
robustness. Variance parameters are log-transformed to ensure unconstrained
optimisation over the positive real line. The deterministic metafrontier of Equation \@ref(eq:deterministic)
is solved via linear programming using the
\CRANpkg{lpSolveAPI} package [@lpSolveAPI], with LP object reuse across
observations for improved performance on large datasets. The squared-deviations
variant (`objective = "qp"`) is solved exactly by the dual method of
Goldfarb and Idnani as implemented in \CRANpkg{quadprog} [@quadprog], which
is listed in `Suggests` to keep the import footprint unchanged. Numerical
stability is maintained through a safe Mills ratio computation
(`.safe_mills()`) that prevents `NaN` values in extreme tail regions of the
normal distribution, relevant for both panel SFA and latent class models.

The package keeps its dependency footprint minimal. The seven `Imports`
(`stats`, `graphics`, `grDevices`, and `methods` from base \proglang{R};
`Formula` for multi-part formula parsing; `numDeriv` for Hessian
computation; and `lpSolveAPI` for LP solving) are each exercised by core
functionality. \CRANpkg{lpSolveAPI} was chosen over alternatives such as
\CRANpkg{Rglpk} because it is a self-contained C library with no external
system dependencies, so it installs without additional system libraries
on all supported platforms. Packages used
only in optional features (\CRANpkg{ggplot2} for `autoplot()` methods,
\pkg{parallel} for bootstrap parallelisation, \CRANpkg{quadprog} for the QP
objective, and the three upstream SFA/DEA engines for interoperability and
backend delegation) are listed as `Suggests` and loaded conditionally.

Correctness is supported by a test suite of 22 \CRANpkg{testthat} files
(over 125 test blocks) covering core SFA/DEA estimation, Monte Carlo
recovery, bootstrap, Murphy--Topel correction, panel SFA, latent class,
DDF, and ggplot2 `autoplot` methods, plus snapshot tests of S3
outputs. Continuous integration runs the suite on Ubuntu, Windows, and
macOS under R-devel, release, and oldrel. The custom MLE is
additionally benchmarked against \CRANpkg{frontier}
(`test-frontier-equivalence.R`): on a 3-group dataset from
`simulate_metafrontier()` with 150 firms per group and technology gaps of
0, 0.3, and 0.5 (seed 42), similar to the simulation introduced in
Section `r if(knitr::is_latex_output()) "\\@ref(sec:simulating-data)" else "[4.1](#sec:simulating-data)"`, the maximum absolute coefficient difference against
`frontier::sfa()` is $1.2 \times 10^{-5}$ for group 2 and
$5.0 \times 10^{-6}$ for group 3. The two estimators use different
internal parameterisations but deliver numerically equivalent slope and
intercept estimates.

**Computational environment.** All timings reported in this article were
obtained on a laptop with an Intel Core Ultra 5 125U processor and 32 GB
of RAM running Windows 11 x64, using `r R.version.string` and
\CRANpkg{metafrontier} `r as.character(utils::packageVersion("metafrontier"))`.
The complete article code, including the Monte Carlo study of Section `r if(knitr::is_latex_output()) "\\@ref(sec:monte-carlo)" else "[6](#sec:monte-carlo)"`,
runs in under ten minutes in this environment.

# Basic usage {#sec:basic-usage}

All examples in this section use simulated data for full reproducibility. The
\CRANpkg{metafrontier} package includes data-generating functions that produce
data with known true parameters, so estimator accuracy can be assessed
directly.

## Simulating metafrontier data {#sec:simulating-data}

The `simulate_metafrontier()` function creates cross-sectional data with
`n_groups` technology groups, each containing `n_per_group` firms. The
data-generating process uses a Cobb--Douglas production function with
group-specific intercepts that induce known technology gaps.

```{r simulate}
sim <- simulate_metafrontier(n_groups = 3, n_per_group = 200, seed = 42)
dat <- sim$data
head(dat[, c("log_y", "log_x1", "log_x2", "group",
             "true_te", "true_tgr")])
```

The returned list contains `$data` (the simulated data frame) and `$params`
(the true parameter values). The data frame includes the true technical
efficiency (`true_te`), true technology gap ratio (`true_tgr`), and true
metafrontier efficiency (`true_te_star`), so estimates can be compared
directly with
estimated values.

## Deterministic metafrontier {#sec:det-metafrontier-usage}

The deterministic metafrontier of @battese2004 is estimated by specifying
`meta_type = "deterministic"`:

```{r deterministic}
fit_det <- metafrontier(log_y ~ log_x1 + log_x2, data = dat,
                        group = "group", method = "sfa",
                        meta_type = "deterministic", dist = "hnormal")
summary(fit_det)
```

The `summary()` method reports the group-specific frontier coefficients, the
metafrontier coefficients obtained from the constrained optimisation of
Equation \@ref(eq:deterministic), and summary statistics for the technology gap ratios
by group.

## Stochastic metafrontier {#sec:stoch-metafrontier-usage}

The stochastic metafrontier of @huang2014 replaces the deterministic envelope
with a second-stage SFA:

```{r stochastic}
fit_sto <- metafrontier(log_y ~ log_x1 + log_x2, data = dat,
                        group = "group", method = "sfa",
                        meta_type = "stochastic", dist = "hnormal")
coef(fit_sto)
confint(fit_sto)
```

Unlike the deterministic approach, the stochastic metafrontier provides
standard errors, $z$-statistics, and confidence intervals for the metafrontier
coefficients directly.

## DEA-based metafrontier {#sec:dea-metafrontier-usage}

The DEA-based metafrontier computes group-specific and pooled DEA frontiers
without parametric assumptions. Because DEA requires strictly positive inputs
and outputs, we use level-scale variables rather than the log-transformed
variables used for SFA:

```{r dea}
dat_lev <- within(dat, {
  y  <- exp(log_y)
  x1 <- exp(log_x1)
  x2 <- exp(log_x2)
})
fit_dea <- metafrontier(y ~ x1 + x2, data = dat_lev,
                        group = "group", method = "dea", rts = "vrs")
tgr_summary(fit_dea)
```

## Extracting results and diagnostics {#sec:extracting-results}

The `efficiencies()` function extracts efficiency scores. Three types are
available: group technical efficiency (`"group"`), metafrontier efficiency
(`"meta"`), and technology gap ratios (`"tgr"`). The fundamental identity
of Equation \@ref(eq:identity) can be verified numerically:

```{r identity-check}
te <- efficiencies(fit_det, type = "group")
tgr <- efficiencies(fit_det, type = "tgr")
te_star <- efficiencies(fit_det, type = "meta")
all.equal(te_star, te * tgr, tolerance = 1e-10)
```

The `tgr_summary()` function provides descriptive statistics for the TGR by
group:

```{r tgr-summary}
tgr_summary(fit_det)
```

The `poolability_test()` function tests the null hypothesis that all groups
share a common frontier against the alternative of group-specific
frontiers. For SFA fits it conducts a likelihood ratio test; for DEA fits,
where no likelihood exists, it dispatches to a permutation test that
exploits the exchangeability of group labels under the pooled-technology
null [@simar2007; @kneip2016]:

```{r poolability}
poolability_test(fit_det)
```

Convergence of every estimation stage can be verified with
`check_convergence()`, which reports one row per group frontier plus the
metafrontier stage. The same information is printed by `summary()`, and a
warning is issued at fit time whenever an optimiser stops at a non-zero
code. Estimation aborts with an error only when no usable estimate exists;
otherwise the codes are recorded so that users can inspect them before
interpreting TGRs, confidence intervals, or productivity decompositions:

```{r convergence-check}
check_convergence(fit_det)
```

The \CRANpkg{ggplot2} [@ggplot2]-based `autoplot()` method provides four plot types.
Figure \@ref(fig:tgr-density) shows the distribution of technology gap ratios
by group, and Figure \@ref(fig:decomposition) displays the efficiency
decomposition.

```{r tgr-density, fig.width=7, fig.height=4.5, fig.cap="Distribution of technology gap ratios by group from the deterministic metafrontier (generated with the package autoplot() method, view: TGR density). Group G1 defines the metafrontier (TGR = 1.000 for all observations, annotated separately); groups G2 and G3 show increasing technology gaps.", fig.alt="A density plot showing two smooth distributions of technology gap ratios: Group G2 centred near 0.74 and Group G3 centred near 0.62. An annotation marks Group G1 as having TGR identically equal to one."}
autoplot(fit_det, which = "tgr")
```

```{r decomposition, fig.width=7, fig.height=4.5, fig.cap="Mean efficiency decomposition by group (generated with the package autoplot() method, view: decomposition). The three bars per group show mean technical efficiency (TE), technology gap ratio (TGR), and metafrontier efficiency TE* = TE x TGR.", fig.alt="A grouped bar chart with three bars per group showing mean TE, TGR, and TE star. TE is similar across groups around 0.75, while TGR decreases from 1.0 in Group 1 to about 0.7 in Group 3, driving the decrease in TE star."}
autoplot(fit_det, which = "decomposition")
```

## Real-data illustration {#sec:real-data}

The examples above use simulated data with known parameters. To demonstrate
the package on empirical data, we use the `utility` dataset from
\CRANpkg{sfaR}, which contains 791 observations on US fossil-fuel steam
electric generating plants over 1986--1996. The binary variable `regu`
equals one for plants in states that enacted retail-access (restructuring)
legislation during the sample period and zero for plants that remained
under traditional regulation. This gives a natural grouping of plants. The
production function relates output (`y`) to capital (`k`), labour (`labor`),
and fuel (`fuel`).

```{r real-data, cache = FALSE}
data("utility", package = "sfaR")
utility$group <- ifelse(utility$regu == 1, "Retail access", "Regulated")
utility$log_y     <- log(utility$y)
utility$log_k     <- log(utility$k)
utility$log_labor <- log(utility$labor)
utility$log_fuel  <- log(utility$fuel)

fit_util <- metafrontier(log_y ~ log_k + log_labor + log_fuel,
                         data = utility, group = "group",
                         method = "sfa", meta_type = "deterministic")
tgr_summary(fit_util)
```

The traditionally regulated group exhibits a slightly higher mean TGR
(0.946) than the group of plants in retail-access states (0.892); the
regulated plants lie closer to the pooled metafrontier on average in this
sample. We attach no structural interpretation to this
illustrative comparison. In the metafrontier framework, a technology gap
reflects restrictions on the feasible input-output set; regulatory
constraints that only affect the choice of input mix would manifest as
allocative rather than technical inefficiency and would not, by
themselves, generate a technology gap.

Because all three estimators share the same interface and object class,
cross-method robustness checks require no additional plumbing. The
following chunk re-estimates the same model with the stochastic
metafrontier and with DEA, then compares group-mean TGRs and the Spearman
rank correlations of the plant-level TGRs across estimators:

```{r real-data-robustness, cache = FALSE}
fit_util_st  <- metafrontier(log_y ~ log_k + log_labor + log_fuel,
                             data = utility, group = "group",
                             method = "sfa", meta_type = "stochastic")
fit_util_dea <- metafrontier(log_y ~ log_k + log_labor + log_fuel,
                             data = utility, group = "group",
                             method = "dea", rts = "vrs")

tgr_by <- function(fit) tapply(efficiencies(fit, type = "tgr"),
                               utility$group, mean)
round(rbind(Deterministic = tgr_by(fit_util),
            Stochastic    = tgr_by(fit_util_st),
            DEA           = tgr_by(fit_util_dea)), 3)
round(c(det_vs_stoch = cor(fit_util$tgr, fit_util_st$tgr,
                           method = "spearman"),
        det_vs_dea   = cor(fit_util$tgr, fit_util_dea$tgr,
                           method = "spearman")), 3)
```

Agreement in the group ranking across a deterministic envelope, a
distributional second stage, and a nonparametric pooled frontier indicates
that the technology gap is a feature of the data rather than an artefact
of any single method; this kind of triangulation is the main practical
payoff of a unified interface.

# Advanced features {#sec:advanced-features}

## Bootstrap confidence intervals {#sec:bootstrap-ci}

Because the deterministic metafrontier provides no standard errors for the
technology gap ratios, the `boot_tgr()` function implements bootstrap inference.
Both parametric (residual resampling) and nonparametric (case resampling)
approaches are supported, with percentile and bias-corrected accelerated (BCa)
interval types.

```{r bootstrap}
boot <- boot_tgr(fit_det, R = 199, type = "parametric",
                 ci_type = "percentile", seed = 1, progress = FALSE)
boot_ci <- confint(boot)
head(boot_ci, 5)
```

We use $R = 199$ here to keep the article reproducible in minutes; in
applied work, $R \geq 999$ is recommended for reliable percentile
intervals, and $R \geq 1999$ for BCa intervals.

The `confint()` method returns observation-level confidence intervals;
Figure \@ref(fig:boot-ci) displays the bootstrap distribution of the
group-level means of the TGRs, and Table \@ref(tab:boot-ci-tab) summarises
the intervals by group. Parallel computation is available via the `ncores`
argument, which uses the \pkg{parallel} package for distributing
replications across cores. Table \@ref(tab:boot-ci-tab) is produced by the
following (echoed) chunk.

```{r boot-ci, fig.width=7, fig.height=4.5, fig.cap="Bootstrap distribution of the group-level mean technology gap ratios, from 199 parametric replications (generated with the package autoplot() method for bootstrap objects, default view). Dashed vertical lines mark the 95 percent percentile confidence intervals for the group means.", fig.alt="A set of histograms showing the bootstrap distribution of mean TGR for each group. Vertical dashed lines mark the 95 percent confidence interval bounds for each group mean. Group G1 has the narrowest interval near 1, while Group G3 has the widest interval at lower values."}
autoplot(boot)
```

```{r boot-ci-tab}
ci_df <- data.frame(
  Group = dat$group,
  Lower = boot_ci[, 1],
  Upper = boot_ci[, 2],
  TGR = efficiencies(fit_det, type = "tgr")
)
ci_summary <- aggregate(cbind(TGR, Lower, Upper) ~ Group,
                        data = ci_df, FUN = mean)
names(ci_summary) <- c("Group", "Mean TGR", "Mean lower bound",
                       "Mean upper bound")
cap <- paste("Bootstrap 95\\% confidence intervals for TGR by group",
             "(group-level means of observation-level intervals).")
knitr::kable(ci_summary, caption = cap, digits = 4,
             align = "lccc", booktabs = TRUE, linesep = "") |>
  ktbl()
```

## Murphy--Topel variance correction {#sec:murphy-topel}

The stochastic metafrontier of @huang2014 uses fitted values from the
first-stage group SFA models as the dependent variable in a second-stage SFA.
This two-step procedure understates the standard errors of the metafrontier
coefficients because it ignores the sampling variability from the first
stage (the generated-regressor problem). The `vcov()` method implements the
correction of @murphy1985, which inflates the second-stage
variance--covariance matrix to account for first-stage estimation
uncertainty.

```{r murphy-topel}
se_uncorrected <- sqrt(diag(vcov(fit_sto)))
se_corrected <- sqrt(diag(vcov(fit_sto, correction = "murphy-topel")))
comparison <- data.frame(
  Uncorrected = se_uncorrected,
  `Murphy-Topel` = se_corrected,
  `Ratio` = se_corrected / se_uncorrected,
  check.names = FALSE
)
knitr::kable(comparison, digits = 4, booktabs = TRUE, linesep = "",
  caption = paste("Comparison of uncorrected and",
    "Murphy--Topel corrected standard errors",
    "for the stochastic metafrontier coefficients.")
) |> ktbl()
```

The corrected standard errors in Table \@ref(tab:murphy-topel) are
uniformly larger; the increase is the sampling uncertainty contributed by
the generated regressors. Confidence intervals computed with
`confint(fit_sto, correction = "murphy-topel")` use the corrected standard
errors. The correction is computed via numerical differentiation of the
second-stage log-likelihood with respect to the first-stage parameters,
requiring $O(p_1 \times p_2)$ likelihood evaluations per group, where
$p_1$ is the number of first-stage parameters of that group's frontier
(frontier coefficients plus variance parameters) and $p_2$ is the number
of second-stage metafrontier parameters. For typical metafrontier
specifications (3--5 regressors) this completes in seconds, but the cost
scales quadratically and may become noticeable with high-dimensional
translog models.

## Heteroscedastic SFA {#sec:heteroscedastic-sfa}

The package supports heteroscedastic group frontiers through the formula
interface. A two-part formula `y ~ x1 + x2 | z1` models the mean of
$\sigma_{u,i}$ (for half-normal and exponential) or $\mu_i$ (for
truncated-normal) as a function of environmental variables:

```{r heteroscedastic}
fit_het <- metafrontier(log_y ~ log_x1 + log_x2 | log_x1, data = dat,
                        group = "group", method = "sfa",
                        meta_type = "deterministic", dist = "hnormal")
coef(fit_het)
```

This specification allows the degree of inefficiency to vary with observable
firm characteristics within each group, which is relevant when the variance
of inefficiency is not constant across producers. Heteroscedasticity enters
only the group TE scores: the metafrontier envelope depends on
$\mathbf{x}'\widehat{\boldsymbol{\beta}}_j$, not on variance parameters, so
TGRs are unaffected while TE and TE* reflect firm-specific inefficiency.

## Latent class metafrontier {#sec:latent-class}

When group membership is unobserved, the `latent_class_metafrontier()` function
discovers technology classes endogenously and estimates the metafrontier
simultaneously using an EM algorithm. Unlike `metafrontier()`, this function
does not accept a `group` argument: class membership is determined entirely by
the data. The number of latent classes can be selected using BIC via
`select_n_classes()`.

```{r latent-class}
lc <- latent_class_metafrontier(log_y ~ log_x1 + log_x2, data = dat,
                                n_classes = 3, n_starts = 5, seed = 42)
coef(lc)
```

As with all latent class stochastic frontier models, estimation may be
sensitive to starting values and local optima. The EM algorithm terminates
when the relative change in marginal log-likelihood between iterations
falls below a user-controllable tolerance (`tol`, default $10^{-6}$) or
after `max_iter` iterations (default 200). The package therefore supports
multiple random starts via the `n_starts` argument (default 10) and
information-criteria-based model selection via `select_n_classes()`, which
fits models for each class count in `n_classes_range` and returns the
BIC-optimal specification. For applied work we recommend at least $n\_starts = 10$ for
two or three classes, increased to 20--30 for four or more classes or when
the likelihood surface is known to be multimodal.

## Panel SFA with time-varying inefficiency {#sec:panel-sfa}

For panel data, the package supports the time-varying inefficiency
specifications of @battese1992 (BC92) and @battese1995 (BC95). The BC92 model
parameterises inefficiency as
$u_{it} = u_i \cdot \exp\!\left[-\eta(t - T)\right]$, where $u_i$ is a
firm-specific truncated-normal draw, $T$ is the final period of the panel,
$\eta > 0$ implies declining inefficiency over time, and $\eta < 0$ implies
increasing inefficiency. Unbalanced panels are supported: each firm
contributes its own set of periods $\mathcal{T}_i$ to the likelihood, and
the decay term is still measured from the panel-wide final period $T$,
following the unbalanced formulation of @battese1992.

The BC95 model instead specifies observation-specific inefficiency
$u_{it} \sim N^+(\mathbf{z}_{it}'\boldsymbol{\delta}, \sigma_u^2)$, where
$\mathbf{z}_{it}$ is a vector of inefficiency determinants and
$\boldsymbol{\delta}$ is estimated jointly with the frontier in a single
stage by maximum likelihood. Unlike BC92, the BC95 model treats
observations as independent draws (no within-firm time dependence). It is
selected with `panel_dist = "bc95"` and requires the two-part formula
syntax `y ~ x1 + x2 | z1 + z2` introduced for the heteroscedastic models
above.

```{r panel-simulate}
psim <- simulate_panel_metafrontier(
  n_groups = 2, n_firms_per_group = 30, n_periods = 5,
  eta = 0.05, seed = 42
)
pdat <- psim$data
```

```{r panel-estimate}
fit_panel <- metafrontier(log_y ~ log_x1 + log_x2, data = pdat,
                          group = "group", method = "sfa",
                          meta_type = "deterministic",
                          panel = list(id = "firm", time = "year"),
                          panel_dist = "bc92")
coef(fit_panel)
tgr_summary(fit_panel)
```

All estimated parameters, including the decay parameter $\eta$ and the
variance parameters, are available with
`coef(fit_panel, which = "group", extraPar = TRUE)`, and the group tables
printed by `summary()` report $\eta$ with its standard error.

Panel SFA can be combined with either `meta_type = "deterministic"` (as
above) or `meta_type = "stochastic"`; in the latter case the group-specific
time-varying fits are fed into the second-stage stochastic metafrontier
exactly as in the cross-sectional case.

## Malmquist productivity decomposition {#sec:malmquist-usage}

The `malmquist_meta()` function computes the metafrontier Malmquist TFP index
with the three-way decomposition of Equation \@ref(eq:malmquist). Firms are matched
across periods by the identifier supplied through the `id` argument;
observations without a match in the adjacent period (unbalanced panels,
entry and exit, or firms that switch groups) are dropped from the affected
transition with a warning that reports the count. When `id` is omitted the
function falls back to matching by row position within each group, which is
only valid for balanced panels sorted identically in every period, and says
so in a message. Under variable, decreasing, or increasing returns to
scale, cross-period DEA programs can be genuinely infeasible when an
observation lies outside the comparison technology; such cases yield `NA`
(never infinite values), are excluded from reported means, and are counted
in a consolidated warning and in the printed output. Constant returns to
scale (used below) avoids the issue. The SFA path (`method = "sfa"`)
approximates metafrontier distances by the pointwise maximum over the
estimated group frontiers and announces this in a message; the DEA path
computes exact distances to the pooled frontier and is the default. The
identity $M^* = \text{TEC} \times \text{TGC} \times \text{TC}^*$ can be
verified numerically.

```{r malmquist}
pdat_lev <- within(pdat, {
  y  <- exp(log_y)
  x1 <- exp(log_x1)
  x2 <- exp(log_x2)
})
malm <- malmquist_meta(y ~ x1 + x2, data = pdat_lev,
                       group = "group", time = "year", id = "firm",
                       method = "dea", rts = "crs")
print(malm)
```

Figure \@ref(fig:malmquist-plot) displays the Malmquist decomposition over
time. Each point is plotted at the end period of its transition, so the
first point, at period 2, is the change from period 1 to period 2.

```{r malmquist-plot, fig.width=7, fig.height=4.5, fig.cap="Mean metafrontier Malmquist productivity index (MPI) by period and group (generated with the package autoplot() method, view: MPI trend). Each point sits at the end period of its transition; values above the dashed line indicate improvement relative to the previous period.", fig.alt="A line plot showing the mean MPI over periods 2 to 5 for each group, with a dashed reference line at 1. Points are marked at each period, with separate coloured lines for each group."}
autoplot(malm, which = "mpi_trend")
```

# Monte Carlo evidence {#sec:monte-carlo}

A simulation study evaluates the finite-sample performance of the
deterministic, stochastic, and DEA metafrontier estimators.

## Design {#sec:mc-design}

We study two data-generating processes. **Scenario A** follows the group
SFA specification of Equation \@ref(eq:sfa) with two groups ($J = 2$), 100 firms per
group, and known true parameters:
$\boldsymbol{\beta}^* = (1.0, 0.5, 0.3)'$ (intercept, $\beta_1$, $\beta_2$),
technology gap $\delta = (0, 0.4)$, inefficiency standard deviations
$\sigma_u = (0.2, 0.3)$, and noise standard deviation $\sigma_v = 0.15$.
Log inputs are drawn independently and uniformly on $[0, 5]$, identically
across groups, and the groups differ only in their frontier intercepts, so
a common log-linear metafrontier exists and coefficient recovery is well
defined. **Scenario B** relaxes the features of Scenario A that are
unrealistic in applications: the three groups have group-specific *slope*
coefficients (not merely intercept shifts), log inputs are correlated
within firm (correlation 0.6, a firm-size effect), and the group input
distributions differ in location (one group is intensive in each input).
Under group-specific slopes no single log-linear frontier tightly envelops
all groups; the true envelope is the pointwise maximum
$\max_j \mathbf{x}_i'\boldsymbol{\beta}_j$, the true TGR
$\exp(\mathbf{x}_i'\boldsymbol{\beta}_g - \max_j
\mathbf{x}_i'\boldsymbol{\beta}_j)$ varies across observations within a
group, and the estimated metafrontier coefficients converge to a
pseudo-true tightest linear envelope that depends on the input
distribution. Scenario B is therefore evaluated on the recovery of the
efficiency quantities (TE, TE$^*$, TGR) rather than on coefficient bias.
Each scenario is replicated 500 times, and in each replication we estimate
the deterministic SFA, stochastic SFA, and DEA metafrontiers. The complete
simulation code for this section, including both scenario loops and the
table construction, is provided in the supplementary replication script
(Targets 3 and 7 to 10).

```{r monte-carlo, cache=TRUE, echo=FALSE}
set.seed(123)
n_reps <- 500
beta_true <- c(1.0, 0.5, 0.3)
mc_results <- vector("list", n_reps)
mc_eff <- vector("list", n_reps)

# Efficiency-recovery statistics for one fitted metafrontier object
eff_stats <- function(fit, dat, method_label) {
  te  <- efficiencies(fit, type = "group")
  tes <- efficiencies(fit, type = "meta")
  tgr <- efficiencies(fit, type = "tgr")
  data.frame(
    method = method_label,
    te_true = mean(dat$true_te),   te_est = mean(te),
    rho_te = cor(te, dat$true_te, method = "spearman"),
    tes_true = mean(dat$true_te_star), tes_est = mean(tes),
    rho_tes = cor(tes, dat$true_te_star, method = "spearman"),
    tgr_true = mean(dat$true_tgr), tgr_est = mean(tgr),
    rho_tgr = suppressWarnings(
      cor(tgr, dat$true_tgr, method = "spearman"))
  )
}

for (r in seq_len(n_reps)) {
  sim_mc <- simulate_metafrontier(
    n_groups = 2, n_per_group = 100,
    beta_meta = beta_true,
    tech_gap = c(0, 0.4),
    sigma_u = c(0.2, 0.3),
    sigma_v = 0.15,
    seed = r * 1000
  )

  fit_d <- tryCatch(
    metafrontier(log_y ~ log_x1 + log_x2, data = sim_mc$data,
                 group = "group", method = "sfa",
                 meta_type = "deterministic"),
    error = function(e) NULL
  )
  fit_s <- tryCatch(
    metafrontier(log_y ~ log_x1 + log_x2, data = sim_mc$data,
                 group = "group", method = "sfa",
                 meta_type = "stochastic"),
    error = function(e) NULL
  )
  fit_dea <- tryCatch(
    metafrontier(log_y ~ log_x1 + log_x2, data = sim_mc$data,
                 group = "group", method = "dea",
                 orientation = "output", rts = "vrs"),
    error = function(e) NULL
  )

  if (!is.null(fit_d) && !is.null(fit_s)) {
    beta_d <- coef(fit_d)
    beta_s <- coef(fit_s)
    tgr_d <- efficiencies(fit_d, type = "tgr")
    tgr_s <- efficiencies(fit_s, type = "tgr")
    true_tgr <- sim_mc$data$true_tgr

    mc_results[[r]] <- data.frame(
      rep = r,
      method = rep(c("Deterministic", "Stochastic"), each = 3),
      parameter = rep(c("Intercept", "beta_1", "beta_2"), 2),
      true = rep(beta_true, 2),
      estimate = c(beta_d[1:3], beta_s[1:3]),
      mean_tgr_est = rep(c(mean(tgr_d), mean(tgr_s)), each = 3),
      mean_tgr_true = rep(mean(true_tgr), 6),
      tgr_corr = rep(c(cor(tgr_d, true_tgr, method = "spearman"),
                        cor(tgr_s, true_tgr, method = "spearman")), each = 3)
    )
  }
  if (!is.null(fit_d) && !is.null(fit_s) && !is.null(fit_dea)) {
    mc_eff[[r]] <- rbind(
      eff_stats(fit_d,   sim_mc$data, "Deterministic SFA"),
      eff_stats(fit_s,   sim_mc$data, "Stochastic SFA"),
      eff_stats(fit_dea, sim_mc$data, "DEA (VRS)")
    )
  }
}

mc_df <- do.call(rbind, mc_results)
mc_df$bias <- mc_df$estimate - mc_df$true
mc_eff_df <- do.call(rbind, mc_eff)
```

## Parameter recovery {#sec:parameter-recovery}

Table \@ref(tab:mc-results) reports the bias, Monte Carlo standard error
(MC SE), standard deviation, and root mean squared error (RMSE) of the
metafrontier coefficient estimates across the 500 replications.

```{r mc-results, echo=FALSE}
n_successful <- length(unique(mc_df$rep))
mc_summary <- aggregate(
  cbind(bias, estimate) ~ method + parameter,
  data = mc_df,
  FUN = function(x) c(mean = mean(x), sd = sd(x))
)
mc_tab <- data.frame(
  Parameter = mc_summary$parameter,
  Method = mc_summary$method,
  True = beta_true[match(mc_summary$parameter,
                         c("Intercept", "beta_1", "beta_2"))],
  Bias = mc_summary$bias[, "mean"],
  `MC SE` = mc_summary$estimate[, "sd"] / sqrt(n_successful),
  SD = mc_summary$estimate[, "sd"],
  RMSE = sqrt(mc_summary$bias[, "mean"]^2 + mc_summary$estimate[, "sd"]^2),
  check.names = FALSE
)
mc_tab <- mc_tab[order(mc_tab$Parameter, mc_tab$Method), ]
knitr::kable(mc_tab, digits = 4, row.names = FALSE,
             caption = paste0("Monte Carlo results for metafrontier coefficient estimation (", n_successful, " successful replications out of 500). MC SE is the Monte Carlo standard error of the mean estimate; RMSE is the root mean squared error."),
             align = "llccccc",
             booktabs = TRUE, linesep = "") |>
  ktbl()
```

```{r mc-lpqp, cache=TRUE, echo=FALSE}
# LP vs QP objective: max abs TGR difference across Scenario A designs
lpqp_reps <- 50
lpqp_diff <- numeric(lpqp_reps)
for (r in seq_len(lpqp_reps)) {
  sim_lq <- simulate_metafrontier(
    n_groups = 2, n_per_group = 100, beta_meta = c(1.0, 0.5, 0.3),
    tech_gap = c(0, 0.4), sigma_u = c(0.2, 0.3), sigma_v = 0.15,
    seed = r * 1000
  )
  f_lp <- metafrontier(log_y ~ log_x1 + log_x2, data = sim_lq$data,
                       group = "group", objective = "lp")
  f_qp <- metafrontier(log_y ~ log_x1 + log_x2, data = sim_lq$data,
                       group = "group", objective = "qp")
  lpqp_diff[r] <- max(abs(f_lp$tgr - f_qp$tgr))
}
max_lpqp_diff <- max(lpqp_diff)
```

All 500 replications complete without error and with optimiser convergence
code zero at every stage, which is expected given the well-behaved
Cobb--Douglas DGP with moderate noise levels; convergence failures are more
likely with empirical data featuring collinear regressors or very small
samples. In applied work, the `check_convergence()` function and the
convergence block printed by `summary()` let users verify all estimation
stages before interpreting TGRs, confidence intervals, or productivity
decompositions. Figure \@ref(fig:mc-bias) displays the distribution of
coefficient bias across replications for both estimators.

Because the `objective` argument offers both identification criteria of
@battese2004, we also refit the deterministic metafrontier under the
squared-deviations QP objective on the first 50 Scenario A designs. The
maximum absolute difference between LP- and QP-based TGRs is
`r signif(max_lpqp_diff, 2)`: on this design the binding-constraint
pattern coincides and the two criteria are numerically indistinguishable,
although they can diverge at extreme input mixes.

```{r mc-bias, echo=FALSE, fig.width=7, fig.height=4.5, fig.cap="Distribution of metafrontier coefficient bias across 500 Monte Carlo replications for the deterministic and stochastic estimators (drawn with ggplot2; code in the supplementary materials). Boxes show the interquartile range and median; whiskers extend to 1.5 times the IQR.", fig.alt="A grouped boxplot showing the distribution of bias for three metafrontier coefficients (intercept, beta 1, beta 2) under both deterministic and stochastic estimation. Both methods show bias centred near zero, with the stochastic method showing slightly wider dispersion for the intercept."}
ggplot(mc_df, aes(x = parameter, y = bias, fill = method)) +
  geom_boxplot(position = position_dodge(0.8), alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  scale_fill_manual(values = c("Deterministic" = "#2166AC",
                                "Stochastic" = "#D6604D")) +
  labs(x = "Parameter", y = "Bias (estimate - true)",
       fill = "Metafrontier method") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

Both estimators recover the true slope coefficients ($\beta_1$, $\beta_2$)
with negligible bias. The stochastic metafrontier intercept shows a
persistent downward bias because the composed error $v^* - u^*$ has
non-zero mean $-\sigma^*\sqrt{2/\pi}$ under a half-normal specification,
so the second-stage SFA absorbs part of the intercept into the
inefficiency term. This is a structural feature of the two-step
stochastic metafrontier [@huang2014]: slope coefficients remain
consistent as $n \to \infty$, but the intercept does not. Applied users
interested in level-unit intercept interpretation should add back
$\hat\sigma^*\sqrt{2/\pi}$ or rely on the deterministic metafrontier,
whose intercept is unbiased in our Monte Carlo.

## Efficiency recovery {#sec:efficiency-recovery}

Table \@ref(tab:mc-eff) reports, for each of the three estimators under
Scenario A, the mean estimated technical efficiency against the own-group
frontier (TE), the mean metafrontier efficiency (TE$^*$), and the mean
technology gap ratio (TGR), together with the corresponding true means and
the Spearman rank correlations between estimated and true values.

```{r mc-eff, echo=FALSE}
eff_agg <- aggregate(cbind(te_true, te_est, rho_te,
                           tes_true, tes_est, rho_tes,
                           tgr_true, tgr_est, rho_tgr) ~ method,
                     data = mc_eff_df, FUN = mean)
eff_agg <- eff_agg[match(c("Deterministic SFA", "Stochastic SFA",
                           "DEA (VRS)"), eff_agg$method), ]
names(eff_agg) <- c("Estimator",
                    "TE (true)", "TE (est)", "Corr(TE)",
                    "TE* (true)", "TE* (est)", "Corr(TE*)",
                    "TGR (true)", "TGR (est)", "Corr(TGR)")
knitr::kable(eff_agg, digits = 3, row.names = FALSE,
             caption = "Efficiency recovery under Scenario A (means over 500 replications). TE is efficiency against the own-group frontier, TE* against the metafrontier, and TGR the technology gap ratio; Corr denotes the Spearman rank correlation between estimated and true values within a replication.",
             align = "lccccccccc",
             booktabs = TRUE, linesep = "") |>
  ktbl()
```

Two features of Table \@ref(tab:mc-eff) require care in interpretation.
First, the DEA rows should be read as robustness of the framework under a
noisy DGP: with $\sigma_v > 0$, observations with favourable noise draws
define the DEA frontiers, so DEA efficiency levels are contaminated by
construction and only their rankings are informative. Second, under the
intercept-only Scenario A the true TGR takes only two distinct values (one
per group), so the TGR rank correlation mostly reflects the between-group
ordering; Scenario B below provides within-group TGR variation and hence a
sharper test.

```{r mc-scenario-b, cache=TRUE, echo=FALSE}
set.seed(321)
beta_G <- rbind(G1 = c(1.0, 0.55, 0.25),
                G2 = c(0.9, 0.40, 0.40),
                G3 = c(0.8, 0.30, 0.55))
input_mu <- rbind(G1 = c(3.5, 2.0),
                  G2 = c(2.5, 2.5),
                  G3 = c(2.0, 3.5))
input_R <- matrix(c(1, 0.6, 0.6, 1), 2, 2)

mc_eff_b <- vector("list", n_reps)
n_b_fail <- 0L
for (r in seq_len(n_reps)) {
  sim_b <- simulate_metafrontier(
    n_groups = 3, n_per_group = 100,
    beta_groups = beta_G,
    input_means = input_mu,
    input_corr = input_R,
    sigma_u = c(0.2, 0.25, 0.3),
    sigma_v = 0.15,
    seed = r * 3000
  )
  fit_bd <- tryCatch(
    metafrontier(log_y ~ log_x1 + log_x2, data = sim_b$data,
                 group = "group", method = "sfa",
                 meta_type = "deterministic"),
    error = function(e) NULL
  )
  fit_bs <- tryCatch(
    metafrontier(log_y ~ log_x1 + log_x2, data = sim_b$data,
                 group = "group", method = "sfa",
                 meta_type = "stochastic"),
    error = function(e) NULL
  )
  fit_bdea <- tryCatch(
    metafrontier(log_y ~ log_x1 + log_x2, data = sim_b$data,
                 group = "group", method = "dea",
                 orientation = "output", rts = "vrs"),
    error = function(e) NULL
  )
  if (!is.null(fit_bd) && !is.null(fit_bs) && !is.null(fit_bdea)) {
    mc_eff_b[[r]] <- rbind(
      eff_stats(fit_bd,   sim_b$data, "Deterministic SFA"),
      eff_stats(fit_bs,   sim_b$data, "Stochastic SFA"),
      eff_stats(fit_bdea, sim_b$data, "DEA (VRS)")
    )
  } else {
    n_b_fail <- n_b_fail + 1L
  }
}
mc_eff_b_df <- do.call(rbind, mc_eff_b)
n_b_success <- sum(!vapply(mc_eff_b, is.null, logical(1)))
```

Table \@ref(tab:mc-eff-b) reports the same statistics under Scenario B
(three groups with group-specific slopes, correlated inputs, and
group-specific input distributions), where the true TGR varies across
observations within every group. Of the 500 replications,
`r n_b_success` complete successfully for all three estimators.

```{r mc-eff-b, echo=FALSE}
eff_agg_b <- aggregate(cbind(te_true, te_est, rho_te,
                             tes_true, tes_est, rho_tes,
                             tgr_true, tgr_est, rho_tgr) ~ method,
                       data = mc_eff_b_df, FUN = mean)
eff_agg_b <- eff_agg_b[match(c("Deterministic SFA", "Stochastic SFA",
                               "DEA (VRS)"), eff_agg_b$method), ]
names(eff_agg_b) <- c("Estimator",
                      "TE (true)", "TE (est)", "Corr(TE)",
                      "TE* (true)", "TE* (est)", "Corr(TE*)",
                      "TGR (true)", "TGR (est)", "Corr(TGR)")
knitr::kable(eff_agg_b, digits = 3, row.names = FALSE,
             caption = "Efficiency recovery under Scenario B: three groups with group-specific slopes, correlated log inputs, and group-specific input means (means over successful replications). The true TGR is computed against the pointwise maximum over the group frontiers; column definitions follow the Scenario A table.",
             align = "lccccccccc",
             booktabs = TRUE, linesep = "") |>
  ktbl()
```

Under Scenario B the estimated metafrontier coefficients converge to a
pseudo-true tightest linear envelope rather than to a data-generating
parameter vector, so no coefficient-bias table is reported. The efficiency
rows in Table \@ref(tab:mc-eff-b) show what this implies in practice. A
single log-linear envelope cannot bend around three crossing group
frontiers, so it lies above the pointwise-maximum envelope over much of
the input space, and the deterministic TGR levels are understated by
construction (mean 0.608 against a true mean of 0.925); the residual-based
stochastic TGR centres near one regardless of the truth (mean 1.010).
Rankings survive in both cases, with TGR rank correlations between 0.76
and 0.79. DEA imposes no functional form and recovers the TGR levels
almost without bias (mean 0.902) at similar rank correlations. This
contrast is the practical content of Scenario B: when group frontiers
differ in slope and not merely in level, the DEA-based metafrontier is the
safer choice for TGR levels, while the SFA-based variants still order
firms and groups reliably.

## Small-sample and unbalanced-group robustness {#sec:robustness}

To assess performance under more challenging conditions, we repeat the
Scenario A simulation with 30 firms per group (vs.\ 100 in the baseline),
keeping all other DGP parameters unchanged.

```{r mc-small-n, cache=TRUE, echo=FALSE}
set.seed(456)
n_reps_small <- 500
mc_small <- vector("list", n_reps_small)

for (r in seq_len(n_reps_small)) {
  sim_s <- simulate_metafrontier(
    n_groups = 2, n_per_group = 30,
    beta_meta = c(1.0, 0.5, 0.3),
    tech_gap = c(0, 0.4),
    sigma_u = c(0.2, 0.3),
    sigma_v = 0.15,
    seed = r * 2000
  )
  fit_ds <- tryCatch(
    metafrontier(log_y ~ log_x1 + log_x2, data = sim_s$data,
                 group = "group", method = "sfa",
                 meta_type = "deterministic"),
    error = function(e) NULL
  )
  if (!is.null(fit_ds)) {
    mc_small[[r]] <- list(
      det_coef = coef(fit_ds),
      det_tgr_cor = cor(efficiencies(fit_ds, type = "tgr"),
                        sim_s$data$true_tgr, method = "spearman")
    )
  }
}
mc_small <- mc_small[!sapply(mc_small, is.null)]
n_small_ok <- length(mc_small)
det_bias_small <- rowMeans(sapply(mc_small, function(x) x$det_coef)) - beta_true
det_tgr_small <- mean(sapply(mc_small, function(x) x$det_tgr_cor))
```

With only 30 observations per group, `r n_small_ok` of 500 replications
complete successfully (here and below, "successfully" means the estimation
finished without error; non-zero optimiser codes are recorded separately
and can be inspected with `check_convergence()`). The deterministic
estimator shows slope biases of
`r round(det_bias_small[2], 3)` ($\beta_1$) and
`r round(det_bias_small[3], 3)` ($\beta_2$), modestly larger than the
baseline. The TGR rank correlation remains high at
`r round(det_tgr_small, 3)`. The framework ranks group technologies
reliably even with small samples.

Empirical metafrontier applications frequently involve groups of markedly
different sizes, and the group frontiers are estimated separately, so a
small group's frontier (and hence its TGR) can be unstable even when the
total sample is large. To assess this, we simulate a strongly unbalanced
design with 200 firms in the frontier-defining group and only 20 firms in
the lagging group (a 10:1 ratio, with the technology gap on the small
group), keeping all other Scenario A parameters unchanged.

```{r mc-unbalanced, cache=TRUE, echo=FALSE}
set.seed(789)
n_reps_unb <- 500
mc_unb <- vector("list", n_reps_unb)

for (r in seq_len(n_reps_unb)) {
  sim_u <- simulate_metafrontier(
    n_groups = 2, n_per_group = c(200, 20),
    beta_meta = c(1.0, 0.5, 0.3),
    tech_gap = c(0, 0.4),
    sigma_u = c(0.2, 0.3),
    sigma_v = 0.15,
    seed = r * 4000
  )
  fit_u <- tryCatch(
    metafrontier(log_y ~ log_x1 + log_x2, data = sim_u$data,
                 group = "group", method = "sfa",
                 meta_type = "deterministic"),
    error = function(e) NULL
  )
  if (!is.null(fit_u)) {
    small_idx <- sim_u$data$group == "G2"
    tgr_u <- efficiencies(fit_u, type = "tgr")
    codes <- c(vapply(fit_u$group_models,
                      function(m) m$convergence, integer(1)),
               fit_u$meta_convergence)
    mc_unb[[r]] <- list(
      coef = coef(fit_u),
      all_zero = all(codes == 0L),
      tgr_cor = cor(tgr_u, sim_u$data$true_tgr, method = "spearman"),
      tgr_bias_small = mean(tgr_u[small_idx]) -
        mean(sim_u$data$true_tgr[small_idx])
    )
  }
}
mc_unb <- mc_unb[!sapply(mc_unb, is.null)]
n_unb_ok <- length(mc_unb)
n_unb_zero <- sum(sapply(mc_unb, function(x) x$all_zero))
unb_bias <- rowMeans(sapply(mc_unb, function(x) x$coef)) - c(1.0, 0.5, 0.3)
unb_tgr_cor <- mean(sapply(mc_unb, function(x) x$tgr_cor))
unb_tgr_bias_small <- mean(sapply(mc_unb, function(x) x$tgr_bias_small))
```

Of the 500 replications, `r n_unb_ok` complete without error and
`r n_unb_zero` additionally have optimiser convergence code zero at every
stage. The deterministic estimator's slope biases
(`r round(unb_bias[2], 3)` for $\beta_1$, `r round(unb_bias[3], 3)` for
$\beta_2$) remain modest, and the overall TGR rank correlation is
`r round(unb_tgr_cor, 3)`. The instability concentrates where expected: the
mean TGR bias in the 20-firm group is `r round(unb_tgr_bias_small, 3)`.
The small group's own frontier estimate drives this bias, since the pooled
metafrontier is estimated from all 220 observations. The large group shows no comparable bias.
Applied users with severely unbalanced groups should therefore interpret
the small groups' TGRs with caution and lean on the bootstrap intervals of
`boot_tgr()`, which reflect the group-specific sample sizes.

# Comparison with related packages {#sec:comparison}

Table \@ref(tab:comparison) provides a feature-level comparison of
\CRANpkg{metafrontier} with related \proglang{R} packages for efficiency
analysis. All compared packages are on CRAN; \CRANpkg{smfa} is the CRAN
release (April 2026) of the GitHub package \pkg{metafrontieR} cited in the
introduction. The table deliberately includes rows on which
\CRANpkg{metafrontier} offers nothing: the standard (single-technology)
Malmquist index, the smoothed DEA score bootstrap of Simar and Wilson, and
the sample selection metafrontier are all better served elsewhere.

```{r comparison, echo=FALSE}
comp <- data.frame(
  Feature = c("Deterministic metafrontier",
              "Stochastic metafrontier",
              "DEA-based metafrontier",
              "Latent class metafrontier",
              "Metafrontier Malmquist (O'Donnell et al. 2008)",
              "Standard Malmquist index",
              "Bootstrap TGR inference",
              "DEA score bootstrap (Simar--Wilson)",
              "Murphy--Topel correction",
              "Panel SFA (time-varying)",
              "Unbalanced panel SFA",
              "Sample selection MF",
              "Directional distance functions",
              "Hyperbolic (graph) efficiency",
              "FDH technology",
              "Slack analysis",
              "ggplot2 visualisation",
              "External model import",
              "S3 model methods"),
  metafrontier = c("LP + QP", "Yes", "Yes", "Yes", "Yes", "No",
                   "Yes", "No", "Yes", "Yes", "Yes", "No",
                   "Yes (b)", "Yes", "Yes", "Yes", "Yes",
                   "SFA fits", "38"),
  smfa = c("LP + QP", "Yes", "No", "Yes", "No", "No",
           "No", "No", "No", "No", "No", "Yes",
           "No", "No", "No", "No", "No", "sfaR only", "10"),
  deaR = c("No", "No", "Manual (a)", "No", "No", "Yes",
           "No", "Yes", "No", "No", "No", "No",
           "Yes", "No", "Yes", "Yes", "Yes", "No", "Yes"),
  Benchmarking = c("No", "No", "Manual (a)", "No", "No", "Yes",
                   "No", "Yes", "No", "No", "No", "No",
                   "Yes", "Yes", "Yes", "Yes", "No", "No", "Yes"),
  frontier = c("No", "No", "No", "No", "No", "No",
               "No", "No", "No", "Yes", "Yes", "No",
               "No", "No", "No", "No", "No", "N/A", "Yes"),
  sfaR = c("No", "No", "No", "No", "No", "No",
           "No", "No", "No", "No", "No", "No",
           "No", "No", "No", "No", "No", "N/A", "Yes"),
  sfa = c("No", "No", "No", "No", "No", "No",
          "No", "No", "No", "Yes", "Yes", "No",
          "No", "No", "No", "No", "No", "N/A", "Basic"),
  check.names = FALSE
)
knitr::kable(comp,
             caption = "Feature comparison across R packages for efficiency and metafrontier analysis. N/A means not applicable (the package is a model engine, not a metafrontier framework); notes (a) and (b) are explained in the text below.",
             align = "lccccccc",
             booktabs = TRUE, linesep = "") |>
  ktbl()
```

Two cells of Table \@ref(tab:comparison) carry notes. Note (a): pooled-reference
scoring can be assembled manually in \CRANpkg{deaR} and \CRANpkg{Benchmarking}
by scoring against a user-supplied reference set, but neither package forms,
summarises, or provides inference for technology gap ratios. Note (b): custom
direction vectors and firm-specific direction matrices are supported;
\CRANpkg{deaR} (`model_dir()`) and \CRANpkg{Benchmarking} (`DIRECT`) also
accept custom directions.

The \CRANpkg{metafrontier} package is, to our knowledge, the only \proglang{R}
package that combines SFA-based and DEA-based metafrontier analysis with
the metafrontier Malmquist decomposition and TGR-specific inference in a
single framework. Several complementary strengths of the other packages
deserve acknowledgement. The \CRANpkg{smfa} package [@smfa] offers both
deterministic criteria and two stochastic formulations, implements latent
class group frontiers via \CRANpkg{sfaR}'s `sfalcmcross()`, and uniquely
provides the sample selection metafrontier of @greene2010 for non-random
group composition. The \CRANpkg{deaR} package [@deaR] provides a broad DEA
toolkit (including fuzzy DEA and the Simar--Wilson bootstrap) that extends
well beyond metafrontier analysis. The \CRANpkg{sfaR} [@sfaR] and
\CRANpkg{frontier} [@frontier] packages remain leading engines for
stochastic frontier estimation, with \CRANpkg{sfaR} in particular
supporting ten inefficiency distributions against the three implemented
here; the newer \CRANpkg{sfa} package [@sfa] provides a recent alternative
with flexible panel specifications. The \CRANpkg{metafrontier} package's
interoperability and engine layers are designed to complement rather than
replace these estimators.

What does unification add beyond convenience? Two things. First, the three
estimators return identical S3 objects, so cross-method robustness checks,
as demonstrated in the real-data illustration of Section `r if(knitr::is_latex_output()) "\\@ref(sec:real-data)" else "[4.6](#sec:real-data)"`, require no
glue code; discrepancies between a deterministic envelope, a stochastic
second stage, and a nonparametric pooled frontier are themselves
diagnostic. Second, and more fundamentally, the technology gap ratio is a
*generated* quantity: it compares two estimated frontiers, so its sampling
uncertainty is not available from any single-frontier package. The
inference layer built around it (the TGR bootstrap of `boot_tgr()` and the
Murphy--Topel correction for the stochastic metafrontier's
generated-regressor problem) exists, to our knowledge, in no other
\proglang{R} package, and is the part of the workflow that cannot be
assembled from the engines in Table \@ref(tab:comparison).

# Summary and future directions {#sec:summary}

The \CRANpkg{metafrontier} package provides a unified \proglang{R} framework
for metafrontier production function analysis. Its principal contribution
is the integration of three metafrontier approaches (deterministic,
stochastic, and DEA) with the metafrontier Malmquist decomposition,
bootstrap inference, Murphy--Topel correction, and latent class estimation
under a consistent S3 interface. All simulations complete in under ten
minutes in the computational environment described in Section `r if(knitr::is_latex_output()) "\\@ref(sec:computational-details)" else "[3.4](#sec:computational-details)"`, and
Section `r if(knitr::is_latex_output()) "\\@ref(sec:basic-usage)" else "[4](#sec:basic-usage)"` demonstrates the package on real data from \CRANpkg{sfaR};
additional empirical examples are provided in the package vignette
(`vignette("metafrontier")`).

Three limitations should be noted. First, the SFA-based metafrontiers
require the researcher to commit to a parametric functional form (for
example Cobb--Douglas or translog) through the model formula, and the same
form is imposed on every group frontier and on the metafrontier envelope;
the package performs no specification search, so a misspecified form
propagates from the group frontiers into the estimated technology gap
ratios. The DEA-based estimators avoid this choice, at the cost of
ignoring statistical noise. Second, the package estimates production
frontiers only ($\varepsilon = v - u$); cost frontier support
($\varepsilon = v + u$) is planned. Third, in the SFA-based Malmquist
decomposition (`method = "sfa"` in `malmquist_meta()`), the distance of
each observation to the metafrontier is approximated by evaluating all
estimated group frontiers at that observation's inputs and taking the
largest predicted output, rather than by re-estimating the enveloping
metafrontier of Equation \@ref(eq:deterministic) in each period. This
pointwise maximum coincides with the metafrontier of @odonnell2008
wherever one group frontier dominates all others, but can understate it
where group frontiers cross, so the resulting TGC and TC$^*$ components
are approximate; the DEA-based Malmquist (the default) computes exact
distances to the pooled frontier and should be preferred when an exact
decomposition is required. The function documents this approximation and
announces it at run time. Planned extensions include a sample selection
metafrontier, spatial SFA metafrontiers, a Bayesian estimator propagating
first-stage uncertainty, and multiplicatively complete, transitive
(proper) TFP indexes such as the Färe--Primont and Lowe indexes
[@odonnell2012; @odonnell2018].

The package is available on CRAN
(\url{https://CRAN.R-project.org/package=metafrontier}) and is developed
on GitHub at \url{https://github.com/iik1/metafrontier}.

# Acknowledgements and declaration of tool use {-}

The author thanks the editor and three anonymous reviewers for comments
that substantially improved both the package and this article. Large
language model tools (Anthropic Claude) were used to assist with drafting
portions of the manuscript text and package documentation and with code
review. All methods, code, results, and the final wording were written
and verified by the author and remain the author's sole responsibility.
