% !TeX root = RJwrapper.tex
\title{Editorial}
\author{by Catherine Hurley}

\maketitle


On behalf of the editorial board, I am pleased to present Volume 14
Issue 1 of the R Journal. This issue heralds a switch from two issues
per year to four and is my first as Editor-in-Chief. The change to four
issues per year is in response to the increase in published articles in
recent years. As articles will appear more speedily in a published
issue, we will no longer list pdfs for Accepted articles on The R
Journal website.

First, some news about the journal board. Dianne Cook has stepped down
as Editor-in-Chief but continues as an Executive Editor. In her time as
EIC she provided excellent leadership and brought in many advances, most
notably the change to the new modern journal format. One new Associate
Editor, Simone Blomberg, has recently joined the team. We have a new,
slimmed-down Editorial advisory board consisting of Henrik Bengtssen,
Gabriela de Quiroz, Michael Kane and Rececca Killick. The board will
provide continuity across changes in the editorial board, offering
advice and acting as an independent body to handle issues of academic
integrity.

Behind the scenes, several people are assisting with the journal
operations and the new developments. Mitchell O'Hara-Wild continues to
work on infrastructure, and H. Sherry Zhang continues to develop the
\pkg{rjtools} package. In addition, articles in this issue have been
carefully copy edited by Hannah Comiskey.

\hypertarget{in-this-issue}{%
\section{In this issue}\label{in-this-issue}}

News from the CRAN and the R Foundation are included in this issue.

This issue features 22 contributed research articles the majority of
which relate to R packages for modelling tasks. All packages are
available on CRAN. Topics covered are:

\begin{itemize}
\tightlist
\item
  \textbf{Temporal and longitudinal methods}

  \begin{itemize}
  \tightlist
  \item
    \pkg{cpsurvsim}: An R Package for Simulating Data from Change-Point
    Hazard Distributions
  \item
    The \pkg{smoots} Package in R for Semiparametric Modeling of Trend
    Stationary Time Series
  \item
    \pkg{starvars}: An R Package for Analysing Nonlinearities in
    Multivariate Time Series
  \item
    \pkg{FMM}: An R package for modeling rhythmic patterns on
    oscillatory systems
  \item
    \pkg{tvReg}: Time-varying Coefficients in Multi-Equation Regression
    in R
  \item
    Power and Sample Size for Longitudinal Models in R - The
    \pkg{longpower} Package and Shiny App
  \end{itemize}
\item
  \textbf{Estimation and inference}

  \begin{itemize}
  \item
    \pkg{dglars} A Software Tool For Sparse Estimation Of A General
    Class Of High-dimensional GLMs
  \item
    \pkg{bayesanova}: An R package for Bayesian inference in the
    analysis of variance via Markov Chain Monte Carlo in Gaussian
    mixture models
  \item
    \pkg{RKHSMetaMod}: An R package to estimate the Hoeffding
    decomposition of a complex model by solving RKHS ridge group sparse
    optimization problem
  \item
    \pkg{PSweight}: An R Package for PropensityScore Weighting Analysis
  \end{itemize}
\item
  \textbf{Machine learning}

  \begin{itemize}
  \item
    \pkg{RFpredInterval}: An R Package for Prediction Intervals with
    Random Forests and Boosted Forests
  \item
    \pkg{fairmodels}: A Flexible Tool For Bias Detection, Visualization,
    And Mitigation Graphics and Visualisation, Machine Learning \&
    Statistical Learning
  \item
    \pkg{spherepc}: An R Package for Dimension Reduction on a Sphere
  \end{itemize}
\item
  \textbf{Other topics}

  \begin{itemize}
  \tightlist
  \item
    \pkg{blindrecalc}: An R Package for Blinded Sample Size
    Recalculation
  \item
    \pkg{rmonad}: Pipelines you can compute on
  \item
    \pkg{etrm}: Energy Trading and Risk Management in R
  \item
    \pkg{fcaR}, Formal Concept Analysis with R
  \item
    Advancing reproducible research by publishing R markdown notebooks
    as interactive sandboxes using the \texttt{learnr} package
  \end{itemize}
\item
  \textbf{Applications}

  \begin{itemize}
  \tightlist
  \item
    Palmer Archipelago Penguins Data in the \pkg{palmerpenguins} R
    Package - An Alternative to Anderson's Irises
  \item
    A Computational Analysis of the Dynamics of R Style Based on 94
    Million Lines of Code from All CRAN Packages in the Past 20 Years
  \item
    Measuring the Extent and Patterns of Urban Shrinkage for Small Towns
    Using R
  \item
    Revisiting Historical Bar Graphics on Epidemics in the Era of R
    ggplot2
  \end{itemize}
\end{itemize}


\address{%
Catherine Hurley\\
Maynooth University\\%
\\
%
\url{https://journal.r-project.org}\\%
%
\href{mailto:r-journal@r-project.org}{\nolinkurl{r-journal@r-project.org}}%
}
