% !TeX root = RJwrapper.tex
\title{SSNbayes: An R Package for Bayesian Spatio-Temporal Modelling on Stream Networks}


\author{by Edgar Santos-Fernandez, Jay M. Ver Hoef, James McGree, Daniel J. Isaak, Kerrie Mengersen, and Erin E. Peterson}

\maketitle

\abstract{%
Spatio-temporal models are widely used in many research areas from ecology to epidemiology. However, a limited number of computational tools are available for modeling river network datasets in space and time. In this paper, we introduce the \texttt{R} package \href{https://CRAN.R-project.org/package=SSNbayes}{SSNbayes} for fitting Bayesian spatio-temporal models and making predictions on branching stream networks. \href{https://CRAN.R-project.org/package=SSNbayes}{SSNbayes} provides a linear regression framework with multiple options for incorporating spatial and temporal autocorrelation. Spatial dependence is captured using stream distance and flow connectivity while temporal autocorrelation is modelled using vector autoregression approaches. \href{https://CRAN.R-project.org/package=SSNbayes}{SSNbayes} provides the functionality to make predictions across the whole network, compute exceedance probabilities, and other probabilistic estimates, such as the proportion of suitable habitat. We illustrate the functionality of the package using a stream temperature dataset collected in the Clearwater River Basin, USA.
}

\input{RJ-2023-061-src.tex}

\hypertarget{refs}{}

\bibliography{SSNbayes.bib}

\address{%
Edgar Santos-Fernandez\\
School of Mathematical Sciences, Queensland University of Technology\\%
Australian Research Council Centre of Excellence for Mathematical and
Statistical Frontiers (ACEMS)\\ Y Block, Floor 8, Gardens Point Campus. GPO Box 2434. Brisbane, QLD
4001.\\ Australia\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0001-5962-5417}{0000-0001-5962-5417}}\\%
\href{mailto:santosfe@qut.edu.au}{\nolinkurl{santosfe@qut.edu.au}}%
}

\address{%
Jay M. Ver Hoef\\
Marine Mammal Laboratory, NOAA-NMFS Alaska Fisheries Science Center.\\%
Seattle, WA and Fairbanks, AK, USA\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0003-4302-6895}{0000-0003-4302-6895}}\\%
\href{mailto:jay.verhoef@noaa.gov}{\nolinkurl{jay.verhoef@noaa.gov}}%
}

\address{%
James McGree\\
School of Mathematical Sciences, Queensland University of Technology\\%
\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0003-2997-8929}{0000-0003-2997-8929}}\\%
\href{mailto:james.mcgree@qut.edu.au}{\nolinkurl{james.mcgree@qut.edu.au}}%
}

\address{%
Daniel J. Isaak\\
Rocky Mountain Research Station, US Forest Service\\%
\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0002-8137-325X}{0000-0002-8137-325X}}\\%
\href{mailto:Daniel.isaak@usda.gov}{\nolinkurl{Daniel.isaak@usda.gov}}%
}

\address{%
Kerrie Mengersen\\
School of Mathematical Sciences, Queensland University of Technology\\%
Australian Research Council Centre of Excellence for Mathematical and
Statistical Frontiers (ACEMS)\\ Y Block, Floor 8, Gardens Point Campus. GPO Box 2434. Brisbane, QLD
4001.\\ Australia\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0001-8625-9168}{0000-0001-8625-9168}}\\%
\href{mailto:k.mengersen@qut.edu.au}{\nolinkurl{k.mengersen@qut.edu.au}}%
}

\address{%
Erin E. Peterson\\
EP Consulting\\%
School of Mathematical Sciences, Queensland University of Technology\\ Australian Research Council Centre of Excellence for Mathematical and
Statistical Frontiers (ACEMS)\\ Y Block, Floor 8, Gardens Point Campus. GPO Box 2434. Brisbane, QLD
4001.\\ Australia\\
%
%
\textit{ORCiD: \href{https://orcid.org/0000-0003-2992-0372}{0000-0003-2992-0372}}\\%
\href{mailto:erin@peterson-consulting.com}{\nolinkurl{erin@peterson-consulting.com}}%
}
