% !TeX root = RJwrapper.tex
\title{Kernel Heaping - Kernel Density Estimation from regional aggregates via measurement error model}


\author{by Lorena Gril, Laura Steinkemper, Marcus Groß, and Ulrich Rendtel}

\maketitle

\abstract{%
The phenomenon of ``aggregation'' often occurs in the regional
dissemination of information via choropleth maps. Choropleth maps
represent areas or regions that have been subdivided and color-coded
proportionally to ordinal or scaled quantitative data. By construction
discontinuities at the boundaries of rigid aggregation areas, often of
administrative origin, occur and inadequate choices of reference areas
can lead to errors, misinterpretations and difficulties in the
identification of local clusters. However, these representations do
not reflect the reality. Therefore, a smooth representation of
georeferenced data is a common goal. The use of naive non-parametric
kernel density estimators based on aggregates positioned at the
centroids of the areas result also in an inadequate representation of
reality. Therefore, an iterative method based on the Simulated
Expectation Maximization algorithm was implemented in the
Kernelheaping package. The proposed approach is based on a partly
Bayesian algorithm treating the true unknown geocoordinates as
additional parameters and results in a corrected kernel density
estimate.
}

\input{RJ-2024-026-src.tex}

\bibliography{KH.bib}

\address{%
Lorena Gril\\
Freie Universität Berlin, FB Wirtschaftswissenschaft\\%
Garystr. 21, D-14195 Berlin\\ Germany\\ \url{lorena.gril@fu-berlin.de}\\
%
%
%
%
}

\address{%
Laura Steinkemper\\
Freie Universität Berlin, FB Wirtschaftswissenschaft\\%
Garystr. 21, D-14195 Berlin\\ Germany\\ \url{steinkel98@zedat.fu-berlin.de}\\
%
%
%
%
}

\address{%
Marcus Groß\\
INWT Statistics GmbH\\%
Hauptstr.8, D-10827 Berlin\\ Germany\\ \url{Marcus.Gross@inwt-statistics.de}\\
%
%
%
%
}

\address{%
Ulrich Rendtel\\
Freie Universität Berlin, FB Wirtschaftswissenschaft\\%
Garystr. 21, D-14195 Berlin\\ Germany\\ \url{Ulrich.Rendtel@fu-berlin.de}\\
%
%
%
%
}
