---
title: 'Exploring Image Analysis in R: Applications and Advancements'
date: '2025-10-22'
abstract: This review offers an overview of image processing packages in R, covering
  applications such as multiplex imaging, cell tracking, and general-purpose tools.
  We found 38 R packages for image analysis, with adimpro and EBImage being the oldest,
  published in 2006, and biopixR among the newest, released in 2024. Of these packages,
  over 90 % are still active, with two-thirds receiving updates within the last 1.5
  years. The pivotal role of bioimage informatics in life sciences is emphasized in
  this review, along with the ongoing advancements of R's functionality through novel
  code releases. It focuses on complete analysis pipelines for extracting valuable
  information from biological images and includes real-world examples. Demonstrating
  how researchers can use R to tackle new scientific challenges in image analysis,
  the review provides a comprehensive understanding of R's utility in this field.
draft: no
author:
- name: Tim Brauckhoff
  affiliation: Brandenburg University of Technology Cottbus - Senftenberg
  address:
  - Institute of Biotechnology
  - Senftenberg, Germany
  - Eberhard Karls Universität Tübingen
  - Department of Computer Science
  - Tübingen, Germany
  orcid: 0009-0002-0142-7017
  email: brauctile@disroot.org
- name: Julius Rublack
  affiliation: University of Münster
  address:
  - Faculty of Biology
  - Münster, Germany
  email: juliusrublack@web.de
- name: Stefan Rödiger (corresponding author)
  affiliation: Brandenburg University of Technology Cottbus - Senftenberg
  address:
  - Institute of Biotechnology
  - Senftenberg, Germany
  email: stefan.roediger@b-tu.de
  orcid: 0000-0002-1441-6512
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
always_allow_html: yes
bibliography: RImageReview.bib
header-includes: \usepackage{float}
date_received: '2025-05-30'
volume: 17
issue: 3
slug: RJ-2025-030

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE, 
                      message = FALSE,
                      fig.align = 'center',
                      fig.pos = 'H',
                      fig.show = 'hold')
```

# Introduction

```{r, child="Introduction.Rmd"}

```

```{r, child="Applications.Rmd"}

```

```{r, child="Essentials.Rmd"}

```

```{r, child="Discussion.Rmd"}

```

# Conclusion

```{r, child="Conclusion.Rmd"}

```

# Funding

This review was partially funded by the project Rubin: NeuroMiR (03RU1U051A,
federal ministry of education and research, Germany).

# Conflict of interest

The authors declare no conflict of interest.

# Acknowledgements

We would like to express our gratitude to Dr. Coline Kieffer for providing the
microbead images used in this review. We thank Robert M Flight at codeberg.org
for reading and improving the manuscript.
