---
knit: "rmarkdown::render('detourr.rmd')"
---

# Radial CDF of hyperspheres projected to 3 dimensions {.appendix}

In order to implement the 3D variant of the sage tour (@laa2021burning), we need an expression for the relative projected volume of a __sphere__ of radius $R$. This is then used as a scaling factor for point radii in the visualisation to prevent points from being crowded towards the centre.

First we denote the volume of a $p$-dimensional hypersphere by:

$$
\frac{2\pi^{p/2}R^p}{p\Gamma(p/2)}
$$

In the appendix of @laa2020hole (equations 7--10) is a derivation for the relative projected volume of a ball of radius $r$.

\begin{align}
    F(r; p, R) &= \frac{V_{inside}(r; p, R)}{ V(p, r) } \\
    &= 1 - \frac{V_{outside}(r; p, R)}{ V(p, r) }
    (\#eq:F2)
\end{align}

And the formula for $V_{outside}(r; p, R)$ for a __circle__ is given as:

\begin{equation}
V_{outside}(r; p, R) =
\int_r^R V(p-2, \sqrt{R^2 - x^2})2\pi x dx 
(\#eq:v-outside)
\end{equation}

To extend this to the 3-dimensional case, we can modify \@ref(eq:v-outside) to express the  volume outside a __sphere__ of radius $r$ as:

\begin{equation}
  V_{outside}(r; p, R) = \int_r^R V(p-3, \sqrt{R^2 - x^2})4\pi x^2 dx
\end{equation}

and it follows that the relative projected volume for a sphere is

\begin{equation}
  F_3(r; p, R) = 1 - 
  \frac{
      \int_r^R V(p-3, \sqrt{R^2 - x^2})4\pi x^2 dx
  }{
      V(p, R)
  } 
\end{equation}

We know $2\Gamma(3/2) = \Gamma(1/2) = \sqrt{ \pi }$ so with some rearranging this can be reduced to:

\begin{equation}
  F_3(r; p, R) = 1 - \frac{2}{R^p} \frac{\Gamma(p/2+1)}{\Gamma(3/2)\Gamma((p-1)/2)}
  \int_r^R (R^2 - x^2)^{(p-3)/2} x^2 dx \\
\end{equation}

Denoting $u = 1 - \left(\frac{x}{R}\right)^2$ and $dx = \frac{R^2}{-2x}du = \frac{R}{-2\sqrt{1-u}}du$ for a change of variable this becomes

\begin{align}
  F_3(r; p, R) &= 1 - \frac{\Gamma(p/2+1)}{\Gamma(3/2)\Gamma((p-1)/2)}
  \int_0^{1-\frac{r^2}{R^2}} u^{(p-3)/2} (1-u)^{1/2} du \\
  &= 1 - \mathrm{BetaInc}\left(1-\left(\frac{r}{R}\right)^2, \frac{p-1}{2}, \frac{3}{2}\right) \\
  &= \mathrm{BetaInc}\left(\left(\frac{r}{R}\right)^2, \frac{3}{2}, \frac{p-1}{2}\right) 
  (\#eq:betainc1)
\end{align}

where _BetaInc_ is the regularised incomplete beta function (the CDF of a Beta distribution).

We can generalise this to any projection from $p$ to $d$ dimensions using the same steps, but with 

\begin{equation}
  V_{outside}(r; p, R, d) = \int_r^R V(p-d, \sqrt{R^2 - x^2}) 
  \frac{2\pi^{d/2}}{\Gamma(d/2)} x^{d-1} dx
\end{equation}

where $\frac{2\pi^{d/2}}{\Gamma(d/2)} x^{d-1}$ is the surface area of a d-ball.

This results in the relative projected volume of a projection from $p$ to $d$ dimensions being given by:

\begin{equation}
F(r; p, R, d) = \mathrm{BetaInc}\left(\left(\frac{r}{R}\right)^2, \frac{d}{2}, \frac{p-d}{2}+1\right)
(\#eq:betainc2)
\end{equation}

Figures \@ref(fig:radial-cdf-p) and \@ref(fig:radial-cdf-d) compare the theoretical results from equations \@ref(eq:betainc1) and \@ref(eq:betainc2) respectively with simulated values.

```{r}
library(MASS)

F <- function(r, R = 1, p, d = 3) {
  pbeta((r / R)^2, d / 2, (p - d + 2) / 2)
}

get_sim_data <- function(p, d, n = 5000) {
  r_unif_pball <- mvrnorm(n, rep(0, p), diag(rep(1, p)))

  # project points on to surface of unit ball
  r_unif_pball <- t(apply(r_unif_pball, 1, function(x) x / sqrt(sum(x^2))))

  # scale in to form a uniform distibution in the p-ball
  r_unif_pball <- r_unif_pball * runif(n)^(1 / p)
  # rand_unif_p_ball <- t(apply(rand_unif_p_ball, 1, function(x) x * (1 - runif(1)^p)))

  r <- (1:n) / n
  proj <- r_unif_pball[, 1:d]
  radii_actual <- sqrt(rowSums(proj^2))
  F_emperical <- ecdf(radii_actual)(r)

  tibble(
    r = r,
    p = p,
    d = d,
    F_emperical = F_emperical,
    F_theoretical = F(r, p = p, d = d)
  )
}
```

```{r, radial-cdf-p, fig.cap = "Relative projected volume for projections from p dimensions to d=3 dimensions. The solid line is simulated data, and the dashed line is the theoretical CDF", out.width="75%", fig.align="center"}
plot_data <- bind_rows(
  get_sim_data(p = 5, d = 3),
  get_sim_data(p = 10, d = 3),
  get_sim_data(p = 50, d = 3)
) |>
  mutate(p = as.character(p))
ggplot(plot_data, aes(x = r, group = p, colour = p)) +
  geom_line(aes(y = F_theoretical), lty = "dotdash") +
  geom_line(aes(y = F_emperical)) +
  theme_bw() +
  labs(
    title = "Radial projected CDF for a sphere",
    x = "r",
    y = "Relative projected volume"
  )
```

```{r, radial-cdf-d, fig.cap = "Relative projected volume for a projection of p=10 dimensions to d dimensions.  The solid line is simulated data, and the dashed line is the theoretical CDF. This shows the generalisation to d > 3 dimensions", out.width="75%", fig.align="center"}
plot_data <- bind_rows(
  get_sim_data(p = 10, d = 2),
  get_sim_data(p = 10, d = 3),
  get_sim_data(p = 10, d = 4),
  get_sim_data(p = 10, d = 8)
) |>
  mutate(d = as.character(d))
ggplot(plot_data, aes(x = r, group = d, colour = d)) +
  geom_line(aes(y = F_theoretical), lty = "dotdash") +
  geom_line(aes(y = F_emperical)) +
  theme_bw() +
  labs(
    title = "Radial projected CDF for a ball with dimension d (p=10)",
    x = "r",
    y = "Relative projected volume"
  )
```


