---
title: Testing Hypotheses Regarding Covariance and Correlation Matrices with the R
  Package CovCorTest
author:
- name: Paavo Sattler
  affiliation: TU Dortmund University, Department of Statistics
  address: 44221 Dortmund, Germany
  email: paavo.sattler@tu-dortmund.de
  orcid: 0000-0001-8731-0893
- name: Svenja Jedhoff
  affiliation: TU Dortmund University, Department of Statistics
  address: 44221 Dortmund, Germany
  email: svenja.jedhoff@tu-dortmund.de
  orcid: 0009-0008-2939-9103
date: '2026-12-31'
output:
  rjtools::rjournal_article:
    toc: no
    self_contained: yes
bibliography: References.bib
abstract: |
  In addition to the commonly analyzed measures of location, measures of  dispersion and dependence, such as variance and correlation, provide valuable  information. Consequently, they play a crucial role in multivariate statistics, which leads to tests regarding covariance and correlation matrices. Furthermore, the structure of these matrices also leads to important hypotheses of interest, as it contains substantial information about the underlying model. In fact, assumptions regarding the structure of covariance and correlation matrices are often fundamental in statistical modeling and testing. In this context, semi-parametric settings with minimal distributional assumptions and very general hypotheses are essential for enabling manifold usage. The freely available package CovCorTest provides suitable tests that address all issues mentioned above, using bootstrap and related techniques to achieve good performance, particularly in small samples. Additionally, the package offers flexible specification options for the hypotheses under investigation in two central tests, accommodating users with varying levels of expertise. This paper also illustrates the application of CovCorTest for various issues using multiple real-world examples.
editor_options:
  markdown:
    wrap: 72
date_received: '2026-02-07'
volume: 18
issue: 4
slug: RJ-2026-061
draft: no

---





# Introduction

Tests for measures of location are widely used in the analysis of multivariate
data, such as those arising from repeated-measures designs or multiple
measurements on the same subject. While expectation vectors are a classical
choice for such analyses [@konietschke2020], alternative parameters
such as quantiles [@baumeister2024], relative effects
[@rubarth2022; @dobler2020a] and others have been proposed. 
In addition to location parameters, measures of dispersion contain substantial
information and are therefore of central importance in multivariate settings.
In particular, covariance and correlation matrices play a key role as excellent
parameters for analyses for several reasons:

- They are comparatively easy to define and understand.

- They quantify the relationships between the components, which most other
  parameters do not take into account.
  
- In statistical modeling, covariance and correlation matrices and their
  structures (e.g. autoregressive or compound symmetry) are essential for
  understanding and describing the underlying processes.
  
- Conditions on the covariance or correlation matrix, such as homogeneity or a
  special structure/pattern, are among the most frequent assumptions of
  statistical methods.
  
- Finite second moments and, therefore, the existence of variances are usually
  presumed and sufficient to ensure consistency of estimators for the unknown
  covariance or correlation matrices.

The R package `CovCorTest` provides a unified and flexible framework for
testing hypotheses about covariance and correlation matrices in multivariate
data. In contrast to many existing implementations, it supports general linear
hypotheses, requires only minimal distributional assumptions and offers a
consistent user interface across a wide range of testing scenarios. The package
is designed for applied statisticians and data analysts working with
multivariate data, as well as for methodological researchers who require
flexible hypothesis specification beyond standard settings.

Due to their importance, a variety of statistical procedures have been
developed to test hypotheses concerning covariance matrices. Unfortunately,
many existing methods rely on restrictive distributional assumptions, such as
normality in @box1953 and @zhong2017, or specific conditions on
the characteristic function in @gupta2006, which limit their
applicability. Furthermore, several approaches are limited to specific
settings, for example, two-group settings as in @sakaori2002 or
@omelka2012. Moreover, they typically allow the investigation of only
one or two concrete hypotheses, such as the equality of covariance matrices
[e.g., @yang2012; @zhu2002] or equality of two correlation
matrices [e.g., @jennrich1970].

Recent methodological developments address several of these limitations by
formulating covariance and correlation hypotheses within a common
semi-parametric framework. In particular, the approaches introduced in
@sattler2022 and @sattler2023 allow for a broad class of
hypotheses while requiring only weak distributional assumptions and maintaining
good small-sample performance through resampling-based inference procedures.
These developments also include combined tests for the simultaneous assessment
of covariance and correlation matrices.

Structural hypotheses of covariance and correlation matrices are also of
practical relevance, as they provide meaningful insights into the underlying
model and frequently serve as model assumptions. Examples include compound
symmetry, Toeplitz, and autoregressive structures. However, existing tests for
such structures are often available only for selected cases, such as compound
symmetry [@winer1991; @votaw1948] or first-order
autoregressive structures [@mckeown1996], or focus exclusively on
either covariance or correlation matrices [@steiger1980]. Building on
the approaches proposed in @sattler2022 and @sattler2023,
these methods were extended in @sattler2024a to cover a wider range of
structural hypotheses on covariance and correlation matrices.

The methodological limitations of earlier approaches are naturally reflected in
existing software implementations. For instance, R packages such as
\CRANpkg{mvhtests}, \CRANpkg{biotools}, and \CRANpkg{heplots} provide classical
procedures for testing the equality of two or more covariance matrices,
including Box's $M$-test and likelihood-based approaches, typically relying on
multivariate normality. Other packages provide tools for more specific
covariance testing problems. For example, \pkg{covTestR}
[@Barnard2018] implements tests for high-dimensional covariance-matrix
problems, including tests for the homogeneity of two or more covariance matrices
and one-sample tests of whether a covariance matrix equals a specified target
matrix, such as the identity matrix. Further related functionality is available
in \CRANpkg{psych} [@Revelle2019], which provides selected tests and
diagnostics for correlation matrices, such as tests of whether a correlation
matrix is an identity matrix and tests for the equality of two correlation
matrices. Thus, existing packages provide useful procedures for specific
settings and selected hypotheses, but their scope is typically limited with
respect to the range of hypotheses, the considered data settings, or the
underlying assumptions.

The R package \CRANpkg{CovCorTest} [@covcortest] implements the above
recent methodological developments and makes them available within a single and
consistent software framework. It includes the hypotheses mentioned above, such
as the equality of covariance or correlation matrices and identity-type
correlation hypotheses, but also goes substantially beyond them. The package
supports general hypothesis formulations for covariance and correlation
matrices, structural hypotheses with unknown structure parameters, combined
covariance-correlation tests, and manually specified hypotheses through suitable
hypothesis matrices. Hence, \CRANpkg{CovCorTest} provides a coherent workflow for
procedures that are otherwise either unavailable in R or available only as
isolated special cases.


The package, which is freely available from the Comprehensive R Archive Network
(CRAN), provides five main functions—`test_covariance`,
`test_correlation`, `test_covariance_structure`,
`test_correlation_structure`, and `test_combined`. These
functions are designed to follow a uniform interface, facilitating a
consistent and user-friendly workflow across different testing scenarios. This
uniform design allows users to apply analogous procedures to hypotheses
targeting different aspects of the data and different data structures.
Together with the coherent methodological framework, this design enhances
usability and improves the comparability of results. Furthermore, to support users with varying levels of experience, the covariance
and correlation tests provide predefined hypotheses for ease of use, as well as
fully customizable hypotheses—specified via a hypothesis matrix and
corresponding vector—for advanced applications.

Because all underlying approaches and tests are based on empirical covariance and correlation
matrices, our functions should only be used for metric data, although further
assumptions, such as a repeated-measures design, are not required.

In the subsequent section, we will introduce the statistical multivariate
model, the investigated null hypotheses, and the underlying assumptions.
Afterwards, we present the implemented inference procedures, followed by the
applications from the R package `CovCorTest` illustrated by some
examples using a real-world dataset. Finally, a conclusion and outlook
complete the paper.

 
# Statistical model and hypotheses

## Data structure and model assumptions

`CovCorTest` is designed for multivariate data observed in one or more
independent groups. Therefore, we consider a general multigroup setting with
$a \in \mathbb{N}$ groups, where each group $i = 1, \dots, a$ contains
$n_i \in \mathbb{N}$ observations. The groups may be unbalanced and therefore
have different sample sizes, while the total number of observations is denoted
by $N = \sum_{i=1}^a n_i$.

Then, a general semi-parametric model is used, defined by independent
$d$-dimensional random vectors
\begin{equation}
\boldsymbol{X}_{ik}
= \boldsymbol{\mu}_i + \boldsymbol{\epsilon}_{ik},
(\#eq:model)
\end{equation}
where $\boldsymbol{\mu}_i \in \mathbb{R}^d$ represents the group-specific mean
vector, and $\boldsymbol{\epsilon}_{ik} \in \mathbb{R}^d$ denotes the random
deviation for subject $k = 1, \dots, n_i$ from the $i$-th group. Here, splitting
the indices allows considering factorial designs and longitudinal data, as
demonstrated, e.g., in @pauly2015. We assume the residuals
$\boldsymbol{\epsilon}_{i1}, \dots, \boldsymbol{\epsilon}_{in_i}$
to be centered,
$\mathbb{E}(\boldsymbol{\epsilon}_{i1}) = \mathbf{0}_d$, and i.i.d. within each group, while
across different groups, only independence is needed, but differing
distributions are possible. Further assumptions are finite fourth moments
($\mathbb{E}(\lVert \boldsymbol{\epsilon}_{i1} \rVert^4) < \infty$), existence of
group-specific covariance $\boldsymbol{V}_i$ and correlation matrices $\boldsymbol{R}_i$ and that
asymptotically no group vanishes
($\frac{n_i}{N} \longrightarrow\kappa_i \in (0,1]$ for $i = 1, \dots, a$). No parametric
distributional assumptions, such as multivariate normality, are
imposed. These weak assumptions ensure the broad applicability of the
implemented procedures and are sufficient for the consistent estimation of
covariance and correlation matrices. For correlation-based tests, the existence
of strictly positive variances is additionally required, as well as $d > 1$.

This framework allows `CovCorTest` to be applied to a wide range of
practical settings while maintaining robustness in small and moderate sample
sizes.

## Hypotheses on covariance and correlation matrices

For covariance matrices, the upper triangular part including the diagonal is
considered (denoted as $\operatorname{vech}$ and resulting in a $p$-dimensional vector), while
for correlation matrices only the off-diagonal elements are used (denoted as
$\operatorname{vech}^{-}$ and resulting in a $p_{-}$-dimensional vector). 

For illustration, consider the case $d=3$. For a covariance matrix and its corresponding correlation matrix
\begin{equation}
V_1 =
\begin{pmatrix}
v_{11} & v_{12} & v_{13}\\
v_{12} & v_{22} & v_{23}\\
v_{13} & v_{23} & v_{33}
\end{pmatrix},
\qquad
R_1 =
\begin{pmatrix}
1 & r_{12} & r_{13}\\
r_{12} & 1 & r_{23}\\
r_{13} & r_{23} & 1
\end{pmatrix},
(\#eq:matrix-example)
\end{equation}
the corresponding vectors are given by
\begin{equation}
\operatorname{vech}(V_1)
=
(v_{11}, v_{12}, v_{13}, v_{22}, v_{23}, v_{33})^\top,
\qquad
\operatorname{vech}^{-}(R_1)
=
(r_{12}, r_{13}, r_{23})^\top .
(\#eq:vech-example)
\end{equation}
Thus, $\operatorname{vech}(V_1)$ contains all unique entries of the covariance matrix, including the variances on the diagonal, whereas $\operatorname{vech}^{-}(R_1)$ contains only the off-diagonal correlations, since the diagonal entries of a correlation matrix are always equal to one and therefore contain no information for the hypothesis tests.


Now, based on the pooled covariance vector
$\boldsymbol{v}=\bigl(\operatorname{vech}(\boldsymbol{V}_1)^\top, \dots, \operatorname{vech}(\boldsymbol{V}_a)^\top\bigr)^\top$,
respectively, the pooled correlation vector
$\boldsymbol{r}=\bigl(\operatorname{vech}^{-}(\boldsymbol{R}_1)^\top, \dots, \operatorname{vech}^{-}(\boldsymbol{R}_a)^\top\bigr)^\top$,
general linear hypotheses can be formulated.

Using adequate hypothesis matrices $\boldsymbol{C}_{\boldsymbol{v}}$ and
$\boldsymbol{C}_{\boldsymbol{r}}$, respectively, and corresponding vectors
$\boldsymbol{\zeta}_{\boldsymbol{v}}$ and $\boldsymbol{\zeta}_{\boldsymbol{r}}$, respectively,
they are given through
\begin{equation}
\mathcal{H}_0^{\boldsymbol{v}} :
\boldsymbol{C}_{\boldsymbol{v}} \boldsymbol{v} = \boldsymbol{\zeta}_{\boldsymbol{v}}
\quad \text{and} \quad
\mathcal{H}_0^{\boldsymbol{r}} :
\boldsymbol{C}_{\boldsymbol{r}} \boldsymbol{r} = \boldsymbol{\zeta}_{\boldsymbol{r}} .
(\#eq:Hypothesen)
\end{equation}

Popular hypotheses included in this model, among many others, are

- Equal covariance matrices:
  $\boldsymbol{V}_1 = \dots = \boldsymbol{V}_a$
  
- Equal traces of covariance matrices:
  $\operatorname{tr}(\boldsymbol{V}_1) = \dots = \operatorname{tr}(\boldsymbol{V}_a)$
  
- Equal variances:
  $(\boldsymbol{V}_1)_{11} = \dots = (\boldsymbol{V}_1)_{dd}$
  
- Equal correlation matrices:
  $\boldsymbol{R}_1 = \dots = \boldsymbol{R}_a$
  
- Equal correlated components:
  $\operatorname{vech}^{-}(\boldsymbol{R}_1)_1 = \dots = \operatorname{vech}^{-}(\boldsymbol{R}_1)_{p_-}$
  
- Uncorrelated components:
  $\operatorname{vech}^{-}(\boldsymbol{R}_1) = \mathbf{0}_{p_-}$
  
- Trace of the covariance matrix has a special value
  $\gamma \in \mathbb{R}_+$:
  $\operatorname{tr}(\boldsymbol{V}_1) = \gamma$
  
- Specific covariance matrix
  $\boldsymbol{V} \in \mathbb{R}^{d \times d}$:
  $\boldsymbol{V}_1 = \boldsymbol{V}$




These kinds of hypotheses arise both as research questions and as modeling assumptions in applied multivariate analyses. For instance, homogeneity of covariance matrices is a classical assumption in MANOVA-type analyses and discriminant analysis; see, for example, @FriendlySigal2020. Hypotheses on correlation matrices are relevant when comparing dependence structures between populations or experimental conditions, for example, in gene expression studies comparing healthy and tumor tissue; see @CaballeEtAl2018. Moreover, hypotheses on equal variances or equal traces can be used to compare marginal or total multivariate variability across groups or populations; see @LeMaitreMitteroecker2019. Finally, hypotheses of uncorrelated components are closely related to assessing whether the correlation matrix differs from the identity matrix, a common diagnostic step before factor analysis; see @Bartlett1951 and @Revelle2019.



For ease of use, `CovCorTest` provides a set of predefined hypotheses that
can be selected via simple character arguments. At the same time, advanced
users may specify custom hypotheses directly through hypothesis matrices,
enabling highly flexible and non-standard testing scenarios. For example, using the centering matrix
$\boldsymbol{P}_a = \boldsymbol{I}_a - \mathbf{1}_{a \times a} / a$ and the Kronecker product, hypotheses
such as equality of covariance or correlation matrices can be expressed in terms
of linear constraints on the vectorized parameters, e.g.,
$(\boldsymbol{P}_a \otimes \boldsymbol{I}_p)\boldsymbol{v} = \mathbf{0}_{ap},$
and similarly for other hypotheses.
 

Note that the same hypothesis can often be expressed in multiple ways through
different choices of $\boldsymbol{C}$ and $\boldsymbol{\zeta}$
[see @sattler2023a; @sattler2025 for a detailed discussion]. In the case of
$\boldsymbol{\zeta} = \mathbf{0}$, unique projection matrices exist to check the underlying
hypotheses, while in general, the considered hypothesis matrices do not have to
be square, idempotent or symmetrical.

## Structural hypotheses

In many applications, assumptions on the *structure* of covariance or
correlation matrices are of primary interest. For covariance matrices, common
structures include

- Compound symmetry

- Diagonality

- First-order autoregressive (AR(1))

- Sphericity

- Toeplitz structure,

see @kincaid2005 for more details regarding these structures.
For correlation matrices, analogous—but typically fewer—structures exist, which
often arise as heterogeneous versions of other structures such as heterogeneous
autoregressive, heterogeneous compound symmetry, or heterogeneous Toeplitz
structures. `CovCorTest` supports a broad class of such structural
hypotheses by expressing them as linear constraints on vectorized covariance or
correlation matrices, including both homogeneous and heterogeneous versions of
common structures. Except for autoregressive structures and their variants, for
which a transformation is necessary, these covariance and correlation
structures can be expressed within the above general hypothesis framework using
appropriate choices of $\boldsymbol{C}$ and $\boldsymbol{\zeta}$, as shown in
@sattler2024a. The package additionally implements banded and banded
Toeplitz structures. For banded structures, all entries outside a specified
bandwidth $q$ are constrained to be zero, so only the diagonal and the first
$q$ off-diagonals may contain non-zero entries. Depending on the selected
structure, the entries within this band may either remain unrestricted or be
further constrained, for example by a Toeplitz pattern.

All supported structures can be tested using dedicated functions with a unified
interface, allowing users to assess structural assumptions in a transparent and
reproducible manner. Also, structures from a linear covariance structure model,
comparable to @anderson1973 or @szatrowski1980, can be
expressed through a hypothesis matrix and vector. Although these expressions are
not unique, the package includes a function `get_hypothesis`, which, for
given vectors specifying the model, returns a possible hypothesis matrix $\boldsymbol{C}$
and corresponding vector $\boldsymbol{\zeta}$.
This highlights the versatility and generality of the presented model in also capturing a wide range of relevant
covariance and correlation structures.

# Inference

To evaluate the hypotheses formulated in the previous section, we develop
inference procedures based on the asymptotic distributions of vectorized
covariance and correlation estimators. For our hypotheses regarding covariance
matrices, we consider the vectorized empirical group covariances
$\widehat{\boldsymbol{v}}_i$ as the counterpart of $\boldsymbol{v}_i$. Stacking
these vectors across all $a$ groups yields the pooled vector
$\widehat{\boldsymbol{v}}=\bigl(
\widehat{\boldsymbol{v}}_1^\top, \dots,
\widehat{\boldsymbol{v}}_a^\top
\bigr)^\top.$ Then, for developing a statistical test, we use the fact that under our
assumptions of Section 2 and the null hypothesis
$\mathcal{H}_0^{\boldsymbol{v}}$, it holds that
\begin{equation}
\sqrt{N}
\bigl(
\boldsymbol{C}_{\boldsymbol{v}} \widehat{\boldsymbol{v}}
- \boldsymbol{\zeta}_{\boldsymbol{v}}
\bigr)
\stackrel{\mathcal D}{\longrightarrow}
\mathcal{N}
\bigl(
\mathbf{0},
\boldsymbol{C}_{\boldsymbol{v}} \boldsymbol{\Sigma} \boldsymbol{C}_{\boldsymbol{v}}^\top
\bigr),
(\#eq:cov-limit)
\end{equation}
with the block-diagonal covariance matrix $\boldsymbol{\Sigma}$.

In the same way, $\widehat{\boldsymbol{r}}_i$ denotes the empirical counterpart
of the upper-half-vectorized correlation matrix of group $i$ and
$\widehat{\boldsymbol{r}}$ is the corresponding pooled vector. Following the
results in @sattler2023 (where the exact form can also be found), there
exists a continuous matrix-valued function
$\boldsymbol{M} \in \mathbb{R}^{p_- \times p}$ such that
\begin{equation}
\sqrt{n_i}
\bigl(
\widehat{\boldsymbol{r}}_i - \boldsymbol{r}_i
\bigr)
=
\boldsymbol{M}(\boldsymbol{v}_i, \boldsymbol{r}_i)
\cdot
\sqrt{n_i}
\bigl(
\widehat{\boldsymbol{v}}_i - \boldsymbol{v}_i
\bigr)
+ {\scriptstyle \mathcal{O}_P(1)}.
(\#eq:relation)
\end{equation}

Using this result and the previous asymptotic expansion, we obtain the joint
asymptotic distribution under $\mathcal{H}_0^{\boldsymbol{r}}$ as
\begin{equation}
\sqrt{N}
\bigl(
\boldsymbol{C}_{\boldsymbol{r}} \widehat{\boldsymbol{r}}
- \boldsymbol{\zeta}_{\boldsymbol{r}}
\bigr)
\stackrel{\mathcal D}{\longrightarrow}
\mathcal{N}
\bigl(
\mathbf{0},
\boldsymbol{C}_{\boldsymbol{r}} \boldsymbol{\Upsilon} \boldsymbol{C}_{\boldsymbol{r}}^\top
\bigr),
(\#eq:cor-limit)
\end{equation}
using a block-diagonal matrix based on the transformation matrix
$\boldsymbol{M}(\boldsymbol{v}_i, \boldsymbol{r}_i)$ and $\boldsymbol{\Sigma}$.

Based on these limit distributions, to test our multivariate hypotheses, we use
the ATS (ANOVA-type statistic), initially developed in
@brunner1997, which in the present setting are given by

\begin{equation}
ATS_{\boldsymbol{v}}
=
N
\bigl[
\boldsymbol{C}_{\boldsymbol{v}}
\widehat{\boldsymbol{v}} - \boldsymbol{\zeta}_{\boldsymbol{v}}
\bigr]^\top
\bigl[
\boldsymbol{C}_{\boldsymbol{v}}
\widehat{\boldsymbol{v}} - \boldsymbol{\zeta}_{\boldsymbol{v}}
\bigr]
\Big/
\operatorname{tr}
\bigl(
\boldsymbol{C}_{\boldsymbol{v}} \widehat{\boldsymbol{\Sigma}} \boldsymbol{C}_{\boldsymbol{v}}^\top
\bigr),
(\#eq:ATS1)
\end{equation}

and

\begin{equation}
ATS_{\boldsymbol{r}}
=
N
\bigl[
\boldsymbol{C}_{\boldsymbol{r}}
\widehat{\boldsymbol{r}} - \boldsymbol{\zeta}_{\boldsymbol{r}}
\bigr]^\top
\bigl[
\boldsymbol{C}_{\boldsymbol{r}}
\widehat{\boldsymbol{r}} - \boldsymbol{\zeta}_{\boldsymbol{r}}
\bigr]
\Big/
\operatorname{tr}
\bigl(
\boldsymbol{C}_{\boldsymbol{r}} \widehat{\boldsymbol{\Upsilon}} \boldsymbol{C}_{\boldsymbol{r}}^\top
\bigr).
(\#eq:ATS2)
\end{equation}

Here, for the unknown covariance, we use the estimators
$\widehat{\boldsymbol{\Sigma}}$ and $\widehat{\boldsymbol{\Upsilon}}$, respectively, built from the
groupwise estimators and a plug-in version of $\boldsymbol{M}$ (see
@sattler2022 and @sattler2023 for the explicit formulas).
This quadratic form is essentially the norm of the asymptotically normally
distributed vector. A high value, therefore, leads to rejection, while
standardization enhances the statistical stability of the results and ensures
invariance under scalar multiplications, such as unit changes.

There also exist alternative versions of the ATS, which, for example, allow the
approximation through an $F$-distribution (see, e.g.,
@brunner2019 or @thiel2024). However, for the statistics
defined in \@ref(eq:ATS1) and \@ref(eq:ATS2), the limiting null distribution can be characterized as a weighted sum of independent chi-square distributed random variables. Although this distribution is generally not available in closed form, this characterization enables valid inference without stringent distributional assumptions.

For both ATS, the asymptotic distribution under the respective null hypothesis
is a weighted $\chi^2$ distribution, i.e.,
\begin{equation}
ATS
\stackrel{\mathcal D}{\longrightarrow}
\sum_{\ell=1}^{m} \lambda_\ell B_\ell,
\qquad
B_\ell \stackrel{i.i.d.}{\sim} \chi_1^2,
(\#eq:weighted-limit)
\end{equation}
where $m$ denotes the number of rows of the corresponding hypothesis matrix
$\boldsymbol{C}$. The weights $\lambda_\ell$ are given by the eigenvalues of the matrices
$\boldsymbol{C}_{\boldsymbol{v}} \boldsymbol{\Sigma} \boldsymbol{C}_{\boldsymbol{v}}^\top$ and
$\boldsymbol{C}_{\boldsymbol{r}} \boldsymbol{\Upsilon} \boldsymbol{C}_{\boldsymbol{r}}^\top$,
respectively, divided by the trace of the corresponding matrix, and
therefore reflect the covariance structure of the underlying estimators.


Since no explicit formula is available for this weighted distribution, several
approaches can be used to determine the required critical values. One such
approach is a Monte-Carlo procedure, which we describe first. For the covariance and correlation statistics, the consistent estimators
$\widehat{\boldsymbol{\Sigma}}$ and $\widehat{\boldsymbol{\Upsilon}}$,
respectively, yield estimators $\hat{\lambda}_1,\dots,\hat{\lambda}_m$
for the unknown normalized eigenvalues. With
these weights, we generate appropriate realizations
$B_1, \dots, B_m \stackrel{i.i.d.}{\sim} \chi_1^2$. These realizations and our
estimators allow calculating a weighted sum
\begin{equation}
\sum_{\ell=1}^{m} \hat{\lambda}_\ell B_\ell,
(\#eq:mc-weighted-sum)
\end{equation}
which converges in distribution to the above limit distribution. Repeating
this procedure many times allows calculating the required Monte-Carlo
quantiles, and with an analogous procedure for the correlation for
$\alpha \in (0,1)$, we get asymptotic level-$\alpha$ tests through
\begin{equation}
\begin{array}{lrl}
\varphi_{\boldsymbol{v}}^{MC}
&=&
1\hspace{-0.7ex}1
\bigl(
ATS_{\boldsymbol{v}}
>
q_{\boldsymbol{v}, 1-\alpha}^{MC}
\bigr),
\\[0.5em]
\varphi_{\boldsymbol{r}}^{MC}
&=&
1\hspace{-0.7ex}1
\bigl(
ATS_{\boldsymbol{r}}
>
q_{\boldsymbol{r}, 1-\alpha}^{MC}
\bigr).
\end{array}
(\#eq:mc-tests)
\end{equation}

Since the covariance estimators and the corresponding estimated eigenvalues
have to be calculated only once and all involved random variables are
univariate, this procedure is computationally efficient even for large numbers
of repetitions or large values of $m$.

A further technique for obtaining quantiles that requires more computational
effort but usually yields better small-sample performance is a bootstrap
approach. For hypotheses regarding covariance matrices, the parametric
bootstrap mimics the asymptotic distribution of the test statistic by
generating bootstrap vectors
$\boldsymbol{Z}_{i1}^\star, \dots, \boldsymbol{Z}_{in_i}^\star
\stackrel{i.i.d.}{\sim}
\mathcal{N}(\mathbf{0}, \widehat{\boldsymbol{\Sigma}}_i)$
for each group, using the covariance estimators
$\widehat{\boldsymbol{\Sigma}}_i$ obtained from the data. For each group, the empirical covariance matrix of the bootstrap sample is
calculated and therewith $\widehat{\boldsymbol{\Sigma}}^\star$. The bootstrap analogue of
the ATS is given through $ATS_{\boldsymbol{v}}^\star=
N
\bigl[
\boldsymbol{C}_{\boldsymbol{v}} \overline{\boldsymbol{Z}}^\star
\bigr]^\top
\bigl[
\boldsymbol{C}_{\boldsymbol{v}} \overline{\boldsymbol{Z}}^\star
\bigr]
\Big/
\operatorname{tr}
\bigl(
\boldsymbol{C}_{\boldsymbol{v}}
\widehat{\boldsymbol{\Sigma}}^\star
\boldsymbol{C}_{\boldsymbol{v}}^\top
\bigr),$ using the bootstrap means
$\overline{\boldsymbol{Z}}_i^\star
=
n_i^{-1}
\sum_{k=1}^{n_i} \boldsymbol{Z}_{ik}^\star$
and
$\overline{\boldsymbol{Z}}^\star
=
\bigl(
{\overline{\boldsymbol{Z}}_1^\star}^\top, \dots,
{\overline{\boldsymbol{Z}}_a^\star}^\top
\bigr)^\top$. Now, based on a large number of repetitions, the bootstrap quantile
$q_{\boldsymbol{v}, 1-\alpha}^{\star}$ is determined.

In a similar way, by replacing $\widehat{\boldsymbol{\Sigma}}$ and $\widehat{\boldsymbol{\Sigma}}_i^\star$ with
$\widehat{\boldsymbol{\Upsilon}}$ and $\widehat{\boldsymbol{\Upsilon}}_i^\star$, the bootstrap quantile
for the correlation
$q_{\boldsymbol{r}, 1-\alpha}^{\star}$ can be determined. This allows
formulating asymptotic bootstrap tests for the respective hypothesis and
$\alpha \in (0,1)$ by
\begin{equation}
\begin{array}{lrl}
\varphi_{\boldsymbol{v}}^{\star}
&=&
1\hspace{-0.7ex}1
\bigl(
ATS_{\boldsymbol{v}}
>
q_{\boldsymbol{v}, 1-\alpha}^{\star}
\bigr),
\\ [0.5em]
\varphi_{\boldsymbol{r}}^{\star}
&=&
1\hspace{-0.7ex}1
\bigl(
ATS_{\boldsymbol{r}}
>
q_{\boldsymbol{r}, 1-\alpha}^{\star}
\bigr).
\end{array}
(\#eq:bootstrap-tests)
\end{equation}

Since for each bootstrap run, all observations have to be generated and all
empirical covariance matrices have to be calculated, this approach can be
time-consuming, especially for higher dimensions. However, it usually leads to 
improved finite-sample performance, particularly for small sample sizes; see 
the simulation studies in @sattler2022 and @sattler2023.

For the investigation of hypotheses regarding correlations, an additional
approach exists to obtain the required quantiles. It is called a Taylor-based
Monte-Carlo technique and combines elements of Monte-Carlo and bootstrap, using
a refined version of the expansion in equation \@ref(eq:relation) to simulate the
asymptotic distribution directly. Further details are provided in
@sattler2023. Since for this approach, only a single observation per
group is generated, and standardization is based on the trace term in
\@ref(eq:ATS2), this results in a computationally efficient procedure with good
finite-sample performance.

In `CovCorTest`, a common interface is used, where all three inference
strategies—Monte-Carlo, parametric bootstrap, and Taylor-based Monte-Carlo—can
be chosen. This design allows users to explicitly select the inference method
depending on computational constraints, sample size considerations, or desired
accuracy, while keeping the specification of hypotheses unchanged.

**Remark.**
Despite their high flexibility, the linear hypotheses introduced in
\@ref(eq:Hypothesen) do not allow the formulation of all relevant structures of
covariance and correlation matrices. In particular, the autoregressive
structure and all of its versions include some proportionality between the
matrix components that cannot be expressed through linear constraints on the
vectorized covariance or correlation matrices. To accommodate such nonlinear structures, suitable transformations of the
parameter vectors are considered. Specifically, let $\boldsymbol{f}$ denote a
differentiable transformation of the vectorized covariance parameters and
$\boldsymbol{g}$ the corresponding transformation for correlation parameters. This leads
to hypotheses of the form
$\mathcal{H}_0^{\boldsymbol{v},s}
:
\widetilde{\boldsymbol{C}}_{\boldsymbol{v}} \boldsymbol{f}(\boldsymbol{v})
=
\widetilde{\boldsymbol{\zeta}}_{\boldsymbol{v}}$
or
$\mathcal{H}_0^{\boldsymbol{r},s}
:
\widetilde{\boldsymbol{C}}_{\boldsymbol{r}} \boldsymbol{g}(\boldsymbol{r})
=
\widetilde{\boldsymbol{\zeta}}_{\boldsymbol{r}}$. By replacing the original parameter vectors with their transformed empirical
counterparts and incorporating the corresponding Jacobian matrix, the
above-mentioned tests can be adapted accordingly. A detailed discussion of
this approach, along with a useful transformation, referred to as the
subdiagonal-mean-ratio, is provided in @sattler2024a.


 
 Although these tests are highly versatile due to
 weak distributional assumptions and a broad class of admissible hypotheses, they do not by themselves provide insight into the specific reasons for rejecting a null hypothesis. In particular, a rejection does not directly reveal whether group differences are driven by changes in marginal variances, in dependence structures, or in both components. Since correlation matrices reflect dependency structures but contain no information about the variances of individual components, whereas covariance matrices include variances but lack a standardized measure of dependence, neither parameter is sufficient on its own.
 
 To address this limitation, @sattler2023 propose a combined test for simultaneously assessing group differences in both covariance and correlation structures.
 This approach is based on the concept of multiple contrast testing [see, e.g., @bretz2001] and is applicable to two-group comparisons ($a=2$). In addition to testing the null hypothesis of equal covariance matrices, the procedure allows identifying whether detected differences are attributable to marginal variances, to dependence structures, or to both components.
 
 In `CovCorTest`, this combined testing procedure is implemented in a dedicated function that automates the construction of the test statistic, the simulation-based calibration of critical values, and the decomposition of the global decision into variance- and correlation-specific components. The combined test is based on a vector-valued test statistic

\begin{equation}
\boldsymbol{T}
=
\begin{pmatrix}
T_1\\
\vdots\\
T_p
\end{pmatrix}
=
\sqrt{N}
\left(
\begin{pmatrix}
\widehat{V}_{1,11}\\
\widehat{V}_{1,22}\\
\vdots\\
\widehat{V}_{1,dd}\\
\widehat{\boldsymbol{r}}_1
\end{pmatrix}
-
\begin{pmatrix}
\widehat{V}_{2,11}\\
\widehat{V}_{2,22}\\
\vdots\\
\widehat{V}_{2,dd}\\
\widehat{\boldsymbol{r}}_2
\end{pmatrix}
\right),
(\#eq:combined-statistic)
\end{equation}

where the first $d$ components correspond to differences in marginal variances and the remaining components capture differences in correlation structures. Under the null hypothesis $\mathcal{H}_0^{\boldsymbol{v}}:\boldsymbol{V}_1=\boldsymbol{V}_2$, the statistic $\boldsymbol{T}$ converges in distribution to a multivariate normal distribution. The explicit form of the corresponding asymptotic covariance matrix is provided in @sattler2023. 
 
 To obtain critical values, a simulation-based procedure related to the Taylor-based Monte-Carlo approach is employed to approximate the limiting distribution. Specifically, for both groups $i=1,2$, independent realizations

\begin{equation}
\boldsymbol{\Lambda}_i
\sim
\mathcal{N}(\mathbf{0},\widehat{\boldsymbol{\Sigma}}_i),
(\#eq:combined-simulation)
\end{equation}

are generated and subsequently transformed into $\boldsymbol{\Lambda}_i^{\mathrm{Tay}}$. Based on these transformed quantities, a simulated version of the test statistic, denoted by $\boldsymbol{T}^{\mathrm{Tay}}$, is constructed. Repeating this procedure $B \in \mathbb{N}$ times yields empirical component-wise $\beta$-quantiles $q_{\ell,\beta}$. The local significance level $\widetilde{\beta}$ is calibrated to control the family-wise type-I error rate at level $\alpha$ using the approach of @munko2024, which assigns equal local error probabilities across all components. 
 
An asymptotic level-$\alpha$ test of the null hypothesis of equal covariance matrices is then defined by
\begin{equation}
\max\left[
\max_{\ell=1,\dots,d}
\left(
1-1\hspace{-0.7ex}1\!\left(
q_{\ell,\widetilde{\beta}/2}
\leq T_\ell \leq
q_{\ell,1-\widetilde{\beta}/2}
\right)
\right),
\;
\max_{\ell=d+1,\dots,p}
\left(
1-1\hspace{-0.7ex}1\!\left(
q_{\ell,\widetilde{\beta}/2}
\leq T_\ell \leq
q_{\ell,1-\widetilde{\beta}/2}
\right)
\right)
\right].
(\#eq:combined-test)
\end{equation}

The components responsible for rejection provide additional interpretative information. Rejection in the first group of components indicates differences in marginal variances, whereas rejection in the second group indicates differences in correlation structures. From a software perspective, this decomposition is reflected in the returned output of `CovCorTest`. Specifically, the implementation provides three $p$-values: a variance-specific $p$-value, a correlation-specific $p$-value, and a global $p$-value defined as the minimum of the two component-wise $p$-values. This structure allows users to assess both the overall test decision and interpretable component-wise results within a single, coherent workflow.
 
 From a reproducibility standpoint, all inference procedures implemented in `CovCorTest` are fully deterministic conditional on the state of the random number generator. When a random seed is set via `set.seed`, Monte-Carlo, bootstrap, and Taylor-based Monte-Carlo methods yield reproducible results across repeated runs. The default values for the number of repetitions follow methodological recommendations, while they remain user-accessible. 
 
# Software and examples

In this section, we analyze datasets under different hypotheses to illustrate
the usage of the testing procedures implemented in the `CovCorTest` 
package. To this end, we first explain the general syntax of the functions,
thereby facilitating their application to a real-world dataset. Since most of
the implemented tests follow a similar construction and share a common
interface, their application is largely consistent across different testing
scenarios. We therefore begin with a general overview of the syntax and then
highlight the main differences between the functions. In particular,
`test_combined` differs from the other functions because it is designed
for a more specific hypothesis and setting, resulting in fewer user-specified
options.

## Syntax 

The implemented test functions follow a common workflow. First, the user
provides the data in one of the supported input formats. Second, the hypothesis
or structure of interest is specified. Third, the inference method used to
approximate the critical value or p-value is chosen. The following schematic
call illustrates this workflow for testing equality of covariance matrices
across three groups:

```r
X <- list(group1, group2, group3)
nv <- vapply(X, nrow, integer(1))

res <- test_covariance(
  X = X,
  nv = nv,
  hypothesis = "equal",
  method = "MC",
  repetitions = 1000
)
```

Here, the data object `X` contains the group-specific data matrices or data 
frames, `nv` contains the corresponding group sample sizes, 
`hypothesis = "equal"` specifies the null hypothesis of equal covariance
matrices, `method = "MC"` selects the Monte-Carlo approximation, and 
`repetitions = 1000` determines the number of Monte-Carlo repetitions. The same
general structure is used for the other test functions, with minor differences 
depending on whether a covariance matrix, a correlation matrix, a structure, or 
the combined covariance-correlation hypothesis is tested. The most important 
arguments are described below.



-  *X*: The dataset for which the test should be conducted. Two input
   formats are supported. First, `X` may be a list, where each list element is
   a numeric matrix or numeric data frame corresponding to one group. In this
   format, observations are stored in rows and variables in columns. Hence,
   all list elements must have the same number of columns, while the number of
   rows may differ between groups. If *nv* is omitted for list input with
   multiple groups, the group sample sizes are inferred from the numbers of
   rows of the list elements. Second, `X` may be a single numeric matrix or
   numeric data frame containing all groups, again with observations in rows
   and variables in columns. In this case, the observations have to be ordered
   by group, and the accompanying vector `nv` specifies the group sample sizes.
   For single-group tests, including covariance and correlation structure
   tests, `nv` is not required.

-  *hypothesis*: For increased user-friendliness, predefined hypotheses can be selected in
   `test_covariance`, `test_correlation`, `test_covariance_structure` and
   `test_correlation_structure`. The available options depend on the
   respective test function:
   
   **Covariance matrix:** Here, in the case of one group,
   the options are "equal" (denoting equal variances between the components),
   "given-trace", "given-matrix" and "uncorrelated".
   In the case of multiple groups, the predefined hypotheses are
   "equal" (now denoting equal covariance matrices in the different
   groups), "equal-trace" and "equal-diagonals".
   
   **Correlation matrix:** For one group, the hypotheses
   are "equal-correlated", denoting equal correlation between the
   components and "uncorrelated", which means that all components
   have a correlation of zero. In the case of multiple groups, the
   only hypothesis is "equal-correlated", which here denotes the
   equality of correlation matrices.
   
   **Covariance structure:** For covariance structures, the
   selectable hypotheses are autoregressive ("autoregressive" or
   "ar"), first-order autoregressive ("fo-autoregressive" or
   "fo-ar"), diagonal ("diagonal" or "diag"), sphericity
   ("sphericity" or "spher"), compound symmetry ("compoundsymmetry"
   or "cs"), Toeplitz ("toeplitz" or "toep"), constant off-diagonal
   ("constant-offdiagonal" or "const-offdiag"), standard Toeplitz
   ("standard-toeplitz" or "std-toep"), banded ("banded" or
   "band"), and banded Toeplitz ("banded-toeplitz" or
   "band-toep"). For the banded structures, the additional argument
   `bandwidth` specifies the number $q$, $1 \leq q \leq d-2$, of
   off-diagonals that are allowed to be non-zero.
   
   **Correlation structure:** The predefined structures of
   the correlation matrices are heterogeneous autoregressive
   ("hautoregressive" or "har"), heterogeneous Toeplitz
   ("htoeplitz" or "htoep"), heterogeneous compound symmetry
   ("hcompoundsymmetry" or "hcs"), diagonal ("diagonal" or
   "diag"), banded ("banded" or "band"), and banded Toeplitz
   ("banded-toeplitz" or "band-toep"). For the latter two, the same
   `bandwidth` argument is used.

    - For `test_covariance` and `test_correlation`, more experienced
      users can specify custom hypotheses through a hypothesis
      matrix together with a corresponding vector, as introduced in
      @sattler2022 or @sattler2023, especially for
      less standard hypotheses, which increases flexibility. Both refer
      to the corresponding kind of upper triangular vectorization, and
      unlike in other procedures, the hypothesis matrix here need
      be neither a projection matrix nor square. This especially
      allows structures from the linear covariance structure model and
      the linear correlation structure model by using the function
      `get_hypothesis`.

    - The function `test_combined`, which jointly assesses the
      equality of covariance and correlation matrices, is restricted to
      one specific hypothesis. Consequently, this argument is not
      available in that context.
-  *AM*: Specifies whether the internal calculations are performed using an
   alternative hypothesis matrix. With *AM = 1*, a companion matrix of
   the original hypothesis matrix is used. This matrix has fewer rows
   but yields the same value of the ATS and can therefore reduce the
   computational effort. The companion matrix is constructed using the
   approach implemented in @HypoShrink.
-  *method*: Specifies the method by which the quantile for the test is
   calculated. With *BT*, the quantiles are determined through a
   parametric bootstrap and with *MC* through a Monte-Carlo technique,
   both based on @sattler2022 and @sattler2023. In
   the latter, there is even another method, a Taylor-based Monte-Carlo
   approach, which leads to the option *Tay* for `test_correlation`.
   Again, the combined test works only for a Taylor-based Monte-Carlo
   approach, which makes any options here unnecessary.

-  *repetitions*: The number of Monte-Carlo or bootstrap repetitions
   used in the approximation of the quantiles. This number
   substantially influences both the test's performance and its
   computation time. Since a small number of repetitions increases the
   influence of the randomness on the results, a predefined value of
   *1000* is a good trade-off, and substantially smaller numbers are not
   recommended.
    
The conducted test returns an object containing both the numerical test result 
and additional information on the specified hypothesis, the chosen method, and 
the data structure. The most relevant output components are described in the 
following subsection.

## Output

After conducting a test, the result is stored in an object, such as `res` in
the schematic example above. For the output, we distinguish between the
combined test and all other implemented test functions.

For the latter functions, the returned object consists of several components.
Some of them provide descriptive information about the conducted test, such as
the investigated null hypothesis, the method used to approximate the critical
value or p-value, the number of repetitions, and the vector containing the group
sample sizes. If the hypothesis was chosen from the predefined options, it is reported through the corresponding label. In addition, the returned object contains the hypothesis matrix $\boldsymbol{C}$ and vector $\boldsymbol{\zeta}$ used in the calculation, as well as the corresponding original quantities $\boldsymbol{C}_{\mathrm{original}}$ and $\boldsymbol{\zeta}_{\mathrm{original}}$, which contain the hypothesis before the optional companion-matrix transformation.

Moreover, the estimated covariance matrix used for the test statistic, the
value of the test statistic, and the resulting p-value are included in the
output.

The print method gives a short and compact summary of the conducted test. For
instance, in the case of a correlation test, it reports that a correlation test
was performed, the number of groups, the tested hypothesis, for example, equality
of correlation matrices in a multi-group setting, the value of the test
statistic, and the resulting p-value. Thus, the user can directly identify the
tested hypothesis and interpret the result based on the reported p-value. The
printed output focuses on the quantities usually needed for interpreting the
result, while the full returned object contains additional information that can
be inspected if required. In the examples below, we therefore display the
printed output and interpret the resulting p-values in the context of the
corresponding hypotheses.

For the combined test, the hypothesis, the number of groups, and the
approximation method are fixed, since this function is designed for one specific
two-group hypothesis. Moreover, due to the Taylor-based Monte-Carlo approach
used in this test, no covariance estimator of the test statistic is needed and
therefore none is returned. In addition, the test statistic is vector-valued and
non-standardized, making its components difficult to compare directly. Since
conclusions based only on the numerical value of this test statistic are usually
not meaningful, these values are not printed in order to avoid
misinterpretation.

Instead, the combined test returns and prints three p-values:
`pvalue_Variances` for the local hypothesis of equal variances,
`pvalue_Correlations` for the local hypothesis of equal correlations, and
`pvalue_Total` for the global hypothesis combining the variance and correlation
parts. These p-values allow the user to distinguish whether possible differences
are mainly related to the marginal variances, to the standardized dependence
structure, or to the combined covariance-correlation hypothesis.

## Examples
```{r setup-examples, include=FALSE}
library(CovCorTest)
```


To demonstrate the usage of the main functions and to illustrate typical
outputs, we present several example analyses based on real-world data. For this
purpose, we use the dataset EEG, which includes multivariate measurements from
a neurological study conducted at the University Clinic of Salzburg, Department
of Neurology. This dataset is part of the \CRANpkg{MANOVA.RM} package
[@Friedrich2019] and contains $N = 160$ patients from both genders,
with three different diagnoses of impairments (subjective cognitive complaints
(SCC), mild cognitive impairment (MCI), and Alzheimer's disease (AD)). Thereby,
two neurological parameters (z-score of the brain rate and Hjorth complexity)
were measured at three different locations of the head: temporal, frontal, and
central. This design results in a multivariate factorial structure with multiple
response variables and multiple independent groups and was further described in
@bathke2018. In the following sections, we apply the core testing
procedures from `CovCorTest` to this dataset, showcasing the practical
implementation and interpretation of results.
 
 
### Covariance matrix test

We first consider the equality of covariance matrices across diagnostic groups.
Due to the group sample sizes, we restrict this example to female patients with
the three diagnoses SCC, MCI, and AD, leading to $a = 3$ unbalanced groups.
The six selected variables correspond to the two neurological parameters
measured at the three locations. The data are stored in the list `X`, where
each list element contains the observations from one diagnosis group, and `nv`
contains the corresponding group sample sizes.

The null hypothesis is
\begin{equation}
H_0:\boldsymbol{V}_{\mathrm{SCC}}
=\boldsymbol{V}_{\mathrm{MCI}}
=\boldsymbol{V}_{\mathrm{AD}}.
(\#eq:eeg-cov-hypothesis)
\end{equation}
We first test this hypothesis using the predefined option
`hypothesis = "equal"`. To compare the available approximation methods, the
same test is performed once with the Monte-Carlo approach and once with the
parametric bootstrap. Afterwards, we specify the same null hypothesis manually
through a hypothesis matrix and a corresponding vector.

```{r covariance-equal-tests}
data("EEGwide", package = "MANOVA.RM")
vars <- colnames(EEGwide)[1:6]
SCCFemale <- EEGwide[EEGwide$sex == "W" & EEGwide$diagnosis == "SCC", vars]
MCIFemale <- EEGwide[EEGwide$sex == "W" & EEGwide$diagnosis == "MCI", vars]
ADFemale  <- EEGwide[EEGwide$sex == "W" & EEGwide$diagnosis == "AD", vars]

X <- list(SCCFemale, MCIFemale, ADFemale)
nv <- unlist(lapply(X, nrow))

set.seed(123)
test_covariance(X = X, nv = nv, hypothesis = "equal",
                method = "MC", repetitions = 1000)

set.seed(123)
test_covariance(X = X, nv = nv, hypothesis = "equal",
                method = "BT", repetitions = 1000)
```

For the manual specification of the hypothesis matrix, $P_3$, the
three-dimensional centering matrix, is used. This is the standard projection
matrix for comparing three groups. Since in this example each group is
represented by a 21-dimensional covariance vector, the hypothesis matrix is
constructed as the Kronecker product of the centering matrix and the
corresponding identity matrix. The corresponding hypothesis vector is the zero
vector. For further details on this type of hypothesis formulation, see
@sattler2018. This manual formulation is included to illustrate that
the same hypothesis can either be selected through a predefined option or
specified explicitly by the user.

```{r covariance-manual-hypothesis}
d <- 6
p <- d * (d + 1) / 2
P3 <- diag(3) - matrix(1 / 3, 3, 3)

Hypothesismatrix <- P3 %x% diag(1, p, p)
Hypothesisvector <- rep(0, 3 * p)

set.seed(123)
test_covariance(
  X = X, nv = nv, C = Hypothesismatrix, Xi = Hypothesisvector,
  method = "BT", repetitions = 1000
)
```

The two tests based on the predefined hypothesis lead to the same value of the
test statistic, since the test statistic itself does not depend on the chosen
approximation method. The p-values differ only slightly because the critical
values are approximated by different simulation-based procedures. With p-values 
of 0.026 and 0.023, respectively, the null hypothesis of equal covariance 
matrices across the three diagnosis groups is rejected at the usual
$5\%$ level. The manually specified hypothesis leads to the same test statistic and
p-value as the predefined formulation, confirming that both calls test the same
null hypothesis.


### Correlation matrix test 

After comparing the covariance matrices for women with these diagnoses, the correlation matrices are also investigated for equality. In addition to the parametric bootstrap, the test is also conducted with the Taylor-based Monte-Carlo approach, specifically developed for testing vectorized correlation matrices.

```{r correlation-equal-tests}
set.seed(123)
test_correlation(X = X, nv = nv, hypothesis = "equal-correlated",
                 method = "BT", repetitions = 1000)

set.seed(123)
test_correlation(X = X, nv = nv, hypothesis = "equal-correlated",
                 method = "Tay", repetitions = 1000)
```
 
 
Here, the p-values are considerably higher than for the comparison of covariance matrices. This is plausible, since equality of covariance matrices implies equality of the corresponding correlation matrices, whereas the converse does not hold in general. Hence, the rejection of equal covariance matrices may be driven by differences in marginal variances rather than by differences in the standardized dependence structure. Therefore, the hypothesis of equal correlation matrices is not rejected at the usual $\alpha$ levels.

### Combined test 
Instead of testing equality of covariance matrices and equality of correlation
matrices for two groups separately, the combined test from @sattler2023 allows both
aspects to be investigated within one procedure. This test 
decomposes the comparison into variance-related and correlation-related components. It can therefore be used to assess whether
differences between two covariance matrices are mainly driven by differences in
the marginal variances, by differences in the standardized dependence structure,
or by both.
In this example, we compare male patients with AD and MCI. The six selected
variables are the same as in the previous examples and are stored in the list
`X`, while `nv` contains the two group sample sizes.

```{r combined-test-example}
data("EEGwide", package = "MANOVA.RM")
vars <- colnames(EEGwide)[1:6]

X <- list(
  EEGwide[EEGwide$sex == "M" & EEGwide$diagnosis == "AD", vars],
  EEGwide[EEGwide$sex == "M" & EEGwide$diagnosis == "MCI", vars]
)
nv <- unlist(lapply(X, nrow))

set.seed(123)
test_combined(X, nv)
```

The output contains three p-values. The p-value for the variance part is
0.388 and therefore gives no evidence against equality of the marginal
variances. In contrast, the p-value for the correlation part is 0.026, indicating
evidence against equality of the correlation matrices at the usual (5%)
level. The global p-value is also 0.026, so the combined null hypothesis is
rejected. In this example, the detected difference between the two groups is
therefore mainly attributable to the standardized dependence structure rather
than to the marginal variances.

### Structure tests 
In this analysis, we want to investigate whether the position of the measuring
points influences the measured values. While \CRANpkg{MANOVA.RM} could
investigate a related question regarding mean effects, it does not assess the
covariance or dependence structure between the components. The present approach
therefore addresses a different aspect of the data by testing whether the
covariance matrix is compatible with a compound symmetry structure. This
structure assumes that all variances are equal and that all covariances between
different components are equal, as would be expected if the measurement
positions were exchangeable with respect to their variability and covariance
pattern. If this hypothesis is rejected, this indicates that the variability and
covariance pattern depend on the position of the measuring points. Moreover, to
check whether the standardized dependencies between the measuring points are
equal, we additionally test for a heterogeneous compound symmetry structure of
the correlation matrix.

```{r structure-tests}
X <- as.matrix(EEGwide[EEGwide$sex == "W" & EEGwide$diagnosis == "AD",
  c("brainrate_temporal", "brainrate_frontal", "brainrate_central",
    "complexity_temporal", "complexity_frontal", "complexity_central")])

set.seed(123)
test_covariance_structure(
  X = X, structure = "compoundsymmetry", method = "MC"
)

set.seed(123)
test_correlation_structure(
  X = X, structure = "hcompoundsymmetry", method = "MC"
)
```
 
The covariance structure hypothesis is not rejected at the usual (5%)
significance level, since the corresponding p-value is 0.088. Thus, the test
does not provide sufficient evidence against a compound symmetry structure of
the covariance matrix in this example. In contrast, the heterogeneous compound
symmetry structure of the correlation matrix is rejected, with a p-value of
0.010. This indicates that the standardized dependencies between the components
are not all equal. In the present setting, this provides evidence that the correlation structure
of the measurements depends on the considered components.

We can also test the compound symmetry structure of the covariance matrix using a hypothesis formulation based on `get_hypothesis` and `test_covariance`. The function `get_hypothesis` constructs a hypothesis matrix and a corresponding vector for linear covariance or correlation structure models based on a vector $v_0$ and a matrix $V$. The columns of $V$ are the vectorized basis components $v_1,\ldots,v_m$ of the model. Hence, admissible parameter vectors are of the form
\begin{equation}
v_0 + \theta_1 v_1 + \cdots + \theta_m v_m,
(\#eq:linear-structure-model)
\end{equation}
with unknown coefficients $\theta_1,\ldots,\theta_m$. For the compound symmetry example, $v_0$ is zero and the two columns of $V$ encode the diagonal and off-diagonal components. With the resulting hypothesis matrix and vector, the test for this structure can be conducted.


```{r manual-structure-test}
X <- as.matrix(EEGwide[EEGwide$sex == "W" & EEGwide$diagnosis == "AD",
  c("brainrate_temporal", "brainrate_frontal", "brainrate_central",
    "complexity_temporal", "complexity_frontal", "complexity_central")])

v0 <- rep(0, 21)
v_auxiliary <- c(1, rep(0, 5), 1, rep(0, 4), 1, rep(0, 3),
                 1, rep(0, 2), 1, 0, 1)
V <- cbind(v_auxiliary, 1 - v_auxiliary)
h <- get_hypothesis(v0, V)

set.seed(123)
test_covariance(
  X = X, C = h$hypothesis_matrix, Xi = h$hypothesis_vector,
  method = "MC", repetitions = 1000
)
```
 
Although this is based on the same dataset, structure, and seed, the p-value
differs from the predefined structure test above. This illustrates that, for the
ATS, the test decision can depend on the representation of the null hypothesis,
as also discussed in @sattler2023a. The manual formulation therefore
serves two purposes: it demonstrates how users can specify structures through
`get_hypothesis`, and it shows that different but equivalent formulations of a
hypothesis may lead to different finite-sample behavior for this type of test
statistic. 

# Conclusion and outlook

In the presented paper, we illustrate the functionality and scope of the
R package `CovCorTest`, which provides a comprehensive framework for
hypothesis testing concerning covariance and correlation matrices in
multivariate data analysis. It contains tests for both parameters and different
hypotheses in a uniform way, which excels through its applicability to a large
class of possible hypotheses and fewer distributional assumptions. Apart from
multi-group settings, single groups are also allowed, as well as factorial
designs. Essentially, it contains functions for five different tests: regarding
covariance matrices, correlation matrices, a combined test for equality of
covariance and correlation matrices, and tests regarding their corresponding
structures. To this end, quantiles, based on various techniques including
resampling and Monte-Carlo approaches, are used. Although the Monte-Carlo
approach leads to the least time-consuming procedure, a parametric bootstrap
for covariances and a Taylor-based Monte-Carlo approach for correlations are
preferable for smaller sample sizes.

Future developments will focus on methodological extensions that require 
additional investigation before they can be recommended as routine software
features. In particular, multiple contrast tests for general covariance or 
correlation hypotheses and extensions of the combined covariance-correlation 
test to more than two groups lead to a rapidly increasing number of local 
alternatives. This may result in poor finite-sample performance and difficult
interpretation of the resulting local decisions. We therefore regard these 
extensions as topics for further methodological research rather than as stable
features for the present software release. Since the approach proposed in 
@sattler2026 allows equivalent or redundant hypotheses to be combined 
into smaller hypothesis systems, it may provide a promising starting point for 
addressing this issue. Further possible developments concern usability and 
computational efficiency.

# Acknowledgment 

Svenja Jedhoff gratefully acknowledges the support from TRR 391
Spatio-temporal Statistics for the Transition of Energy and Transport
(520388526), funded by the Deutsche Forschungsgemeinschaft (DFG, German
Research Foundation).

The authors used ChatGPT (OpenAI) for language polishing and editorial
suggestions during the preparation of the manuscript. The authors reviewed
and approved all resulting text and take full responsibility for the content
of the article.
