---
title: Changes on CRAN
abstract: Updates from CRAN for 2024-01-01 to 2024-06-30
date: '2024-03-01'
draft: no
author:
- name: Kurt Hornik
  affiliation: WU Wirtschaftsuniversität Wien
  address: Austria
  orcid: 0000-0003-4198-9911
  email: Kurt.Hornik@R-project.org
- name: Uwe Ligges
  affiliation: TU Dortmund
  address: Germany
  orcid: 0000-0001-5875-6167
  email: Uwe.Ligges@R-project.org
- name: Achim Zeileis
  affiliation: Universität Innsbruck
  address: Austria
  orcid: 0000-0003-0918-3766
  email: Achim.Zeileis@R-project.org
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
preamble: \usepackage{float, longtable}
volume: 16
issue: 1
slug: RJ-2024-1-cran
journal:
  lastpage: 213
  firstpage: 212

---


```{r global_options, include=FALSE, message=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

# CRAN growth

```{r cran_growth_data, include=FALSE, message=FALSE}
from <- as.Date("2024-01-01")
to <- as.Date("2024-06-30")

cran_growth <- readRDS("cran_growth.rds")
cran_zoo <- cran_growth$zoo
loadNamespace("zoo")

nmonths <- cran_growth$nmonths
npackages <- cran_growth$npackages
mirrors <- cran_growth$mirrors
```

In the past `r nmonths` months, `r npackages["new"]` new packages were
added to the CRAN package repository. `r npackages["unarchived"]` packages
were unarchived, `r npackages["archived"]` were archived and 
`r npackages["removed"]` had to be removed. The following shows the
growth of the number of active packages in the CRAN package repository:

```{r cran_growth, fig.height=7, fig.width=10, fig.align='center', out.width = "100%", fig.alt="CRAN growth: Number of CRAN packages over time in levels (left) and in logs (right).", echo=FALSE}
par(mfrow = c(1, 2))
plot(cran_zoo, lwd = 3, xlab = "Year", ylab = "", main = "Number of CRAN Packages")
plot(cran_zoo, lwd = 3, log = "y", xlab = "Year", ylab = "", main = "Number of CRAN Packages (Log-Scale)")
```

\noindent On `r to`, the number of active packages was around `r cran_zoo[to]`.


<!--

# Changes in the CRAN Repository Policy

Start out from

```
cd ~/src/org/R-project/R-dev-web/CRAN/Policy
svn diff -r{2024-01-01}:{2024-06-30}
```

-->


# CRAN package submissions

```{r cran_submissions_data, include=FALSE}
## generate on cran master via cran_submissions(from, to, ...)
cran_sub <- readRDS("cran_submissions.rds")
```

From `r format(cran_sub$from, "%B %Y")` to `r format(cran_sub$to, "%B %Y")`
CRAN received `r cran_sub$nsubmissions` package submissions.
For these, `r cran_sub$nactions` actions took place of which
`r cran_sub$autocheck["TRUE"]` (`r cran_sub$autocheckrel["TRUE"]`%) were auto processed actions and
`r cran_sub$autocheck["FALSE"]` (`r cran_sub$autocheckrel["FALSE"]`%) manual actions.

Minus some special cases, a summary of the auto-processed and manually
triggered actions follows:

```{r cran_submissions_autouser, echo=FALSE}
knitr::kable(cran_sub$autouser[, !grepl("clang", colnames(cran_sub$autouser))], format = "pipe")
```

These include the final decisions for the submissions which were

```{r cran_submissions_autolast, echo=FALSE}
knitr::kable(cran_sub$autolast, format = "pipe", align = "rr")
```

\noindent where we only count those as _auto_ processed whose publication or
rejection happened automatically in all steps.

`r gsub("!", ".", cran_sub$newmember, fixed = TRUE)`
`r gsub("~", "&nbsp;", cran_sub$oldmember, fixed = TRUE)`


# CRAN mirror security

Currently, there are `r mirrors[1]` official CRAN mirrors,
`r mirrors[2]` of which provide both
secure downloads via '`https`' _and_ use secure mirroring from the CRAN master
(via rsync through ssh tunnels). Since the R 3.4.0 release, `chooseCRANmirror()`
offers these mirrors in preference to the others which are not fully secured (yet).


# CRAN Task View Initiative

```{r cran_views_data, include=FALSE}
## generate via cran_views(new = NULL)
cran_views <- readRDS("cran_views.rds")
cran_views$active <- cran_views$ntotal/as.numeric(tail(cran_zoo, 1))
```

```{r cran_views_new, echo=FALSE, results="asis"}
writeLines(cran_views$new)
```

Currently there are `r cran_views$nviews` task views (see <https://CRAN.R-project.org/web/views/>),
with median and mean numbers of CRAN packages covered
`r round(median(cran_views$npackages))` and `r round(mean(cran_views$npackages))`, respectively.
Overall, these task views cover `r cran_views$ntotal` CRAN packages,
which is about `r round(100 * cran_views$active)`% of all active CRAN packages.

Julia Piaskowski (University of Idaho) joined the team of CRAN Task View Editors, welcome!
