---
title: Conformal Prediction Tools in R
date: '2026-07-16'
abstract: |
  Conformal prediction constitutes a flexible family of methods for predictive uncertainty quantification, which has recently drawn significant attention in the statistical literature because of its distribution-free, finite-sample, and model-agnostic coverage guarantees. While numerous methodological advances have extended the original conformal framework to broader settings, the availability of related software in R remains limited. In this paper, we present an overview of several conformal prediction methods and exemplify the use of available R packages, with a focus on regression for continuous responses given covariates and its extensions to time series forecasting. We further identify current limitations and outline opportunities for future improvements.
draft: no
author:
- name: Jesus Uriel Diaz-Martinez
  affiliation: King Abdullah University of Science and Technology (KAUST)
  address:
  - Computer, Electrical and Mathematical Sciences and Engineering Division
  - Department of Statistics
  - Thuwal, 23955-6900, Kingdom of Saudi Arabia
  url: ~
  orcid: ~
  email: jesus.diazmartinez@kaust.edu.sa
- name: Xiaoqian Wang
  affiliation: Academy of Mathematics and Systems Science, Chinese Academy of Sciences
  address:
  - The Center for Forecasting Science (corresponding author)
  - Beijing, China
  url: https://xqnwang.rbind.io/
  email: xiaoqian.wang@amss.ac.cn
  orcid: 0000-0003-4827-496X
- name: Paula Moraga
  affiliation: King Abdullah University of Science and Technology (KAUST)
  address:
  - Computer, Electrical and Mathematical Sciences and Engineering Division
  - Department of Statistics
  - Thuwal, 23955-6900, Kingdom of Saudi Arabia
  url: http://www.paulamoraga.com
  email: paula.moraga@kaust.edu.sa
  orcid: 0000-0001-5266-0201
type: package
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
header-includes: \usepackage{todonotes}
bibliography: RJreferences.bib
editor_options:
  markdown:
    wrap: 72
date_received: '2025-10-16'
volume: 18
issue: 2
slug: RJ-2026-024

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

$$
\newcommand{\spcoord}{\boldsymbol{s}}
\newcommand{\spdomain}{\mathcal{D}}
\newcommand{\exampspace}{\textbf{Z}}
\newcommand{\exampobs}{\textbf{Z}^{(*)}}
$$

# Introduction

Conformal prediction is a family of methods to construct prediction
intervals for point predictions. These methods have been popularized
recently in the statistical literature due to their model-agnostic and
finite-sample coverage guarantees, which require mild assumptions on the
distribution of the data. Such properties make the framework appealing
for a wide range of tasks such as regression and time series
forecasting. The model-agnostic property allows conformal prediction to
be wrapped around several models. On the other hand, the finite-sample
property is related to conservatively marginal coverage guarantees.
Specifically, for a future observation of interest $Y_{n+1}$, the
resulting prediction interval, denoted by $\Gamma^{\alpha}$, will
satisfy $P(Y_{n+1}\in\Gamma^{\alpha})\geq 1-\alpha$ for any significance
level $\alpha\in(0,1)$. These notable features are particularly valuable
for providing predictive uncertainty quantification in models that
inherently lack it, in particular black-box prediction models. Standard
references on conformal prediction include the book *Algorithmic
Learning in a Random World* by @vovk2022 with a recent edition
[@vovk2022], and a tutorial by @shafer2008.

Conformal prediction has emerged as a reliable framework for uncertainty
quantification, with applications spanning a wide range of domains. In
the medical sciences, it has been applied in breast cancer diagnosis,
survivability prediction, and symptom-based condition identification
[@vazquez2022]. In the energy sector, conformal prediction has been used
to measure prediction uncertainty in simulation-based forecasting,
energy and gas demand prediction, and electricity price prediction [e.g.,
@mendil22a]. More broadly, conformal prediction has also been used to construct
prediction intervals for the Direction-of-Arrival (DOA) predictions in
geophysics and underwater acoustics, as well as
enhancing reliability in aircraft landing systems [@vilfroy2024].

On the methodological advances, significant progress has been made to
extend the scope and applicability of conformal prediction. @lei2018
demonstrated that conformal prediction methods are not only
conservatively valid but also admit model-independent upper bounds on
coverage under additional assumptions on the residuals. Many attempts have been made to broaden the framework to more
general settings. For example, @tibshirani2019 proposed a method to
address covariate shift, while @romano2019 tackled the problem of
variance heterogeneity in regression settings using flexible quantile
regression. @barber2023 proposed a general theory based on weighted
quantiles, extending conformal prediction to settings previously far
from the scope of the original formulation. @chen2018 investigated
computationally efficient methods to implement conformal prediction in
regression tasks. In the context of time series forecasting, specialized
methods based on online updating of prediction intervals have been
pioneered by @Gibbs_Candes_2021, @zaffran2022, @Bhatnagar_Wang_Xiong_Bai_2023 and
recently by @Wang_Hyndman_2024.

In the R ecosystem, several CRAN packages provide implementations of conformal prediction for both regression and time series forecasting. Available packages for regression include **pintervals** [@pintervals],
**marginaleffects** [@marginaleffects], and **probably** [@probably]. For time series forecasting, the **caretForecast** [@caretForecast] and
**conformalForecast** [@conformalForecast] are two packages for
conformal prediction of time series forecasting. These packages generally adopt a flexible framework that can be applied to diverse underlying models.

Other available packages offer model-specific
implementations, among them are **conformalInference.multi**
[@conformalmulti], designed for multivariate regression; and
**ConformalSmallest** [@ConformalSmallest], which optimizes conformal
prediction intervals generated with quantile and ridge regression. While
these packages provide valuable contributions, they also present some
limitations. For instance, some implementations restrict conformal
prediction to models from specific packages due to dependencies on
custom objects generated by those ecosystems.

In other languages, like Python, libraries such as **MAPIE**
[@Cordier_Flexible_and_Systematic_2023], **crepes** [@bostrom2024] and
**torchCP** [@huang2024torchcp] provide implementations of conformal
prediction algorithms that augment regression and classification
algorithms available in **scikit-learn** [@scikit-learn] and **pytorch**
[@pytorch_2019], which are two major machine learning libraries.

In this paper, we demonstrate the use of the **pintervals**,
**marginaleffects**, **probably** and **conformalForecast** packages in R to
the problem of constructing uncertainty intervals in two settings:
regression with covariates and time series forecasting. In the regression
context, the methods are applied to predict a continuous response
variable using a number of covariates, exemplified through a wind farm
energy production dataset. For the time series domain, where
multi-step-ahead forecasting is of particular interest, we demonstrate
the approach using data on the number of active cases of Dengue in
Brazil.

The remainder of the paper is as follows. In Section 2 we review the
basics of conformal prediction, from the classical perspective, and its
adaptations to regression and time series forecasting. In Sections 3 and
4, we provide code examples for regression and time series forecasting,
respectively. Section 5 provides a discussion from the perspective of
our examples, while Section 6 states our conclusions.

# Background on conformal prediction

In this section, we provide an overview of the technical aspects of
conformal prediction required to understand the examples and the code.
For an in-depth overview, the readers are encouraged to consult the work
by @vovk2022, @shafer2008, @lei2018,
@romano2019, and @Wang_Hyndman_2024, just to mention a few.

## The setting for conformal prediction

The basic setting for conformal prediction assumes that there exists an
*example space* $\textbf{Z}$, from which data points
$Z_1, Z_2, Z_3, \dots$ can be drawn according to a data-generating
process. In practice, only a finite sample $\{Z_1, Z_2, \dots, Z_n\}$ is
available; this collection will be denoted by $\textbf{Z}^{(*)}$. The
goal of conformal prediction methods is to construct prediction sets,
$\Gamma^{\alpha}$, for a future observation $Z_{n+1}\in \textbf{Z}$ such
that the coverage guarantee
$P(Z_{n+1}\in \Gamma^{\alpha}) \geq 1-\alpha$ holds for a predefined
significance level $\alpha\in(0,1)$. This guarantee is known as
*conservative validity*. A key distributional assumption to achieve this
goal is that the data generating process is exchangeable, i.e., the
process produces an exchangeable collection of random variables.
Specifically, a collection of (possibly vector-valued) random variables,
$(Z_1, Z_2, \ldots, Z_n)$, is said to be exchangeable if for any
permutation $\pi$ of the set $\{1,\ldots, n\}$, the joint distribution
of $(Z_{\pi(1)}, Z_{\pi(2)}, \ldots, Z_{\pi(n)})$ is the same as the
joint distribution of the original collection. This assumption is
weaker, and hence more flexible, than the usual i.i.d. assumption since
exchangeable random variables are not necessarily independent.

The practical construction of the prediction set relies on the so-called
*nonconformity measure*, which is a function
$S:\textbf{Z}^{(*)}\times \textbf{Z}\to [-\infty,+\infty]$, to quantify
how different a new example $z\in\textbf{Z}$ is relative to the
available examples in $\textbf{Z}^{(*)}$. From this function, one
obtains nonconformity scores; the corresponding score for each
observation $Z_i$ is defined as $V_i := S(\textbf{Z}^{(*)}, Z_i)$. In
several tasks, such as regression and time series forecasting, there are
*natural* choices of the nonconformity measure, for example the absolute
residuals for regression and the signed residuals for time series
forecasting. This does not imply that they must be the unique choice, as
the choice usually depends on the context of the problem [@vovk2022]. A
detailed discussion and review of nonconformity measures in the context
of regression is provided in @kato2023review.

## Full conformal prediction

One of the earliest and more general approaches is full conformal
prediction, which serves as a conceptual starting point for other
methods in this framework for uncertainty quantification. In this
method, the collection of available examples is augmented with a
potential value, $z\in\mathbf{Z}$, for $Z_{n+1}$. Let
$\textbf{Z}^{(*)}_{z} := \{Z_1 , Z_2, \ldots, Z_n, z\}$ be the
collection of augmented examples; note that for each potential value in
$\textbf{Z}$ the augmented set of available examples will be different.
Define $V_j := S\left(\textbf{Z}^{(*)}_{z}, Z_j\right)$ for
$j=1,2,\ldots, n$, $V_{n+1}^{z} := S(\textbf{Z}^{(*)}_{z}, z)$, and let
$\alpha$ be a significance level. The p-value for the new potential
observation will be defined as

\begin{equation}
p_{n+1}^{z} = \dfrac{\lvert \{j = 1\ldots, n+1: V_j \geq V_{n+1}^z\} \rvert}{n+1},
(\#eq:fcp-pvalnext)
\end{equation} and the full conformal prediction set is defined as

\begin{equation}
\Gamma^{\alpha} := \left\{ z\in \textbf{Z}: p_{n+1}^{z}>\alpha \right\}.
(\#eq:fcp-predset)
\end{equation}

Intuitively, full conformal prediction evaluates how plausible a
candidate future observation would be if it were included together with
the training data. Candidate values that produce nonconformity scores
similar to those observed in the data will have larger p-values and will
be included in the prediction set, whereas those that appear atypical
relative to the observed data will have smaller p-values and will be
excluded. Therefore, p-values play a crucial role in conformal
prediction, determining whether a candidate value is included in the
prediction set $\Gamma^{\alpha}$. As noted by
@vovk2022, the choice of nonconformity measure
directly affects the resulting prediction set.

A practical difficulty with full conformal prediction is that, to
construct the prediction set, one needs to compute the nonconformity
scores for all candidate elements in the set $\textbf{Z}$, as every
element can be potentially included. In terms of computations, there are
particular algorithms for which the expensive computations can be
avoided due to the structure of the prediction algorithm, such as
conformalized ridge regression and nearest neighbor regression
[@vovk2022]. Nevertheless, as highlighted by @lei2018,
the construction of full conformal prediction sets remains
computationally expensive in general. This challenge has motivated the
development of alternative methods with lower computational cost, a
widely used alternative is described in the following subsection.

\newtheorem{result}{Result}

## Inductive conformal prediction

A widely used and computationally efficient variant of conformal
prediction is inductive conformal prediction, also known as split
conformal prediction. It is conceptually similar to the full conformal
framework; however, the key difference lies in how p-values are
constructed. In split conformal prediction, model fitting and prediction
set construction are separated, eliminating the need to recompute scores
for each candidate value, thereby substantially reducing the
computational cost. Instead of augmenting the available examples
$\textbf{Z}^{(*)}$ with potential values from the example space, the
available data is split into two disjoint subsets: a proper training set
$\textbf{Z}^{(*)}_\text{Train}$ and a proper calibration set
$\textbf{Z}^{(*)}_\text{Cal}$, containing $l$ and $m$ elements,
respectively, such that $l + m = n$. The key idea in split conformal is
to use only the data in the proper training set to fit the model and
then use the examples in the calibration set to construct the prediction
intervals. Therefore, the nonconformity scores $V_j$ are only defined
for calibration samples, with
$V_j := S(\textbf{Z}^{(*)}_\text{Train},Z_j)$,
$Z_j\in \textbf{Z}^{(*)}_\text{Cal}$, and for a candidate
$z\in \textbf{Z}$ representing a potential value of $Z_{n+1}$,
$V_{n+1}^{z} := S(\textbf{Z}^{(*)}_\text{Train},z)$. Using these
nonconformity scores, nonconformity of a candidate value is assessed by
comparing its score against the empirical distribution of calibration
scores. The p-value for inductive (split) conformal prediction is then
defined by

\begin{equation}
p_{n+1}^{z} = \dfrac{\lvert \{j\in I_\text{Cal}: V_j\geq V_{n+1}^{z}\} \rvert}{m+1},
(\#eq:scp-pval)
\end{equation} where $I_\text{Cal}$ is the index set of the
calibration set, with $|I_\text{Cal}|=m$. The corresponding
inductive prediction interval is constructed as

\begin{equation}
\Gamma^{\alpha}:=\{ z\in \textbf{Z}: p_{n+1}^{z} > \alpha \}.
(\#eq:scp-predset)
\end{equation}

For split conformal prediction, an alternative expression to Equation
\@ref(eq:scp-predset) exists in terms of the empirical quantiles induced
by the nonconformity scores on the calibration set. The prediction
interval is constructed according to Equation
\@ref(eq:scp-predsetquant), which is particularly advantageous for
computer implementations and theoretical analysis.

\begin{equation}
\Gamma^{\alpha} = \left\{ z\in\textbf{Z}: V_{n+1}^z\leq \text{Quantile}\left(1-\alpha; \{V_j\}_{j\in I_\text{Cal}}\right) \right\}.
(\#eq:scp-predsetquant)
\end{equation}

## Applications to regression and time series forecasting

In general, the regression setting can be described as the task of
estimating a function $\hat f: X \to Y$, where
$X$ corresponds to the space where the predictors take values,
typically a subset of $\mathbb{R}^{d}$ with $d$ representing the number
of *predictors*; the space $Y$, which is usually a subset of
$\mathbb{R}$, contains the *response* of interest. For the application
of conformal prediction to regression, the example space is defined as
$\textbf{Z}=X \times Y$. It is assumed that a future
value of the predictor $X_{n+1}$ is observed in order to produce a
prediction for the response $Y_{n+1}$ given by
$\hat Y_{n+1} = \hat f(X_{n+1})$.

In the particular case of full conformal prediction, the data
augmentation step will be performed as
$\textbf{Z}^{(*)}_{y} = \{(x_1,y_1), (x_2,y_2),\ldots,(x_n,y_n), (x_{n+1},y)\}$,
for some $y\in Y$. The nonconformity scores will be computed as
$V_i=S(\textbf{Z}^{(*)}_{y},y_i)$ and
$V_{n+1}^{y}=S(\textbf{Z}^{(*)}_{y},y)$; then the p-value and the
prediction interval will follow a similar definition as in Equations
\@ref(eq:fcp-pvalnext) and \@ref(eq:fcp-predset), respectively, with
minor notational changes to make explicit the dependence only on the
potential value $y$ for $Y_{n+1}$:

\begin{equation}
\Gamma^{\alpha}:=\{y\in Y: p_{n+1}^{y} \geq \alpha\}.
(\#eq:fcp-regpredset)
\end{equation}

The theoretical properties of this approach have been investigated by
@lei2018, who established an upper bound on the coverage probability in
finite sample settings. This property, also known as
anti-conservativeness, is independent of the predictive model or any
distributional assumptions, depending only on the sample size of the
dataset. The result is stated below.

\begin{result}
Under the exchangeability assumption the conformal prediction method for regression satisfies that for a new pair of observation $(X_{n+1},Y_{n+1})$ it holds that $1-\alpha \leq P(Y_{n+1}\in\Gamma^{\alpha})\leq 1-\alpha + \dfrac{1}{n+1}$.
\end{result}

It is important to remark that this result was proven using the absolute
deviation in Equation \@ref(eq:absdevncm) as nonconformity measure,
although the result holds with any valid nonconformity measure as long
as the data is exchangeable and the algorithm is symmetric.

\begin{equation}
S(Y_i,\hat{Y}_i) := \lvert Y_i - \hat{Y}_i \rvert.
(\#eq:absdevncm)
\end{equation}

Split conformal prediction extends naturally to the regression setting
by using the same ideas as full conformal prediction, but with
nonconformity scores computed exclusively on the calibration set. This
yields the following coverage guarantee.

\begin{result}
Under the exchangeability assumption the split conformal prediction method for regression, constructed with a calibration set with $m$ examples, satisfies that for a new pair of observations $(X_{n+1}, Y_{n+1})$, $1-\alpha \leq P(Y_{n+1}\in\Gamma^{\alpha})\leq 1- \alpha + \dfrac{1}{m+1}$.
\end{result}

Using Result 2, one can solve for the size of the calibration set
$m$ to get

\begin{equation}
m\geq \dfrac{2-(\nu + \alpha)}{\nu + \alpha - 1},
(\#eq:scp-calsize)
\end{equation} where $\nu$ is a positive scalar that satisfies
$1-\alpha<\nu\leq1$. The equation above does not imply that using the
derived calibration set size in practice we will observe an empirical
coverage around $\nu$, as this upper bound is derived based on a
worst-case analysis and not a coverage calibration. However, we can use
it to choose the size of the calibration set and we will use it in the
next section.

When the nonconformity measure defined in Equation \@ref(eq:absdevncm)
is used in the split conformal prediction framework, the resulting
prediction intervals take the following simple form:

\begin{equation}
[\hat{Y}_{n+1}-\hat{Q}_\alpha, \hat{Y}_{n+1}+\hat{Q}_\alpha],
(\#eq:scp-predset-absdev)
\end{equation} where $\hat{Q}_\alpha$ is the
$\lceil (m+1)(1-\alpha) \rceil$-th empirical quantile of the
nonconformity scores computed on the calibration set. The prediction
interval from Equation \@ref(eq:scp-predset-absdev) therefore has a
constant width across all values of the predictors.

The coverage guarantees of the methods described above rely on the
validity of exchangeability in the data and on the prediction algorithm
being symmetric. However, in the context of time series, the
exchangeability is not a realistic assumption for the data and hence
coverage guarantees of usual conformal prediction methods do not apply.
One of the main features of time series is the existence of temporal
dependence, with observations ordered in time and often displaying
autocorrelation. In addition, several methods for time series modeling
and prediction are not symmetric, again, because of the temporal
structure inherent to the data. In practice, applications of classic
conformal prediction methods to time series forecasting can lead to
systematic under coverage, meaning the obtained prediction interval does not
achieve the nominal coverage level. In time series forecasting, two
scenarios are of particular interest: one-step ahead and multi-step
ahead forecasting. In the following, time series will be denoted by
$\{Y_t\}$, where each $Y_t \in Y$. An $h$-step ahead forecast
is denoted by $\hat Y_{t+h}$, where $H\geq 1$ denotes the forecasting
horizon. If covariates are used to produce the forecast, they will be
denoted by $\textbf x_t \in X$.

Several methods have been developed to extend conformal prediction to
time series forecasting. Of particular interest has been to address the
violation of data exchangeability and to allow for nonsymmetric
prediction algorithms. For example, the NexCP (Non Exchangeable
Conformal Prediction) method of @barber2023 extends full and split
conformal prediction by introducing weights to the construction of
quantiles, thereby allowing more "trusted" observations to play a
greater role in the prediction step. The approach provides approximate
coverage when exchangeability does not hold and exact coverage when it
does. This method is general and applies not only to time series but
other instances where dependence plays a major role in the data and must
not be ignored. Other approaches exist such as the EnbPI method
proposed by @xu2021, which employs ensembles of bootstrap models to
construct prediction intervals, one notable advantage of this method is
the fact that it is suitable for multi-step forecasting tasks.

Beyond weighted or bootstrap-based extensions, another line of research
known as online conformal prediction, is inspired by the sequential
nature of time series. Online conformal prediction methods produce
intervals dynamically as new data becomes available, refining the
significance level over time to adapt to changes in data distribution or
updates to the model based on the most recent information. The first
such method was developed by @Gibbs_Candes_2021, under the assumption that an
optimal significance level exists for constructing the desired
$1-\alpha$ prediction interval for the point forecast $\hat Y_{t+1}$.
Hence the authors propose a method to estimate this time-varying
significance level $\alpha_{t+1}$, and use it to construct prediction
intervals. Further enhancements of the update rule include @zaffran2022,
@Bhatnagar_Wang_Xiong_Bai_2023, @angelopoulos2023, and recently @Wang_Hyndman_2024 to
construct prediction intervals for multi-step ahead predictions.
@Wang_Hyndman_2024 also developed the **conformalForecast** R package
[@conformalForecast], providing a practical implementation of their
approaches.

Online conformal prediction methods take into account the temporal
ordering of the data by introducing the idea of sequential splits. At
each time point $t=N$ where $N >> H$, the available data are split into
a proper training set of size $t_r$ and a proper calibration set of size
$N-t_r>H$. For notational convenience, the data at time $t$ is denoted
by $z_t$, to accommodate both the univariate case, where only the time
series of interest is used to fit the model (hence $z_t = y_t$), and the
multivariate case, where covariates are included so that
$z_t = (\textbf x_t, y_t)$. In the latter, the covariates may include
not only information up to the current time but also information of the
subsequent $h$ steps. Within this framework, the training set is defined
as a moving window of size $t_r$ and the nonconformity measure is
defined for each $h$-step-ahead forecast at time $t+h$ as follows $$
S(\{Z_j\}_{j=t-t_r + 1}^{t}, Y_{t+h}), \quad \text{for } t_r\leq t \leq N; \text{ } h=1,\ldots,H,
$$ with corresponding score expressed as $$
V_{t+h\lvert t} := S(\{z_j\}_{j=t-t_r + 1}^{t}, y_{t+h}), \quad \text{for } t_r\leq t \leq N; \text{ } h=1,\ldots,H.
$$

In the particular case of the work by @Wang_Hyndman_2024, the
nonconformity measure is given by

$$
S(\{Z_j\}_{j=t-t_r + 1}^{t}, Y_{t+h}) := Y_{t+h} - \hat{Y}_{t+h\lvert t},
$$ where $\hat Y_{t+h\lvert t}$ is the prediction of $Y_{t+h}$ using
data in the training window corresponding to time $t$, i.e,
$\{Z_j\}_{j=t-t_r + 1}^{t}$.

The steps for the multi-step-ahead conformal prediction are outlined as
follows:

1.  Initialization: Obtain the initial proper training set
    $\{z_k\}_{k=1 + t - t_r}^{t}$ with $t = t_r$, and fit the
    forecasting model. Then obtain $H$-step-ahead point forecasts
    $\{\hat{y}_{t+h\lvert t}\}_{h=1}^H$ and compute the corresponding
    nonconformity scores $\{V_{t+h\lvert t}\}_{h=1}^{H}$.
2.  Rolling procedure: Roll the training window of size $t_r$ forward by
    one observation, i.e., set $t\to t+1$, for $t_r\leq t\leq N$. Then
    repeat Step 1 until the collection of nonconformity scores,
    $\{V_{t+h\lvert t}\}_{t=t_r}^{N-H}$, $h=1,\ldots,H$, have been
    computed across the entire calibration set.
3.  Prediction interval construction: Use the nonconformity scores
    obtained from Step 2 to compute quantile estimates
    $\hat{q}_{t+h\lvert t}$ for each forecasting horizon $h=1,\ldots,H$.
    Conformal prediction intervals are then given by
    $\Gamma_{t+h\lvert t}(\hat{q}_{t+h\lvert t}):=\{y\in Y: V_{t+h\lvert t}^y \leq \hat{q}_{t+h\lvert t}\}$,
    with $t=N$ and for each horizon $h=1,\ldots, H$.
4.  Online update: As new observations arrive, update the calibration
    and the training sets by rolling the data forward one data point at
    a time. Recompute the nonconformity scores based on the new
    calibration set and then repeat Step 3 to update prediction
    intervals across time.

The quantile $\hat{q}_{t+h\lvert t}$ in Step 3 can be estimated using
the methods proposed by @Wang_Hyndman_2024. In particular, the update
rule for the proposed Autocorrelated Multi-step Conformal Prediction
(AcMCP) method is defined as

\begin{equation}
\hat{q}_{t+h\lvert t}=\hat{q}_{t+h-1\lvert t-1}+\eta\left(\operatorname{err}_{t\lvert t-h}-\alpha\right)+r_t\left(\sum_{i=h+1}^t\left(\operatorname{err}_{i\lvert i-h}-\alpha\right)\right) + \tilde e_{t+h \lvert t},
(\#eq:updtrule-wang)
\end{equation} where $\tilde e_{t+h\lvert t}$ is a term that accounts
for autocorrelation in the nonconformity scores and is the combination
of a MA($h-1$) model trained on the $h$-step-ahead forecast errors and a
linear regression model trained to regress the $h$-step-ahead error from
the past $h-1$ errors. $\operatorname{err}_{t\lvert t-h}$ is defined as

$$
\operatorname{err}_{t\lvert t-h}:=\begin{cases}
1,\quad \text{if } y_t\notin \Gamma_{t\lvert t-h}(\hat{q}_{t\lvert t-h})\\
0, \quad\text{otherwise}
\end{cases}.
$$ $\eta>0$ is a constant learning rate, and $r_t$ is a saturation
function that satisfies the conditions

$$
x\geq c\cdot g(t-h)\Rightarrow r_t(x)\geq b,\quad \text{and}\quad x\leq -c\cdot g(t-h)\Rightarrow r_t(x)\leq-b,
$$for some constant $b,c>0$ and a function $g$ that is sublinear,
nonnegative and nondecreasing. These conditions are required to
establish the asymptotic coverage guarantees.

In the following sections, we show how to use available implementations
of conformal prediction methods for regression and time series
forecasting. It is important to note that for the regression part, the
space $Y = \{y\in \mathbb{R}: y\geq 0\}$, and the covariate
space is $X = \{(1,x): x\geq 0 \}$. For time series
forecasting, the response $Y$ remains the same.

# Conformal prediction methods for regression in R

In this section, split conformal prediction will be illustrated using
the [SCADA dataset](https://github.com/uriel278/tutorials-cp/tree/main)
obtained from Kaggle [@windtur]. SCADA, short for Supervisory Control
and Data Acquisition systems, is widely used to monitor wind energy
production. The dataset consists of 39, 692 registries of electricity
production (KWh), wind speed (m/s), theoretical electricity production
(KWh), and wind direction collected from a wind turbine's SCADA system
in Turkey. Our objective is to predict the production of a windmill
using wind speed as a covariate. Such datasets present challenges for
modeling, as system failures can generate anomalous observations that
deviate from expected behavior. Therefore, the models used for
prediction should be robust and capable of quantifying uncertainty in
the presence of outliers, which may otherwise compromise the reliability
of prediction intervals.

## Data preparation

First, we load the dataset and inspect its first rows. Then, we create a
data frame `d_scada` for the analysis, containing the response `y`,
representing active power (`active_power`), and the covariate `x`,
representing wind speed (`wind_speed`). A simple summary of the data
indicates that active power (`y`) ranges from 0 to 3618.7 (kW), while
wind speed ranges from 1.2 to 25.2 (m/s).

```{r data-load, message=FALSE, echo=TRUE}
library(readr)
library(curl)
url_scada <- "https://raw.githubusercontent.com/uriel278/tutorials-cp/main/data/scada.csv"
ds_scada <- read_csv(curl(url_scada))
head(ds_scada)
d_scada <- data.frame(y = ds_scada$active_power, x = ds_scada$wind_speed)
```

As split conformal prediction requires a calibration set and we need a
proper training set for model fitting, we generate the data splits by
using the function `split(d_scada, idx)`, where the second argument
`idx` is a vector of the same length as the number of observations and
contains one of the strings from the vector
`c("train", "calib", "test")` selected randomly by the `sample()`
function with arguments `prob = c(0.6, 0.252, 0.148)`,
`size = nrow(d_scada)`, and `replace = TRUE`; this results in a split
that allocates 60% of the data to the proper training set, 25.2% to the
proper calibration set, and the remaining 14.8% to the testing set. In this example, the choice
of allocating 60% of the data for training is expected to preserve the
response-covariate relationship for reliable model training, the size of
the calibration set was determined by setting $\nu = 0.95001$ in
Equation \@ref(eq:scp-calsize).

```{r splits-data, echo=TRUE}
set.seed(141019)
idx <- sample(x = c("train", "calib", "test"), prob = c(0.6, 0.252, 0.148),
  size = nrow(d_scada), replace = TRUE)
scada <- split(d_scada, idx)
```

Next, we visualize in Figure \@ref(fig:plot-splits) the obtained data
splits.

```{r plot-splits, layout = "l-page", echo=TRUE, fig.show = "hold", fig.dim=c(12, 4), fig.width = "100%", out.width="100%", fig.cap = "Splits of the original dataset. Left: Proper training data to fit each model. Middle: Proper calibration data for computation of nonconformity scores. Right: Test split to evaluate empirical coverage of each prediction interval."}
cols <- c(
  palette.colors(3, palette = "Paired"),
  palette.colors(2, palette = "Okabe-Ito")
)
par(mfrow = c(1, 3))
with(scada, {
  plot(y ~ x, data = train,
     col = cols[1], xlab = "x", ylab = "y", main = "Training", pch = 20)
plot(y ~ x, data = calib,
     col = cols[2], xlab = "x", ylab = "y", main = "Calibration", pch = 20)
plot(y ~ x, data = test,
     col = cols[3], xlab = "x", ylab = "y", main = "Testing", pch = 20)
})
```

## Modeling

In this subsection the models to perform predictions for the data set
will be introduced. Three models will be fitted to the data to exemplify
different workflows for conformal prediction. The first model is a
linear model with quadratic term and will be implemented with the well
known `lm()` function, the second model will be a monotone generalized
additive model fitted using the `scam()` function from the **scam**
package, and for users familiar with the tidymodels framework, the third
model will be a random forest regression model fitted using functions
from the ecosystem.

### Model: Linear model

For simplicity, we first model the response variable using a simple
linear regression (SLR) model with a quadratic term for the covariate
using the training data.

$$
Y  = \beta_0 + \beta_1\text{X} + \beta_2\text{X}^2 + \epsilon,\quad \epsilon\sim N(0,\sigma^2).
$$ In R, the following call fits the model described above.

```{r baseR-lm, echo=TRUE, warning=FALSE, message=FALSE}
linear_model <- lm(y ~ x + I(x^2), data = scada$train)
```

### Model: SCAM

Given the nonlinear shape of the relationship between the response and
the covariate of interest, we also fit an alternative model, namely a
Shape Constrained Additive Model (SCAM), which offers greater
flexibility in capturing the form of the regression function and
includes the option to add monotonicity constraints to the fitted
regression function. The model is of the form

$$
\text{Y} = f(\text{X}) + \epsilon, \quad \epsilon \sim N(0,\sigma^2),
$$ where $f(\text{X}) = \sum_{j=1}^{k}b_j(\text{X})\beta_j$ and
$b_j(\cdot)$ denotes the basis function, taken here to be cubic splines.

To fit the model above, we use the functions `s()`and `scam()` from the
**scam** package [@scam] to specify the smooth term and fit the model to the
training data, respectively. To specify the monotone increasing smooth
term, we call the `s(x, bs = "mpi")` in the right hand side of the
formula; here the argument `bs = "mpi"` indicates the monotonicity
constraint. Additionally, we have specified the argument
`family = Gamma(link = "identity")` to select the Gamma distribution as
the likelihood for the data with identity link, as energy production is
non-negative.

```{r reg-scam, echo=TRUE, message=FALSE, warning=FALSE}
library(scam)
scam_model <- scam(y ~ s(x, bs = "mpi"), family = Gamma(link = "identity"),
                   data = scada$train)
```

### Model: Random forest regression

As a non-parametric alternative that makes fewer assumptions about the
underlying distribution and functional form, we also consider a Random
Forest (RF) regression model. Random Forest is an ensemble learning
method that captures complex, non-linear interactions by aggregating the
predictions of multiple decision trees. This model can be expressed as
$$
\text{Y} = \dfrac{1}{B}\sum_{b=1}^{B}T_{b}(X,\theta_b) + \epsilon
$$ Here, the regression function has the form
$f(X) = \dfrac{1}{B}\sum_{b=1}^{B}T_{b}(X,\theta_b)$ where each
$T_{b}(\cdot)$ denotes an individual decision tree fitted using a
bootstrap sample of the training data. To fit this model, we will
illustrate the usage of the tidymodels framework for those users
familiar with it. In this framework, we will use the **parsnip** package to
specify the model, we will set the engine as ranger to specify such
package as backend for model fitting.

```{r rf-model-spec, echo=TRUE, warning=FALSE, message=FALSE}
library(tidymodels)
library(parsnip)
rf_spec <- rand_forest(trees = 154, min_n = 813) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")
rf_wf <- workflow() %>%
  add_model(rf_spec) %>%
  add_formula(y ~ x)
rf_fit <- rf_wf %>% fit(data = scada$train)
```

The parameters `trees = 154` and `min_n = 813` were selected via 5-fold cross-validation on the training split. The argument `min_n` was optimized first by setting the `trees` argument to its default (`trees = 500`). After the optimal value was found, the `trees` parameter was optimized by setting `min_n = 813`, the optimal value from the previous step.

### Prediction intervals

For the computation of prediction intervals powered by conformal
prediction, we will use the packages **pintervals** and
**marginaleffects** for all the fitted models, while the **probably**
package will be used for the model fitted with the tidyverse framework.
We will focus on split conformal prediction to reduce the computational
load. This will help us illustrate the differences between the two
packages.

The **pintervals** package follows a package agnostic implementation for
the computation of the intervals. The function for split conformal
prediction is `pinterval_conformal()`, which requires as arguments a
vector of predictions made by a model on the data for which the
intervals will be computed, a vector of predictions on the calibration
set, the vector of true values of the response on the calibration set,
and the confidence level. These arguments correspond to the argument
`pred`, `calib`, `calib_truth`, and `alpha`. The `pinterval_conformal`
handles the computation of the nonconformity scores and returns the
prediction intervals as a tibble.

```{r pintervals-cp, echo=TRUE, eval=FALSE}
# This step takes on average 2 minutes on a MacBook Pro machine with 16 GB of 
# RAM and a 2.3 GHz 8-Core Intel(R) Core(TM) i9-9889H CPU.
library(pintervals)
lm_pintervals <- pinterval_conformal(
  pred = predict(linear_model, newdata = scada$test),
  calib = predict(linear_model, newdata = scada$calib),
  calib_truth = scada$calib$y
)
scam_pintervals <- pinterval_conformal(
  pred = predict(scam_model, newdata = scada$test),
  calib = predict(scam_model, newdata = scada$calib),
  calib_truth = scada$calib$y
)
rf_pintervals <- pinterval_conformal(
  pred = predict(rf_fit, new_data = scada$test)$.pred,
  calib = predict(rf_fit, new_data = scada$calib)$.pred,
  calib_truth = scada$calib$y
)
```

```{r load-pintervals, echo=FALSE, message=FALSE, warning=FALSE}
library(here)
# saveRDS(lm_pintervals, here("data", "intermediate", "lm-pintervals.rds"))
# saveRDS(scam_pintervals, here("data", "intermediate", "scam-pintervals.rds"))
# saveRDS(rf_pintervals, here("data", "intermediate", "rf-pintervals.rds"))
lm_pintervals <- readRDS("data/intermediate/lm-pintervals.rds")
scam_pintervals <- readRDS("data/intermediate/scam-pintervals.rds")
rf_pintervals <- readRDS("data/intermediate/rf-pintervals.rds")
```

When using the **marginaleffects** package to compute the prediction
intervals, we need to use the `predictions` and `inferences` functions
from the package. The arguments to the first function should be a fitted
model and covariate values at which predictions will be computed in our
case they should be specified by `model = scam_model` and
`newdata = scada$test`. The `inferences` function is used to specify the
method employed to compute uncertainty estimates around the predictions
from the model. The three required arguments by this function are the
method, in our case `method = "conformal_split"`, the training data and
the calibration data specified by `data_train = scada$train` and
`data_calib = scada$calib`.

```{r me-intervals, echo=TRUE}
library(marginaleffects)
linear_meint <- predictions(model = linear_model, newdata = scada$test) |>
  inferences(
    method = "conformal_split",
    data_train = scada$train,
    data_calib = scada$calib
  )

scam_meint <- predictions(model = scam_model, newdata = scada$test) |>
  inferences(
    method = "conformal_split",
    data_train = scada$train,
    data_calib = scada$calib
  )
```

Finally, the **probably** package can be used to obtain prediction
intervals for models fitted with the tidymodels framework. The
nonconformity scores for split conformal prediction are computed using
the `int_conformal_split()` function. The arguments of this function
include the fitted model workflow and the calibration split. Then, to
retrieve the prediction intervals we use the `predict()` function by
providing the nonconformity scores obtained by `int_conformal_split()`
function, the predictor values for which we want to get prediction
intervals, and the desired confidence level.

```{r probably-intervals, echo=TRUE}
library(probably)
rf_printervals <- rf_fit |>
  int_conformal_split(cal_data = scada$calib) |>
  predict(new_data = scada$test, level = 0.95)
```

### Plot predictions and prediction intervals

After the results for each method have been obtained, plots of the
estimated regression lines together with their corresponding prediction
intervals are shown in Figure \@ref(fig:plot-predintcp). The code used
to generate the figure is provided below.

```{r plot-bands-me-function, echo=TRUE}
plot_pred <- function(pred, pred.low, pred.high, newdata, label ="Split", model = "LM") {
  p <- data.frame("x" = newdata$x, "y" = newdata$y, "pred" = pred, 
                  "pred.low" = pred.low, "pred.high" = pred.high)

  p <- p[order(p$x), ]
  with(p, {
    coverage <- mean(y >= pred.low & y <= pred.high) * 100
    width <- median(pred.high - pred.low)
    caption <- sprintf("%s + %s\nCoverage = %.2f%%\nMedian width = %.2f",
                       model, label, coverage, width)
    plot(y~x, col = cols[3], pch = 19, cex =.5, main = caption)
    lines(pred~x)
    lines(x, pred.high, lwd = 1.5, col = cols[5])
    lines(x, pred.low, lwd = 1.5, col = cols[5])
  })
}
```

```{r plot-predintcp, echo=FALSE, layout = "l-page", fig.show="hold", fig.dim=c(8, 12), fig.width="80%", out.width="100%", fig.cap= "Point predictors depicted with black color and its corresponding 95% prediction intervals in dark orange. Top row: Prediction intervals computed with the pintervals package for the linear model with quadratic term (left), and the marginaleffects package (right). Middle row: Conformal prediction intervals for the SCAM model using pintervals (left), and marginaleffects (right). Bottom row: Prediction intervals for the random forest model produced by the pintervals package (left), and the probably package (right)."}
par(mfrow = c(3, 2))
# Intervals LM Model
plot_pred(pred = lm_pintervals$pred, 
          pred.low = lm_pintervals$lower_bound,
          pred.high = lm_pintervals$upper_bound, 
          newdata = scada$test)
plot_pred(pred = linear_meint$estimate, 
          pred.low = linear_meint$pred.low,
          pred.high = linear_meint$pred.high, 
          newdata = scada$test)
# Intervals SCAM
plot_pred(pred = scam_pintervals$pred, 
          pred.low = scam_pintervals$lower_bound,
          pred.high = scam_pintervals$upper_bound, 
          newdata = scada$test, model = "SCAM")
plot_pred(pred = scam_meint$estimate, 
          pred.low = scam_meint$pred.low,
          pred.high = scam_meint$pred.high, 
          newdata = scada$test, model = "SCAM")
# Intervals random forest
plot_pred(pred = rf_pintervals$pred, 
          pred.low = rf_pintervals$lower_bound,
          pred.high = rf_pintervals$upper_bound, 
          newdata = scada$test, model = "Random forest")
plot_pred(pred = rf_printervals$.pred, 
          pred.low = rf_printervals$.pred_lower,
          pred.high = rf_printervals$.pred_upper, 
          newdata = scada$test, model = "Random forest")
```

# Conformal prediction methods for time series forecasting

The R package **conformalForecast** [@conformalForecast] implements the
AcMCP method developed in @Wang_Hyndman_2024 and some extensions of
several other popular conformal prediction methods for multi-step
forecasting. An attractive feature of the package design is the attempt
to separate the model specification and fitting from the conformal
prediction step. This structure reflects the core principle of conformal
prediction--- its flexibility and ability to wrap around almost any
model. In particular, the functions for conformal prediction require the
user to provide a function capable of generating predictions (a model).
This function is then used to obtain predictions on the (online)
calibration set, from which conformal prediction intervals are
constructed using any of the methods implemented in the package.

In this example, a time series of the weekly number of [dengue
cases](https://github.com/uriel278/tutorials-cp) in São Paulo, Brazil is
considered. The data has been obtained from the surveillance system
[InfoDengue](https://info.dengue.mat.br) [@codeco2018] and contains the
number of dengue cases spanning 678 weeks from 2010 to 2022 (52 weeks
per year, with the exception of 2012 and 2017, which each contains 53
weeks).

## Data preparation {#data-prep-section}

We begin by reading the dengue time series data with `read.csv()`, and
then we create a time series object with the `ts()` function using the
weekly number of cases. As mentioned above, there are two years in which
the number of observations is 53 instead of 52. For visualization
purposes, we specify `frequency = 52` in the `ts()` function, which
simplifies the visualization despite these irregularities. An
alternative approach could involve using `frequency = 52.28571` to
account for leap years, or imputing the 53rd week's data by averaging it
with the 52nd week, if appropriate for the context of the problem and
data at hand. For this illustration, however, we prioritize simplicity
and clarity in illustrating how to calculate intervals, so we proceed
with `frequency = 52`. Importantly, in the model building step (see
section [Modeling and prediction](#model-pred-sect)), the `frequency`
argument is not used, as seasonality is explicitly defined in the model.
If a model were to depend on this argument, the user would need to
adjust the frequency accordingly to ensure valid results.

A summary of the data shows weekly dengue cases range from 38 to 12, 367
cases.

```{r load-data-ts, message=FALSE, echo=TRUE}
library(readr)
library(curl)
url_dengue <- "https://raw.githubusercontent.com/uriel278/tutorials-cp/main/data/dengue.csv"
d_dengue <- read_csv(curl(url_dengue))
d_dengue <- ts(d_dengue$x, start = c(2010, 1), frequency = 52)
summary(d_dengue)
```

For this time series, the data from 2010-2013 will play the role of the
initial training set, the data from 2014-2017 will be used for
calibration, and the data from 2018-2022 will serve as the testing set
for evaluation of empirical coverage. The **conformalForecast** package
splits the data automatically, as described in section [Prediction
intervals](#prediction-bands-section) below, ensuring consistency with
the setting illustrated in Figure \@ref(fig:plot-ts-splits).

For visualization purposes, the initial training, calibration, and
testing splits can be created manually using the `window()` function by
providing the start and end dates of each segment. The resulting initial
splits are displayed in the Figure \@ref(fig:plot-ts-splits), each
represented with a different color.

```{r dengue-ts-time-split, echo=TRUE}
dtrain_dengue <- window(d_dengue, start = c(2010, 1), end = c(2013, 52))
dcal_dengue <- window(d_dengue, start = c(2014, 1), end = c(2017, 52))
dtest_dengue <- window(d_dengue, start = c(2018, 1))
```

```{r plot-ts-splits, layout = "l-page", fig.show = "hold", fig.dim=c(8, 4), fig.width = "100%", out.width="100%", fig.cap = "Splits of the dengue time series for the example. The initial training split (in light blue), calibration set (in dark blue) and the testing set in dashed green lines.", echo=FALSE}
cols <- c(
  palette.colors(3, palette = "Paired"),
  palette.colors(2, palette = "Okabe-Ito")
)
plot(d_dengue*NA, ylim = c(0, 12500), ylab = "# Cases")
lines(dtrain_dengue, col = cols[1], lwd = 2)
lines(dcal_dengue, col = cols[2], lwd = 2)
lines(dtest_dengue, col = cols[3], lwd = 2, lty = 5)
legend('topright', legend=c('Initial training','Calibration', 'Testing'),
       col = cols, lty = c(1, 1, 3), lwd = 2)
```

## Modeling and prediction {#model-pred-sect}

The **conformalForecast** package separates the model specification and
fitting from the process of computing conformal prediction intervals.
Users provide a forecasting function that generates point forecasts for
the calibration set. These forecasts are then used to compute
nonconformity scores and to construct conformal prediction intervals.

It is important to remark that the main function from the
**conformalForecast** package, `cvforecast()`, does not require the user
to pass pre-defined data splits. The splits generated above are only for
visualization purposes. Instead, the `cvforecast()` function generates
splits following the sequential split framework described in Section
[Multi-step Conformal Prediction](#multi-step-section).

### The `cvforecast()` function

The `cvforecast()` function is the main function of the
**conformalForecast** package. As arguments it requires a time series
object via the `y` argument, a forecasting function with the
`forecastfun` argument, the forecasting horizon `h`, and the confidence
level(s) `level` for the prediction interval. The `level` argument
accepts either a single numerical value or a vector of numerical values
to compute prediction intervals at multiple confidence levels. However,
caution is required when providing a vector, as some conformal
prediction methods available in the package cannot handle multiple
significance values and may return `NA` values for prediction intervals
without warning. All methods work correctly when a single numerical
input is provided, although this limitation is not documented.
Additional arguments are available, and readers are encouraged to see
the package documentation for details.

The `cvforecast()` function applies the `forecastfun` to the data,
computes predictions for each forecasting horizon given sequential
information based on sequential splits, and computes nonconformity
scores (in this case, residuals). The returned object is of class
`cvforecast` and `forecast`.

This function only computes predictions that are required by the
conformal prediction methods available in the package, it does not
compute any conformal prediction intervals directly. However, if the
forecasting function supplied via `forecastfun` produces its own
prediction intervals, these will be stored in the returned object.

An example of a call to the `cvforecast()` function is as follows.

```{r cvforecast-norun, echo=TRUE, eval = FALSE}
pred <- cvforecast(y = d_dengue, forecastfun = forecastfunction,
                   h = 4, level = 95, window = 52*4)
```

Here, the data `d_dengue` contains the dengue time series. The
forecasting function is provided through the argument `forecastfun`
(details are explained in the following subsection). The horizon `h = 4`
requests four-step-ahead forecasts, and `level = 95` specifies 95%
prediction intervals from the forecasting function. The argument
`window = 52*4` restricts the training window to four years of data.

### Model specification and the `forecastfun` argument

Here, we specify the function used to obtain point forecasts. In this
illustration example, we use a Seasonal ARIMA (SARIMA) model, as in the
plots we can observe some temporal patterns that repeat over time. The
SARIMA model for a time series $Y_t$ is of the form:

$$
\Phi_P(B)\phi_p(B)\nabla^{d}\nabla^{D}_{S}Y_t=\Theta_{Q}(B)\theta_q(B)W_t,
$$ where $B$ is a lag operator, $\Phi(\cdot)$ and $\phi(\cdot)$ are the
seasonal autoregressive and autoregressive polynomials of orders $P$ and
$p$, respectively; $\Theta_Q(\cdot)$ and $\theta_q(\cdot)$ are the
seasonal moving average and moving average polynomials of orders $Q$ and
$q$, respectively; $\nabla^D_S$ and $\nabla^d$ are the seasonal
difference and difference operators of orders $(D,S)$ and $d$,
respectively, and $W_t$ is a white noise process.

The code below shows how to define a forecasting function,
`forecastfunction()`, using the `Arima()` function of the **forecast** R
package. This function will later be passed to `cvforecast()` to compute
forecasts as the `forecastfun` argument. To be compatible, any
forecasting function defined here must accept three arguments: a time
series `x`, the forecasting horizon `h`, and the confidence level
`level`. Failure to do so will result in errors when calling the
`cvforecast()` function. Note that this requirement does not pose a
limitation in the forecasting models that can be used, as the user can
always create a wrapping function mapping these arguments to the
corresponding inputs of the model of interest. In our example, the
wrapper combines the `Arima()` function, which estimates the model given
a dataset and specification arguments, with the `forecast` function,
which produces multi-step forecasts and confidence intervals.
Furthermore, note that before the prediction/forecast step we fit the
model. This is not a mandatory step, we may have an already fitted model
and can call only the corresponding prediction function. A complete
example of the `forecastfunction()` is shown below.

```{r sarima-spec, echo=TRUE, warning=FALSE}
library(forecast)
forecastfunction <- function(x, h, level){
  Arima(x, order = c(1, 0, 1), 
    seasonal = list(order = c(0, 1, 0), period = 52),
    method = "ML") |>
    forecast(h = h, level = level)
}
```

We can verify that this function is working well by applying it to the
training set `dtrain_dengue` to obtain four-step-ahead forecasts with
95% prediction intervals. These intervals are generated by the
`forecast()` function of the **forecast** package, applied to the object
returned by `Arima()`, and are not conformal prediction intervals.

```{r sarima-trial-train-split, echo=TRUE}
forecastfunction(x = dtrain_dengue, h = 4, level = 95)
```

Despite being very flexible, the **conformalForecast** requires objects
of type `forecast` with a certain structure to be able to work properly.
In the following, using the **caretForecast**, we will illustrate how to
go over the issues by extracting the required fields from a model, and
creating a similar object for the **conformalForecast** package to work
with. The second model will be fitted using nonnegative least squares,
which restrict the fitted regression function to be nonnegative; in the
**caretForecast** package all the models have an autoregressive
structure.

To fit the model, we specify the `predfun` as follows

```{r parse-fun-eval, echo = FALSE, eval=TRUE}
caretToforecast <- function(caretForecastObj){
  level <- caretForecastObj$level
  colnames(caretForecastObj$lower) <- paste0(level, "%")
  colnames(caretForecastObj$upper) <- paste0(level, "%")
  standardize_fc <- structure(
    list(
      "method" = caretForecastObj$method,
      "level" = caretForecastObj$level,
      "mean" = caretForecastObj$mean,
      "lower" = caretForecastObj$lower,
      "upper" = caretForecastObj$upper,
      "fitted" = caretForecastObj$fitted
    ),
    class = "forecast"
  )
  return(standardize_fc)
}
```

```{r predfun-nnls, echo=TRUE, warning=FALSE, message=FALSE}
library(caretForecast)
nnlsreg <- function(x, h, level = 0.95){
  fit <- ARml(x, max_lag = 52, caret_method = "nnls", 
              verbose = FALSE, seasonal = FALSE, fixed_window = TRUE, 
              cv_horizon = h, initial_window = 52)
  fc <- forecast(fit, h = h, level = level)
  caretToforecast(fc)
}
```

The function `caretToforecast()` is specified in the following, where
the fields of the forecast object produced by the resulting fit from the
**caretForecast** are parsed to the expected format of the
**conformalForecast**, in this case the `lower` and `upper` fields must
have as column names the coverage specified by the user.

```{r parser-fun-echo, echo = TRUE, eval=FALSE}
caretToforecast <- function(caretForecastObj){
  level <- caretForecastObj$level
  colnames(caretForecastObj$lower) <- paste0(level, "%")
  colnames(caretForecastObj$upper) <- paste0(level, "%")
  standardize_fc <- structure(
    list(
      "method" = caretForecastObj$method,
      "level" = caretForecastObj$level,
      "mean" = caretForecastObj$mean,
      "lower" = caretForecastObj$lower,
      "upper" = caretForecastObj$upper,
      "fitted" = caretForecastObj$fitted
    ),
    class = "forecast"
  )
  return(standardize_fc)
}
```

## Prediction intervals {#prediction-bands-section}

Now that the `cvforecast()` function and the `forecastfun` argument have
been introduced, we proceed to obtain prediction intervals with
conformal prediction. The first step is to call the `cvforecast()`
function with the required arguments, as shown below.

```{r cvforecast-forecast, echo=TRUE, eval=FALSE, warning=FALSE}
# This step takes on average 4 minutes and 20 seconds on a MacBook Pro machine with 16 GB of 
# RAM and a 2.3 GHz 8-Core Intel(R) Core(TM) i9-9889H CPU.
library(conformalForecast)
pred_arima <- cvforecast(d_dengue, forecastfunction,
                   h = 4, level = 95, initial = 1, window = 52*4)
pred_nnls <- cvforecast(d_dengue, nnlsreg, 
                        h = 4, level = 95, window = 52*4)
```

```{r load-cvforecast, echo=FALSE, eval = TRUE}
library(here, verbose = FALSE)
library(conformalForecast)
# saveRDS(pred_arima, file = here("data", "intermediate", "cvpred-arima.rds"))
# saveRDS(pred_nnls, file = here("data", "intermediate", "cvpred-nnls.rds"))
pred_arima <- readRDS("data/intermediate/cvpred-arima.rds")
pred_nnls <- readRDS("data/intermediate/cvpred-nnls.rds")
```

Once forecasts have been obtained using `cvforecast()`, prediction
intervals can be generated using conformal prediction methods
implemented in **conformalForecast** package, see @Wang_Hyndman_2024 for
more details and a complete list of available algorithms.

In this tutorial, we focus on the AcMCP method developed by the authors,
implemented via the `acmcp()` function, and the extension of the Adaptive Conformal Prediction (ACP) method implemented in the function `acp()`. 

The AcMCP method applies the update rule described in Equation \@ref(eq:updtrule-wang). Its first argument
is the object of class `cvforecast` returned by the `cvforecast()`
function, which in our case is stored as `pred`. The calibration set
size should also be specified; we set `ncal = 52*4` to match the four
years of calibration data (2014–2017). In addition, we set
`rolling = TRUE` to use a rolling window for quantile construction. With
these settings, the initial training set covers 2010-2013, while the
initial calibration set corresponds to 2014-2014, as described earlier
in subsection [Data preparation](#data-prep-section).

The `acmcp()` function also accepts several additional parameters, which
are documented in detail in @Wang_Hyndman_2024 and in the package
documentation. Briefly, the learning rate parameter `lr` controls the
step size of the update and is set here to `lr = 0.1`. The parameter
`KI` --- specific to this conformal method --- is a positive constant in
order to place the integrator on the same scale as the nonconformity
scores. If prior knowledge about the scale of the scores is available,
it can be used to set `KI`; in this example, we set `KI = 2000`.
Finally, the parameter `Csat` is a positive constant ensuring that by
time $T$, the absolute coverage guarantee is at least $1-\alpha-\delta$,
where $\delta$ reduces the desired coverage guarantee $1-\alpha$
because, unlike traditional conformal prediction methods that are
conservative in finite samples, the coverage guarantees here are
asymptotic. This reduction $\delta$ is *controlled* by the argument
`delta` and can be interpreted as the amount of coverage one is willing
to forgo given that we cannot observe the process for an infinite
period of time. The expression for `Csat` in the `acmcp` function is
defined in Appendix C of @angelopoulos2023.

```{r acmcp-bands, echo=TRUE, eval=TRUE}
# This step takes on average 16 seconds on a MacBook Pro machine with 16 GB of RAM and a 
# 2.3 GHz 8-Core Intel(R) Core(TM) i9-9889H CPU.
arima_acmcp <- acmcp(pred_arima, ncal = 52*4, rolling = TRUE,
              lr = 0.1, Tg = length(d_dengue), delta = 0.01, KI = 2000)
nnls_acmcp <- acmcp(pred_nnls, ncal = 52*4, rolling = TRUE,
              lr = 0.1, Tg = length(d_dengue), delta = 0.01, KI = 2000)
```

For the ACP method, we require fewer parameters. In addition to the nonconformity scores, only two parameters are necessary: a learning rate `gamma`, similar to `lr`, and the size of the calibration set `ncal`.

```{r acp-bands, echo=TRUE, eval=TRUE}
# This step takes on average 7 seconds on a MacBook Pro machine with 16 GB of RAM and a 
# 2.3 GHz 8-Core Intel(R) Core(TM) i9-9889H CPU.
arima_acp <- acp(pred_arima, gamma = 0.1, ncal = 52*4)
nnls_acp <- acp(pred_nnls, gamma = 0.1, ncal = 52*4)
```

### Plot predictions and prediction intervals

We plot the prediction intervals for each forecasting horizon as
follows, and the result is displayed in Figure
\@ref(fig:mcp-bands-plot01). We can observe that the width of the
prediction intervals grows as the forecasting horizon increases, this
is an expected behavior as extending the forecast horizon generally
increases forecast uncertainty.

```{r plot-function-conformalForecast, echo=TRUE}
plot_cpForecast <- function(obj, plot_h, ts_train, ts_cal, ts_test, ts,
                            model, label = "AcMCP"){
  coverage_mean <- coverage(obj, level = 95, window = 52)
  width_mean <- width(obj, window = 52)
  for(h in plot_h){
    caption <- sprintf("%s + %s,h = %s\nCoverage = %.2f%%\nMean width = %.2f",
                       model, label, h, coverage_mean$mean[h], width_mean$mean[h])
    plot(ts*NA, ylim = c(-500, 12500), ylab = "# Cases",
     main = caption)
    lines(ts_train, col = cols[1], lwd = 2)
    lines(ts_cal, col = cols[2], lwd = 2)
    lines(ts_test, col = cols[3], lwd = 2, lty = 3)
    lines(obj$LOWER$`95%`[, h], col = cols[5])
    lines(obj$UPPER$`95%`[, h], col = cols[5])
  }
  }
```

```{r mcp-bands-plot01, layout = "l-page", fig.show = "hold", fig.dim=c(8, 6), fig.width = "100%", out.width="100%", fig.cap = "Prediction intervals (in orange) using AcMCP on the dengue data for two forecasting horizon for the ARIMA (first row) and NNLS models (second row).", echo = FALSE}
par(mfrow = c(2, 2))
plot_cpForecast(arima_acmcp, , plot_h = c(1,4), 
                ts_train = dtrain_dengue, ts_cal = dcal_dengue,
                ts_test = dtest_dengue, ts = d_dengue, 
                model = "ARIMA")
plot_cpForecast(nnls_acmcp, , plot_h = c(1,4), 
                ts_train = dtrain_dengue, ts_cal = dcal_dengue,
                ts_test = dtest_dengue, ts = d_dengue, 
                model = "NNLS")
```

```{r acp-bands-plot01, layout = "l-page", fig.show = "hold", fig.dim=c(8, 6), fig.width = "100%", out.width="100%", fig.cap = "Prediction intervals (in orange) using ACP on the dengue data for two forecasting horizon for the ARIMA (first row) and NNLS models (second row).", echo = FALSE}
par(mfrow = c(2, 2))
plot_cpForecast(arima_acp, plot_h = c(1,4), 
                ts_train = dtrain_dengue, ts_cal = dcal_dengue,
                ts_test = dtest_dengue, ts = d_dengue, 
                model = "ARIMA", label = "ACP")
plot_cpForecast(nnls_acp, plot_h = c(1,4), 
                ts_train = dtrain_dengue, ts_cal = dcal_dengue,
                ts_test = dtest_dengue, ts = d_dengue, 
                model = "NNLS", "ACP")
```

# Discussion

The examples presented in this work demonstrate how the packages
**pintervals**, **marginaleffects**, **probably**, and
**conformalForecast** enable the practical implementation of conformal
prediction for regression and time series forecasting, respectively. The
main strength of these packages is their modular design, which
facilitates their integration with existing R packages, and broader
modeling ecosystems such as **tidymodels**. This modularity allows
practitioners to incorporate conformal prediction into existing
workflows with minimal changes, making distribution-free uncertainty
quantification more accessible in applied settings.

This modularity allows practitioners to incorporate conformal prediction
into existing workflows with minimal disruption, making
distribution-free uncertainty quantification more accessible in applied
settings. The **pintervals** package shows progress in this direction by
including multiple nonconformity measures and allowing users to define
custom ones, providing a pathway toward more general and adaptable
tools, according to their documentation. During the development of this
tutorial, however, we identified practical limitations. In particular,
computing split conformal prediction intervals can be computationally
expensive in some settings, and the package may return missing values
for prediction intervals without providing clear warnings or diagnostic
messages. Addressing these issues in future releases would further
improve reliability for both practitioners and researchers.

The **marginaleffects** and **probably** packages provide efficient
implementations of split conformal prediction through functions designed
to integrate naturally into their respective frameworks. These tools
fill an important gap by enabling conformal prediction within widely
used modeling pipelines. The**marginaleffects** package may be
particularly well suited for users working in local computing
environments, while **probably** may be more suitable for
production-oriented workflows. Because both packages are part of larger
modeling ecosystems with standardized objects and interfaces, they
support a wide range of models and simplify the computation of
nonconformity scores and prediction intervals. However, this
framework-based design can also introduce limitations. Models outside
these ecosystems may require additional development effort to be
compatible, which can pose challenges for users without experience
extending these frameworks. Future developments could focus on improving
extensibility and reducing the technical barrier for integrating custom
models.

For time series forecasting, the **conformalForecast** package offers a broad
set of algorithms for online conformal prediction. The package includes
standardized functions for computing, storing, and retrieving results,
facilitating reproducible workflows. Currently, the package relies
primarily on the forecast modeling framework. While this does not
represent a major practical limitation, as users can construct
compatible objects manually as demonstrated in our example, the
development of native interfaces to additional ecosystems such as
**fable** [@fable] would further enhance usability and adoption.

Beyond the packages considered in this tutorial, several specialized
conformal prediction packages are available on CRAN, including
**conformalbayes** [@conformalbayes],
**conformalInference.fd** [@conformalfd],
**conformalInference.multi** [@conformalmulti],
**conformalpvalue** [@conformalpval],
**ConformalSmallest** [@ConformalSmallest], **fabPrediction** [@fabpred],
**ClusTorus** [@ClusTorus]. These packages often provide computationally
efficient implementations tailored to specific conformal prediction
variants, though sometimes at the expense of model-agnostic flexibility.
Together, they reflect the rapid evolution of conformal prediction
methodology and software tools.

From a practical standpoint, users seeking seamless integration of
conformal prediction into existing regression workflows may benefit most
from packages embedded in large modeling ecosystems. In contrast, users
requiring specialized conformal prediction variants or optimized
implementations for large-scale problems may prefer dedicated conformal
prediction packages, even if this reduces model flexibility. The choice
of tool therefore depends strongly on the intended application,
computational constraints, and existing modeling infrastructure. Looking
forward, further progress will likely depend on the development of
standardized data structures for storing intermediate results across the
multiple stages of conformal prediction. Equally important will be the
design of interfaces that clearly define interactions between predictive
models, nonconformity measures, score computation, and prediction
interval construction. Advances in these areas could significantly
improve flexibility while maintaining coherence across software
implementations.

# Conclusions

In this paper, we showcase the **pintervals**, **marginaleffects**,
**probably**, and **conformalForecast** packages for regression and time
series forecasting, respectively. A key contribution of these tools is
the separation of predictive model fitting from conformal inference
procedures, allowing users to construct prediction intervals for a wide
range of models without relying on distributional assumptions. These
packages are designed for practical use and integrate naturally with
existing tools in the R ecosystem.

Together, these packages highlight the importance of modular software
development for statistical methods and conformal prediction naturally
lends itself to such design principles, since multiple interacting
components --- the predictive model, the choice of nonconformity
measure, the computation of nonconformity scores, and the construction
of prediction intervals --- must work together in a flexible yet
coherent way, their design sets them apart from other existing packages
that restrict the applicability of conformal prediction to a narrow
class of models and offers practitioners a straightforward path to add
distribution-free uncertainty quantification to their existing
prediction and forecasting workflows.

A current limitation shared by these packages, and by many conformal
prediction tools more broadly, is the restricted set of built-in
nonconformity measures. Future development could focus on expanding the
available measures and improving support for user-defined nonconformity
functions, further increasing flexibility across application domains.

Overall, these four packages represent an important step toward making
reliable, model-agnostic prediction intervals more accessible to the R
community. As software ecosystems continue to mature, conformal
prediction is likely to become an increasingly standard component of
predictive modeling workflows, lowering the barrier for statisticians
and data scientists to incorporate principled uncertainty quantification
into applied prediction and forecasting tasks.

# Acknowledgments

Xiaoqian Wang was supported by the Presidential Foundation of the Academy of Mathematics and Systems Science, Chinese Academy of Sciences, China (No. E555930101).
