---
title: "Supplementary information"
toc: yes
toc_depth: 3
number_sections: yes
output:
  pdf_document:
    highlight: default
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{colortbl}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage{threeparttablex}
- \usepackage{makecell}
- \usepackage{amsthm}
- \usepackage{fancyhdr} 
- \usepackage{hyperref}
- \usepackage{setspace}
- \usepackage{geometry}
 \geometry{
 a4paper,
 total={170mm,200mm},
 left=20mm,
 top=20mm,
 bottom=30mm,
 }
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(puzzle)
library(knitr)
library(kableExtra)
library(magrittr)
```

<br />

\newpage

# Example of dataset for bolus administration

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_start)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(0),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(0),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset for oral administration

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_start)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset for parallel zero and first order absorption

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_start)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(0,1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(0,1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset for sequential zero and first order absorption

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_start)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(0,1),
       parallel = F,
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(0,1),
             parallel = F,
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset for complex absorption

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_start)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1,1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1,1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset with adding extra times

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_start)),
          dose = as.data.frame(puzzle::df_dose_start),
          cov = as.data.frame(puzzle::df_cov_start))
extra_times = as.data.frame(puzzle::df_extra_times)
readr::write_csv(extra_times,"extra_times.csv")
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov),
       extratimes=list(name="extra_times.csv"))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov),
       extratimes=list(name="extra_times.csv")) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset with time dependent covariates

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_start)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_time_dependent_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(0),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(0),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset with parent drug and metabolite

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_parent),
                    metabolite=as.data.frame(puzzle::df_pk_metabolite)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_time_dependent_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset with PK and PD

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_start)),
          pd = list(response=as.data.frame(puzzle::df_pd_start)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_time_dependent_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       pd=list(data=nm$pd), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk),
             pd=list(data=nm$pd), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset with PK and PD with multiple entities and responses

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(parent=as.data.frame(puzzle::df_pk_parent), 
                    metabolite=as.data.frame(puzzle::df_pk_metabolite)),
          pd = list(response1=as.data.frame(puzzle::df_response1),
                    response2=as.data.frame(puzzle::df_response2),
                    response3=as.data.frame(puzzle::df_response3)),
          dose=as.data.frame(puzzle::df_dose_start),
          cov=as.data.frame(puzzle::df_cov_time_dependent_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       pd=list(data=nm$pd), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk),
             pd=list(data=nm$pd), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov)) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset passing optional columns

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(pk=as.data.frame(puzzle::df_pk_optional_columns)),
          dose=as.data.frame(puzzle::df_dose_optional_columns),
          cov=as.data.frame(puzzle::df_cov_time_dependent_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov),
       optionalcolumns=c("OCC","TIMEPOINT","TRT"))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov),
             optionalcolumns=c("OCC","TIMEPOINT","TRT")) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset filling columns

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(pk=as.data.frame(puzzle::df_pk_optional_columns)),
          dose=as.data.frame(puzzle::df_dose_optional_columns),
          cov=as.data.frame(puzzle::df_cov_time_dependent_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov),
       optionalcolumns=c("OCC","TIMEPOINT","TRT"), 
       fillcolumns="TRT")
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov),
             optionalcolumns=c("OCC","TIMEPOINT","TRT"), 
             fillcolumns="TRT") %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset using coercion

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(pk=as.data.frame(puzzle::df_pk_optional_columns)),
          dose=as.data.frame(puzzle::df_dose_optional_columns),
          cov=as.data.frame(puzzle::df_cov_time_dependent_start))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov),
       coercion=list(name="coercion_file.txt"),
       optionalcolumns=c("OCC","TIMEPOINT","TRT"), 
       fillcolumns="TRT")
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov),
             coercion=list(name="coercion_file.txt"),
             optionalcolumns=c("OCC","TIMEPOINT","TRT"), 
             fillcolumns="TRT") %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10)%>%
  row_spec(0, angle = 90)
```


\newpage

# Example of dataset with EVID = 4

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(pk=as.data.frame(puzzle::df_pk)),
          dose=as.data.frame(puzzle::df_dose_evid4),
          cov=as.data.frame(puzzle::df_cov_evid4))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov),
       coercion=list(name="coercion_file.txt"),
       optionalcolumns=c("PERIOD","TRT"), 
       fillcolumns=c("PERIOD","TRT"))
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov),
             coercion=list(name="coercion_file.txt"),
             optionalcolumns=c("PERIOD","TRT"), 
             fillcolumns=c("PERIOD","TRT")) %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10) %>% 
  landscape()%>%
  row_spec(0, angle = 90)

```

\newpage

# Example of dataset using arrange

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(pk=as.data.frame(puzzle::df_pk)),
          dose=as.data.frame(puzzle::df_dose_evid4),
          cov=as.data.frame(puzzle::df_cov_evid4))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       cov=list(data=nm$cov),
       coercion=list(name="coercion_file.txt"),
       optionalcolumns=c("PERIOD","TRT"), 
       fillcolumns=c("PERIOD","TRT"),
       arrange="ID,PERIOD,TIME,desc(MDV),CMT")
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
             order=c(1),
             pk=list(data=nm$pk), 
             dose=list(data=nm$dose), 
             cov=list(data=nm$cov),
             coercion=list(name="coercion_file.txt"),
             optionalcolumns=c("PERIOD","TRT"), 
             fillcolumns=c("PERIOD","TRT"),
             arrange="ID,PERIOD,TIME,desc(MDV),CMT") %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 10) %>% 
  landscape()%>%
  row_spec(0, angle = 90)
```

\newpage

# Example of dataset using datetime

## Puzzle input
```{r, warning=F, message=F, error=F, results='asis', echo=T}
nm = list(pk = list(pk=as.data.frame(puzzle::df_pk_datetime)),
          dose=as.data.frame(puzzle::df_dose_datetime))
```

## Puzzle code
```{r, warning=F, message=F, error=F, results='hide', echo=T}
puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       datetimeformat="%Y-%m-%d %H:%M")
```

## Puzzle output
```{r, warning=F, message=F, error=F, results='asis', echo=FALSE}
df = puzzle(directory=file.path(getwd()),
       order=c(1),
       pk=list(data=nm$pk), 
       dose=list(data=nm$dose), 
       datetimeformat="%Y-%m-%d %H:%M") %>% 
  mutate_if(is.numeric, round, digits=1)
df[is.na(df)] <- "."
kable(df,
      "latex", booktabs = T,longtable=T)%>% 
  kable_styling(full_width = T,latex_options=c("scale_down","striped"), font_size = 8) %>% 
  landscape()%>%
  row_spec(0, angle = 90)
```

