---
title: "Importance benchmark results"
output:
  html_document:
    self_contained: false
    toc: true
    toc_float: true
---

```{r setup-appendix-01}
#| echo: false
knitr::opts_chunk$set(
	cache = FALSE,
	echo = FALSE,
	fig.path = "figures/importance/",
	fig.height = 9,
	fig.width = 10,
	fig.align = "center"
)
# Shrink rendered PNGs losslessly; skipped where optipng is unavailable
if (nzchar(Sys.which("optipng"))) {
	knitr::knit_hooks$set(optipng = knitr::hook_optipng)
	knitr::opts_chunk$set(optipng = "-o2 -quiet")
}

library(ggplot2)
library(data.table)
source("R/plotting.R")
source("R/clean-results.R")
importances <- readRDS("results/importance/importances.rds")
# Earlier SAGE runs were superseded by a re-run; keep the newest run per cell.
importances <- dedup_latest_run(importances)
```

Importance results in the form of boxplots split across FI method and learner type, separately for each dataset / DGP.
Table 1 lists all DGPs used in the importance benchmark.

```{r tab-importance-problems-supplement}
#| echo: false
library(kableExtra)
tibble::tribble(
	~DGP           , ~p , ~description                                          , ~model                                                                            ,
	"ewald"        ,  5 , "DGP used by Ewald et al. (2024)"                     , r"($Y = X_4 + X_5 + X_4 X_5 + \varepsilon$)"                                      ,
	"correlated"   ,  4 , "$X_1$ and $X_2$ correlated ($r = 0.25, 0.75$)"       , r"($Y = 2 X_1 + X_3 + \varepsilon$)"                                              ,
	"interactions" ,  5 , "$X_1$ and $X_2$ interact but have no direct effects" , r"($Y = 2 X_1 X_2 + X_3 + \varepsilon$)"                                          ,
	"friedman1"    , 10 , "Regression benchmark from mlbench"                   , r"($Y = 10 \sin(\pi X_1 X_2) + 20 (X_3 - 0.5)^2 + 10 X_4 + 5 X_5 + \varepsilon$)" ,
	"independent"  ,  5 , "Uncorrelated features with direct effects"           , r"($Y = 2 X_1 + X_2 + 0.5 X_3 + \varepsilon$)"                                    ,
	"confounded"   ,  4 , "Unobserved confounder $H$ with proxy"                , r"($Y = H + X_1 + \varepsilon$)"                                                  ,
	"mediated"     ,  4 , "Mediator masks exposure variable"                    , r"($Y = 1.5 \cdot \mathrm{mediator} + 0.5 \cdot \mathrm{direct} + \varepsilon$)"  ,
	"bike sharing" , 12 , "Real-world dataset              "                    , "N/A"
) |>
	kbl(
		col.names = c("Dataset", "p", "Description", "Model"),
		caption = "DGPs used for comparison, alongside their number of features (p) and a brief description. Unless otherwise specified, features in simulated data were drawn from a univariate normal distribution and do not have an effect on the target.",
		escape = FALSE,
		booktabs = TRUE
	) |>
	kable_styling()
```

## Importance (scaled) {.tabset}

Importances are scaled to percentages such that 100 is the highest importance value assigned by the corresponding method on a dataset in a given replication.

### Confounded

```{r plot-confounded}
#| fig.cap: Scaled feature importance scores for the Confounded task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "scaled",
	problem = "confounded",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Correlated

```{r plot-correlated-25}
#| fig.cap: Scaled feature importance scores for the Correlated (r = 0.25) task by FI method and learner, colored by implementing package.
plot_importance(
	importances[correlation == 0.25],
	type = "scaled",
	problem = "correlated",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

```{r plot-correlated-75}
#| fig.cap: Scaled feature importance scores for the Correlated (r = 0.75) task by FI method and learner, colored by implementing package.
plot_importance(
	importances[correlation == 0.75],
	type = "scaled",
	problem = "correlated",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Ewald

```{r plot-ewald}
#| fig.cap: Scaled feature importance scores for the Ewald task by FI method and learner, colored by implementing package.
#| #| fig.height: 10
#| fig.width: 10
plot_importance(
	importances,
	type = "scaled",
	problem = "ewald",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Friedman1

```{r plot-friedman1}
#| fig.cap: Scaled feature importance scores for the Friedman1 task by FI method and learner, colored by implementing package.
#| fig.height: 14
plot_importance(
	importances,
	type = "scaled",
	problem = "friedman1",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Independent

```{r plot-independent}
#| fig.cap: Scaled feature importance scores for the Independent task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "scaled",
	problem = "independent",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Interactions

```{r plot-interactions}
#| fig.cap: Scaled feature importance scores for the Interactions task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "scaled",
	problem = "interactions",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Mediated

```{r plot-mediated}
#| fig.cap: Scaled feature importance scores for the mediated task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "scaled",
	problem = "mediated",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Bike sharing

```{r plot-bikesharing}
#| fig.height: 13
#| fig.cap: Scaled feature importance scores for the bike sharing task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "scaled",
	problem = "bike_sharing",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

## Importance (ranks) {.tabset}

Importance scores are converted to ranks with 1 (leftmost) being the highest importance score, i.e., the most important feature as judged by the respective method.
Boxplots across ranks illustrate how consistent the rankings are.

### Confounded

```{r plot-confounded-ranks}
#| fig.cap: Ranked feature importance scores for the Confounded task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "rank",
	problem = "confounded",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Correlated

```{r plot-correlated-25-rank}
#| fig.cap: Ranked feature importance scores for the Correlated (r = 0.25) task by FI method and learner, colored by implementing package.
plot_importance(
	importances[correlation == 0.25],
	type = "rank",
	problem = "correlated",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

```{r plot-correlated-75-rank}
#| fig.cap: Ranked feature importance scores for the Correlated (r = 0.75) task by FI method and learner, colored by implementing package.
plot_importance(
	importances[correlation == 0.75],
	type = "rank",
	problem = "correlated",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Ewald

```{r plot-ewald-rank}
#| fig.cap: Ranked feature importance scores for the Ewald task by FI method and learner, colored by implementing package.
#| #| fig.height: 10
#| fig.width: 10
plot_importance(
	importances,
	type = "rank",
	problem = "ewald",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Friedman1

```{r plot-friedman1-rank}
#| fig.cap: Ranked feature importance scores for the Friedman1 task by FI method and learner, colored by implementing package.
#| fig.height: 14
plot_importance(
	importances,
	type = "rank",
	problem = "friedman1",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Independent

```{r plot-independent-rank}
#| fig.cap: Ranked feature importance scores for the Independent task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "rank",
	problem = "independent",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Interactions

```{r plot-interactions-rank}
#| fig.cap: Ranked feature importance scores for the Interactions task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "rank",
	problem = "interactions",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Mediated

```{r plot-mediated-rank}
#| fig.cap: Ranked feature importance scores for the mediated task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "rank",
	problem = "mediated",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Bike sharing

```{r plot-bikesharing-rank}
#| fig.height: 13
#| fig.cap: Ranked feature importance scores for the bike sharing task by FI method and learner, colored by implementing package.
plot_importance(
	importances,
	type = "rank",
	problem = "bike_sharing",
	method = NULL,
	color = "package",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

## Pairwise differences to xplainfi (scaled) {.tabset}

The plots above show the distribution of importance scores across replications for each implementation separately.
As suggested by a reviewer, the plots below instead show the distribution of the *paired difference* in importance between `xplainfi` and each reference implementation, where both scores are computed within the same replication, i.e., on identical train-test splits.
For each feature, method, learner, and replication, we subtract the reference package's importance from `xplainfi`'s; boxplots centered on the dashed line at zero therefore indicate agreement on a given dataset, which is more informative than comparing marginal distributions.
Differences are computed on the scaled importances (scaled to `[0, 1]` per implementation and replication, as in the figures above), so they are expressed in percentage points and share a common axis.

The `vip` package is omitted here because it reports importance on the RMSE scale rather than MSE, so the comparison is not directly meaningful; this is addressed separately.

### Confounded

```{r plot-confounded-diff}
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the Confounded task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances,
	problem = "confounded",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Correlated

```{r plot-correlated-25-diff}
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the Correlated (r = 0.25) task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances[correlation == 0.25],
	problem = "correlated",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

```{r plot-correlated-75-diff}
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the Correlated (r = 0.75) task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances[correlation == 0.75],
	problem = "correlated",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Ewald

```{r plot-ewald-diff}
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the Ewald task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances,
	problem = "ewald",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Friedman1

```{r plot-friedman1-diff}
#| fig.height: 14
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the Friedman1 task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances,
	problem = "friedman1",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Independent

```{r plot-independent-diff}
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the Independent task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances,
	problem = "independent",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Interactions

```{r plot-interactions-diff}
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the Interactions task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances,
	problem = "interactions",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Mediated

```{r plot-mediated-diff}
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the mediated task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances,
	problem = "mediated",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```

### Bike sharing

```{r plot-bikesharing-diff}
#| fig.height: 13
#| fig.cap: Paired differences in scaled importance (xplainfi − reference, percentage points) for the bike sharing task by FI method and learner, colored by reference package.
plot_importance_diff(
	importances,
	problem = "bike_sharing",
	facets = c("method", "learner_type"),
	ncol = 4,
	title = TRUE
)
```
