---
title: 'SLCARE: An R Package for Semiparametric Latent Class Analysis of Recurrent
  Events'
date: '2025-05-20'
abstract: |
  Recurrent event data frequently arise in biomedical follow-up studies. The concept of latent classes  enables researchers to characterize complex population heterogeneity in a plausible and parsimonious way. This article introduces the R package SLCARE, which implements a robust and flexible algorithm  to  carry out Zhao, Peng, and Hanfelt (2022)’s latent class analysis method for recurrent event data, where  semiparametric multiplicative intensity modeling is adopted. SLCARE returns estimates for non-functional model parameters along with the associated variance estimates and $p$ values. Visualization tools are provided to depict the estimated functional model parameters and related functional quantities of interest. SLCARE also delivers a model checking plot  to help assess the adequacy of the fitted model.
draft: no
author:
- name: Qi Yu
  affiliation: Emory University
  address:
  - Department of Biostatistics and Bioinformatics
  - 1518 Clifton Rd. NE, Atlanta, GA 30322, USA
  email: qi.yu2@emory.edu
  orcid: 0009-0007-7585-1183
- name: Limin Peng
  affiliation: Emory University
  address:
  - Department of Biostatistics and Bioinformatics
  - 1518 Clifton Rd. NE, Atlanta, GA 30322, USA
  email: lpeng@emory.edu
  orcid: 0000-0002-8042-3112
type: package
output:
  rjtools::rjournal_pdf_article:
    toc: no
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: yu-peng.bib
preamble: \usepackage{longtable} \usepackage{float} \floatplacement{figure}{H} \floatplacement{table}{H}
date_received: '2024-01-26'
volume: 16
issue: 3
slug: RJ-2024-030
journal:
  lastpage: 149
  firstpage: 134

---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(dplyr)
library(tidyr)
library(table1)
library(plotly)
```

# Introduction

Recurrent event data frequently arise in biomedical follow-up studies where the event of interest, such as infection or hospitalization, occurs repeatedly over time. The event recurrence  often demonstrates different patterns across individuals. To accommodate such heterogeneity, many regression methods have been developed with implementation available as  R packages. To name a few, analyses based on the proportional intensity model [@andersen1982cox] can be carried out with function `coxph()` from R package \CRANpkg{survival} [@survival-package] or `cph()` from R package \CRANpkg{rms} [@rms].
Fitting @wang2001analyzing 's method based on a multiplicative intensity model is implemented by  function `reReg()` from R package \CRANpkg{reReg} [@reReg].  Modeling the gap time between recurrent events, @clement2009conditional proposed  conditional generalized estimating equation, and developed R package \CRANpkg{condGEE} implementing this approach.
@fine2004temporal 's temporal regression strategy can be applied to  regress the mean function of recurrent events through using function `tpr()`  from R package \CRANpkg{tpr} [@yan2004estimating]. 

More recently, viewing the observed data as a manifestation of latent classes or subgroups, @zhao2022semiparametric proposed a semiparametric latent class analysis (LCA) method to help reveal  more realistic  heterogeneity structure of recurrent event data that may not be adequately captured by a single regression model.
@zhao2022semiparametric adopted latent variable mixture modeling (LVMM)  while avoiding stringent parametric assumptions commonly imployed in LCA literature. Note that several R packages are available for fitting LVMM for data types other than recurrent event data. For example, R package \CRANpkg{flexmix} [@flexmix] delivers a general tool for tackling finite mixtures of regression models, such as the standard linear model and the generalized linear model, with  the expectation-maximization (EM) algorithms. R package \CRANpkg{lcmm} [@lcmm] allows for fitting joint mixture models with  longitudinal and single time-to-event outcomes based on the Newton-Raphson algorithm. However, none of these packages offer options to conduct LCA oriented to recurrent event outcomes.

In this paper, we introduce the R package \CRANpkg{SLCARE}, which delivers robust and flexible implementation of @zhao2022semiparametric 's LCA method for recurrent event data and is available from the Comprehensive R Archive Network (CRAN) at
<https://CRAN.R-project.org/package=SLCARE> and Github at
<https://github.com/qyxxx/SLCARE>. The package \CRANpkg{SLCARE}
 offers a user-friendly software for conducting LCA of recurrent event data in R with the core function `SLCARE()`. The function `SLCARE()` enables a variety of
options for specifying and estimating the flexible semiparametric latent class model of @zhao2022semiparametric . For example, the number of latent classes can be pre-specified or chosen based on an entropy-based measure. Users can opt to built-in or customized initializer. Algorithm convergence criterion can be set through specifying the maximum number of iterations or the minimum change in parameter estimates. The function `SLCARE()` also provides visual tools to aid in result presentation and interpretation as well as model checking upon \CRANpkg{ggplot2} plotting environment [@ggplot2].

The remainder of this paper is organized  as follows. Next section describes the methodological background for the R package \CRANpkg{SLCARE}, including model assumptions, estimation procedure and algorithm, inference, and model evaluation. In the third section, we elaborate the structure of the package. The utility of the package is illustrated via simulated data examples in Section 4. A real application is presented in Section 5. We conclude with a few remarks in the last section.

# Methodological background

## Recurrent event data and notation
For subject \(i\),  let \(T_i^{(j)}\) denote time to the \(j^{th}\)
recurrent event (\(i = 1, \dots , n\), $j=1,2,\ldots$). The underlying counting process of recurrent events  is defined as
\(N_{i} ^{*}(t) = \sum _{j = 1} ^{\infty} I(T_{i} ^{(j)} \le t)\)
,representing the number of events occurred before or at time \(t\) for subject \(i\),  where \(I(\cdot)\) denotes the indicator function.
Suppose the observation of subject $i$ is ended at  time \(C_i\). Then the observed counting process is given by
\(N_{i} (t) = N_{i} ^{*}(\min(t, C_i)) = \sum _{j = 1} ^{\infty} I(T_{i} ^{(j)} \le \min(t, C_i))\) (\(i = 1, \dots , n\)). Let \(\tilde{Z}_i\) be a $p$-dimensional
vector of time-independent covariates for subject \(i\). The observed
data consist of \(\{ N_{i}(t), C_i, \tilde{Z}_{i} \}_{i = 1} ^{n}\). 

## Models and assumptions
Suppose the whole population consists of \(K\) latent classes, and assume that how covariates are related to the intensity function  of recurrent events are the same within each latent class while being allowed to vary across different latent classes. @zhao2022semiparametric proposed a latent class multiplicative intensity model, which assumes that $N_i^*(t)$ is  a nonstationary Poisson process with the intensity function,
\begin{equation} 
(\#eq:1)
\lambda _{i} (t) = \sum _{k = 1} ^{K} I (\xi _{i} = k) \times \lambda _{0} (t) \times W_{i} \times \eta _{0,k} \times \exp(\tilde{Z} _{i} ^{\top} \tilde{\beta} _{0,k}) 
\end{equation}
where \(I(\cdot)\) denotes the indicator function, \(\xi_i\) denotes the unobserved latent class membership, \(\lambda _{0} (t)\) is an unspecified, continuous, nonnegative baseline intensity function shared by all latent classes,  \(W_{i}\) is a positive subject-specific latent variable independent of \((\xi_i, \tilde{Z}_i, C_i)\). The latent variable $W_i$  plays the same role as individual frailty reflecting more or less frequent occurrences of recurrent events. Here \(\eta _{0,k}\) is a positive number that captures the class-$k$ scale shift from the baseline intensity function, and \(\tilde{\beta} _{0,k}\) is the regression coefficient that represents the class-$k$ covariate effects on the intensity function. For the identifiability of \(\lambda_{0}(t)\) and \(\eta _{0,k}\), it is assumed that \(E(W_{i} | \tilde {Z} _{i} , \xi _{i} = k) = 1\) for \(k = 1, \dots, K\) and \(\int _{0}^{\nu^{*}} \lambda_{0} (u) du = 1\), where \(\nu^{*}\) is a predetermined constant. As commented in @zhao2022semiparametric, one may choose \(\nu^{*}\) to be slightly smaller than the upper bound of \(C_i\)'s support. A different choice of \(\nu^{*}\) only results in a scale shift to \(\lambda_{0}(t)\) and has no influence on the regression coefficients, \(\tilde{\beta} _{0,k}\)'s.

@zhao2022semiparametric modeled the distribution of the latent class membership \(\xi_i\)  by a
logistic regression model:
\begin{equation} 
(\#eq:2)
P(\xi _{i} = k | \tilde{Z} _{i}) = p_{k} (\alpha _{0} , \tilde{Z} _{i}) \doteq \frac{\exp(\tilde{Z} _{i} ^{\top} \alpha _{0,k})}{\sum_{k = 1}^{K}\exp(\tilde{Z} _{i} ^{\top} \alpha _{0,k}) } , \quad k = 1, \cdots, K 
\end{equation}
where $\alpha_0=(\alpha_{0,1}^\top,\ldots, \alpha_{0, K}^\top)^\top$ with \(\alpha_{0,1} = \textbf{0}_{p\times 1}\).

## Estimation and inference procedures
### Estimation procedure
Let \(Z_i = (1, \tilde{Z}_{i}^{\top})^{\top}\) and \(\beta _{0,k} = (\log{\eta_{0,k}}, \tilde{\beta}_{0,k} ^{\top})^{\top}\). By the model assumptions stated in Section 2, the following equalities hold:
\begin{equation} 
(\#eq:3)
E[ I(\xi _{i} = k) Z_{i} \{ \frac{N_{i} ^{*} (C _{i})}{\mu_{0} (C_{i})}  - \exp(Z_{i} ^{\top} \beta _{0,k}) \}] = 0,\ \ k=1,\ldots, K, 
\end{equation}

\begin{equation} 
(\#eq:4)
E[\tilde Z_i\{ I(\xi_i=k)-\frac{\exp(\tilde{Z} _{i} ^{\top} \alpha _{0,k})}{\sum_{k = 1}^{K}\exp(\tilde{Z} _{i} ^{\top} \alpha _{0,k}) } \}]= 0,\ \ k=1,\ldots, K,
\end{equation}
where \(\mu_0(t)=\int_0^t \lambda_0(s)ds\), representing the cumulative baseline intensity function. 

Equalities in  \@ref(eq:3) and \@ref(eq:4) cannot be directly used to construct
estimating equations for \(\alpha_0\) and \(\beta_{0,k}\)'s because 
\(\xi _{i}\)'s are not observable and $\mu_0(\cdot)$ is unknown. Adapting the principle of
conditional score [@stefanski1987conditional] for handling missing covariates, @zhao2022semiparametric proposed to address the unobserved \(I(\xi _i = k )\) by substituting it with \(\tau _{ik} \doteq E[I(\xi _i = k) | Z_i, C_i, D_i]\), where \(D_i \doteq N_i(C_i)\), capturing the number of the observed recurrent events from subject \(i\).  
First, it follows from the definition that
\begin{equation}
(\#eq:7)
\tau _{ik} = \frac{P(D_{i} = d_{i} | \xi _{i} = k, Z _{i}, C_{i}) P(\xi _{i} = k | Z_{i}, C_{i})}{\sum _{l = 1} ^{K} P(D_{i} = d_{i} | \xi _{i} = l, Z _{i}, C_{i}) P(\xi _{i} = l | Z_{i}, C_{i})}. 
\end{equation}
As implied by model \@ref(eq:2) and the assumption that
\(C_i \perp (N^{*} _i (\cdot) , \xi _i ) | Z_i\), 
\begin{equation}
(\#eq:8)
P(\xi _i = k | Z_i, C_i) = p_k (\alpha _0, \tilde{Z} _i) = \frac{\exp(\tilde{Z} _{i} ^{\top} \alpha _{0,k})}{\sum _{l = 1} ^{K} \exp(\tilde{Z} _{i} ^{\top} \alpha _{0,l})} 
\end{equation}
and 
\begin{align} 
(\#eq:9)
&P ( D _{i} = d_{i} | \xi _{i} = k , Z _{i}, C _{i})\\\nonumber
&= \int _{0} ^{\infty} \frac{ \{\exp(Z_{i} ^{\top} \beta _{0,k}) w \cdot \mu_{0}(C _{i}) \} ^{d_{i}} }{d_{i} !} \exp \{ - \exp(Z_{i} ^{\top} \beta _{0,k}) w \cdot \mu _{0} (C _{i}) \} \cdot f _{W} (w) dw,   
\end{align}
where \(f_W (\cdot)\) denotes the density function of frailty \(W\).
Combining the results in \@ref(eq:7)-\@ref(eq:9), $\tau_{ik}$ can be explicitly expressed as a function of $\alpha_0\doteq (\alpha_{0,1}^\top\ldots, \alpha_{0, K}^\top)^\top, \beta_0\doteq (\beta_{0,1}^\top,\ldots, \beta_{0,k}^\top)^\top$ and $\mu_0(\cdot)$, which is denoted by $\tau_{ik}(\alpha_0, \beta_0, \mu_0)$. 

In addition, @zhao2022semiparametric proposed a
Nelson-Aalen type estimator of $\mu_0(\cdot)$, given by
\begin{equation} 
(\#eq:10)
\hat{\mu}(t) = \exp \{  \hat{H}(t) \} \qquad \text{with} \quad \hat{H}(t) = - \int _{t} ^{\nu ^{*}} \frac{\sum _{i = 1} ^{n} d N_i(s)}{\sum _{i = 1} ^{n} I(C_i \ge s) N_i (s)}.  
\end{equation}

Replacing $I(\xi_i=k)$  and $\mu_0(\cdot)$ by $\tau_{ik}(\alpha, \beta, \hat\mu)$ and $\hat\mu(\cdot)$ respectively in the empirical counterparts of \@ref(eq:3)-\@ref(eq:4) then leads to the following estimating equations:
\begin{equation} 
(\#eq:11)
n^{1/2} S_{1,n}(\alpha, \beta, \hat{\mu}) = 0,
\end{equation}
\begin{equation} 
(\#eq:12)
n^{1/2} S_{2,n}(\alpha, \beta, \hat{\mu}) = 0,
\end{equation}
where
\(S_{j,n} (\alpha, \beta, \hat{\mu}) = ( S_{j,n,1} (\alpha, \beta, \hat{\mu})^{\top}, \cdots , S_{j,n,K} (\alpha, \beta, \hat{\mu})^{\top} )^{\top}\) 
(\(j = 1,2\))
with
\begin{equation}
S_{1,n,k}(\alpha, \beta, \mu) = \frac{1}{n}\sum _{i = 1} ^{n} \tau_{ik}(\alpha, \beta, \mu) Z_{i} \{ \frac{N_i ^{*} (C _{i})}{\hat\mu(C_{i})} - \exp(Z_{i} ^{\top} \beta_{k})\},\ \ k=1,\ldots, K;
\end{equation}

\begin{equation}
S_{2,n,k}(\alpha, \beta, \mu) = \frac{1}{n}\sum _{i = 1} ^{n}  \tilde{Z_{i}} \{ \tau_{ik}(\alpha, \beta, \mu) - \frac{\exp(\tilde{Z} _{i} ^{\top} \alpha _{k})}{\sum _{j = 1} ^{K} \exp(\tilde{Z} _{i} ^{\top} \alpha _{j})}\},\ \ k=1,\ldots, K.
\end{equation}

Solving equations \@ref(eq:11) and \@ref(eq:12) renders estimators of $\alpha_0$ and $\beta_0$, denoted by $\hat\alpha$ and $\hat\beta$. Detailed derivations of estimating equations as well as the asymptotic
properties of $\hat\alpha$ and $\hat\beta$ can be found in @zhao2022semiparametric .

### Estimation algorithm
Based on the result in \@ref(eq:10) and estimating equations in \@ref(eq:11), and \@ref(eq:12), we propose the following algorithm to obtain $\hat\alpha$ and $\hat\beta$, the parameter estimates for  models \@ref(eq:1) and \@ref(eq:2).

> Step 1: Compute \(\hat{\mu} (\cdot)\) based on formula \@ref(eq:10). Let $L_{iter}=0$ and set initial values of \(\alpha _0\) and \(\beta _0\) as \(\hat{\alpha}^{*}\) and
\(\hat{\beta}^{*}\) respectively. Calculate
\(\hat{\tau}^*_{ik} \doteq \tau _{ik} (\hat{\alpha}^{*}, \hat{\beta}^{*}, \hat{\mu})\)
based on equations \@ref(eq:7), \@ref(eq:8) and \@ref(eq:9).
>
> Step2: Repeat in the loop:
>
> > (i) Solve estimating equations \@ref(eq:11) and \@ref(eq:12) with
  \(\tau _{ik} (\alpha, \beta, \hat{\mu})\) fixed as
  \(\hat{\tau}^*_{ik}\). Denote the solutions by \(\check{\alpha}\) and
  \(\check{\beta}\).
> >
> > (ii) \(\hat{\alpha}^{*} := \check{\alpha}; \: \hat{\beta}^{*} := \check{\beta}; \: \hat{\tau}^* _{ik} := \tau _{ik} (\hat{\alpha}^{*}, \hat{\beta}^{*}, \hat{\mu})\). Increase $L_{iter}$ by 1.
> >
> > (iii) If
  \(\max{ ( \| \frac{\check{\beta} - \hat{\beta}^{*}}{\hat{\beta}^{*}} \| _{\infty} , \: \| \frac{\check{\alpha} - \hat{\alpha}^{*}}{\hat{\alpha}^{*}} \| _{\infty} ) }\)
  is smaller than a pre-specified threshold value or $L_{iter}$ is greater than a pre-specified maximum number of iteration,  then exit the loop. Here \(\| \cdot \| _{\infty}\)
  denotes the \(L _{\infty}\) norm and the fraction between vectors
  \(\alpha\) and \(\beta\) is component-wise fraction.
>
> > End repeat loop.
> >
> > Return  \(\check{\alpha}\) and \(\check{\beta}\) as the final estimates for $\alpha_0$,and $\beta_0$ (i.e., $\hat\alpha$ and $\hat\beta$).

### Algorithm implementation
In the following, we present some details related to the algorithm implementation, such as how to set initial values, $\hat\alpha^*$ and $\hat\beta^*$, and how to solve the estimating equations involved in the presented algorithm.

#### Setting initial values for the estimation algorithm
The function `SLCARE()` in package \CRANpkg{SLCARE}  allows users to specify the initial values, $\hat\alpha^*$ and $\hat\beta^*$, by their own choice. Unless users manually specify the initial values,
`SLCARE()` implements an automated initializer, which obtains $\hat\alpha^*$ and $\hat\beta^*$ through the following procedure:

> Step 1: Perform $K$-means clustering [@lloyd1982least] of the observed covariates and number of recurrent events, i.e., $\{\tilde Z_i, N_i(C_i)\}_{i=1}^n$, with R function `kmeans()`, and divide all subjects into $K$ group, where \(K\) stands for the number of latent classes. Denote the assigned group membership for subject $i$ by $G_i (\in\{1,\ldots, K\})$.
>
> Step 2: Obtain $\hat\alpha^*$, the initial estimate for \(\alpha _0\), as the regression coefficients from fitting the
multinomial regression of $\{G_i\}_{i=1}^n$ on covariates \(\{\tilde{Z}_i\}_{i=1}^n\).
>
> Step 3: Obtain $\hat\beta^*$, the initial estimate of \(\beta _0\), as $(\hat\beta_1^{*\top},\ldots, \hat\beta_K^{*\top})$, where $\hat\beta_k^{*\top}$ is the regression coefficient estimate from fitting @wang2001analyzing 's multiplicative intensity model to the subset with $G_i=k$ using `reReg()` from R package \CRANpkg{reReg} [@reReg].

#### Specifying the distribution of frailty $W$
The `SLCARE()` function allows users to specify the distribution of frailty $W$. While @zhao2022semiparametric 's method allows the frailty distribution to be a general parametric distribution, the implementation by `SLCARE()` confines to settings where $W=1$ or $W$ follows a distribution that is parameterized as $Gamma(k, k)$. These choices of frailty distributions cover a variety of density forms.

#### Solving estimating equations
In Step 2(i) of the presented algorithm,  an updated estimate for $\beta_0$, $\check\beta$, is obtained from
solving equations,
\(\sum _{i = 1} ^{n} \hat{\tau} _{ik} \cdot Z_i \cdot \Bigl\{  \frac{N _{i} ^{*} (C _i )}{\hat{\mu} (C_i)} - \exp (Z_i ^{\top} \beta _k )   \Bigr\} = 0 , \quad k = 1, \dots, K. \)
As suggested by @zhao2022semiparametric, we solve these equations by fitting a
'pseudo' weighted Poisson regression model to an augmented dataset which includes responses,
\(\{\frac{N _{i} ^{*} (C _i )}{\hat{\mu} (C_i)}\}_{i=1}^n \), and covariates, \(\{Z_i\}_{i=1}^n\),
along with weights \(\hat{\tau}_{ik}\). \CRANpkg{SLCARE} performs such Poisson regression  with R function `glm.fit()` from R package \pkg{stats}.

Meanwhile, in Step 2(i) of the presented algorithm, an updated estimate for $\alpha_0$,  \(\check{\alpha}\), is
obtained by solving equations,
\(\frac{1}{\sqrt{n}} \sum _{i =1} ^{n} \tilde{Z}_i \Bigl\{  \hat{\tau} _{ik} - \frac{\exp(\tilde{Z} _{i} ^{\top} \alpha _{k}) }{\sum _{j = 1} ^{K} \exp(\tilde{Z} _{i} ^{\top} \alpha _{j})} \Bigr\} = 0, \quad k = 2, \cdots, K \).
To solve these equations,  we use an alternative and yet equivalent approach that fits weighted multinomial regression model on an augmented dataset, which consists of $nK$ response-covariate pairs,  $\{(1, \tilde Z_i),\ldots, (K, \tilde Z_i)\}_{i=1}^n$ with weight $\hat\tau_{ik}$ assigned to the pair $(k, \tilde Z_i)$ ($k=1,\ldots, K$). \CRANpkg{SLCARE} implements the weighted multinomial regression with function `multinom()` from R package \CRANpkg{nnet} [@nnet].

#### Determination of the number of latent classes
@zhao2022semiparametric proposed to determine $K$, the number of latent classes, based on  a relative entropy measure
[@ramaswamy1993empirical]. That is, upon fitting models \@ref(eq:1) and \@ref(eq:2) that assume $M$ latent classes, the corresponding relative entropy measure is defined as
\begin{equation} 
(\#eq:15)
Entropy(M) = 1 - \frac{\sum _{i = 1} ^{n} \sum _{k = 1} ^{M} - \hat{\tau} _{ik} \log(\hat{\tau} _{ik})}{n \log(M)} 
\end{equation}
where 
\(\hat{\tau} _{ik} = \tau _{ik} (\hat{\alpha}, \hat{\beta}, \hat{\mu})\) ($k=1,\ldots, M$).
By definition, \(Entropy(M)\) is bounded between \(0\) and \(1\) with a larger value
indicating a better fit to the data [@Celeux1996]. Following the rationale, one may choose $K$ as the maximizer of \(Entropy(\cdot)\) over a sequence of candidate values for $K$. 
Given a pre-specified $K$, the value of \(Entropy(K)\) can be generated by S3 method `print(x, type = "Entropy")`.

#### Model checking
@zhao2022semiparametric proposed a graphical method for checking the overall fit of models \@ref(eq:1) and \@ref(eq:2). The key idea is to compare the
observed recurrent events \(D_i \doteq N_i(C_i)\) versus the expected number of recurrent events under models \@ref(eq:1)-\@ref(eq:2). Specifically, models \@ref(eq:1) and \@ref(eq:2) imply
\(E \{ N_{i} (C_i) | Z_i \} = E \{ N_{i} ^{*} (C_i) | Z_i \} = \sum _{k = 1} ^{K} \tau _{ik} \cdot \mu _{0} (C_i) \exp (Z_{i} ^{\top} \beta _{0, k}). \)
Then the expected number of recurrent events under the assumed models can be approximated by
\begin{equation} 
(\#eq:16)
\hat{D}_i \doteq \sum_{i = 1} ^{K} \hat{\tau} _{ik} \cdot \hat{\mu}(C_i) \cdot \exp(Z_{i} ^{\top} \hat{\beta}_k ).
\end{equation}
Therefore, in the scatter plot of \(\hat D_i\) versus \({D} _i\), a major departure
from the identity line \(D_i = \hat{D} _i\) may suggest a lack-of-fit of the assumed models.
Model checking plot can be generated by S3 method `plot(x, type = "ModelChecking")`.

#### Class-specific mean function of recurrent events
To help illustrate heterogeneity in recurrent event occurrence across different latent classes, 
\CRANpkg{SLCARE} computes crude estimates for the class-specific mean functions of recurrent events. Specifically, the crude estimate for the class-specific mean function of recurrent event is computed as
\begin{equation} 
(\#eq:17)
\frac{ \sum _{i = 1} ^{n} I(\hat{\xi} _i = k) \sum _{j = 1} ^{K} \hat{\tau} _{ij} \cdot \hat{\mu}(t) \exp (Z_i ^{\top} \hat{\beta} _j) }{ \sum _{i = 1} ^{n} I (\hat{\xi} _i = k)}, \quad k = 1, \dots, K,
\end{equation}
where 
\(\hat{\xi} _i = \text{arg} \max _{1 \le k \le K} \hat{\tau} _{ik}\).
A plot of the estimated mean functions stratified by
latent groups can be generated by S3 method `plot(x, type = "EstMeans")`.

# Package structure

The main function of R package \CRANpkg{SLCARE}  is `SLCARE()`. The dataset imported to function `SLCARE()` should take  the long format, where each row corresponds to one time point of a subject at which a recurrent event is observed or censored. With a dataset coded in the wide format, where each row contains the timing information of all recurrent events observed or censored within one subject, users may convert such a dataset into the required long format by using function `pivot_longer()` from package  \CRANpkg{tidyr} [@tidyr] or function `melt()` from package  \CRANpkg{reshape2} [@reshape2]. In addition to allowing users to specify  initial parameter estimates manually, 
`SLCARE()` offers a default initializer that implements the informative selection of initial values described in Section **Estimation procedure**.
The built-in initializer performs $K$-means clustering of the observed data  using function `kmeans()` from package
\pkg{stats} and fitting multiplicative intensity models using `reReg()` from package \CRANpkg{reReg}.
When solving the estimating equations involved in the iterative estimation algorithm, 
`SLCARE()` performs weighted Poisson regression using
`glm.fit()` from  package \pkg{stats} and weighted multinomial regression using `multinom()` from package \CRANpkg{nnet}. 
After running `SLCARE()`, model fitting results can be
easily extracted from the output with S3 methods `summary()`, `print()`, `predict()`, and `plot()`.
Graphical results generated by `SLCARE()` are presented via \CRANpkg{ggplot2} environment and are fully customizable. 

Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:SLCAREfunctions-interactive)', '\\@ref(tab:SLCAREfunctions-static)'))` lists the main function `SLCARE()`, corresponding S3 methods and other functions called by this function, along with brief descriptions of their purposes.

```{r SLCAREfunctions-interactive, eval = knitr::is_html_output(), layout = "l-body-outset"}
table1 <- data.frame(
  `FUNCTION` = c("SLCARE()", "summary()", "print()", "predict()", "plot()",
                 "reReg()", "multinom()", "kmeans()", "glm.fit()"),
  `PURPOSE` = c("Conduct LCA with recurrent events data based on semiparametric multiplicative intensity modeling.", 
                "Generic function; used to summarize estimates for non-functional model parameters along with the associated variance estimates and $p$ values.", 
                "Generic function; used to initial estimates for the estimation algorithm, convergence criterion, latent class membership probability and predicted number of recurrent events.", 
                "Generic function; used to predict the posterior number of recurrent events.", 
                "Generic function; used to generate cumulative baseline intensity function, estimated mean function, model checking plot.",
                "Imported from  package reReg; used to find an initial estimate for $\\beta_0$.", 
                "Imported from  package nnet; used to find an initial estimate for $\\alpha_0$; used to solve the estimating equation to update the estimate for $\\alpha_0$.", 
                "Imported from package stats; used to find initial estimates for $\\alpha_0$ and $\\beta_0$.", 
                "Imported from package stats; used to solve the estimating equation to update the estimate for $\\beta_0$.")
)

knitr::kable(table1, format = "html", caption = "An overview of the main function and imported functions in SLCARE", booktabs=TRUE) |>
  kableExtra::kable_styling(position = "center")
```

```{r SLCAREfunctions-static, eval = knitr::is_latex_output()}
table1 <- data.frame(
  `FUNCTION` = c("SLCARE()", " ", "summary()", " ", "print()", " ", " ", "predict()", "plot()", " ",
                 "reReg()", "multinom()", " " , "kmeans()", "glm.fit()", " "),
  `PURPOSE` = c("Conduct LCA with recurrent events data based on semiparametric",
                "multiplicative intensity modeling.",
                "Generic function; used to summarize estimates for model parameters",
                "along with the associated variance estimates and $p$ values.", 
                "Generic function; used to initial estimates for the estimation algorithm",
                "convergence criterion, latent class membership probability and",
                "predicted number of recurrent events.",
                "Generic function; used to predict the posterior number of recurrent events.", 
                "Generic function; used to generate cumulative baseline intensity function",
                "estimated mean function, model checking plot.",
                "Imported from  package reReg; used to find an initial estimate for $\\beta_0$.",
                "Imported from  package nnet; used to find an initial estimate for $\\alpha_0$;",
                "used to solve the estimating equation to update the estimate for $\\alpha_0$.",
                "Imported from package stats; used to find initial estimates for $\\alpha_0$ and $\\beta_0$.",
                "Imported from package stats; used to solve the estimating equation to",
                "update the estimate for $\\beta_0$.")
)

knitr::kable(table1, format = "latex", caption = "An overview of the main function and imported functions in SLCARE", booktabs=TRUE, escape = FALSE) |>
  kableExtra::kable_styling(position = "center")
```

# Illustrations

The package \CRANpkg{SLCARE} is designed to conduct LCA of recurrent events data based on semiparametric multiplicative intensity modeling [@zhao2022semiparametric]. 
The main function `SLCARE()` has function arguments that enable flexible model specification and method implementation. In the below, we illustrate the use of `SLCARE()` through simulated datasets.

## Input dataset

An input dataset for `SLCARE()` is a data frame containing time to recurrent events or censoring along with time-independent covariates of interest.
The input dataset should take the long format, where each row contains the information in one time interval in which a recurrent event is observed or censored.
As such,  multiple rows in the dataset may correspond to one subject.
As mentioned in Section **Package structure**, a dataset in the wide format data can be readily converted into the long format  required by `SLCARE()` with the use of function `pivot_longer()` from package \CRANpkg{tidyr} or function `melt()` from package \CRANpkg{reshape2}.

The package \CRANpkg{SLCARE} has a build-in dataset, `SimData`, which is a simulated dataset perturbed from a real
dataset. `SimData` contains 48 subjects. For each subject,
`SimData` contains values for the following six variables:
`id`, which is the unique identifier for each subject;
`start`, which records the starting time in minutes of the counting process interval;
`stop`, which records the ending time in minutes of the counting process interval. If the subject entered the study at time zero, this column represents the time in minutes from the baseline visit to the occurrence of event or censoring;
`event`, which indicates  whether a recurrent event is observed (`event = 1`) or not (`event = 0`) at the end of the counting process interval;
`x1`, which is a binary covariate;
`x2`, which is a continuous covariate.
Below we display how the data from one example subject are recorded in `SimData`:

```{r}
library(SLCARE)
subset(SimData, id %in% c("G052") )
```
As shown above, subject `G052` experienced two current events at time 1950 and time 2580
before the censoring time  7085. The time-independent covariates, `x1` and `x2`, remain unchanged across different time points within the same subject.

## Command line and function arguments

The LCA of recurrent event data outlined in Section **Methodological background** can be carried out with a single command line `SLCARE()`. `SLCARE()` provides flexible function arguments, which are shown below:

```{r}
args(SLCARE)
```
At minimum, `SLCARE()` requires the following three arguments:

> `data`: a dataframe with the format similar to `SimData`.
>
> `K`: pre-determined number of latent classes.
>
> `formula`: a string containing the covariates of interest in `data`.

The optional arguments of `SLCARE()` are:

> `alpha`: initial estimate for \(\alpha_0\) in the estimation procedure. The default is NULL, which represents the initial estimate for $\alpha_0$ resulted from   the automated initializer described in Section **Estimation and inference procedures**.
>
> `beta`: initial estimate for  \({\beta_0}\) in the estimation procedure.
The default is NULL, which represents the initial estimate for $\alpha_0$ resulted from the automated initializer described in
Section **Estimation and inference procedures**.
>
> `id_col`: name of the column that includes subject identifiers.
>
> `start_col`: name of the column that records the start time of each at-risk time interval.
>
> `stop_col`: name of the column that records the start time of each at-risk time interval.
>
> `event_col`: name of the column that indicates whether a recurrent event is observed or not (i.e, 1=observed; 0=otherwise).
>
> `gamma`: parameter that indicates the distribution  of frailty $W$. The
default is 0 which indicates model \@ref(eq:1) holds without the
subject-specific frailty (i.e., $W=1$); `gamma` $= k$ indicates that $W$ follows the $Gamma(k,k)$ distribution.
>
> `max_epochs`: maximum number of iterations for the estimation algorithm.
>
> `conv_threshold`: convergence threshold for the estimation algorithm.
>
> `boot`: number of bootstrap replicates used to obtain the standard
error estimation. The default is NULL which indicates bootstrap is not conducted.

## Output and illustration with a sample dataset

The following code is used to perform LCA of the recurrent event dataset `SimData` with two latent classes (i.e., $K=2$), 20 bootstrap replicates, and covariates `x1` and `x2`.

```{r, echo = TRUE}
set.seed(0)
model1 <- SLCARE(formula = "x1 + x2", data = SimData,
                 id_col = "id", start_col = "start", stop_col = "stop", 
                 event_col = "event",  K=2, boot = 20)
```

The default output/print of `SLCARE()` includes parameter estimates and the associated variance estimates and $p$-values, along with the model's relative entropy measure. The same results can also be obtained using the generic method `summary(model1, digits = 3)`, which allows for manually setting the minimum number of significant digits.

```{r, echo = TRUE}
model1
```

Recall that  in model \@ref(eq:2), \(\alpha_{0,1} = \textbf{0}\), indicating that `class1` is used as the reference group for evaluating the odds ratio of being in another latent class. The numbers in the row corresponding to `class2` represent covariate effects on  the log odds ratio of belonging to `class2` versus `class1`.

These results are also can be extracted separately with the generic method function `print(model1, type = "Est")`, `print(model1, type = "SE")`, `print(model1, type = "PValue")`, and `print(model1, type = "Entropy")`.

### Initial estimates for the estimation algorithm

Unless manually specifying the arguments `alpha` and `beta`, `SLCARE()` implements the automated initializer described in Section **Estimation and inference procedures**.
The resulting initial estimates for \({\beta_0}\) and \({\alpha_0}\) can be found through the following generic method function `print()`:

```{r, echo = TRUE}
print(model1, type = "Init")
```
### Convergence criterion

`SLCARE()` provides arguments to control the termination of the iterative estimation procedure. By default, `max\_epochs=500` and `conv_threshold=0.01`. This means that the iterations would stop when the  number of iterations reaches 500 or the magnitude of  the change in parameter estimates is below 0.01. In addition, users are allowed to check the change in parameter estimates in the last
iteration using the following code:

```{r, echo = TRUE}
print(model1, type = "ConvergeLoss")
```

If the output is greater than the pre-determined convergence threshold, users may consider increasing the maximum number of iterations and/or setting a larger convergence threshold.

### Latent class membership probability

`SLCARE()`  generates output on the estimated probability that a subject belong to different latent classes. The following are the example code and output in the case where one is interested in the class membership probabilities of the \(8^{th}\) to \(10^{th}\) subjects in `SimData`.

```{r, echo = TRUE, eval = knitr::is_latex_output()}
print(model1, type = "ClassProb")[8:10,]
```

```{r, eval = knitr::is_html_output(), layout = "l-body-outset"}
knitr::kable(print(model1, type = "ClassProb")[8:10,], format = "html", booktabs=TRUE) |>
  kableExtra::kable_styling(position = "center") 
```

### Predicted number of recurrent events

`SLCARE()` computes the predicted number of recurrent events for each subject according to equation \@ref(eq:16). The code to extract such a result for  the \(8^{th}\) to \(10^{th}\) subjects in `SimData` is shown below, along with the corresponding output:

```{r, echo = TRUE, eval = knitr::is_latex_output()}
predict(model1)[8:10,]
```

```{r, eval = knitr::is_html_output(), layout = "l-body-outset"}
knitr::kable(predict(model1)[8:10,], format = "html", booktabs=TRUE) |>
  kableExtra::kable_styling(position = "center") 
```

Note that `SLCARE()` returns the estimated numbers of recurrent
events as float rather than integer. Users may use the argument `integer = TRUE` in the genetic function `preduct()` to round those numbers to integers.

```{r, echo = TRUE, eval = knitr::is_latex_output()}
predict(model1, integer = TRUE)[8:10,]
```

```{r, eval = knitr::is_html_output(), layout = "l-body-outset"}
knitr::kable(predict(model1, integer = TRUE)[8:10,], format = "html", booktabs=TRUE) |>
  kableExtra::kable_styling(position = "center") 
```

### Model checking plot

`SLCARE()` generates a model checking plot following the procedure described in Section **Methodological background** via \CRANpkg{ggplot2} environment. The following shows the `S3` method for plotting a model checking plot:

```{r modelchecking, echo = TRUE, fig.height = 3, fig.cap = "Predicted numbers of events versus the observe numbers of events.", eval = knitr::is_latex_output()}
plot(model1, type = "ModelChecking")
```

```{r modelchecking-html, fig.height = 5, fig.cap = "Predicted numbers of events versus the observe numbers of events.", eval = knitr::is_html_output(), layout = "l-body-outset"}
ggplotly(plot(model1, type = "ModelChecking"))
```

### Estimated cumulative baseline intensity function

By equation \@ref(eq:10), `SLCARE()` calculates \(\hat{\mu}(t)\), which is an estimates for \(\mu _{0} (t)\), cumulative baseline intensity function. A plot of \(\hat{\mu}(t)\) versus $t$ can be generated with the following code:

```{r cumbaseline, echo = TRUE, fig.height = 3, fig.cap = "Estimated cumulative baseline intensity function.", eval = knitr::is_latex_output()}
plot(model1, type = "mu0")
```

```{r cumbaseline-html, fig.height = 5, fig.cap = "Cumulative baseline intensity function.", eval = knitr::is_html_output(), layout = "l-body-outset"}
cumbaseline <- model1$Estimated_mu0t_gg +
    geom_line(linewidth = 1) +
    theme(aspect.ratio = 1) +
    ggtitle("Plot of Cumulative Baseline Intensity Function") +
    xlab("t") +
    ylab("Cumulative Baseline Intensity Function") +
    expand_limits(x = 0, y = 0)
ggplotly(cumbaseline)
```

Evaluation of \(\hat{\mu}(t)\) at specific time points, for example,  $t=100, 1000$ and $5000$,
can be obtained by the following command:

```{r, echo = TRUE}
model1$est_mu0(c(100, 1000, 5000))
```

### Estimated mean function plot

Estimated mean function plots are useful for describing and comparing the expected number of recurrent events across different latent groups over time. `SLCARE()` computes the estimated mean function over time according to equation \@ref(eq:17), which can be plotted with the following code:

```{r estmeans, echo = TRUE, fig.height = 3, fig.cap = "Estimated mean function for two latent classes.", eval = knitr::is_latex_output()}
plot(model1, type = "EstMeans")
```

```{r estmeans-html, fig.height = 5, fig.cap = "Estimated mean function for two latent classes.", eval = knitr::is_html_output(), layout = "l-body-outset"}
ggplotly(plot(model1, type = "EstMeans"))
```

# A real application

We apply \CRANpkg{SLCARE} to a dataset from a randomized phase III clinical trial FFCD 2000-05 conducted between 2002 and 2007 in patients diagnosed with advanced colorectal cancer
[@ducreux2011sequential]. The dataset is publicly available in \CRANpkg{frailtypack} package [@rondeau2012frailtypack] and is also included in the \CRANpkg{SLCARE} package. One recurrent event of interest is the appearance of new colorectal lesions. The colorectal lesion is the region in a colon that has suffered damage through colorectal cancer. The appearance of new colorectal lesions indicates colorectal cancer progression. The main interest of our analysis is to explore the heterogeneity in the recurrence patterns of colorectal lesions.

To address this interest, we extract the dataset `colorectalSLCARE` from the `colorectal` dataset. The dataset `colorectalSLCARE is arranged in the long format,  consisting of 150 patients and 289 records. 

Each record stores the value of the six variables described as follows. Variable `id` is the unique patient identifier. Variable `time0` is the start time (in years) of the recurrent event interval. Variable `time1` records the interval end time (in years) of the detection of new colorectal lesions or terminal event (death or right-censoring). Variable `new.lesions = 1` indicates that the appearance of a new colorectal lesion is observed and `new.lesions = 0` indicates that a terminal event is observed. The other two variables represent covariates of interest:  (1) treatment, which is coded as  `treatment = 1` if the patient received a combination chemotherapy and as `treatment = 0` if the patient received sequential chemotherapy; 
(2) previous resection indicator, which is coded as `prev.resection = 1` if the patient had a previous resection and as `prev.resection = 0` otherwise. 

The construction and the structure of the `colorectalSLCARE`
dataset are presented below.

```{r, echo = TRUE}
data("colorectal", package = "SLCARE")
colorectalSLCARE <- colorectal |>
  dplyr::select(id, time0, time1, new.lesions, treatment, prev.resection) |>
  dplyr::mutate(id = paste0("patient", id), treatment = as.numeric(treatment) -1,
                prev.resection = as.numeric(prev.resection) -1)
str(colorectalSLCARE, vec.len = 3)
```
We fit models \@ref(eq:1) and \@ref(eq:2) to the dataset `colorectal_SLCARE` with $T_i^{(j)}$ corresponding to time to the $j$-th detection of new colorectal lesions. We consider
three candidate distributions for \(W\), the frailty term in model \@ref(eq:1), which are
\(W = 1\) and \(Gamma(r,r)\), \(r = 1,~3\). For each candidate
\(f_W(\cdot)\), we calculate the relative entropy measure \(Entropy\)
as described in Section **Estimation and inference procedures** using the command illustrated in Section **Illustrations**. The results are presented in Table
`r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:realentropytable-interactive)', '\\@ref(tab:realentropytable-static)'))`.

```{r}
table2 <- data.frame(
  `x` = c("Gamma(1,1)", "Gamma(3,3)", "W = 1"),
  `K=2` = c(0.785, 0.802, 0.462),
  `K=3` = c(0.788, 0.793, 0.459)
)
```

```{r realentropytable-interactive, eval = knitr::is_html_output(), layout = "l-body-outset"}
knitr::kable(table2, col.names = c("", "K = 2", "K = 3"), format = "html", caption = "Real data example: Relative entropy calculated with different
choices of $K$ and $f_{W}(w)$.", booktabs=TRUE) |>
  kableExtra::kable_styling(position = "center") 
```

```{r realentropytable-static, eval = knitr::is_latex_output()}
knitr::kable(table2, col.names = c("", "K = 2", "K = 3"), format = "latex", caption = "Real data example: Relative entropy calculated with different
choices of $K$ and $f_{W}(w)$.", booktabs=TRUE) |>
  kableExtra::kable_styling(position = "center") 
```

As shown in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:realentropytable-interactive)', '\\@ref(tab:realentropytable-static)'))`,  the maximum relative entropy is achieved with the combination of $K=2$ and $Gamma(3,3)$. Therefore, we set \(K=2\) and select \(f_W(\cdot)\) as the density of \(Gamma(3,3)\) for the rest of the analysis.

```{r, echo = TRUE}
set.seed(66)
finalmodel <- SLCARE(formula = "treatment + prev.resection", data = colorectalSLCARE,
       id_col = "id", start_col = "time0", stop_col = "time1", event_col = "new.lesions",
       K = 2, gamma = 3, boot = 200)
summary(finalmodel, digits = 3)
```

Based on the results from fitting models \@ref(eq:1) and \@ref(eq:2), we classify patients into two groups according to the modal class assignment rule, i.e., \(\hat{\xi} _i = \text{arg} \max _{1 \le k \le K} \hat{\tau} _{ik}\). Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:realclasstable-interactive)', '\\@ref(tab:realclasstable-static)'))`  (generated by R package \CRANpkg{table1} [@table1]) summarizes the characteristics of the resulting two groups, Class 1 of size 127 and Class 2 of size 23. Comparing the total number of the observed colorectal lesions, captured by $N_i(C_i)$, we note that patients belong to Class 1 tend to experience more colorectal lesions as compared to those belong to Class 2. There is also notable separation in covariates `treatment` and `prev.resection`. That is, over 50\% patients in Class 1 received combination chemotherapy in contrast to 0\% patients in Class 2 on combination chemotherapy, and over 70\% patients in Class 1 has previous resection, while all patients in Class 2 did not have resection in the past. These observations are consistent with the estimation results for $\alpha_0$ (see model summary), which suggest that patients receiving combination chemotherapy or with previous resection are much less likely to be classified to Class 2. These observations entail a plausible characterization of the two latent classes. That is, Class 1 consists of patients with more aggressive  progression that incurs  more intensive past or ongoing  treatment, while Class 2 is featured by relatively mild disease that requires less complicated treatment regimens.

We next examine the covariate effects within each latent class. The coefficient estimates presented in model summary indicate a reasonable direction of how `treatment` and `prev.resection` influence the intensity of lesion recurrence.  That is, combination chemotherapy or previous resection may help reduce the risk of lesion recurrence. However, the estimated effects which conform to our intuition are associated with p values that do not reach statistical significance.

```{r}
# Assign latent class membership.
latent_class <- apply(finalmodel$EstimatedTau[,-1],1, function(x) which.max(x))
# Extract covariates, number of observed recurrent events etc. from colorectal_SLCARE.
long_format <- colorectalSLCARE |> filter(new.lesions == 1)
wide_format <- colorectalSLCARE |> filter(new.lesions == 0)
# Calculate number of observed recurrent events
ID_d <- long_format |> group_by(id) |> summarise(observed.events = n())
# Construct a DataFrame with covariates, number of observed recurrent events, latent class membership assigned by SLCARE(). 
wide_format <- wide_format |> left_join(ID_d, by = "id") |> replace_na(list(observed.events = 0))
Z_d <- wide_format |> select(-c("id","time0","time1","new.lesions"))
bind_data_class <- cbind(Z_d, latent_class)
bind_data_class$treatment <- 
  factor(bind_data_class$treatment, 
         levels=c(0,1),
         labels=c("sequential chemotherapy", 
                  "combination chemotherapy"))
bind_data_class$prev.resection <- 
  factor(bind_data_class$prev.resection, 
         levels=c(0,1),
         labels=c("no", # Reference
                  "yes"))
bind_data_class$latent_class <- 
  factor(bind_data_class$latent_class, 
         levels=c(1,2),
         labels=c("Class 1", # Reference
                  "Class 2"))
pvalue <- function(x, ...) {
    # Construct vectors of data y, and groups (strata) g
    y <- unlist(x)
    g <- factor(rep(1:length(x), times=sapply(x, length)))
    if (is.numeric(y)) {
        # For numeric variables, perform a standard 2-sample t-test
        p <- t.test(y ~ g)$p.value
    } else {
        # For categorical variables, perform a chi-squared test of independence
        p <- chisq.test(table(y, g))$p.value
    }
    # Format the p-value, using an HTML entity for the less-than sign.
    # The initial empty string places the output on the line below the variable label.
    # c("", sub("<", "&lt;", format.pval(p, digits=3, eps=0.001)))
    # use "<" for kable
    c("", sub("<", "<", format.pval(p, digits=3, eps=0.001)))
}
table3 <- table1(~ observed.events + treatment + prev.resection | latent_class, data=bind_data_class, overall=F, extra.col=list(`P-value`=pvalue), caption="Real data example: Characteristics of the two latent classes.")
```

```{r realclasstable-interactive, eval = knitr::is_html_output(), layout = "l-body-outset"}
knitr::kable(as.data.frame(table3), booktabs=TRUE, format = "html", caption = "Real data example: Characteristics of the two latent classes.") |>
  kableExtra::kable_styling(position = "center")
```

```{r realclasstable-static, eval = knitr::is_latex_output()}
table1::t1kable(table3) |>
  kableExtra::kable_styling(position = "center")
```

In Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:realestmean-plotly)', '\\@ref(fig:realestmean-ggplot)'))`, we plot the average estimated mean function of experiencing new colorectal lesions for the two latent classes. Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:realestmean-plotly)', '\\@ref(fig:realestmean-ggplot)'))` shows that Class 1 is associated with a lower frequency of experiencing new colorectal lesions than Class 2. Such a distinction may be contributed by the underlying difference in disease severity between the two latent classes, which is also manifested by the different distributions of receiving combination chemotherapy and having previous resection between these two classes demonstrated in Table `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(tab:realclasstable-interactive)', '\\@ref(tab:realclasstable-static)'))`.

```{r realestmean-plotly, fig.height = 5, fig.cap = "Real data example: Estimated mean function of experiencing new colorectal lesions for two classes.", include=knitr::is_html_output(), eval=knitr::is_html_output(), layout = "l-body-outset"}
ggplotly(plot(finalmodel, type = "EstMeans"))
```

```{r realestmean-ggplot, fig.height = 3, fig.cap = "Real data example: Estimated mean function of experiencing new colorectal lesions for two classes.", include=knitr::is_latex_output(), eval=knitr::is_latex_output()}
plot(finalmodel, type = "EstMeans")
```

```{r realmodelcheck-plotly, fig.height = 5, fig.cap = "Real data example: Predicted numbers of new colorectal lesions versus the observed numbers of new colorectal lesions.", include=knitr::is_html_output(), eval=knitr::is_html_output(), layout = "l-body-outset"}
ggplotly(plot(finalmodel, type = "ModelChecking") +
  xlim(0, 5) + ylim(0, 5))
```

```{r realmodelcheck-ggplot, fig.height = 3, fig.cap = "Real data example: Predicted numbers of new colorectal lesions versus the observed numbers of new colorectal lesions.", include=knitr::is_latex_output(), eval=knitr::is_latex_output()}
plot(finalmodel, type = "ModelChecking") +
  xlim(0, 5) + ylim(0, 5)
```

We further check the overall fit of the latent class models adopted by `SLCARE()` to the
`colorectalSLCARE` dataset using the model checking plot described
in Section **Methodological background**. 
As shown in Figure `r knitr::asis_output(ifelse(knitr::is_html_output(), '\\@ref(fig:realmodelcheck-plotly)', '\\@ref(fig:realmodelcheck-ggplot)'))`, the dots representing the predicted and the observed numbers of new colorectal lesions are generally balanced around the 45-degree line when there are two observed colorectal lesions.  However, over-prediction and under-prediction are noted when the observed number of of colorectal lesions   is one or three respectively.  These observations may suggest a moderate lack-of-fit of the fitted model to this colorectical dataset.

# Discussion
The R package \CRANpkg{SLCARE} provides an easy-to-use software
to conduct latent class analysis of recurrent events based on a flexible
semiparametric multiplicative intensity modeling proposed by @zhao2022semiparametric. A practical automated initializer is embedded in \CRANpkg{SLCARE} to help users set initial estimates  in an informative way. \CRANpkg{SLCARE} provides a single command line function to implement full analysis with optional arguments for model customization. \CRANpkg{SLCARE} also provides visualization tools for result summary and model evaluation with \CRANpkg{ggplot2} plotting environment.

The R package \CRANpkg{SLCARE} is maintained in both Comprehensive R Archive Network
(CRAN) at <https://CRAN.R-project.org/package=SLCARE> and Github at
<https://github.com/qyxxx/SLCARE>. There are several interesting directions to further improve this package. For example, computational speed may be improved with more efficient coding. A new option may be added to perform automated selection of $K$ and frailty distribution. Such updates of package \CRANpkg{SLCARE} will be made available at CRAN and Github.
