---
title: "Supplementary Materials for ldmppr: Location Dependent Marked Point Processes in R"
date: "2026-08-14"
author:
  - name: Lane Drew
    affiliation: Colorado State University
    address:
      - Department of Statistics, Colorado State University
      - 851 Oval Dr, Fort Collins, CO 80524
    url: https://lanedrew.com/
    orcid: 0009-0006-5427-4092
    email: lane.drew@colostate.edu
  - name: Andee Kaplan
    affiliation: Colorado State University
    address:
      - Department of Statistics, Colorado State University
      - 851 Oval Dr, Fort Collins, CO 80524
    url: https://andeekaplan.com/
    email: andee.kaplan@colostate.edu
    orcid: 0000-0002-0912-0225
output:
  pdf_document:
    keep_tex: true
    template: "RJwrapper_sup.tex"
geometry: margin=1in
header-includes:
  - \usepackage{mathrsfs}
  - \newcommand{\defeq}{\stackrel{\tiny def}{=}}
  - \newcommand{\fct}[1]{\texttt{\detokenize{#1}()}}
  - \usepackage{pdflscape}
bibliography: ldmppr.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  fig.align = "center",
  out.width = "100%"
)
library(knitr)
library(readr)
library(dplyr)

safe_read_csv <- function(path) {
  if (!file.exists(path)) return(NULL)
  readr::read_csv(path, show_col_types = FALSE)
}
```

\appendix

\section{Additional details on initialization defaults for \\ \fct{estimate_process_parameters}}\label{appA}

We initialize self-correcting model parameters using data-adaptive heuristics designed to place the optimizer in a stable region of the parameter space. The baseline log-intensity is anchored by the empirical event rate via $\alpha_{\text{base}} = \log(n / \Delta t)$. The self-correction strength $\gamma_1$ is initialized on a conservative $\log(n) / n$-type scale to reduce unstable behavior early in optimization. To capture monotone temporal trends, $\beta_1$ is initialized from a Poisson regression of binned event counts on time (with an offset for bin width), with a fallback that induces a modest multiplicative change across the observation window.

Spatial inhibition range $\alpha_2$ is initialized using the median nearest-neighbor distance (or a conservative geometric fallback). Spatio-temporal interaction scales are bounded by the observation window: the spatial scale parameter is capped by the window diagonal, and the temporal interaction scale is capped by the observed time span (equal to 1 under the default time mapping). These choices provide stable starting values and natural finite bounds for optimization.

\section{Stage-wise runtime benchmarks}\label{appB}

To quantify fitting cost, we benchmarked runtime for \fct{estimate_process_parameters} and \fct{train_mark_model} on three contiguous datasets: (i) \texttt{medium\_example\_data}, (ii) a contiguous window with approximately 200 points, and (iii) a contiguous window with approximately 400 points. The larger datasets were selected as contiguous windows from the full domain (rather than random point subsamples) to preserve local dependence structure.

Stage-wise timing artifacts can be regenerated by running: \newline
\texttt{LDMPPR\_STAGE\_PROFILE=full} \newline
\texttt{LDMPPR\_STAGE\_REPS=10} \newline
\texttt{LDMPPR\_STAGE\_CORES=1,7} \newline
\texttt{LDMPPR\_STAGE\_DATASETS=medium,win200,win400} \newline
\texttt{Rscript "scripts/supplement\_stage\_timing\_study.R"}.

If \texttt{LDMPPR\_RASTER\_DIR} is not set, required external rasters are downloaded automatically from ESS-DIVE into \texttt{data/ess\_dive\_rasters}.

```{r stage_timing_table}
stage_manifest <- safe_read_csv("timing_outputs/stage_timing_datasets_full.csv")
if (is.null(stage_manifest)) stage_manifest <- safe_read_csv("timing_outputs/stage_timing_datasets_quick.csv")

stage_summary <- safe_read_csv("timing_outputs/stage_timing_summary_full.csv")
if (is.null(stage_summary)) stage_summary <- safe_read_csv("timing_outputs/stage_timing_summary_quick.csv")

if (is.null(stage_summary)) {
  cat("No stage timing outputs found. Run scripts/supplement_stage_timing_study.R to generate these tables.")
} else {
  if (!is.null(stage_manifest)) {
    knitr::kable(stage_manifest, caption = "Datasets used in stage-wise timing benchmarks.")
  }

  tbl <- stage_summary |>
    mutate(
      estimate_mean = round(estimate_mean, 2),
      estimate_sd = round(estimate_sd, 2),
      train_mean = round(train_mean, 2),
      train_sd = round(train_sd, 2),
      total_mean = round(total_mean, 2),
      total_sd = round(total_sd, 2)
    ) |>
    select(dataset_id, n_points, window_side, cores, runs,
           estimate_mean, estimate_sd, train_mean, train_sd, total_mean, total_sd)

  knitr::kable(tbl, caption = "Stage-wise runtime summary (seconds): process estimation and mark-model training.",
               col.names = c("Dataset", "Points", "Window Side", "Cores", "Runs",
                             "Est. Mean", "Est. SD", "Train Mean", "Train SD", "Total Mean", "Total SD"))
}
```

```{r stage_timing_interp, results='asis'}
if (!is.null(stage_summary)) {
  s1 <- stage_summary |> filter(cores == 1) |> arrange(n_points)
  if (nrow(s1) >= 2) {
    n_min <- s1$n_points[1]
    n_max <- s1$n_points[nrow(s1)]
    t_est_min <- s1$estimate_mean[1]
    t_est_max <- s1$estimate_mean[nrow(s1)]
    t_tr_min <- s1$train_mean[1]
    t_tr_max <- s1$train_mean[nrow(s1)]
    cat(sprintf("Using one core, process-estimation time increases from %.2fs (n=%d) to %.2fs (n=%d), while mark-model training increases from %.2fs to %.2fs over the same range.\n\n",
                t_est_min, n_min, t_est_max, n_max, t_tr_min, t_tr_max))
  }
}
```

\section{Simulation-check runtime benchmarks}\label{appC}

We benchmarked runtime for \fct{check_model_fit} using \texttt{n\_sim} values of 200, 500, 1000, and 2500, with fixed seeds and matched settings otherwise. This isolates simulation-based goodness-of-fit cost and provides practical guidance for exploratory versus final checks.

Simulation-check timing artifacts can be regenerated by running: \newline
\texttt{LDMPPR\_TIMING\_MODE=full} \newline
\texttt{LDMPPR\_TIMING\_N\_SIM=200,500,1000,2500} \newline
\texttt{LDMPPR\_TIMING\_REPS=10} \newline
\texttt{LDMPPR\_TIMING\_CORES=1,2} \newline
\texttt{Rscript "scripts/supplement\_sim\_timing\_study.R"}.

```{r sim_timing_table}
sim_summary <- safe_read_csv("timing_outputs/sim_timing_summary_full.csv")
if (is.null(sim_summary)) sim_summary <- safe_read_csv("timing_outputs/sim_timing_summary_quick.csv")

if (is.null(sim_summary)) {
  cat("No simulation timing outputs found. Run scripts/supplement_sim_timing_study.R with LDMPPR_TIMING_MODE=full to generate this table.")
} else {
  tbl <- sim_summary |>
    mutate(
      elapsed_mean = round(elapsed_mean, 2),
      elapsed_sd = round(elapsed_sd, 2),
      p_combined_mean = round(p_combined_mean, 3)
    ) |>
    select(n_sim, cores, runs, elapsed_mean, elapsed_sd, p_combined_mean)

  knitr::kable(tbl, caption = "Runtime summary (seconds) for simulation-based model checking.",
               col.names = c("Simulations", "Cores", "Runs", "Elapsed Mean", "Elapsed SD", "Combined $p$-value Mean"))
}
```

```{r sim_timing_interp, results='asis'}
if (!is.null(sim_summary)) {
  s1 <- sim_summary |> filter(cores == 1) |> arrange(n_sim)
  s2 <- sim_summary |> filter(cores == 2) |> arrange(n_sim)
  if (nrow(s1) > 0) {
    cat(sprintf("At one core, mean check time grows from %.2fs at n\\_sim=%d to %.2fs at n\\_sim=%d.\n\n",
                s1$elapsed_mean[1], s1$n_sim[1], s1$elapsed_mean[nrow(s1)], s1$n_sim[nrow(s1)]))
  }
  if (nrow(s1) > 0 && nrow(s2) > 0 && all(s1$n_sim == s2$n_sim)) {
    gain <- 100 * (1 - s2$elapsed_mean[nrow(s2)] / s1$elapsed_mean[nrow(s1)])
    cat(sprintf("For the largest run (n\\_sim=%d), moving from one core to two cores reduces mean wall-clock time by %.1f%%.\n\n",
                s1$n_sim[nrow(s1)], gain))
  }
}
```

For iterative model development, moderate simulation counts can reduce turnaround time. For final reporting, larger simulation counts provide more stable Monte Carlo $p$-value estimates.

For a direct, reproducible benchmark of the improved paper workflow (estimation + mark-model training + model check, matching the script in the main paper), run: \newline
\texttt{LDMPPR\_BENCH\_REPS=10} \newline
\texttt{Rscript "scripts/benchmark\_improved\_workflow\_timing.R"}.

\section{Description and interpretation of the \texttt{spatstat} comparison framework}\label{appD}

The comparison script \texttt{scripts/spatstat\_comp.R} implements a two-stage baseline intended to approximate the improved manuscript workflow as closely as possible while utilizing the \texttt{spatstat} modeling framework:

\begin{enumerate}
\item A spatial point process is fit with \texttt{spatstat::ppm}, using raster-derived covariates and a Strauss interaction term for regularity/inhibition.
\item A separate XGBoost mark model is fit from location/raster features with optional competition-index features. To align with the improved paper workflow, tuning uses MAE with 10-fold cross-validation and a tuning grid size of 150.
\item Simulated spatial patterns from the fitted \texttt{ppm} model are marked using the fitted XGBoost model then evaluated with the same LGFJEV and combined GET-rank envelope machinery used in the package-facing checks.
\end{enumerate}

For the direct comparison below we use \texttt{n\_sim = 2500} for both methods, matching the improved-paper model-check setting. For the \texttt{ldmppr} workflow, we match the paper's specifications exactly: estimation and model-check steps run without parallelization, while mark-model training uses \texttt{parallel = TRUE}. Comparison artifacts can be regenerated by running: \newline
\texttt{LDMPPR\_COMP\_N\_SIM=2500} \newline
\texttt{LDMPPR\_COMP\_PARALLEL=true} \newline
\texttt{Rscript "scripts/ldmppr\_comp\_timing.R"}
and \newline
\texttt{LDMPPR\_SPATSTAT\_N\_SIM=2500} \newline
\texttt{LDMPPR\_SPATSTAT\_CV\_FOLDS=10} \newline
\texttt{LDMPPR\_SPATSTAT\_TUNING\_GRID=150} \newline
\texttt{LDMPPR\_SPATSTAT\_FG\_CORRECTION=km} \newline
\texttt{LDMPPR\_SPATSTAT\_PARALLEL=true} \newline
\texttt{LDMPPR\_SPATSTAT\_NUM\_CORES=7} \newline
\texttt{Rscript "scripts/spatstat\_comp.R"}.

If \texttt{LDMPPR\_RASTER\_DIR} is not set, required external rasters are downloaded automatically from ESS-DIVE into \texttt{data/ess\_dive\_rasters}.

```{r comp_timing_table}
ld_summary <- safe_read_csv("timing_outputs/ldmppr_comp_summary.csv")
comp_summary <- safe_read_csv("timing_outputs/spatstat_comp_summary.csv")

if (is.null(ld_summary) || is.null(comp_summary)) {
  cat("No comparison summary outputs found. Run scripts/ldmppr_comp_timing.R and scripts/spatstat_comp.R to generate these tables.")
} else {
  ld_tbl <- ld_summary |>
    transmute(
      method,
      n_sim,
      fit_process_sec = fit_sec,
      fit_mark_sec,
      check_sec = check_sec,
      total_sec,
      p_combined
    )

  sp_tbl <- comp_summary |>
    transmute(
      method = "spatstat two-stage",
      n_sim,
      fit_process_sec = fit_loc_sec,
      fit_mark_sec,
      check_sec = sim_sec + check_sec,
      total_sec,
      p_combined
    )

  comp_tbl <- bind_rows(ld_tbl, sp_tbl) |>
    mutate(
      across(c(fit_process_sec, fit_mark_sec, check_sec, total_sec), ~round(.x, 2)),
      p_combined = round(p_combined, 4)
    ) |> select(-n_sim)

  knitr::kable(comp_tbl, caption = "Side-by-side timing and combined $p$-value comparison (ldmppr vs spatstat two-stage) for $n = 2500$ simulations in seconds.",
               col.names = c("Method", "Process fit", "Mark fit", "Check fit", "Total (s)", "Combined $p$"))
  # cat("\\n*For spatstat 2-stage, Check includes simulation + envelope-check time.\n")
}
```

```{r comp_timing_interp, results='asis'}
if (!is.null(ld_summary) && !is.null(comp_summary)) {
  ld <- ld_summary[1, ]
  sp <- comp_summary[1, ]
  cat(sprintf("Under matched comparison settings (n\\_sim=%d), the combined GET $p$-value is %.4f for ldmppr and %.4f for the two-stage spatstat baseline.\n\n",
              ld$n_sim, ld$p_combined, sp$p_combined))
  cat(sprintf("Total runtime is %.2fs for ldmppr versus %.2fs for the two-stage spatstat baseline; in the table above, the spatstat check time reports simulation plus envelope-check computation, while the ldmppr check time reports the corresponding integrated model-check step.\n\n",
              ld$total_sec, sp$total_sec))
}
```

```{r comp_pvals_table}
ld_p <- safe_read_csv("timing_outputs/ldmppr_comp_p_by_stat.csv")
sp_p <- safe_read_csv("timing_outputs/spatstat_comp_p_by_stat.csv")

if (is.null(ld_p) || is.null(sp_p)) {
  cat("No $p$-value by statistic outputs found. Run scripts/ldmppr_comp_timing.R and scripts/spatstat_comp.R.")
} else {
  p_tbl <- ld_p |>
    rename(p_ldmppr = p_value) |>
    left_join(sp_p |> rename(p_spatstat_two_stage = p_value), by = "stat")

  if (!is.null(ld_summary) && !is.null(comp_summary)) {
    p_tbl <- bind_rows(
      p_tbl,
      tibble(
        stat = "Combined",
        p_ldmppr = ld_summary$p_combined[1],
        p_spatstat_two_stage = comp_summary$p_combined[1]
      )
    )
  }

  p_tbl <- p_tbl |>
    mutate(across(starts_with("p_"), ~round(.x, 4)))

  knitr::kable(p_tbl, caption = "Side-by-side $p$-values by summary function, including the combined test $p$-value.",
               col.names = c("Statistic", "ldmppr $p$-value", "spatstat two-stage $p$-value"))
}
```

```{r comp_pvals_interp, results='asis'}
if (!is.null(ld_p) && !is.null(sp_p)) {
  p_tbl2 <- ld_p |>
    rename(p_ldmppr = p_value) |>
    left_join(sp_p |> rename(p_sp = p_value), by = "stat")
  sig_ld <- p_tbl2$stat[p_tbl2$p_ldmppr < 0.05]
  sig_sp <- p_tbl2$stat[p_tbl2$p_sp < 0.05]
  cat("At the 0.05 level, ldmppr shows significant departures for: ",
      if (length(sig_ld)) paste(sig_ld, collapse = ", ") else "none",
      ". The two-stage spatstat baseline shows significant departures for: ",
      if (length(sig_sp)) paste(sig_sp, collapse = ", ") else "none",
      ".\n\n", sep = "")
}
```

\begin{landscape}

```{r comp_check_plots, fig.height = 8, fig.width = 16}
library(ldmppr)
library(patchwork)
library(GET)

if (!file.exists("timing_outputs/ldmppr_comp_artifacts.rds") ||
    !file.exists("timing_outputs/spatstat_comp_artifacts.rds")) {
  knitr::include_graphics("ldmppr_supplement_files/figure-latex/comp_check_plots-1.pdf")
} else {
ld_art <- readRDS("timing_outputs/ldmppr_comp_artifacts.rds")
sp_art <- readRDS("timing_outputs/spatstat_comp_artifacts.rds")

p_ld <- plot(ld_art$check$combined_env) +
  ggplot2::ggtitle(paste0("ldmppr: Combined envelope (p = ", round(ld_summary$p_combined[1], 6),")")) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11),
    axis.title = ggplot2::element_text(size = 9),
    axis.text = ggplot2::element_text(size = 8)
  )
p_sp <- plot(sp_art$check$combined_env) +
  ggplot2::ggtitle(paste0("spatstat two-stage: Combined envelope (p = ", round(comp_summary$p_combined[1], 6),")")) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 11),
    axis.title = ggplot2::element_text(size = 9),
    axis.text = ggplot2::element_text(size = 8)
  )

p_ld + p_sp + patchwork::plot_layout(ncol = 2)
}
```

\end{landscape}

```{r comp_data_plots, fig.height = 10, fig.width = 14}
library(ldmppr)
library(ggplot2)
library(patchwork)
library(spatstat.geom)

if (!file.exists("timing_outputs/ldmppr_comp_artifacts.rds") ||
    !file.exists("timing_outputs/spatstat_comp_artifacts.rds")) {
  knitr::include_graphics("ldmppr_supplement_files/figure-latex/comp_data_plots-1.pdf")
} else {
ld_art <- readRDS("timing_outputs/ldmppr_comp_artifacts.rds")
sp_art <- readRDS("timing_outputs/spatstat_comp_artifacts.rds")

to_df_ppp <- function(pp) {
  xy <- spatstat.geom::coords(pp)
  data.frame(
    x = xy[, 1],
    y = xy[, 2],
    marks = as.numeric(spatstat.geom::marks(pp))
  )
}

ref_df <- to_df_ppp(ld_art$reference)
ld_df <- to_df_ppp(ld_art$sim$mpp)
sp_df <- to_df_ppp(sp_art$sim_one)

ref_df$marks <- as.numeric(ref_df$marks)
ld_df$marks <- as.numeric(ld_df$marks)
sp_df$marks <- as.numeric(sp_df$marks)

xy_bounds <- c(
  min(ref_df$x, na.rm = TRUE),
  max(ref_df$x, na.rm = TRUE),
  min(ref_df$y, na.rm = TRUE),
  max(ref_df$y, na.rm = TRUE)
)

sp_mpp <- generate_mpp(
  locations = sp_df[, c("x", "y")],
  marks = sp_df$marks,
  xy_bounds = xy_bounds
)

ref_plot <- plot_mpp(
  mpp_data = ld_art$reference,
  pattern_type = "reference"
)

ld_sim_plot <- plot(
  ld_art$sim,
  pattern_type = "simulated"
)

sp_sim_plot <- plot_mpp(
  mpp_data = sp_mpp,
  pattern_type = "simulated"
) +
  ggplot2::labs(title = "spatstat Simulated Data")

hist_df <- bind_rows(
  data.frame(marks = ref_df$marks, source = "Reference Mark Distribution"),
  data.frame(marks = ld_df$marks, source = "ldmppr Simulated Mark Distribution"),
  data.frame(marks = sp_df$marks, source = "spatstat Simulated Mark Distribution")
)

hist_df <- hist_df |>
  mutate(source = factor(source, levels = c(
    "Reference Mark Distribution",
    "ldmppr Simulated Mark Distribution",
    "spatstat Simulated Mark Distribution"
  )))

hist_plot <- ggplot(hist_df, aes(x = marks)) +
  geom_histogram(binwidth = 100, position = "identity", alpha = 0.75, color = "black", linewidth = 0.25) +
  facet_wrap(~source, nrow = 1) +
  labs(x = "Mark", y = "Count") +
  theme(
    plot.title = element_text(size = 11),
    axis.title = element_text(size = 9),
    axis.text = element_text(size = 8),
    strip.text = element_text(size = 9)
  )

plot_design <- c("
                 111122223333
                 111122223333
                 111122223333
                 111122223333
                 444444444444
                 555555555555
                 555555555555
                 555555555555
                 555555555555")

data_plot <- (ref_plot +
                ggplot2::scale_size(breaks = c(0, 200, 400, 600, 800, 1000, 1200)) +
                ggplot2::theme(
                  legend.position = "bottom",
                  plot.title = ggplot2::element_text(size = 11),
                  axis.title = ggplot2::element_text(size = 9),
                  axis.text = ggplot2::element_text(size = 8),
                  legend.title = ggplot2::element_text(size = 9),
                  legend.text = ggplot2::element_text(size = 8)
                )) +
  (ld_sim_plot +
     ggplot2::scale_size(breaks = c(0, 200, 400, 600, 800, 1000, 1200)) +
     ggplot2::theme(
       legend.position = "none",
       plot.title = ggplot2::element_text(size = 11),
       axis.title = ggplot2::element_text(size = 9),
       axis.text = ggplot2::element_text(size = 8)
     ) +
     ggplot2::labs(title = "ldmppr Simulated Data")) +
  (sp_sim_plot +
     ggplot2::scale_size(breaks = c(0, 200, 400, 600, 800, 1000, 1200)) +
     ggplot2::theme(
       legend.position = "none",
       plot.title = ggplot2::element_text(size = 11),
       axis.title = ggplot2::element_text(size = 9),
       axis.text = ggplot2::element_text(size = 8)
     )) +
  patchwork::guide_area() +
  hist_plot +
  patchwork::plot_layout(design = plot_design, guides = "collect") +
  patchwork::plot_annotation(
    tag_levels = "a",
    tag_suffix = ")",
    theme = ggplot2::theme(plot.tag = ggplot2::element_text(size = 10))
  )

data_plot
}
```

This baseline provides useful context but is not a fully comparable benchmark for the mechanistic \texttt{ldmppr} workflow. In particular, temporal/self-correcting dynamics are not estimated in a single unified model, and mark generation is modularized as a second stage. We therefore interpret these results as contextual sensitivity evidence rather than a definitive head-to-head comparison.
