---
title: 'SIREN: A Hybrid CFA-EFA R Package for Controlling Acquiescence in Restricted
  Factorial Solutions'
abstract: |
  The **siren** package implements a two-step procedure that allows
  restricted (confirmatory) factor analytic (FA) solutions to be fitted
  in data matrices that have been previously 'cleaned' of the biasing
  effects of acquiescent responding (AR) by using an unrestricted
  (exploratory) FA specification. So, the procedure which is implemented
  is hybrid: i.e. (a) an unrestricted acquiescence (ACQ) factor is first
  fitted to the data, (b) the residual data (or covariance) matrix after
  the impact of ACQ has been partialled-out is obtained, and (c) a
  restricted FA solution is fitted to the residual matrix. Although the
  basic foundations of the procedure are known, it contains new
  methodological developments that are, all of them, implemented in the
  package. So, provided that fully or partially balanced scales are
  available, the researcher will be able to: (a) calibrate a
  multidimensional CFA solution which is free from AR, (b) assess the
  goodness of model-data fit of this solution, and (c) obtain individual
  score estimates in the content as well as in the ACQ factors. The
  functioning of the program is assessed by means of a simulation study,
  and illustrated with a toy example. Its usefulness is also
  demonstrated by using an illustrative example in the personality
  domain. **siren** is submitted to be a valuable tool for use in item
  CFA applications when AR is expected to be operating.
author:
- name: David Navarro-Gonzalez
  affiliation: Department of Psychology
  address:
  - University of Lleida
  - Spain
  - 0000-0002-9843-5058
  - |
    [david.navarro@udl.cat](david.navarro@udl.cat){.uri}
- name: Pere J. Ferrando
  affiliation: Department of Psychology
  address:
  - University Rovira i Virgili
  - Spain
  - 0000-0002-3133-5466
  - |
    [perejoan.ferrando@urv.cat](perejoan.ferrando@urv.cat){.uri}
- name: Fabia Morales-Vives
  affiliation: Department of Psychology
  address:
  - University Rovira i Virgili
  - Spain
  - 0000-0002-2095-0244
  - |
    [fabia.morales@urv.cat](fabia.morales@urv.cat){.uri}
- name: Ana Hernandez-Dorado
  affiliation: Department of Psychology
  address:
  - University Rovira i Virgili
  - Spain
  - 0000-0001-9502-9735
  - |
    [ahernandezd@professor.universidadviu.com](ahernandezd@professor.universidadviu.com){.uri}
date: '2025-08-07'
date_received: '2024-01-11'
journal:
  firstpage: 5
  lastpage: 23
volume: 17
issue: 1
slug: RJ-2025-001
citation_url: https://rjournal.github.io/
packages:
  cran:
  - lavaan
  - siren
  bioc: []
preview: preview.png
bibliography: hernandez-dorado.bib
CTV: ~
legacy_pdf: yes
legacy_converted: yes
output:
  rjtools::rjournal_article:
    self_contained: yes
    toc: no
    mathjax: https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-chtml-full.js
    md_extension: -tex_math_single_backslash
draft: no

---


:::::::::: article
## Introduction

Valid interpretation of typical-response or non-cognitive (personality,
attitude, interest, etc.) test scores requires that the item responses
that are to be calibrated and scored meet a series of conditions. Of
these, one of the more basic is that the responses truly reflect the
influence of the content variables intended to be measured and are not
affected by other systematic determinants unrelated to this content. In
particular, this article is concerned with Acquiescent Responding (AR):
the tendency to agree or endorse an item regardless of its content
(Messick 1966) as a response determinant. When AR is operating and is
not properly controlled, a series of invalidating effects can be
expected, both at the calibration level (biased item parameter estimates
and spurious evidence of multidimensionality) and at the scoring level
(scores that reflect a mixture of content and AR and so they cannot be
univocally interpreted).

Test designers and practitioners are generally aware of the potential
invalidating effects of AR, and use procedures for controlling them. Of
these, the most common is to develop balanced scales. Provided that the
content variables can be considered as continuous dimensions with two
end poles, in a fully balanced scale half of the items are keyed toward
one of the poles while the other half are keyed toward the opposite pole
(Savalei and Falk 2014; Vigil-Colet, Navarro-Gonzalez, and Morales-Vives
2020).

Statistical control of AR in test designs that use balanced scales is
generally based on factor analytic (FA) procedures, and essentially
entails explicitly modeling AR as an additional non-content factor. This
FA-based control operates at two levels (see Ferrando, Lorenzo-Seva, and
Chico 2003): first, at the level of the factor structure obtained in the
calibration stage (thus avoiding or minimizing the invalidating effects
mentioned above); and second, at the level of the factor score estimates
derived from the calibration structure (thus, providing "cleaner"
content score estimates that have a more univocal interpretation).

Within the general FA modeling, two main approaches exist at present
(Savalei and Falk 2014; Fuente and Abad 2020). The first is fully
confirmatory, and the solution is identified by restricting all the
loadings on the additional Acquiescence (ACQ) factor to have the same
unit value (Billiet and McClendon 2000). The second is exploratory or
semi-confirmatory ( (Ferrando, Lorenzo-Seva, and Chico 2003)): First,
(a) an unrestricted ACQ factor with (possibly) different loadings and
(b) an also unrestricted (EFA) direct "content" solution are obtained.
Second, the direct content solution is either analytically rotated
(fully exploratory solution) or rotated against a specified or
semi-specified target (semi-confirmatory solution). The pros and cons of
both approaches have been discussed and compared by Savalei and Falk
(2014) and Fuente and Abad (2020). Both studies concluded that the
confirmatory approach is more robust and user-friendly than the EFA with
target rotation. However, it is also more sensitive to violation of the
unit-weight loading assumption for the ACQ factor.

Within the general FA-based controlling procedure, two main approaches
exist at present (Savalei and Falk 2014; Fuente and Abad 2020). The
first is fully confirmatory, and, in it, the structural FA solution is
identified by restricting all the loadings on the additional
Acquiescence (ACQ) factor to have the same unit value (Billiet and
McClendon 2000). The second is exploratory or semi-confirmatory
(Ferrando, Lorenzo-Seva, and Chico 2003): First, an unrestricted ACQ
factor with (possibly) different loadings together with a direct (i.e.
non-rotated) unrestricted or exploratory (EFA) "content" solution are
obtained. Second, the direct content solution is either analytically
rotated (in a fully exploratory solution) or rotated against a specified
or semi-specified target (in a semi-confirmatory solution). The pros and
cons of both approaches have been discussed by Savalei and Falk (2014)
and Fuente and Abad (2020). Both studies concluded that the confirmatory
approach is more robust and user-friendly than the semi-confirmatory EFA
with target rotation. However, it is also more sensitive to violation of
the unit-weight loading assumption for the ACQ factor.

The aim of this paper is to propose and implement a "hybrid" approach,
named SIREN, that combines features of both the CFA and EFA approaches.
Furthermore, the procedure is comprehensive in that it is intended for
fitting multiple content solutions and can also be used with scales that
are not fully balanced (see below). Because we are using the same name
for the proposed procedure and the package that implements it, in the
remainder of the paper, we shall use the distinction "SIREN procedure"
and "**siren** package" when necessary so as to avoid confusion.

Most of the basic foundations of SIREN have been discussed in the FA
literature (e.g. Nunnally 1978). Furthermore, the approach we shall
propose is multi-step (see below), and the first step: estimating an
unrestricted ACQ factor, is, essentially, the same as that used in the
exploratory/semi-confirmatory approaches summarized below. So, this part
of the proposal will not be discussed in detail but relevant references
will be provided to the interested reader. On the other hand, the full
proposal contains new developments, and these are the ones that will be
discussed here in more detail.

### Basic general results and rationale of the proposal

Consider a set of $n$ items intended to measure $p$ common content
factors (e.g. personality dimensions). The basic FA model equation in
the population is:
$$\label{EQ1}
  \mathbf{Z} = \boldsymbol{\Lambda} \boldsymbol{\theta} + \boldsymbol{\Psi} \mathbf{E}   (\#eq:EQ1)$$
where **Z** is an *n × 1* random vector of observed item scores;
$\boldsymbol{\Lambda}$ is an $n \times p$ factor pattern matrix;
$\boldsymbol{\theta}$ is an $p\times1$ random vector of 'true' common
factor scores; $\boldsymbol{\Psi}$ is an $n\times n$ diagonal matrix of
unique-factor loadings, and $\mathbf{E}$ is an $n\times1$ random vector of unique
factor scores. With regards to scaling and assumed relations, the
observed item scores are in reduced form (centered around the mean), the
common factor scores and the unique scores are in standard scale (zero
mean and unit variance), and, finally, the unique scores are assumed to
be uncorrelated with the common factors and among them. In these
conditions, the reproduced covariance matrix among the $n$ item scores
as implied by model \@ref(eq:EQ1) is given by the structural equation:
$$\label{EQ2}
  \boldsymbol{\Sigma} = \boldsymbol{\Lambda} \boldsymbol{\Phi} \boldsymbol{\Lambda'}  + \boldsymbol{\Psi}^{2}   (\#eq:EQ2)$$
where $\boldsymbol{\Phi}$ is *p × p* correlation matrix containing the
correlations between the 'true' common factor scores. Generally, in the
applications considered here, the **Z** scores will not only be
mean-centered, but standardized scores, and so, the implied covariance
matrix $\boldsymbol{\Sigma}$ in Equation \@ref(eq:EQ2) will be a correlation matrix.

The main difference between an unrestricted (exploratory) and a
restricted (confirmatory) solution within the general model \@ref(eq:EQ2) is in the
constraints that are imposed to the pattern matrix
$\boldsymbol{\Lambda}$. In an unrestricted solution, only minimal
identification constraints are imposed, so that the common space:
$\boldsymbol{\Lambda} \boldsymbol{\Phi} \boldsymbol{\Lambda'}$ in Equation \@ref(eq:EQ2)
is not restricted
and multiple solutions of the same type, that fit all equally well, can
be obtained from each other by rotation. In a restricted solution, the
number of imposed restrictions makes the specified solution
$\boldsymbol{\Lambda} \boldsymbol{\Phi} \boldsymbol{\Lambda'}$ unique,
in the sense that it cannot be obtained by rotation of another solution
(see Joreskog 1969). Although a restricted solution can be obtained by
using different sets of constraints, the most usual consist of imposing
an independent-cluster structure (e.g. McDonald 2000): each item has
only a non-zero loading in one factor, having zero loadings in all the
others.

At this point, we will start to develop a small, artificial toy example
to help clarify the explanations that will follow. Suppose a
questionnaire made up of 8 factorially simple items that measure two
moderately correlated factors, so that the independent-cluster structure
in the population is:

::: {#tab1}
  ----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- ------------------------------------------------------------------------------------------------------------
   $\boldsymbol{\Lambda} =  \begin{bmatrix}  \begin{array}{rr}  0.7 & 0.0 \\ -0.7 & 0.0 \\ 0.7 & 0.0 \\ -0.7 & 0.0 \\  0.0  &  0.6 \\  0.0  & -0.6 \\  0.0  &  0.6 \\  0.0  & -0.6 \end{array}\end{bmatrix}$   $\boldsymbol{\Phi} = \begin{bmatrix} \begin{array}{rr} 1.0 &  0.3 \\ 0.3 &  1.0 \end{array} \end{bmatrix}$

  ----------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------------- ------------------------------------------------------------------------------------------------------------

  : (#tab:T1) Toy example: restricted CFA solution with two
  correlated content factors.
:::

A CFA estimation of this structure can be specified by constraining to
zero the 8 elements of $\boldsymbol{\Lambda}$ that should be zero and
freely estimating the remaining 8 loadings and the interfactor
correlation based on the sample correlation matrix **R**. Note that, for
a solution of this type to be defined, the practitioner must be able to
specify first, the number of content factors that the questionnaire
intends to measure (two in the example), and second, the specific items
that define each factor. Furthermore, the items are supposed to be all
factorially simple, so that each item is a marker of the factor it
measures and has negligible loadings on the remaining factors. These
conditions are not easy to achieve, but can be feasible at advanced
stages of test development.

Suppose now that the **content** structure of our example is that in
[Table 1](#tab1){reference-type="ref" reference="tab1"} but, at the same time,
the item responses are also partly affected by AR, conceptualized as an
additional non-content factor (see [Table 2](#tab2){reference-type="ref"
reference="tab2"} below). Now, even though the content structure was
correct, if the specified two-factor structure above was directly fitted
to **R**, two types of distorted results would be expected. First, the
goodness of model-data fit would not be good. Second, the loading and
inter-factor correlation estimates would be biased with respect to the
parameter values in (2) (see e.g. DeMars 2014; Ferrando and Lorenzo-Seva
2010).

The rationale of SIREN can now be explained from the results above. At
the calibration level, the basic idea is to obtain a corrected or
cleaned covariance/correlation matrix $\bf{R_{corr}}$ in which the
impact of the ACQ factor has been partialled-out. If this is done
correctly, a specified restricted CFA solution can be next fitted to
$\bf{R_{corr}}$ instead of $\bf{R}$ . This solution will now fit well,
and the 'true' content parameters in [Table 1](#tab1){reference-type="ref"
reference="tab1"} will be well recovered. At the scoring level, once the
ACQ and the content structures have been properly estimated, ACQ and
content individual score estimates (i.e. factor scores) can be next
obtained based on the unrestricted ACQ pattern and the restricted, CFA
content solution.

As mentioned above, in order to control for the impact of AR, the items
of the questionnaire have to be fully or partially balanced. In the
present scenario, the condition of full balance implies that, within
each factor, half of the items that define this factor are positively
keyed and the other half are negatively keyed. The condition of partial
balance implies here that all the factors contain positively and
negatively keyed items, but that the number of positive and negative
items is not the same at least in one factor (e.g. Lorenzo-Seva and
Ferrando 2009)

We shall now illustrate the points so far discussed with our toy
example. Suppose now that the full content plus ACQ structure in the
population is that in [Table 2](#tab2){reference-type="ref" reference="tab2"}.
As the content pattern loadings show, however, the practitioner has done
her work well and, within each factor, the items are fully balanced:
within each content factor, half of the loadings are positive and half
negative. As for the ACQ factor, (a) all the loadings are positive, and
(b) they are smaller in magnitude than the content loadings. Both
features are expected in empirical applications. First, AR is the
tendency to agree with the item regardless of the direction in content
(hence all loadings are expected to be positive). Second, in a
well-designed measure, the item responses are expected to be far more
determined by the content they measure than by ACQ.

::: {#tab2}
  ------------------------------- ------------------------------
      $\boldsymbol{\Lambda} =          $\boldsymbol{\Phi} =
            \begin{bmatrix}                \begin{bmatrix}
           \begin{array}{rrr}              \begin{array}{rrr}
           0.7  &  0.0 &  0.1 \\           1.0 &  0.3 &  0.0 \\
           -0.7 &  0.0 &  0.2 \\           0.3 &  1.0 &  0.0 \\
           0.7  &  0.0 &  0.3 \\           0.0 &  0.0 &  1.0 \\
           -0.7 &  0.0 &  0.3 \\              \end{array}
           0.0  &  0.6 &  0.3 \\            \end{bmatrix}$
           0.0  & -0.6 &  0.3 \\
           0.0  &  0.6 &  0.2 \\
           0.0  & -0.6 &  0.1 \\
               \end{array}
            \end{bmatrix}$

  ------------------------------- ------------------------------

  : (#tab:T2) Toy example: complete solution when ACQ is operating.
  Balanced items.
:::

Once a complete solution such as that in [Table 2](#tab2){reference-type="ref"
reference="tab2"} has been estimated, it can be next taken as a basis
for obtaining ACQ and content score estimates (i.e. factor scores) for
each individual. As discussed above, these scores will now be cleaner
and have a more univocal interpretation.

### General description of the procedure and relation with previous approaches

We propose a general multi-stage procedure in which the number of stages
depends on whether the test is fully or only partially balanced. So,
within each of the two conditions (full balance or partial balance), the
stages will be described separately. Conceptually, however, it is useful
to view the overall procedure as based on three general stages that are
common in both conditions. In the first stage, an ACQ factor is
estimated from the properties of the (partially or fully) balanced set
of items, and the impact of this factor on the inter-item correlation
matrix is partialled-out. In the second stage, a specified CFA solution
is fitted to the 'cleaned' correlation matrix. Finally, in the third
stage, individual score estimates are obtained from the hybrid solution
(i.e. the unrestricted ACQ factor and the restricted content-factor
solution).

Although the sequential rationale just described is conceptually the
clearest, the structural solution at the second stage above can actually
be specified and fitted in two ways. The first way directly follows from
the corrected-correlation-matrix concept: to fit a CFA solution to a
reduced correlation matrix which is free from ACQ. In the **siren**
package, this way of fitting the model corresponds to the 'resid' method
option (i.e. use the residual or corrected matrix as input for the CFA)
as defined below. The second way is to take the estimated ACQ loadings
obtained at the first stage as if they were fixed and known, and next to
specify a full CFA solution that includes an additional ACQ factor with
loadings fixed at the obtained values. In the **siren** package, this
second choice corresponds to the 'fixed' method option defined below. As
we shall see, the results from both approaches must be the same. In
either case, the corrected or residual correlation matrix is provided in
the output of the SIREN package under the heading "rresidmatrix".

The explanation above allows the relation between SIREN and previous
approaches to be discussed in more detail. If the solution is specified
in the first way above ('resid' method option), SIREN can be viewed as a
particular application of a residual covariance analysis approach, i.e.
to initially correct a covariance or a correlation matrix for unwanted
effects before it is used as input for further structural analyses (e.g.
Andrews 1984; Asparouhov and Muthen 1984; DeCastellarnau and Saris 2021;
Berge 2020)

If the second equivalent specification above: full CFA solution that
includes the additional fixed ACQ factor ('fixed' method option), is
used instead, then SIREN can be regarded as a modification of the CFA
approach initially proposed by Billiet and McClendon (2000). In effect,
in the latter, the ACQ loadings are fixed all of them to unity for
identification purposes (which is generally unrealistic). In contrast,
in SIREN these loadings are fixed at the (possibly different) values
estimated at the first stage.

### Background results and required general conditions

Consider again a fully-balanced questionnaire made up of $n$ items
(where n is even) that measure a set of (possibly related) traits
$\theta_1...\theta_l...\theta_m$, so that each item is a factorially
pure measure of one of the $m$ content factors plus of an acquiescence
factor $\theta_a$ which is unrelated to the content factors. For an
individual $i$ that responds to an item $j$ that measures content factor
$l$, the structural model in a $z$-score metric (mean 0 and variance 1)
is
$$\label{EQ3}
  z_{ij} = \lambda_{jl}\theta_{il}+\alpha_{ja}\theta_{ia}+\varepsilon_{ij}   (\#eq:EQ3)$$
This is a scalar specification of matrix equation \@ref(eq:EQ1)
based on an
independent-cluster pattern. Only a single content loading per item is
specified because the remaining content loadings are zero. The
$\alpha_{ja}$ loading is the loading item $j$ has on the ACQ factor.
Finally, the residual terms $\varepsilon$s have zero means, and are
uncorrelated with the factors or with one another.

The $z$ item responses in Equation \@ref(eq:EQ3) can be treated as categorical or (approximately)
continuous. In the first case, the standardized scores would correspond
to the continuous-unbounded "strength" latent variable that are assumed
to underlie the observed scores and generate them according to a
threshold mechanism (see Muthen 1993). In the second case, they are
directly the standardized item scores. From this general modeling, it
follows that the inter-item correlations are polychoric correlations in
the first case, and product-moment correlations in the second case (see
e.g. Ferrando and Lorenzo-Seva 2013, for further details). In the
**siren** package, the user decides how the item responses are to be
treated by using the 'corr' argument defined below, which has two
options: \"Pearson\" (product-moment correlations) or "Polychoric"
(Polychoric/Tetrachoric correlations). Once the correlation matrix has
been obtained, the results that follow are common for both treatments.

Consider now the (polychoric or product-moment) reduced inter-item
correlation matrix with communalities in the main diagonal (see Ferrando
and Lorenzo-Seva 2010). If (a) all the assumptions so far
(independent-cluster structure, full balance within factors) were met,
(b) the specified FA model was correct, and (c) the item communalities
were known, then the first centroid loading (e.g. Lawley 1960) for item
$j$ would correspond to the loading this item has on the ACQ factor
(i.e. $\alpha_{ja}$) in the population. In the unidimensional content
case, further details on this result can be obtained from Ferrando,
Lorenzo-Seva, and Chico (2003), and, provided that full within-factor
balance applies, the result also holds directly in the multidimensional
case. Indeed, the conditions above are only approximately met at best
(in particular, true communalities as assumed in the centroid method,
are never known; (see McDonald 1978). And, furthermore, the first
centroid loading is a sample estimate. For these reasons, our choice in
SIREN is to first obtain the first principal-axis or canonical factor by
using an efficient EFA estimation procedure: Minimum Rank Factor
Analysis (MRFA, Berge and Kiers 1991), and next rotate this factor
against the centroid vector that is used as a criterion (see Eysenck
1950). The final factor so obtained is taken in SIREN as an estimate of
the ACQ factor. As for the MRFA choice, it is because of its robustness
and because it provides estimates of the proportion of common variance
accounted for by the different factors, and this information is useful
for assessing the relevance of the ACQ factor in terms of explained
common variance. The proportion of ACQ explained common variance is
provided under the heading 'rACQvariance' in the SIREN package output.

Provided that the preliminary general conditions discussed above are
met, the correct functioning of the SIREN procedure once the ACQ factor
has been partialled-out depends on two main points. The first point is
of a mathematical nature, and is critical if unbiased loadings
(especially those of content factors) are to be obtained. The second is
of a statistical nature, and is important if the goodness-of-fit
assessment needs to be correct.

Obtaining unbiased loading estimates in SIREN mainly depends on
achieving full balance (either for the total test or for a core balanced
set), in principle, within each factor. In more detail, what is required
is that the sum of content loadings within each factor be equal to zero.
If it is, then the first centroid (or MRFA factor) will reflect only AR,
so partializing it from the correlation matrix will remove only this
response bias and leave a 'clean' corrected correlation that will
reflect only content. However, if balance is not achieved, then, to a
greater or lesser extent, the first centroid factor will reflect a
mixture of ACQ and content. So, some content will be removed by
partializing, and the resulting content loadings will be biased (see
Berge 2020). Having said that, however, we also note that the strict
condition of full balance within each factor (or the core set within
each factor) is probably too strong. Preliminary results by the writers
suggest that, provided that the content CFA solution is correct and full
balance holds for the entire set (or core set) of items, but not
necessarily for each factor, then SIREN would still perform correctly in
most cases. The extent to which the content loading estimates will
degrade as imbalance increases is best addressed using simulation, and
this will be done in the next section.

We turn now to the second, statistical point. Our proposal can be viewed
as a particular application of what (Nunnally 1978) called an ad-lib
factorial process, in which successive factors are fitted to residual
matrices by using different methods. We consider our proposal to be
indeed legitimate, however, its multi-stage nature necessarily entails a
loss of information and efficiency, because the successive estimates are
taken as fixed and known, and their uncertainty is not taken into
account. In agreement with authors such as (DeCastellarnau and Saris
2021; Nunnally 1978; Berge 2020) and the empirical results provided by
(Oberski and Satorra 2013), however, we believe that the impact of the
loss of efficiency discussed above on the point estimates and indices of
goodness of fit will be relatively minor in practice provided that the
proposed solution is correct and the basis conditions are reasonably
met. The issue, however, needs to be, and will be, assessed by using
simulation.

### Multi-stage approach with fully balanced scales

The fulfillment of the full balance condition is checked automatically
by the **siren** package by using the information provided in the
'target' argument (described below), in which the content pattern matrix
with the signed dominant loading of each item on its corresponding
factor is provided as input by the user. An example of a target would be
the first two columns of $\boldsymbol{\Lambda}$ in
[Table 2](#tab2){reference-type="ref" reference="tab2"}. The specific stages
in this case are:

Stage 1: for the $n$ test items, the ACQ factor loadings are estimated
by: (a) obtaining the direct (canonical) MRFA solution from the
inter-item correlation matrix, by specifying $m+1$ factors (b) obtaining
the first centroid from the inter-item correlation matrix (to be used as
criterion), and (c) rotating the MRFA solution to the position in which
the first canonical MRFA factor has maximal congruence with respect to
the criterion. Essentially, the process is a target rotation against a
single target vector (the first centroid). However, because the
canonical MRFA solution is orthogonal, the rotation is univocally
defined. Finally, the criterion of maximal congruence is defined in
least squares terms: i.e. the first rotated MRFA factor is that is the
closest to the target centroid in the least-squares sense.

Stage 2: obtain the corrected (i.e. ACQ free) inter-item residual matrix
as $\bf{R_{corr}}$ = $\bf{R}$
$- \boldsymbol{\alpha}$$\boldsymbol{\alpha}^{'}$. The $\bf{R_{corr}}$
matrix is, and should be treated as, a residual covariance matrix (not a
correlation matrix).

Stage 3: The prescribed CFA solution can be specified and fitted in two
alternative ways. The first is to input $\bf{R_{corr}}$ specified as a
covariance matrix to the SEM program, and request a standardized
solution. The output will consist of the standardized content pattern
with loadings that are free of ACQ. The second way is to input the raw
data to the SEM program, by specifying the prescribed CFA content
solution, plus an additional ACQ factor in which all the loadings are
specified as fixed and known. In this second specification, the ACQ
loadings $\boldsymbol{\alpha}$ obtained in stage 2 cannot be imputed
directly because $\bf{R_{corr}}$ is a covariance matrix (i.e.
unstandardized) while the loadings on $\boldsymbol{\alpha}$ are
standardized. The correct values to be imputed are thus those obtained
by multiplying each standardized loading on $\boldsymbol{\alpha}$ by the
corresponding item standard deviation:
$\boldsymbol{\alpha_{x}}$(scaled)$=\boldsymbol{\alpha_{x}}\bf{sx}$. This
scaling transforms the standardized loadings $\boldsymbol{\alpha}$ into
unstandardized loadings (e.g. Bollen 1989). As in the first approach, a
standardized solution will next be requested in the output. If this is
done, the output will now provide the standardized content pattern with
loadings free from ACQ and an additional column containing the
standardized ACQ loadings. Indeed, the standardized content pattern must
be the same in both specifications. In the pkg siren programming, the
CFA in stage 3 is done by using the cfa function from the
[**lavaan**](https://CRAN.R-project.org/package=lavaan) and one of the
two options described above. However, of the multiple choices that the
program allows for estimating the structural item content parameters, we
have chosen the one that is most congruent with the previous stages.
Thus, whether the variables are treated as continuous or discrete, the
estimation procedure is robust ULS, in agreement with the choice of
MRFA-ULS for obtaining the ACQ estimates.

Stage 4: testing model-data fit at the structural level. Again, we have
made choices within
[**lavaan**](https://CRAN.R-project.org/package=lavaan) that are in
accordance with the limited-information nature of the procedure as well
as with the results of the simulation study. The chosen indices are: (a)
the RMSR and GFI as overall measures of misfit (McDonald and Mok 1995),
(b) the RMSEA as a measure of relative fit with respect to the degrees
of freedom (i.e. model complexity), and (c) the CFI as a measure of
comparative fit with respect to the null independence model (see Tanaka
1993, for points b and c). These indices are provided in the SIREN
package output under the heading 'rfit_indices'.

Stage 5: obtaining individual score estimates. In the basic FA equation
discussed above, the factor score estimates for each individual are the
estimates of the 'true' scores $\boldsymbol{\theta}$ in Equation \@ref(eq:EQ1), which, of course,
are unknown. For both, the linear and the nonlinear models, the factor
score estimates are Bayes Modal a Posteriori (MAP), which, in the
continuous (linear) model, are known as regression estimates. In both
cases, Bayesian scoring provides finite and plausible estimates for all
the respondents under study (see Ferrando and Lorenzo-Seva 2016). For
each participant, the output information consists of the point estimate
of his/her level on the content factors plus his/her factor score
estimate on the ACQ factor. This last estimate can be interpreted as the
predisposition of the individual to engage in AR. The factor score
estimates are provided under the heading 'rp_factors' (see below).

Stages 1 to 4 in the fully-balanced procedure will be now illustrated
with our toy example. Furthermore, the effects of ignoring the secondary
ACQ factor will be illustrated by fitting directly the content solution
in table 1 to the uncorrected (i.e observed) correlation matrix. To
perform the illustration, we generated a random sample of $N=500$
simulees from a population in which the complete solution in
[Table 2](#tab2){reference-type="ref" reference="tab2"} holds. The results are
in [Table 3](#tab3){reference-type="ref" reference="tab3"}.


```{r tab3, echo = FALSE, results='asis'}
library(kableExtra)
library(dplyr)

# Create the table data structure using HTML entities
toy_data <- data.frame(
  `Siren` = c(
    "$$\\boldsymbol{\\Lambda} = \\begin{bmatrix} 0.758 & 0.000 & 0.118 \\\\ -0.664 & 0.000 & 0.175 \\\\ 0.620 & 0.000 & 0.311 \\\\ -0.723 & 0.000 & 0.257 \\\\ 0.000 & 0.605 & 0.382 \\\\ 0.000 & -0.583 & 0.272 \\\\ 0.000 & 0.681 & 0.137 \\\\ 0.000 & -0.582 & 0.165 \\end{bmatrix}$$",
    "$$\\boldsymbol{\\Phi} = \\begin{bmatrix} 1 & .33 & 0 \\\\ .33 & 1 & 0 \\\\ 0 & 0 & 1 \\end{bmatrix}$$",
    "CFI = 1<br/>GFI = .998<br/>RMSEA = 0<br/>SRMR = .024"
  ),
  `Direct` = c(
    "$$\\boldsymbol{\\Lambda} = \\begin{bmatrix} 0.761 & 0.000 \\\\ -0.693 & 0.000 \\\\ 0.631 & 0.000 \\\\ -0.727 & 0.000 \\\\ 0.000 & 0.549 \\\\ 0.000 & -0.552 \\\\ 0.000 & 0.720 \\\\ 0.000 & -0.586 \\end{bmatrix}$$",
    "$$\\boldsymbol{\\Phi} = \\begin{bmatrix} 1 & .35 \\\\ .35 & 1 \\end{bmatrix}$$",
    "CFI = .922<br/>GFI = .974<br/>RMSEA = .088<br/>SRMR = .049"
  ),
  stringsAsFactors = FALSE
)

# Create the table with proper R Markdown integration
tab3 <- kable(toy_data,
      align = "cc",
      escape = FALSE,
      caption = "Toy Example Results",
      col.names = c("Siren", "Direct")) |>
  kable_styling(full_width = FALSE, position = "center") |>
  pack_rows("Loading Matrices", 1, 1, bold = TRUE) |>
  pack_rows("Phi Matrices", 2, 2, bold = TRUE) |>
  pack_rows("Goodness of Fit Indices", 3, 3, bold = TRUE) |>
  footnote(general = "GOF=Goodness of Fit Indices")
tab3 <- gsub("&amp;", "&", tab3)
tab3
```

When an operating common factor is left unmodeled, both, biased
structural estimates of the remaining parameters and deterioration of
the GOF indicators are expected. So, in general terms, results in
[Table 3](#tab3){reference-type="ref" reference="tab3"} are predictable.
Ferrando and Lorenzo-Seva (2010) and DeMars (2014) can be summarized as
follows. With regards to bias, SIREN does a good job, and recovers quite
acceptably (given both the sample and model size) all the parameters in
the toy solution: content loadings, ACQ loadings, and inter-factor
correlation. On the other hand, the loading estimates in the
uncontrolled solution tend to be slightly more biased, and the
inter-factor correlation slightly over-estimated. In terms of GOF, that
of the SIREN solution is almost perfect by all standards, which is only
to be expected, as the specified solution is correct. The GOF of the
uncorrected solution, however, clearly deteriorates. The amount of
misfit is not terribly bad here, which is also expected in such a small
model. In a larger-sized example, however, the deterioration of fit
would have been much stronger.

### Multi-stage approach with partially balanced scales

The basic idea in this case is to first obtain a fully balanced sub-set
of items, which we shall denote as the core set, and estimate the ACQ
loadings of the items belonging to this core by using the procedure
described above. Next, the ACQ loadings of the remaining items are
estimated by using a type of extension analysis based on the method of
moments. Once the ACQ loading estimates are available for all the test
items, the rest of the procedure can be carried out exactly as in the
fully balanced case. So, the only two points that require specific
discussion are: first, how to determine which items will be included in
the core set, and second, how the ACQ loadings of the remaining items
will be determined.

Stage 1: Choosing the core set. Within each specified factor, the
positive and negative items, as specified in the 'target' argument are
separated into two groups, and a centroid FA is performed separately in
each of the two resulting inter-item correlation matrices. The loadings
in the smaller set (usually that containing the negative items) are
taken as fixed, and each of them is paired with the positive loading
with the most similar value. The aim is for the absolute value of the
sums of the positive and negative loadings to be as similar as possible.
The rationale is that the effect of ACQ will be in the same direction if
items are all worded in the same direction (i. e. both the positive and
the negative loadings will be upwardly biased).

Stage 2: For the $n_c$ items in the core subset, the loadings on the ACQ
factor are estimated using the procedure described in the fully balanced
case.

Stage 3: Denote by $X_o$ an item outside the core set, and let
$j=1,\cdots n_c$ be the items in the core. Let $\sum_{j=1}^{n_c}r_{oj}$
be the sum of the correlations of item $X_o$ with the remaining items in
the core set (see Equation \@ref(eq:EQ5)
for deriving these correlations). If the core items are balanced, all
the terms involving the sum of content loadings will vanish, and it will
then follow that then follows that
$$\label{EQ4}
  \sum_{j=1}^{n_c}r_{oj}=\alpha_{oa}\sum_{j=1}^{n_c}\alpha_{ja}   (\#eq:EQ4)$$
So
$$\label{EQ5}
  \alpha_{oa}=\frac{\sum_{j=1}^{n_c}r_{oj}}{\sum_{j=1}^{n_c}\alpha_{ja}}   (\#eq:EQ5)$$
In words, if full balance holds for the core set, then the quotient
between (a) the sum of correlations of item $X_o$ with the remaining
items in the core set, and (b) the sum of ACQ loadings in the core set
provides a simple estimate of the loading of item $X_o$ on the ACQ
factor. Note that the sum in the denominator of Equation \@ref(eq:EQ5)
is taken as fixed
and known and has been obtained in Stage 2 above. The estimate described
above can be viewed as an extension-analysis estimate (e.g. McDonald
1978) obtained by the method of moments.

The extension estimate Equation \@ref(eq:EQ5) is computed on an item-by-item basis for each of the
items outside the core set in Stage 3. So, at the end of this stage, ACQ
loading estimates are available for all the test items under study. This
is the same situation as at the end of Stage 1 in the
fully-balanced-case approach. Therefore, from this point on, the
procedure is the same in both cases.

In closing this section, it is important to note that the **siren**
package automatically detects if the scales are only partially balanced
through the information provided by the 'target' argument, and, if so,
also automatically carries out the three-step procedure described in
this section.

## The **siren** package details

Available through CRAN, the
[**siren**](https://CRAN.R-project.org/package=siren) package contains
one main function (and additional internal functions) called
`acquihybrid`, which implements the procedures described in the sections
above.

The function usage is the following:

``` r
  acquihybrid(x, content_factors, target, corr = "Pearson", raw_data=TRUE,
  method="fixed", display = TRUE)
```

in which the arguments are:

`x`, raw sample item scores or a covariance/correlation matrix, as a
data.frame or a numerical matrix,

`content_factors`, the number of content factors in the CFA solution.
Each factor has to be defined by at least 3 items,

`target`, the pattern loading target matrix, which provides the signed
dominant loading (higher in absolute value) of each item on its
corresponding factor. The target is only used as a reference for
assessing which items have significant loadings on which factors. The
specific loading estimates are not used,

`corr`, determines the type of matrices to be used in the factor
analysis. \"Pearson\": Computes Pearson inter-item correlation matrices
(linear FA model); \"Polychoric\": Computes Polychoric/Tetrachoric
inter-item correlation matrices (non-linear FA/graded model),

`raw_data`, logical argument, if TRUE, the entered data will be treated
as raw item scores (default). If FALSE, the entered data will be treated
as an inter-item covariance/correlation matrix,

`method`, two choices are provided: fixed, which uses the ACQ loadings
obtained in the first step to specify the ACQ factor in the CFA solution
based on the raw scores, and resid, which uses the ACQ-free corrected
covariance matrix as input for the CFA,

`display`, determines if the output will be displayed in the console,
TRUE by default. If it is TRUE, the output is printed in the console and
if it is FALSE, the output is returned silently to the output variable.

The data provided should be a data frame or a numerical matrix for input
vectors and matrices, character variables for `corr` and method
arguments, and logical values for `raw_data` and `display` arguments.

The `acquihybrid` function returns a list variable, containing the
following variables:

`rloadings`, the factor loadings for each content factor and
acquiescence factor.

`rfactor_cor`, content factor correlations.

`rfit_indices`, a sub-list including a variety of popular fit indices as
described above.

`rACQ_variance`, the amount of common variance explained by ACQ.

`rresid_matrix`, residual matrix after partialling-out for ACQ.

`rpfactors`, factor scores for each participant.

The package includes a detailed vignette titled "siren-vignette", which
provides step-by-step explanations of how to use the package, as well as
guidance on interpreting the data. The vignette uses the same dataset as
the illustrative example below. The vignette is accessible through:

`vignette("siren-vignette")`

## Simulation studies

To assess the behavior of the proposal under favorable conditions
(correct population model) and its robustness against slight
misspecifications, we conducted a simulation study which focused on both
the recovery of the 'true' loadings in the ACQ and the content factors,
and the goodness of fit results.

### Method

A bidimensional content model with an additional ACQ factor (see
Equation 1) was generated under the following specifications: (a) all
the factors were orthogonal (this choice was made for simplicity); (b)
the content factors contained positive and negative loadings
(representing the positively and negatively keyed items); and (c) the
loadings on the ACQ factor were all positive. Referring to Equation 1,
the structure of the simulated model can be understood, as it defines
how the observed scores are reproduced from the common factors and
unique loadings, facilitating the interpretation of the factor analysis
results. The number of items per factor was 10, and the sample size was
fixed to 300, a value slightly higher than recommended to find accurate
factor loadings estimates (Fabrigar et al. 1999). To help the reader to
visually understand the design of the simulation study, we provide a
simplified version of the path diagram in
[Figure 1](#fig:1){reference-type="ref" reference="fig:1"}.

![Figure 1: Simplified version of the path diagram of the simulated
model.](toyimage.jpeg){#fig:1 width="80.0%"
alt="graphic without alt text"}

In the content factors, the simulated loadings had an average value of
.6 (in absolute value) and a standard deviation of 0.1. For the ACQ
factor the mean loading value was .2. The behavior of **siren** was
assessed under three general conditions: (1) type of item: ordinal (four
categories; FA based on polychoric correlations), or continuous (FA
based on Pearson correlations); (2) pattern of substantive loadings at
three levels: (a) completely balanced, (b) 60% of positive items, and
(c) 70% of positive items; and (3) ACQ pattern at three levels: (a)
equal ACQ loadings, (b) low heterogeneity (standard deviation of .01),
and (c) high heterogeneity (standard deviation of .1). Thus, a factorial
design with 18 experimental conditions (2 x 3 x 3) was used. These
conditions were chosen according to (a) the most problematic conditions
for the alternative method discussed above, and (b) the degree of
realism in the applied context.

For each experimental condition, 200 replicas were generated. A higher
number of replicas would simply increase the estimation time without
changing the results. All analyses were conducted with R (R Core Team
2024). The quality of the estimates was assessed using the average bias
$$\label{EQ6}
  \text{Average Bias} = \frac{1}{n} \sum_{i=1}^{n} (\theta_i - \hat{\theta}_i)   (\#eq:EQ6)$$
where $\theta_i$ are the observed or true values, $\hat{\theta}_i$ are
the predicted or estimated values, and $n$ is the total number of
observations; and the consistency of the model-data fit results was
assessed using an analysis of variance (ANOVA) for each fit index
considered in the study (CFI, GFI, RMSR and RMSEA).

### Results

The results for the average bias were very stable in all conditions (see
[Table 4](#tab4){reference-type="ref" reference="tab4"} and
[Table 5](#tab5){reference-type="ref" reference="tab5"}). The loadings
on content factors are recovered more accurately than the loadings on
the ACQ factor. In the content factors, the bias is evenly distributed
across both factors, and never exceeds .05 (continuous condition) or .04
(ordinal condition) in absolute value. In contrast, the average bias in
the ACQ factor is greater in the ordinal case.

```{r tab4, echo = FALSE}
library(kableExtra)
library(dplyr)

# Table 4: Average biases of the content factor loadings
content_data <- data.frame(
  Condition1 = c(rep("Bal", 3), rep("LU", 3), rep("HU", 3), rep("Bal", 3), rep("LU", 3), rep("HU", 3)),
  Condition2 = c(rep(c("EQ", "LH", "HH"), 6)),
  Item1 = c(.047, .047, .046, .040, .044, .043, .044, .043, .050, .040, .040, .036, .039, .039, .037, .036, .037, .038),
  Item2 = c(.044, .044, .043, .044, .046, .043, .047, .045, .044, .036, .036, .036, .035, .034, .037, .036, .038, .037),
  Item3 = c(.045, .045, .047, .047, .043, .038, .041, .046, .046, .036, .036, .036, .037, .038, .039, .031, .036, .033),
  Item4 = c(.047, .047, .047, .044, .044, .043, .041, .045, .045, .039, .039, .038, .040, .038, .040, .039, .040, .042),
  Item5 = c(.045, .045, .045, .043, .044, .043, .048, .045, .047, .039, .039, .034, .034, .037, .036, .037, .037, .038),
  Item6 = c(.043, .043, .044, .046, .043, .046, .046, .041, .043, .038, .038, .037, .038, .037, .038, .037, .035, .038),
  Item7 = c(.045, .045, .045, .044, .046, .043, .045, .045, .044, .036, .036, .037, .035, .037, .034, .036, .036, .039),
  Item8 = c(.042, .042, .045, .042, .043, .042, .042, .040, .047, .037, .037, .037, .035, .034, .037, .041, .036, .036),
  Item9 = c(.040, .046, .043, .043, .043, .046, .049, .043, .050, .035, .035, .036, .033, .040, .038, .039, .041, .035),
  Item10 = c(.041, .041, .044, .043, .045, .050, .050, .045, .049, .041, .041, .036, .036, .039, .039, .037, .038, .038),
  Item11 = c(.044, .047, .045, .040, .045, .046, .044, .048, .046, .038, .039, .037, .032, .040, .035, .035, .036, .040),
  Item12 = c(.043, .043, .046, .043, .041, .044, .046, .045, .049, .039, .042, .035, .038, .037, .037, .040, .038, .040),
  Item13 = c(.042, .044, .040, .045, .041, .045, .044, .046, .042, .037, .037, .036, .038, .035, .037, .037, .035, .039),
  Item14 = c(.046, .043, .046, .045, .039, .044, .045, .043, .048, .038, .036, .040, .035, .033, .037, .033, .035, .038),
  Item15 = c(.045, .045, .043, .045, .043, .042, .044, .048, .046, .035, .035, .040, .036, .039, .040, .036, .041, .038),
  Item16 = c(.046, .043, .043, .042, .042, .044, .046, .040, .047, .035, .034, .037, .036, .034, .040, .035, .037, .043),
  Item17 = c(.041, .041, .044, .043, .043, .042, .046, .043, .049, .036, .038, .037, .040, .036, .037, .038, .036, .035),
  Item18 = c(.049, .046, .050, .044, .043, .043, .049, .045, .045, .040, .039, .038, .035, .037, .036, .044, .036, .038),
  Item19 = c(.042, .040, .045, .042, .043, .041, .041, .043, .044, .037, .037, .038, .037, .039, .037, .040, .039, .037),
  Item20 = c(.044, .041, .042, .043, .042, .039, .046, .045, .046, .031, .034, .037, .037, .034, .033, .042, .039, .038),
  stringsAsFactors = FALSE
)

# Add section indicators
content_data$Section <- c(rep("Continuous", 9), rep("Ordinal", 9))

# Reorder columns to put Section first
content_data <- content_data[, c("Section", "Condition1", "Condition2", paste0("Item", 1:20))]

kable(content_data[, -1], # Remove Section column for display
      caption = "Average biases of the content factor loadings",
      col.names = c("", "", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10",
                    "11", "12", "13", "14", "15", "16", "17", "18", "19", "20"),
      escape = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 10) |>
  add_header_above(c(" " = 2, "factor 1" = 10, "factor 2" = 10)) |>
  pack_rows("Continuous", 1, 9, bold = TRUE) |>
  pack_rows("Ordinal", 10, 18, bold = TRUE) |>
  footnote(general = "Bal = Balanced; LU = Low Unbalanced; HU = High Unbalanced; EQ = equal; LH = Low Heterogeneity; HH = High Heterogeneity",
           general_title = "Note: ") |>
  landscape()
```

```{r, echo = FALSE}
# Table 5: Average biases of the ACQ factor loadings
acq_data <- data.frame(
  Condition1 = c(rep("Bal", 3), rep("LU", 3), rep("HU", 3), rep("Bal", 3), rep("LU", 3), rep("HU", 3)),
  Condition2 = c(rep(c("EQ", "LH", "HH"), 6)),
  Item1 = c(.063, .070, .063, .058, .059, .067, .068, .063, .062, .069, .075, .059, .069, .073, .064, .076, .072, .060),
  Item2 = c(.066, .069, .057, .059, .061, .058, .071, .058, .062, .075, .076, .065, .062, .061, .069, .071, .069, .069),
  Item3 = c(.064, .072, .058, .059, .060, .063, .061, .066, .064, .072, .078, .068, .067, .074, .071, .067, .070, .073),
  Item4 = c(.074, .072, .060, .056, .057, .059, .066, .063, .068, .067, .070, .073, .070, .071, .065, .073, .077, .074),
  Item5 = c(.068, .066, .061, .058, .066, .056, .065, .064, .059, .074, .079, .072, .069, .074, .072, .070, .078, .070),
  Item6 = c(.064, .075, .061, .059, .060, .056, .059, .061, .062, .075, .080, .066, .072, .073, .072, .074, .068, .071),
  Item7 = c(.063, .063, .064, .058, .061, .059, .063, .068, .063, .079, .078, .065, .073, .068, .074, .073, .068, .076),
  Item8 = c(.066, .068, .062, .066, .059, .062, .062, .071, .067, .076, .079, .066, .073, .076, .068, .080, .086, .078),
  Item9 = c(.068, .072, .061, .064, .062, .061, .070, .068, .071, .074, .069, .070, .076, .073, .073, .069, .079, .077),
  Item10 = c(.064, .073, .060, .064, .072, .060, .070, .075, .065, .073, .083, .071, .070, .065, .076, .089, .078, .074),
  Item11 = c(.070, .068, .061, .061, .066, .061, .072, .071, .066, .071, .072, .065, .070, .069, .067, .081, .078, .077),
  Item12 = c(.068, .066, .063, .064, .063, .063, .066, .068, .071, .077, .077, .073, .066, .076, .068, .075, .077, .063),
  Item13 = c(.071, .063, .065, .058, .060, .056, .064, .069, .056, .078, .079, .067, .069, .065, .063, .078, .074, .076),
  Item14 = c(.064, .068, .065, .061, .063, .056, .067, .063, .056, .076, .072, .071, .068, .073, .068, .075, .081, .067),
  Item15 = c(.064, .068, .064, .064, .065, .059, .077, .071, .071, .069, .078, .068, .071, .073, .067, .073, .074, .067),
  Item16 = c(.075, .067, .064, .061, .062, .054, .076, .068, .065, .077, .076, .065, .072, .074, .070, .077, .080, .069),
  Item17 = c(.076, .065, .062, .068, .061, .063, .072, .074, .071, .074, .071, .069, .079, .075, .078, .076, .077, .080),
  Item18 = c(.074, .067, .063, .068, .069, .054, .066, .064, .067, .068, .076, .064, .072, .072, .060, .085, .083, .071),
  Item19 = c(.068, .068, .065, .063, .066, .056, .068, .068, .067, .074, .076, .070, .071, .072, .069, .074, .082, .076),
  Item20 = c(.068, .067, .066, .070, .064, .059, .066, .067, .059, .071, .071, .066, .072, .075, .064, .074, .083, .074),
  stringsAsFactors = FALSE
)

# Add section indicators
acq_data$Section <- c(rep("Continuous", 9), rep("Ordinal", 9))

# Reorder columns to put Section first
acq_data <- acq_data[, c("Section", "Condition1", "Condition2", paste0("Item", 1:20))]

kable(acq_data[, -1], # Remove Section column for display
      caption = "Average biases of the ACQ factor loadings",
      col.names = c("", "", "1", "2", "3", "4", "5", "6", "7", "8", "9", "10",
                    "11", "12", "13", "14", "15", "16", "17", "18", "19", "20"),
      escape = FALSE) |>
  kable_styling(bootstrap_options = c("striped", "condensed"),
                full_width = TRUE,
                font_size = 10) |>
  add_header_above(c(" " = 2, "factor 1" = 10, "ACQ factor" = 10)) |>
  pack_rows("Continuous", 1, 9, bold = TRUE) |>
  pack_rows("Ordinal", 10, 18, bold = TRUE) |>
  footnote(general = "Bal = Balanced; LU = Low Unbalanced; HU = High Unbalanced; EQ = equal; LH = Low Heterogeneity; HH = High Heterogeneity",
           general_title = "Note: ") |>
  landscape()
```

No significant changes are observed when imbalance increases. However, a
slight increase in bias can be noticed in those conditions in which the
ACQ pattern is more heterogeneous. This increase, however, does not
substantially impact the average bias of the ACQ factor
[Table 5](#tab5){reference-type="ref" reference="tab5"}. The bias in
ACQ remains at around .06, with a maximum of .077.

With regards to the model-data fit ANOVA results, finally, no
significant effects were observed. However, when the ACQ patterns are
very heterogeneous, the fit of the model tends to worsen, which suggests
that **siren** is more sensitive to the pattern of ACQ loadings.

In closing, it should be noted that the simulated data are very
favorable: each content factor is strong and well-defined, without the
presence of correlated residuals or cross-loadings. In this framework,
**siren** barely suffers from a lack of specification or bias when
evaluated conditions are degraded. As ACQ loadings are not set to 1
(unlike the confirmatory method of Billiet and McClendon 2000), these
loadings are freely estimated, which is why the heterogeneity of the
acquiescence pattern does not affect the estimation results.

## Illustrative example usage

To illustrate how the SIREN program works, we have used an existing
dataset of 1309 participants (55.8% females) between 14 and 19 years old
(M = 16.4, S.D. = 1.1) from three previous studies (Morales-Vives and
Dueñas 2018; Morales-Vives, Camps, and Dueñas 2020; Morales-Vives et al.
in press). Therefore, further details about this data can be obtained
from the original studies. Those participants with missing data were not
included in the present illustrative analyses. All participants answered
the Psychological Maturity Assessment Scale questionnaire
(Morales-Vives, Camps, and Lorenzo-Seva 2013, PSYMAS), which assesses
the psychological maturity of adolescents, understood as the ability to
take responsibility for one's own obligations, taking into account one's
own characteristics and needs, without showing excessive dependence on
others. It consists of 27 items with a five-point response format (1 =
Completely disagree, 5 = Completely agree) and it assesses the following
factors: work orientation, self-reliance, and identity. The study
carried out by (Morales-Vives, Camps, and Lorenzo-Seva 2013), shows that
(a) the content factors are correlated, and (b) some of the items are
affected by the acquiescence response bias. This second feature is the
reason why we chose the data from this questionnaire as an illustration
of how **siren** works and how its outcomes are to be interpreted. In
the current analysis, we have only used ten items from two of the
subscales of this questionnaire (four items of self-reliance subscale
and six items of identity subscale) so that within each subscale half of
the items were in one direction (lack of maturity) and the other half in
the opposite direction (high maturity). Self-reliance refers to
willingness to take the initiative without allowing others to exercise
excessive control, and Identity refers to knowledge about one's
characteristics and needs. [Table 6](#tab7){reference-type="ref"
reference="tab7"} shows the contents of the used items. We would note
that the dataset is available in the **siren** package, so that the
interested reader can run the program and verify the results that are
presented below.

The code required for running this illustrative example is the
following:

``` r
  psymas_target=cbind(c(-9,-9,0,0,0,9,0,0,9,0),c(0,0,-9,9,-9,0,9,-9,0,9))

  acquihybrid(psymas, content_factors = 2, target = psymas_target,
    corr = “Polychoric”, raw_data = TRUE, method = “fixed”, display = TRUE)
```

Following the procedure explained above, the first step was to estimate
the ACQ factor from the fully balanced set of items, in this case
treating the variables as discrete (i.e. using the nonlinear model). As
can be seen in [Table 7](#tab8){reference-type="ref" reference="tab8"},
the ACQ loading estimates ranged between .001 and .566, and the items
with higher ACQ loadings were 3, 9 and 10. These results suggest that
several items are affected by ACQ, as was expected, and justify the need
to control for this response bias.

::: {#tab7}
  ---------------------------------------------------------------------
                                                                    ACQ
  -------------------------------------------------------------- ------
  Item 1. Consult the peer group before buying clothes             .001

  Item 2. Friends' opinions determine what is considered wrong     .001

  Item 6. Doesn't mind doing different things than friends         .231

  Item 9. Facing consequences of one's own mistakes                .206

  Item 3. Not showing the true self                                .380

  Item 4. Feeling accepted and valued                              .079

  Item 5. Feeling empty                                            .070

  Item 7. Good self-knowledge                                      .156

  Item 8. Others do not really know him/her                        .338

  Item 10. Feeling capable of doing many things well               .566


  ---------------------------------------------------------------------

  : (#tab:T4) Loading estimates in the Acquiescence factor obtained
  in the first step
:::

We next fitted a CFA solution consisting of two correlated content
factors with a full IC structure, in which each item only had a non-zero
loading on its own factor, and an additional ACQ factor in which the
corresponding loading was fixed at the estimate obtained in table A (in
the ordinal case there is no need to multiply this loading by the
standard deviation as this has a unit value). The final ULS estimates
for the full solution are in [Table 7](#tab8){reference-type="ref"
reference="tab8"}. As expected, the four items of self-reliance subscale
loaded in one factor, and the six items of the identity subscale loaded
in the other factor. Inspection of the signs of the loadings suggests
that the full condition of balance within each factor is achieved. As
for the strength of the content solution, items 2, 6, 9 had loadings of
.40 or higher (in absolute value) on the self-reliance factor, while
item 1 had the lowest loading, the same result obtained in the study
carried out by Morales-Vives, Camps, and Lorenzo-Seva (2013). All the
items of identity had loadings on this factor higher than .40, being
item 5 the item with the highest loading, which, again, agrees with the
results by Morales-Vives, Camps, and Lorenzo-Seva (2013). Overall, the
procedures included in the **siren** program provide the expected
results, which are congruent with those reported in the previous study,
even though the latter included a greater number of items than in the
present study. Furthermore, the correlation between the two factors is
.436, as was expected, because the study carried out by Morales-Vives,
Camps, and Lorenzo-Seva (2013) already showed that these factors are
positively correlated.

::: {#tab8}
  -------------------------------------------------------------------------------------------
                                                                   Factor 1   Factor 2    ACQ
  -------------------------------------------------------------- ---------- ---------- ------
  Item 1. Consult the peer group before buying clothes                  .29        .00   .001

  Item 2. Friends' opinions determine what is considered wrong          .53        .00   .001

  Item 6. Doesn't mind doing different things than friends             -.56        .00   .231

  Item 9. Facing consequences of one's own mistakes                    -.40        .00   .206

  Item 3. Not showing the true self                                     .00        .54   .380

  Item 4. Feeling accepted and valued                                   .00       -.57   .079

  Item 5. Feeling empty                                                 .00        .66   .070

  Item 7. Good self-knowledge                                           .00       -.53   .156

  Item 8. Others do not really know him/her                             .00        .43   .338

  Item 10. Feeling capable of doing many things well                    .00       -.47   .566


  -------------------------------------------------------------------------------------------

  : (#tab:T5) Estimated loadings in the CFA solution
:::

The fit of the solution on [Table 7](#tab8){reference-type="ref"
reference="tab8"} was quite acceptable: GFI=.99, RMSR=.04, RMSEA=0.04,
and CFI=0.96. This good fit suggests that, once ACQ is controlled, the
structure of the PSYMAS item pool assessed here is remarkably simple and
strong.

## Concluding remarks

There are at present two factor-analytic approaches for calibrating and
scoring typical-response measures after controlling for the biasing
effects of AR. One of them is fully confirmatory, and the complete
solution is identified by fixing all the ACQ loadings to the same value.
The other is unrestricted (i.e. exploratory or semi-confirmatory).
According to the literature, each of the two approaches has its pros and
cons (Savalei and Falk 2014; Fuente and Abad 2020)

In this article we have proposed a hybrid EFA-CFA procedure, called
SIREN, that tries to combine the best features of the two approaches
above. Thus, in SIREN, the ACQ factor can be identified in a first step
without the need to constrain all its loadings to have the same value.
Next, once the ACQ factor is identified, a fully confirmatory
(restricted) solution can be specified for the content factors at the
second step. Finally, for both types of factors (ACQ and content), our
proposed procedure allows factor score estimates for each individual to
be obtained at the third step. The flexibility of what we propose widens
the available options for assessing the structural properties of the
typical-response measure under scrutiny, and also, for obtaining
accurate score estimates for each individual. Regarding this last point,
we would note that most existing factor-analytical developments designed
for controlling ACQ tend to focus solely on the structural properties of
the instrument. However, accurate and "clean" individual score estimates
might be highly relevant in further validity studies or if clinical
decisions have to be taken on the basis of this instrument.

Apart from increased flexibility, the proposal has many features that
considerably increase its range of application. To start with, it allows
solutions to be fitted with the standard linear FA model or with the
non-linear graded-response model. Second, the solution can be fitted
using a "cleaned" residual covariance matrix (the standard approach to
this type of problems) or directly fitted to the raw data using the ACQ
loading estimates as fixed and known. This second option makes it
possible to use a wide range of estimation procedures and goodness of
fit measures for estimating and assessing model data fit.

The main theoretical and potential shortcoming of SIREN is the loss of
efficiency caused by the sequential limited-information procedure which
it uses. So far, the results of the simulation study suggest that this
loss has little impact in practice. However, more extensive simulation
is warranted. Although SIREN controls acquiescence bias, it is necessary
to assess to what extent null or close-to-zero biases are detected.

The R program that implements SIREN (and which has the same name) has
been designed to be as user-friendly as possible, and requires very few
specifications from the user: essentially, the FA model of choice
(linear or nonlinear) and a target matrix, which specifies the content
factor on which each item is expected to load together with the expected
sign of this loading. So, the program can be used by practitioners with
minimal proficiency in FA. Furthermore, **siren** is extremely
versatile, and provides a considerable amount of information in an
output that is simple and clear to interpret. Even so, we plan to extend
the calibration and the scoring choices of the program in future
developments.
::::::::::

:::::::::::::::::::::::::::::::::::: {#refs .references .csl-bib-body .hanging-indent entry-spacing="0"}
::: {#ref-Andrews:1984 .csl-entry}
Andrews, Frank M. 1984. "Construct Validity and Error Components of
Survey Measures: A Structural Modeling Approach." *Public Opinion
Quarterly* 2 (48): 409--42. <https://doi.org/10.1086/268840>.
:::

::: {#ref-Asparouhov:2023 .csl-entry}
Asparouhov, Tihomir, and Bengt Muthen. 1984. "Residual Structural
Equation Models." *Structural Equation Modeling* 1 (30): 1--31.
<https://doi.org/10.1080/10705511.2022.2074422>.
:::

::: {#ref-tenBerge:1999 .csl-entry}
Berge, Jos M. F. ten. 2020. "A Legitimate Case of Component Analysis of
Ipsative Measures, and Partialling the Mean as an Alternative to
Ipsatization." *Multivariate Behavioral Research* 4 (34): 89--102.
<https://doi.org/10.1207/s15327906mbr3401_4>.
:::

::: {#ref-tenBerge:1991 .csl-entry}
Berge, Jos M. F. ten, and Henk A. L. Kiers. 1991. "A Numerical Approach
to the Approximate and the Exact Minimum Rank of a Covariance Matrix."
*Psychometrika* 2 (56): 309--215. <https://doi.org/10.1007/BF02294464>.
:::

::: {#ref-Billiet:2000 .csl-entry}
Billiet, Jaak B., and McKee J. McClendon. 2000. "Modeling Acquiescence
in Measurement Models for Two Balanced Sets of Items." *Structural
Equation Modeling* 4 (7): 608--28.
<https://doi.org/10.1207/S15328007SEM0704_5>.
:::

::: {#ref-Bollen:1989 .csl-entry}
Bollen, Kenneth A. 1989. "A New Incremental Fit Index for General
Structural Equation Models." *Sociological Methods & Research* 3 (17):
303--16. <https://doi.org/10.1177/0049124189017003004>.
:::

::: {#ref-DeCastellarnau:2021 .csl-entry}
DeCastellarnau, Anna, and Willem E. Saris. 2021. "Correcting Correlation
and Covariance Matrices for Measurement Errors Before Further Analysis."
*Structural Equation Modeling: A Multidisciplinary Journal* 4 (28):
572--81. <https://doi.org/10.1080/10705511.2020.1870229>.
:::

::: {#ref-DeMars:2014 .csl-entry}
DeMars, Christine E. 2014. "An Illustration of the Effects of Ignoring a
Secondary Factor." *Applied Psychological Measurement* 38 (5): 406--9.
<https://doi.org/10.1177/0146621614529360>.
:::

::: {#ref-Eysenck:1950 .csl-entry}
Eysenck, H. J. 1950. "Criterion Analysis--an Application of the
Hypothetico-Deductive Method to Factor Analysis." *Psychological Review*
1 (57): 38--53. <https://doi.org/10.1037/h0057657>.
:::

::: {#ref-Fabrigar:1999 .csl-entry}
Fabrigar, Leandre R., Duane T. Wegener, Robert C. MacCallum, and Erin J.
Strahan. 1999. "Evaluating the Use of Exploratory Factor Analysis in
Psychological Research." *Psychological Methods* 3 (4): 272--99.
<https://doi.org/10.1037/1082-989X.4.3.272>.
:::

::: {#ref-Ferrando:2010 .csl-entry}
Ferrando, Pere J., and Urbano Lorenzo-Seva. 2010. "Acquiescence as a
Source of Bias and Model and Person Misfit: A Theoretical and Empirical
Analysis." *British Journal of Mathematical and Statistical Psychology*
2 (62): 427--48. <https://doi.org/10.1348/000711009X470740>.
:::

::: {#ref-Ferrando:2013 .csl-entry}
---------. 2013. "Unrestricted Item Factor Analysis and Some Relations
with Item Response Theory." Department of Psychology, Universitat Rovira
i Virgili, Tarragona. <http://psico.fcep.urv.es/utilitats/factor.>
:::

::: {#ref-Ferrando:2016 .csl-entry}
---------. 2016. "A Note on Improving EAP Trait Estimation in Oblique
Factor-Analytic and Item Response Theory Models." *Psicologica* 2 (37):
235--47. <https://www.redalyc.org/articulo.oa?id=16946248007>.
:::

::: {#ref-Ferrando:2003 .csl-entry}
Ferrando, Pere J., Urbano Lorenzo-Seva, and Eliseo Chico. 2003.
"Unrestricted Factor Analytic Procedures for Assessing Acquiescent
Responding in Balanced, Theoretically Unidimensional Personality
Scales." *Multivariate Behavioral Research* 3 (38): 353--74.
<https://doi.org/10.1207/S15327906MBR3803_04>.
:::

::: {#ref-delaFuente:2020 .csl-entry}
Fuente, Javier de la, and Francisco J. Abad. 2020. "Comparing Methods
for Modeling Acquiescence in Multidimensional Partially Balanced
Scales." *Psicothema* 4 (32): 590--97.
<http://10.7334/psicothema2020.96>.
:::

::: {#ref-Joreskog:1969 .csl-entry}
Joreskog, K. G. 1969. "A General Approach to Confirmatory Maximum
Likelihood Factor Analysis." *Psychometrika* 2 (34): 183--202.
<https://doi.org/10.1007/BF02289343>.
:::

::: {#ref-Lawley:1960 .csl-entry}
Lawley, DN. 1960. "Approximate Methods in Factor Analysis." *British
Journal of Statistical Psychology* 13 (1): 11--17.
:::

::: {#ref-Lorenzo-Seva:2009 .csl-entry}
Lorenzo-Seva, Urbano, and Pere J. Ferrando. 2009. "Acquiescent
Responding in Partially Balanced Multidimensional Scales." *British
Journal of Mathematical and Statistical Psychology* 2 (62): 319--26.
<https://doi.org/10.1348/000711007X265164>.
:::

::: {#ref-McDonald:1978 .csl-entry}
McDonald, Roderick P. 1978. "Some Checking Procedures for Extension
Analysis." *Multivariate Behavioral Research* 13 (3): 319--25.
<https://doi.org/10.1207/s15327906mbr1303_4>.
:::

::: {#ref-McDonald:2000 .csl-entry}
---------. 2000. "A Basis for Multidimensional Item Response Theory."
*Applied Psychological Measurement* 2 (24): 99--114.
<https://doi.org/10.1177/01466210022031552>.
:::

::: {#ref-McDonald:1995 .csl-entry}
McDonald, Roderick P., and Magdalena M. C. Mok. 1995. "Goodness of Fit
in Item Response Models." *Multivariate Behavioral Research* 1 (30):
23--40. <https://doi.org/10.1207/s15327906mbr3001_2>.
:::

::: {#ref-Messick:1966 .csl-entry}
Messick, Samuel. 1966. "The Psychology of Acquiescence: An
Interpretation of Research Evidence 1." *ETS Research Bulletin Series*
1966 (1): i--44.
:::

::: {#ref-Morales-Vives:2020 .csl-entry}
Morales-Vives, Fabia, Elisa Camps, and Jorge Manuel Dueñas. 2020.
"Predicting Academic Achievement in Adolescents: The Role of Maturity,
Intelligence and Personality." *Psicothema* 1 (31): 84--91.
<https://doi.org/10.1027/1015-5759/a000115>.
:::

::: {#ref-Morales-Vives:2013 .csl-entry}
Morales-Vives, Fabia, Elisa Camps, and Urbano Lorenzo-Seva. 2013.
"Development and Validation of the Psychological Maturity Assessment
Scale (PSYMAS)." *European Journal of Psychological Assessment* 1 (29):
12--18. <https://doi.org/10.1027/1015-5759/a000115>.
:::

::: {#ref-Morales-Vives:2018 .csl-entry}
Morales-Vives, Fabia, and Jorge Manuel Dueñas. 2018. "Predicting
Suicidal Ideation in Adolescent Boys and Girls: The Role of
Psychological Maturity, Personality Traits, Depression and Life
Satisfaction." *The Spanish Journal of Psychology* 21: E10.
<https://doi.org/10.1017/sjp.2018.12>.
:::

::: {#ref-Morales-Vives:inpress .csl-entry}
Morales-Vives, Fabia, P. J. Ferrando, Jorge Manuel Dueñas, S.
Martín-Arbós, and E Castarlenas. in press. "Are Older Teens More
Frustrated Than Younger Teens by the COVID-19 Restrictions? The Role of
Psychological Maturity, Personality Traits, Depression and Life
Satisfaction." *Current Psychology*, in press.
<https://doi.org/10.1007/s12144-023-04317-6>.
:::

::: {#ref-Muthen:1993 .csl-entry}
Muthen, Bengt. 1993. "Goodness of Fit with Categorical and Other
Non-Normal Variables." In *Testing Structural Equation Models*, edited
by K. A. Bollen and S. J. Long, 205--43. Sage Publications.
:::

::: {#ref-Nunnally:1978 .csl-entry}
Nunnally, Jum C. 1978. "An Overview of Psychological Measurement." In
*Clinical Diagnosis of Mental Disorders*, edited by B. B Wolman,
97--146. Springer.
:::

::: {#ref-Oberski:2013 .csl-entry}
Oberski, Daniel L., and Albert Satorra. 2013. "Measurement Error Models
with Uncertainty about the Error Variance." *Structural Equation
Modeling: A Multidisciplinary Journal* 3 (20): 409--28.
<https://doi.org/10.1080/10705511.2013.797820>.
:::

::: {#ref-RCore:2024 .csl-entry}
R Core Team. 2024. *R: A Language and Environment for Statistical
Computing*. Vienna, Austria: R Foundation for Statistical Computing.
<https://www.R-project.org/>.
:::

::: {#ref-Savalei:2014 .csl-entry}
Savalei, Victoria, and Carl F. Falk. 2014. "Recovering Substantive
Factor Loadings in the Presence of Acquiescence Bias: A Comparison of
Three Approaches." *Multivariate Behavioral Research* 5 (49): 407--24.
<https://doi.org/10.1037/1082-989X.4.3.272>.
:::

::: {#ref-Tanaka:1993 .csl-entry}
Tanaka, Jeffrey S. 1993. "An Overview of Psychological Measurement." In
*Testing Structural Equation Models*, edited by K. A. Bollen and S. J.
Long, 10--40. Sage Publications.
:::

::: {#ref-Vigil-Colet:2020 .csl-entry}
Vigil-Colet, Andreu, David Navarro-Gonzalez, and Fabia Morales-Vives.
2020. "To Reverse or to Not Reverse Likert-Type Items: That Is the
Question." *Psicothema* 1 (32): 108--14.
<https://doi.org/10.7334/psicothema2019.286>.
:::
::::::::::::::::::::::::::::::::::::
