---
title: "Benchmark blasso p>n lasso prior"
output: rmarkdown::html_vignette
params:
  dataset_name: NULL
  results_dir: "results"
vignette: >
  %\VignetteIndexEntry{Benchmark blasso p>n lasso prior}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 5)
```


```{r load-pkg, include=FALSE}
# pkg_root <- normalizePath(file.path(dirname(knitr::current_input()), ".."))
# devtools::load_all(pkg_root)
```


```{r load-helpers, include=FALSE}
helpers_dir <- file.path(dirname(knitr::current_input()), "helpers")

helper_files <- c(
  "benchmark_PC.R",
  "benchmark_blasso_bayesreg.R",
  "benchmark_blasso_hans.R",
  "effective_sample_size.R",
  "mcmc_stats.R",
  "normalize.R"
)

# ---- actually load them ----
helper_paths <- file.path(helpers_dir, helper_files)

missing <- helper_paths[!file.exists(helper_paths)]
if (length(missing) > 0) {
  stop("Missing helper file(s):\n", paste(missing, collapse = "\n"))
}

invisible(lapply(helper_paths, source, local = knitr::knit_global()))

```


# Load R scripts

Note that each of the benchamrking functions the inputs are

* vy
* mX
* nburn
* nsamples
* a, b, u, v - prior hyperparameters
* trials - the number of times each MCMC sampler is run
* beta_inds - an index set of beta coefficients for plotting - if NA will calculate these.

The outputs are:

* Median efficiencies for beta, sigma2, lambda2 over the number of trials
* beta_inds
* (x,y) grid for plotting the top 10 average betas
* (x,y) grid for plotting posterior for sigma2 and lambda2
* rhat values for sigma2 and lambda2
* sigma2 and lambda2 samples


# Select a dataset name

Note that for fitting the models to a specific data set uncomment the dataset_name for that data set and comment the rest of the data sets in the following chunk.


```{r load-package, include=FALSE}
# library(LassoHiDFastGibbs)

```




```{r dataset, include=FALSE}
dataset_name <- params$dataset_name
results_dir  <- params$results_dir
if (is.null(results_dir) || !nzchar(results_dir)) results_dir <- "results"
dir.create(results_dir, showWarnings = FALSE, recursive = TRUE)

if (is.null(dataset_name) || !nzchar(dataset_name)) {
  stop("params$dataset_name is missing/empty")
}

# Enforce that ONLY paper datasets are allowed
allowed_ngtp <- c("BostonHousing2", "diabetes2", "Hitters2", "Crime", "energy2", "Kakadu2")
allowed_pgtn <- c("eyedata", "cookie", "riboflavin", "covid", "qtl")

this_file <- basename(knitr::current_input())
if (grepl("ngtp", this_file, fixed = TRUE)) {
  if (!(dataset_name %in% allowed_ngtp)) stop("Dataset not in paper (ngtp): ", dataset_name)
} else if (grepl("pgtn", this_file, fixed = TRUE)) {
  if (!(dataset_name %in% allowed_pgtn)) stop("Dataset not in paper (pgtn): ", dataset_name)
}

message("Dataset: ", dataset_name)
message("Results dir: ", results_dir)


```



# Set options

```{r set-up}
set.seed(123)

# Set prior hyperparameter constants
a = 1.0E-2  # Prior shape for sigma2
b = 1.0E-2  # Prior scale for sigma2
u = 1.0E-2  # Prior shape for lambda2
v = 1.0E-2  # Prior scale for lambda2

# Initial values for lambda2 and sigma2
lambda2_init  = 10
lambda_init = sqrt(lambda2_init)
sigma2_init = 1

# Number of samples to run the MCMC
nburn = 1000
nsamples = 11000
inds_use = (nburn + 1):nsamples

# How many times to run each sampler
trials = 5

```

# Generate and normalize dataset

```{r}
# res = generate_data(dataset_name)
# 
# mX = res$mX
# vy = res$vy
# p = res$p
# n = res$n
# 
# rm(res)
```


 

# The modified Park and Casella Gibbs sampler



```{r PC, eval=TRUE}
if(dataset_name != "riboflavin"){
  res_PC = benchmark_PC(vy, mX, lambda_init, sigma2_init, a, b, u, v, nburn, nsamples, trials, beta_inds=NA)
}

```


# Modified Hans sampler


```{r Hans}
res_hans = benchmark_blasso_hans(vy, mX,lambda_init, sigma2_init, a, b, u, v, nburn, nsamples, trials, beta_inds=NA)

```



```{r Bayesreg}
trials <- 5

set.seed(0)



res_bayesreg <- NULL

if (requireNamespace("bayesreg", quietly = TRUE)) {

  res_bayesreg <- benchmark_blasso_bayesreg(
    vy, mX,
    lambda_init, sigma2_init, a, b, u, v,
    nburn, nsamples, trials,
    beta_inds = NA #res_PC$beta_inds
  )



} else {
  stop("Skipping bayesreg benchmark: package 'bayesreg' is not installed.")
}

```


```{r}
if(dataset_name == "riboflavin"){
  save(res_hans,
       res_bayesreg,
       file = file.path(results_dir, paste0(dataset_name,"_results_pgtn.Rdata")))
}else{
  save(res_PC,
       res_hans,
       res_bayesreg,
       file = file.path(results_dir, paste0(dataset_name,"_results_pgtn.Rdata")))
}
```


