% !TeX root = RJwrapper.tex
\title{Editorial}
\author{by Catherine Hurley}

\maketitle


On behalf of the editorial board, I am pleased to present Volume 14
Issue 2 of the R Journal.

First, some news about the journal board. Mark van der Loo has very
kindly agreed to move from Associate Editor to Executive Editor to fill
a temporary gap. One new Associate Editor, Kevin Burke, has recently
joined the team.

Behind the scenes, several people are assisting with the journal
operations and the new developments. Mitchell O'Hara-Wild continues to
work on infrastructure, and H. Sherry Zhang continues to develop the
\pkg{rjtools} package. In addition, articles in this issue have been
carefully copy edited by Hannah Comiskey.

\hypertarget{in-this-issue}{%
\section{In this issue}\label{in-this-issue}}

News from the CRAN, the R Foundation and the Forwards Taskforce are
included in this issue. We also have a report from the Why R? Turkey
2022 conference.

This issue features 18 contributed research articles the majority of
which relate to R packages for modelling tasks. All packages are
available on CRAN. Topics covered are:

\begin{itemize}
\tightlist
\item
  \textbf{Statistical modelling and inference}

  \begin{itemize}
  \tightlist
  \item
    \pkg{APCI}: An R and Stata Package for Age-Period-Cohort Analysis
  \item
    \pkg{refreg}: an R package for estimating conditional reference
    regions
  \item
    From the multivariate Faà di Bruno's formula to unbiased estimates
    of joint cumulant products: the \pkg{kStatistics} package in R
  \item
    The Concordance Test, an Alternative to Kruskal-Wallis Based on the
    Kendall tau Distance: An R Package
  \item
    \pkg{shinybrms}: Fitting Bayesian Regression Models Using a
    Graphical User Interface for the R Package brms
  \item
    \pkg{PDFEstimator}: An R Package for Density Estimation and Analysis
  \item
    \pkg{htestClust}: Hypothesis Tests for Clustered Data under
    Informative Cluster Size in R
  \item
    \pkg{ClusTorus}: An R Package for Prediction and Clustering on the
    Torus by Conformal Prediction
  \item
    \pkg{TensorTest2D}: Fitting Generalized Linear Models with Matrix
    Covariates
  \end{itemize}
\item
  \textbf{Ecological and Environmental analysis}

  \begin{itemize}
  \tightlist
  \item
    Quantifying Population Movement Using a Novel Implementation of
    Digital Image Correlation in the \pkg{ICvectorfields} package
  \item
    \pkg{iccCounts}: an R Package to Estimate the Intraclass Correlation
    Coefficient for Assessing Agreement with Count Data
  \item
    \pkg{rassta}: Raster-based Spatial Stratification Algorithms
  \end{itemize}
\item
  \textbf{Missing data}

  \begin{itemize}
  \tightlist
  \item
    R-miss-tastic: a unified platform for missing values methods and
    workflows
  \item
    \pkg{reclin2}: a Toolkit for Record Linkage and Deduplication
  \end{itemize}
\item
  \textbf{Time Series Analysis}

  \begin{itemize}
  \tightlist
  \item
    \pkg{wavScalogram}: an R package with wavelet scalogram tools for
    time series analysis
  \item
    \pkg{brolgar}: An R package to BRowse Over Longitudinal Data
    Graphically and Analytically in R''
  \end{itemize}
\item
  \textbf{Other}

  \begin{itemize}
  \tightlist
  \item
    \pkg{akc}: A tidy framework for automatic knowledge classification
    in R
  \item
    An Open-Source Implementation of the CMPS Algorithm for Assessing
    Similarity of Bullets
  \end{itemize}
\end{itemize}


\address{%
Catherine Hurley\\
Maynooth University\\%
\\
%
\url{https://journal.r-project.org}\\%
%
\href{mailto:r-journal@r-project.org}{\nolinkurl{r-journal@r-project.org}}%
}
