% !TeX root = RJwrapper.tex
\title{Identifying Counterfactual Queries with the R Package cfid}


\author{by Santtu Tikka}

\maketitle

\abstract{%
In the framework of structural causal models, counterfactual queries describe events that concern multiple alternative states of the system under study. Counterfactual queries often take the form of ``what if'' type questions such as ``would an applicant have been hired if they had over 10 years of experience, when in reality they only had 5 years of experience?'' Such questions and counterfactual inference in general are crucial, for example when addressing the problem of fairness in decision-making. Because counterfactual events contain contradictory states of the world, it is impossible to conduct a randomized experiment to address them without making several restrictive assumptions. However, it is sometimes possible to identify such queries from observational and experimental data by representing the system under study as a causal model, and the available data as symbolic probability distributions. Shpitser and Pearl (2007) constructed two algorithms, called ID* and IDC*, for identifying counterfactual queries and conditional counterfactual queries, respectively. These two algorithms are analogous to the ID and IDC algorithms by Shpitser and Pearl (2006b, 2006a) for identification of interventional distributions, which were implemented in R by Tikka and Karvanen (2017) in the causaleffect package. We present the R package \href{https://CRAN.R-project.org/package=cfid}{cfid} that implements the ID* and IDC* algorithms. Identification of counterfactual queries and the features of cfid are demonstrated via examples.
}

\input{RJ-2023-053-src.tex}

\hypertarget{refs}{}

\bibliography{tikka.bib}

\address{%
Santtu Tikka\\
Department of Mathematics and Statistics, University of Jyväskylä\\%
P.O. Box 35, FI-40014, Finland\\
%
%
%
\href{mailto:santtu.tikka@jyu.fi}{\nolinkurl{santtu.tikka@jyu.fi}}%
}
