---
title: "RGAN Replication"
output: pdf_document
date: '2023-03-03'
editor_options: 
  chunk_output_type: console
---

```{r setup}
install.packages(c("RGAN", "torchvision"))
library(RGAN)
library(here)
```

# Replication of the Example in Section 4.1

The first example with very simple tabular data should serve as an easy entry point
to working with the RGAN package.

```{r Setting up the data}

data <- RGAN::sample_toydata()

#' # Build new transformer to standardize data
transformer <- RGAN::data_transformer$new()

#' # Fit transformer to data
transformer$fit(data)

#' # Transform data and store as new object
transformed_data <-  transformer$transform(data)
```

```{r Figure 1}
# Plot original data and transformed data (Figure 1)

par(mfrow = c(1, 2))
dimensions <- c(1, 2)

plot(
  data[, dimensions],
  bty = "n",
  col = viridis::viridis(2, alpha = 0.7)[1],
  #xlim = c(-50, 50),
  pch = 19,
  xlab = ifelse(
    !is.null(colnames(data)),
    colnames(data)[dimensions[1]],
    paste0("Var ", dimensions[1])
  ),
  ylab = ifelse(
    !is.null(colnames(data)),
    colnames(data)[dimensions[2]],
    paste0("Var ", dimensions[2])
  ),
  main = "(A)",
  las = 1
)

plot(
  transformed_data[, dimensions],
  bty = "n",
  col = viridis::viridis(2, alpha = 0.7)[1],
  #xlim = c(-50, 50),
  pch = 19,
  xlab = ifelse(
    !is.null(colnames(data)),
    colnames(data)[dimensions[1]],
    paste0("Var ", dimensions[1])
  ),
  ylab = ifelse(
    !is.null(colnames(data)),
    colnames(data)[dimensions[2]],
    paste0("Var ", dimensions[2])
  ),
  main = "(B)",
  las = 1
)
```

```{r Training the GAN}
#' # Train the default GAN

torch::torch_manual_seed(20220629)
trained_gan_4_1 <- RGAN::gan_trainer(transformed_data)

# Sample synthetic data from the trained GAN

trained_gan_4_1$settings$eval_dropout <- FALSE

synthetic_data_no_dropout <- sample_synthetic_data(trained_gan_4_1, transformer)

trained_gan_4_1$settings$eval_dropout <- TRUE

synthetic_data_dropout <- sample_synthetic_data(trained_gan_4_1, transformer)
```

```{r Figure 2}
par(mfrow = c(1, 2))
# Plot the results
GAN_update_plot(
  data = data,
  synth_data = synthetic_data_no_dropout,
  main = "(A)"
)

GAN_update_plot(
  data = data,
  synth_data = synthetic_data_dropout,
  main = "(B)"
)
```

# Replication of the Example in Section 4.2

Note that this example is a bit more resource intensive.
With a GPU (NVIDIA GeForce RTX 2070 SUPER) training the following GAN for one epoch takes about 20 minutes. 
Training on CPU takes considerably longer, on my machine (Apple Macbook Air M1, running R and torch through Rosetta) about 6 hours.

```{r Getting the celeba images}
if(!dir.exists(here::here("celeba/img_align_celeba"))){
# Create celeba directory in working directory
dir.create(here::here("celeba"))

# Download the celeba images
invisible(system(
  paste0(
    "wget --load-cookies /tmp/cookies.txt \"https://docs.google.com/uc?export=download&confirm=$(wget --quiet --save-cookies /tmp/cookies.txt --keep-session-cookies --no-check-certificate \'https://docs.google.com/uc?export=download&id=0B7EVK8r0v71pZjFTYXZWM3FlRnM\' -O- | sed -rn 's/.*confirm=([0-9A-Za-z_]+).*/\\1\\n/p')&id=0B7EVK8r0v71pZjFTYXZWM3FlRnM\" -O",
    here::here("celeba/img_align_celeba.zip"),
    " && rm -rf /tmp/cookies.txt"
  ),
  intern = TRUE
))

# Unzip the folder
invisible(system(paste0(
  "unzip -o ",
  here::here("celeba/img_align_celeba.zip"),
  " -d ",
  here::here("celeba")
), intern = TRUE))
}
```

```{r Preparing the data and GAN}

torch::torch_manual_seed(20220629)
dataset <- torchvision::image_folder_dataset(root = here::here("celeba"),
                                             transform = function(x) {
                                               x = torchvision::transform_to_tensor(x)
                                               x = torchvision::transform_resize(x, size = c(64, 64))
                                               x = torchvision::transform_center_crop(x, c(64, 64))
                                               x = torchvision::transform_normalize(x, c(0.5, 0.5, 0.5), 
                                                                                    c(0.5, 0.5, 0.5))
                                               return(x)
                                             })

device <- torch::torch_device(ifelse(torch::cuda_is_available(), "cuda", "cpu"))

g_net <- RGAN::DCGAN_Generator(dropout_rate = 0, noise_dim = 100)$to(device = device)
d_net <- RGAN::DCGAN_Discriminator(dropout_rate = 0, sigmoid = F)$to(device = device)

g_optim <- torch::optim_adam(g_net$parameters, lr = 0.0002, betas = c(0.5, 0.999))
d_optim <- torch::optim_adam(d_net$parameters, lr = 0.0002, betas = c(0.5, 0.999))

noise_dim <- c(100, 1, 1)
fixed_z <-
  torch::torch_randn(c(16, noise_dim))$to(device = device) 
```

```{r Training the GAN}
trained_gan_4_2 <- RGAN::gan_trainer(
  data = dataset,
  noise_dim = noise_dim,
  noise_distribution = "normal",
  data_type = "image",
  value_function = "wasserstein",
  generator = g_net,
  generator_optimizer = g_optim,
  discriminator = d_net,
  discriminator_optimizer = d_optim,
  plot_progress = FALSE,
  plot_interval = 10,
  batch_size = 128,
  synthetic_examples = 16,
  device = device,
  eval_dropout = FALSE,
  epochs = 1
)
```

Now we sample fake images after one epoch with `expert_sample_synthetic_data`; the function allows us to pass a fixed noise sample so that one can observe the evolution of the same images through training.

```{r Figure 3}
GAN_update_plot_image(synth_data = expert_sample_synthetic_data(g_net, fixed_z, device,eval_dropout = FALSE))

```
