---
title: "GB rainfall chemistry"
output: 
  learnr::tutorial:
    progressive: false
    allow_skip: true
    code_folding: hide

runtime: shiny_prerendered
description: >
 Testing this setup.
# https://stackoverflow.com/questions/29830505/error-the-application-took-too-long-to-respond
# > load_all("/data/rticles/rticles/")

---
## Welcome

This is a demo on how we may publish a notebook from DataLabs, with data, packages, and workflow pre-loaded in a coding envrionment, accompanied by rich narratives. This notebook is a companion to the following R journal publication, where we provides details of our approach:

[![RJournal paper](https://img.shields.io/badge/DOI%20(R%20Journal)-https%3A%2F%2F10.32614%2FRJ--2021--097-blue)](https://doi.org/10.32614/RJ-2021-097)

The source code (.Rmd) of this notebook is deposited in the NERC Environmental Information Data Centre (EIDC). Please cite it when you are referring to the app:

[![DOI](https://img.shields.io/badge/EIDC_DOI-10.5285/df57b002--2a42--4a7d--854f--870dd867618c-blue.svg?style=style=flat&labelColor=whitesmoke&logo=data%3Aimage%2Fpng%3Bbase64%2CiVBORw0KGgoAAAANSUhEUgAAABAAAAAQCAYAAAAf8%2F9hAAAAAXNSR0IArs4c6QAAAvVJREFUOE9lk0tonFUYhp%2Fv%2FOefzOWfq0RpEW0XLlx00Z3d1m7ciFe8ZaFErGajtdQIhRZUEKNQtVSigQpiQbDpwoIFpQiCYmwtIiiULqap0lCSSeb2z%2Fy3c47M1MagZ%2FPxLd7nvB%2B8r%2FCf98yHP1X6af8po9XjNqf39Ix168Pkh7%2FCaLHT6n7B8anuVolsXY58dmG6E2dz3YzGwAoxglFCOe%2BxrezzZzhofXX160PR0Tc%2BvaXbBLz5%2BcX3tNYHh0YROw%2FjFFYJ2tfUSj6TFZ8gD2d%2Bv8yXy2%2B9y9Hzr40gY8D0B9893wiKC0GhBKKJLSQOrAhKK8p5n8kgR6Pocb3d5%2FWLh0lcc5ojSydl3%2By31XLdNSfL1XqlWMSJInWQAQbGeyHnUc1rGnmPy%2B0VPr42i02jdTJ2yr2HTr80WSp9dEe5TmEihygh%2BwcwAqGEvFYUPI%2BuXeXc4Dh9ex1iB6l9UWozpxYbQeWRbUGVgq9RnsIgJM6Np68ETzlWcj9yRZ8lcxFYIBrfeVrUcycveaXq7nKxQkl7aE%2FhRBARtBKkeIVe9SxDvUJqIXUyni5yMHS%2FCFMLl5gIdpMvI9obO%2FA9j6DYor79HDr4Azdy6yCxQmSF0AiDoYNoBHhy%2Fgy50sMjACOAb7h7x%2FfcuX0JrbKbYhyxdZviXia0%2BmMXi8IT8zMofYKJChOVkPvv%2B4bbghbWZVgcqbPE1hDajL4xjMRrsbDaAzI7Izx0rEau0CTn15599DfuqQ8J0w6JiUhsysBmhCalkyV0TMpG6lgOFeGADaJs580kPnZiP%2FXV%2BbkpSz1viLMNekmXbhrSTYe0s4SNLKaVpSyHwo0QXGJf4OCFhX%2B7MDX3%2FoMPNF%2Fee9ftKOnRS9dYjzu00gFracSNOOHaAFaHYDN3jFd%2BfnUzyreK4c9O79%2B7y7xTy5tqK1unY0K6o99TQyeBOHFtnMxyYOmT%2F5Vps5Vv76pP5IKnPaX3RZgd1o0CLU2cO4%2BKT3Hg1%2FbWBv8NFZRz6L1fWHsAAAAASUVORK5CYII%3D)](https://catalogue.ceh.ac.uk/documents/df57b002-2a42-4a7d-854f-870dd867618c)

This example is derived from an academic paper I have submitted that is currently under review. More information will follow.

This work is supported by:

- NERC Grant NE/T006102/1, Methodologically Enhanced Virtual Labs for Early Warning of Significant or Catastrophic Change in Ecosystems: Changepoints for a Changing Planet, funded under the Constructing a Digital Environment Strategic Priority Fund. 
- Additional support is provided by the UK Status, Change and Projections of the Environment (UK-SCAPE) programme started in 2018 and is funded by the Natural Environment Research Council (NERC) as National Capability (award number NE/R016429/1). 
- The initial development work of DataLabs was supported by a NERC Capital bid as part of the Environmental Data Services (EDS).


The solution proposed is to redner the Rmarkdown notebook as an R Shiny site. Specifically, it is to render the R markdown document as a Shiny prerendered document with the [learnr](https://rstudio.github.io/learnr/exercises.html#Exercise_Setup) package. This package was designed for interactive online tutorials, with 'sandboxes' (i.e. executable code boxes) to allow users to type in their code, which they can run it and check answers. We can pre-filled the 'sandboxes' with parts of the published workflow and their outputs (and linked different parts if one depends on another). And with the 'start over' button, the pre-filled published workflow will re-appear.

Here are some of the advantages:

- No change to the normal workflow to work with R markdown documents. Only need to supply a few extra code chunk options. Minimal extra training for scientist already 
- This makes sure the published version of the code is not altered, but users can tweek it in a contained way (changes does not affect beyond each sandbox during each session).
- This is a more contained approach than the Jupyter/BinderHub/Colab approach, where users can change the entire notebook . Our approach makes sure the narrative and any bits of codes we don't want the user to change (e.g. the `setup` chunk) cannot be changed. It gives more structure and the user can reset a 'sandbox' rather than the entire notebook if they want to see what's done in the published version.
- This is less involved than writing a full-on Shiny app. No need to worry much about UI appearance and reactivity.
- DataLabs already has the capability to publish Shiny sites. This approach requires only minor re-configuration on DataLabs.
- Currently, DataLabs publish Shiny sites by making a copy from the directory of the Shiny app to a separate container. So all the required packages (right versions), data, and scripts will be most up to date. 

### Example application

The goal of our approach helps users to reproduce *elements* of a research workflow. As a demonstrator, this app allows users to reproduce elements (e.g. figures) of our paper 'The evolving role of weather types on rainfall chemistry under large reductions in pollutant emissions', which is currently under review.


### Setup

I have preloaded the following datasets, packages, and scripts. Note that only the `setup` chunk in the Rmd document can be called by other code chunks by default. Future releases of [learnr](https://rstudio.github.io/learnr/exercises.html#Exercise_Setup) should allow better linking of outputs between chunks.

This code chunk reads all the required datasets and define functions to be used, which includes:

- Reading [ECN](https://ecn.ac.uk) rainfall chemistry data
- Reading [ECN](https://ecn.ac.uk) weather data
- Reading [ERA5 weather reanalysis](https://doi.org/10.24381/cds.adbb2d47) data
- Reading daily [Lamb weather type](https://crudata.uea.ac.uk/cru/data/lwt/) data
- Reading daily [Atlantic multidecadal oscillation (AMO)](https://climatedataguide.ucar.edu/climate-data/atlantic-multi-decadal-oscillation-amo) data
- Various functions to read, aggregate, and transform data in a usable form

```{r setup, warning = FALSE, message = FALSE, error=FALSE}

# https://beta.rstudioconnect.com/content/2671/Combining-Shiny-R-Markdown.html
library(learnr)
tutorial_options(exercise.timelimit = 60, 
                 exercise.completion=TRUE,
                 exercise.eval=TRUE)
knitr::opts_chunk$set(
  tidy = FALSE
)

library(ncdf4)
library(lubridate)
library(plotly)
library(ggplot2)
library(lubridate)
library(stringr)
library(cowplot)
library(dplyr)
library(changepoint)
library(feather)
library(stringi)
library(stringr)
library(readr)
library(tidyr)
library(maps)
library(here)
library(ggrepel)

# color blind friendly palette (with grey):
cbf_1 <- c("#999999", "#E69F00", "#56B4E9", "#009E73", 
           "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

# color_blind_friendly palette (with black): 
cbf_2 <- c("#000000", "#E69F00", "#56B4E9", "#009E73", 
           "#F0E442", "#0072B2", "#D55E00", "#CC79A7")

era5 = function(ncfname){
  era5_single = function(ncfname){
    ncin <- nc_open(ncfname)
    #print(ncin)
    out = tibble(t  = make_datetime(1900,1,1,ncvar_get(ncin,"time"),0,0), 
                 cp = ncvar_get(ncin,"cp"),
                 lsp = ncvar_get(ncin,"lsp"),
                 lspf = ncvar_get(ncin,"lspf"),
                 ERA5_rain = ncvar_get(ncin,"tp"),)
    nc_close(ncin)
    return(out)
  }
  out = tibble()
  for (i in 1:length(ncfname)){
    out = rbind(out,era5_single(ncfname[i]))
  }
  
  return(out)
}

get_datechem <- function(site){
  lamb_type_lookup <-  read_csv(here('data',"lamb-type-lookup.csv"))  
  
  lamb_data = read_csv(here('data','20CR_1871-1947_ncep_1948-2020_12hrs_UK.csv')) %>% 
    dplyr::mutate(DATE = make_date(year,month,day)) %>% 
    dplyr::select(DATE,LWT) %>% 
    dplyr::mutate(LWT = as.character(LWT)) %>% 
    dplyr::mutate(LWT = setNames(lamb_type_lookup$Type,lamb_type_lookup$Num)[LWT])
  
  
  ### get ERA5
  
  # files = list.files(pattern = paste0('download_',site,"_"),path = '/data/rain_pattern/ERA5',full.names = T)
  # ERA5 = era5(files) %>% 
  #   mutate(cp = 1000*cp, lsp = 1000*lsp,ERA5_rain =cp+lsp, frac = cp/ERA5_rain) %>%  # m to mm
  #   group_by(date(t)) %>% 
  #   mutate(frac_daily = mean(frac)) %>% 
  #   ungroup() %>% 
  #   mutate(frac = ifelse(!between(frac,0,1),NA,frac) ) %>% 
  #   mutate(frac = ifelse(frac <1e-6,0,frac) ) %>% 
  #   tidyr::fill(frac,.direction="down")
  
  ### get rain flux
  rain_daily = read_feather(here('data',"rain_hourly.feather")) %>% 
    dplyr::filter(SITECODE == site) %>% 
    dplyr::mutate(SDATE = as.Date(SDATE,format="%d-%b-%y")) %>% 
    dplyr::rename(DATE = SDATE) %>%
    group_by(DATE) %>% 
    summarise(VALUE = sum(VALUE))
  
  ### end get rain flux
  
  # list of PC dates that passes QC
  # QC_date = readr::read_csv('/data/rain_pattern/ECN_PC_chemistry-QC.csv') %>% 
  #   dplyr::filter(OVERALL_CHECK_OK == 1, SITECODE== site) %>% 
  #   dplyr::rename(DATE = SDATE) %>% 
  #   dplyr::filter(SITECODE == site) %>% 
  #   dplyr::mutate(DATE = as.Date(DATE,format="%d-%b-%y")) %>% 
  #   dplyr::select(DATE) %>% pull
  
  
  rain_chem = read_csv(here('data',"ECN_PC1.csv"))   %>% 
    dplyr::rename(DATE = SDATE) %>% 
    dplyr::filter(SITECODE == site) %>% 
    dplyr::filter(!FIELDNAME %in% c("VACUUM","PHAQCS","PHAQCU")) %>%  ## remove variables
    dplyr::filter(!str_detect(FIELDNAME, "Q"))    %>%
    #dplyr::mutate(FIELDNAME = paste(FIELDNAME,"(Rain)")) %>% 
    dplyr::mutate(DATE = as.Date(DATE,format="%d-%b-%y"))
  
  # outlier removal, note 1 mg/L to 1ppm
  # http://inside.mines.edu/~epoeter/_GW/resultsNOV03.pdf
  rain_chem = rain_chem %>% 
    tidyr::spread(FIELDNAME,VALUE) %>% 
    #dplyr::filter(!PO4P > 1.0) %>% 
    #dplyr::filter(between( (SODIUM/23)/(CHLORIDE/35.5), 0.8,1.2)) %>% # seasalt check
    dplyr::mutate(`non-marine sulphate` = (SO4S*3*20.83-28.21*0.104*CHLORIDE)/1000 ) %>% # in meq/L 
    dplyr::mutate(#ALKY = ALKY/61, # as for Hco3, everything to meq/L
      H = 10^(6-PH)/1000, # ph to meq/L
      ALUMINIUM=ALUMINIUM/26.98*3,
      CALCIUM=CALCIUM/40.08*2,
      CHLORIDE=CHLORIDE/35.45,
      IRON=IRON/55.845*2,
      MAGNESIUM=MAGNESIUM/24.3*2,
      POTASSIUM=POTASSIUM/39.1,
      SODIUM=SODIUM/22.99,
      NH4N=NH4N/14.01,
      NO3N=NO3N/14.01,
      PO4P=PO4P/15,
      SO4S=SO4S*3/96*2,
      EC=(H*349.8+SO4S*79.8+NO3N*71.4+NH4N*73.5+SODIUM*50.1+
            MAGNESIUM*53.05+CALCIUM*59.5+CHLORIDE*76.4+POTASSIUM*73.5), # tehorectical EC based on diffusion coefficient (Nernst-Einstein Equation),https://www.aqion.de/site/77
      EC_ratio = CONDY/EC,
      EC_balance = (CONDY-EC)/(CONDY+EC)
    ) %>%  # convert all ions to meq/L
    dplyr::mutate(sum_anion = CHLORIDE+NO3N+PO4P+SO4S, 
                  sum_cation= NH4N+CALCIUM+MAGNESIUM+POTASSIUM+SODIUM,
                  charge_balance = (sum_cation-sum_anion)/(sum_cation+sum_anion)) %>% 
    #dplyr::filter(!abs(charge_balance) > 0.05) %>% 
    #dplyr::filter(!abs(EC_balance) > 0.2) %>%  # measured TDS/ theorectical TDS should be between 1.0 and 1.2 
    tidyr::gather(FIELDNAME,VALUE,-SITECODE,-DATE) %>% 
    dplyr::filter(DATE %in% QC_date)
  
 
  # convert all ions to meq/L
  
  date_chem = distinct(rain_chem,DATE) %>% 
    dplyr::arrange(DATE) %>% 
    dplyr::mutate(SDATE = as.Date(DATE,format="%d-%b-%y")) %>% 
    dplyr::mutate(DIFF = as.numeric(difftime(DATE,lag(DATE,1)))) %>% 
    dplyr::mutate(STARTDATE = lag(DATE,1)+1, ENDDATE = DATE) %>% 
    dplyr::filter(!is.na(DIFF)) %>%
    dplyr::filter(between(DIFF,6,8)) #%>% 
  #dplyr::mutate(lamb_str2 = NA,mean_rain_per_day = NA) # maximum len of continuous rain > 0.1
  
  
  get_lamb_str2 <- function(lamb_data,STARTDATE,ENDDATE){
    lamb_data %>% 
      filter(between(DATE,STARTDATE,ENDDATE)) %>% 
      select(LWT) %>% 
      mutate(LWT = case_when(LWT == "C" | LWT == "W" |
                               LWT == "SW" |LWT == "CW" |
                               LWT == "CSW"  ~ "1" , TRUE ~ "0")) %>% #event generating types
      pull() %>% paste(collapse = '')
  }
  get_lamb_strW <- function(lamb_data,STARTDATE,ENDDATE){
    lamb_data %>% 
      filter(between(DATE,STARTDATE,ENDDATE)) %>% 
      select(LWT) %>% 
      mutate(LWT = case_when(LWT == "W" |LWT == "SW" |LWT == "CW" |LWT == "CSW"  ~ "1" ,
                             TRUE ~ "0")) %>% #event generating types
      pull() %>% paste(collapse = '')
  } 
  get_lamb_strC <- function(lamb_data,STARTDATE,ENDDATE){
    lamb_data %>% 
      filter(between(DATE,STARTDATE,ENDDATE)) %>% 
      select(LWT) %>% 
      mutate(LWT = case_when(LWT=="C"|LWT=="CSW" |LWT=="CW"|LWT=="CNW"|LWT=="CN"|
                               LWT=="CNE"|LWT=="CE"|LWT=="CSE"|LWT=="CS"  ~ "1" ,
                             TRUE ~ "0")) %>% #event generating types
      pull() %>% paste(collapse = '')
  }
  
  date_chem = date_chem %>% 
    rowwise() %>%
    mutate(lamb_str2= get_lamb_str2(get("lamb_data"),STARTDATE,ENDDATE),
           lamb_strW= get_lamb_strW(get("lamb_data"),STARTDATE,ENDDATE),
           lamb_strC= get_lamb_strC(get("lamb_data"),STARTDATE,ENDDATE),
           mean_rain_per_day = filter(rain_daily,between(DATE,STARTDATE,ENDDATE)) %>% 
             select(VALUE) %>% pull() %>% mean(na.rm=T),
           #cp_week=filter(ERA5,between(`date(t)`,STARTDATE,ENDDATE)) %>% select(cp) %>% sum(),
           #era5_week = filter(ERA5,between(`date(t)`,STARTDATE,ENDDATE)) %>% select(ERA5_rain) %>% sum() 
    ) %>% 
    ungroup()
  
  date_chem = date_chem %>% mutate(SITECODE=site) %>% select(-SDATE,-STARTDATE,-ENDDATE) %>% 
    left_join(rain_chem %>% select(-SITECODE) %>% tidyr::spread(FIELDNAME,VALUE), by="DATE") %>%
    gather(FIELDNAME,VALUE,-DATE,-DIFF,-lamb_str2,-lamb_strW,-lamb_strC,-SITECODE,-mean_rain_per_day,
           #-cp_week,-era5_week,
           -VOLUME)
  
  print(paste0('Complete ',site,'!'))
  
  
  return(date_chem)  
}


data0 = rbind(get_datechem('T02'),get_datechem('T04'),get_datechem('T06'),get_datechem('T11'))


date_chem = data0 %>%
   tidyr::spread(FIELDNAME,VALUE)  %>% 
  mutate(SITECODE = case_when(
    SITECODE == 'T02' ~ 'Glensaugh',
    SITECODE == 'T04' ~ 'Moor_House',
    SITECODE == 'T06' ~ 'Rothamsted',
    SITECODE == 'T01' ~ 'Drayton',
    SITECODE == 'T07' ~ 'Sourthrope',
    SITECODE == 'T11' ~ 'Snowdon',
    TRUE ~ SITECODE
  )) %>% 
mutate(lamb2count=sapply(strsplit(lamb_str2, ''), function(x) sum(as.numeric(x))) ,
       lambCcount=sapply(strsplit(lamb_strC, ''), function(x) sum(as.numeric(x))) ,
       lambWcount=sapply(strsplit(lamb_strW, ''), function(x) sum(as.numeric(x))) ,
       yr_grp = case_when(
         between(year(DATE),1992,1995) ~ '1992-1995',
         between(year(DATE),1996,1999) ~ '1996-1999',
         between(year(DATE),2000,2003) ~ '2000-2003',
         between(year(DATE),2004,2007) ~ '2004-2007',
         between(year(DATE),2008,2011) ~ '2008-2011',
         between(year(DATE),2012,2015) ~ '2012-2015',
         TRUE                      ~ NA_character_),
       lamb_grp2 = case_when(
         between(lamb2count,0,2) ~ '0-2',
         #between(lamb2count,2,3) ~ '2-3',
         #         between(lamb2count,3,3) ~ '3',
         #         between(lamb2count,4,4) ~ '4',
         between(lamb2count,3,5) ~ '3-5',
         between(lamb2count,6,8) ~ '6 or more',
         TRUE                      ~ NA_character_),
       lambCcountgrp = case_when(
         between(lambCcount,0,1) ~ '0-1',
         between(lambCcount,2,3) ~ '2-3',
         between(lambCcount,4,5) ~ '4-5',
         between(lambCcount,6,8) ~ '6 or more',
         TRUE                      ~ NA_character_),
       lambWcountgrp = case_when(
         between(lambWcount,0,1) ~ '0-1',
         between(lambWcount,2,3) ~ '2-3',
         between(lambWcount,4,5) ~ '4-5',
         between(lambWcount,6,8) ~ '6 or more',
         TRUE                      ~ NA_character_),
       # conv3grp = case_when(
       #   between(cp_week/era5_week,0,0.2) ~ '0-20%',
       #   between(cp_week/era5_week,0.2,0.5) ~ '20-50%',
       #   between(cp_week/era5_week,0.5,0.8) ~ '50-80%',
       #   between(cp_week/era5_week,0.8,1.0) ~ '80-100%',
       #   TRUE                      ~ NA_character_),
       season = case_when(
         month(DATE) %in% 3:5 ~ 'Spring (MAM)',
         month(DATE) %in% 6:8 ~ 'Summer (JJA)',
         month(DATE) %in% 9:11 ~ 'Autumn (SON)',
         month(DATE) %in% c(12,1,2) ~ 'Winter (DJF)',
         TRUE                      ~ NA_character_),
       season = factor(season,levels=c('Spring (MAM)','Summer (JJA)','Autumn (SON)','Winter (DJF)')),
       #wm = weighted.mean(VALUE,VOLUME)
) #%>% 
  #ungroup() %>% 
  # filter(!(FIELDNAME == 'CONDY' & VALUE > 100)) %>%
  # filter(!(FIELDNAME == 'NO3N' & VALUE > 5)) %>%
  # filter(!(FIELDNAME == 'NH4N' & VALUE > 5)) %>%
  # filter(!(FIELDNAME == 'CHLORIDE' & VALUE > 7.5)) %>%
  # filter(!(FIELDNAME == 'SODIUM' & VALUE > 5)) %>%
  # filter(!(FIELDNAME == 'non-marine sulphate' & VALUE > 50)) %>%
  #filter(!(VALUE < 0))

date_chem = date_chem %>% 
  mutate(SODIUM=SODIUM*1000, CHLORIDE=CHLORIDE*1000, NH4N=NH4N*1000,
         NO3N=NO3N*1000, `non-marine sulphate`=`non-marine sulphate`*1000)
# date_chem = date_chem %>% group_by(SITECODE) %>% 
#   mutate(conv3quant = rank(cp_week/era5_week)/length(cp_week/era5_week)) %>% 
#   mutate( conv3grp = case_when(
#     between(conv3quant,0,0.33) ~ '0-33%',
#     between(conv3quant,0.33,0.67) ~ '33-67%',
#     between(conv3quant,0.67,1.0) ~ '67-100%',
#     TRUE                      ~ NA_character_))
```

```{r license, context="server"}

    # Return the UI for a modal dialog with a checkbox for agreeing license agreement If 'failed' is
    # TRUE, then display a message that the previous value was invalid.
    dataModal <- function(failed = FALSE) {
      modalDialog(
        h4('This is a sandbox application for the publication "GB Rainfall" '),
        p('The applicable data licenses includes:'),
        tags$div(
          tags$ul(
            tags$li(a("ECN rainfall chemistry data: Environmental Information Data Centre", 
                    href='https://catalogue.ceh.ac.uk/documents/18b7c387-037d-4949-98bc-e8db5ef4264c')),
            tags$li(a("ERA5 reanalysis: Copernicus ERA5 API")),
            tags$li(a("NOAA Hysplit backtrajectory analysis: NOAA")),
            tags$li(a("Lamb weather types data data: Climate Research Unit"))
          )
        ),

        
        #span('author: Michael Tso (mtso at ceh.ac.uk)", ',
        #     'UK Centre for Ecology and Hydrology'),
        if (failed)
          div(tags$b("You must agree with the license agreements to proceed.", style = "color: red;")),

        footer = tagList(
          #modalButton("Cancel"),
          checkboxInput("ok", "I agree with the relevant license agreements listed above.", FALSE),
          actionButton("ok2","Submit")
        )
      )
    }

    # Show modal when button is clicked.
    observeEvent(input$show, {
      showModal(dataModal())
    })
    
    # Show the model on start up ...
    showModal(dataModal())

    
    
    # When OK button is pressed, attempt to load the data set. If successful,
    # remove the modal. If not show another modal, but this time with a failure
    # message.
    observeEvent(input$ok2, {
      # Check that data object exists and is data frame.
      if (input$ok ==TRUE) {
        removeModal()
      } else {
        showModal(dataModal(failed = TRUE))
      }
    })

```


## Figure 1: ECN map

This code plots the map of selected ECN stations.

```{r fig1, warning=FALSE, exercise=TRUE, exercise.lines=30,fig.fullwidth=TRUE}
# draw ECN map
library(maps)

library(ggrepel)
UK <- map_data("world") %>% filter(region=="UK")
Ireland <- map_data("world") %>% filter(region=="Ireland")
IoM <- map_data("world") %>% filter(region=="Isle of Man")
data <- world.cities %>% filter(country.etc=="UK")

# 4 ECN sites
ECN_site_info <- data.frame(
  "sitecode" = 1:12, 
  "name" = c("Drayton","Glensaugh","Hillsborough","Moor House",
             "North Wyke","Rothamsted","Sourhope","Wytham",
             "Alice Holt","Porton Down","Snowdon","Cairngorms"),
  "lat" = c(52.193875,56.9092667,54.4534,54.6950416666,
            50.781933335,51.803425,55.4898527778,51.781349999999996,
            51.15457222,51.127175,53.07455,57.11634444), 
  "long" = c(-1.764430555,-2.55337222,-6.0781277778,-2.38785,
             -3.91780555553,-0.372683333,-2.2120333,-1.3360583333333333,
             -0.86321666,-1.63985,-4.03351111,-3.82971667))

ECN_site_info <-ECN_site_info[c(2,4,6,11),]
ggplot() +
  geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="darkgrey", alpha=0.7) +
  geom_polygon(data = Ireland, aes(x=long, y = lat, group = group), fill="grey", alpha=0.5) +
  geom_polygon(data = IoM, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  #geom_point( data=data, aes(x=long, y=lat, alpha=pop)) +
  geom_text_repel( data=ECN_site_info, aes(x=long, y=lat, label=name), size=5) +
  geom_point( data=ECN_site_info, aes(x=long, y=lat), color="red", size=3) +
  theme_void() + ylim(50,61) + coord_map() +
  theme(legend.position="none")

```
## Figure 1: ECN map (with selectors)

This code is a duplicate of the previous one. However, R Shiny selectors are added for easier site selection.

```{r, echo=FALSE}
selectInput("site", "Choose sites:",multiple = TRUE,choices = 1:12,selected = c(2,4,6,11))
plotOutput("ecnMap")
```

```{r, context="server"}

output$ecnMap = renderPlot({

# draw ECN map

UK <- map_data("world") %>% filter(region=="UK")
Ireland <- map_data("world") %>% filter(region=="Ireland")
IoM <- map_data("world") %>% filter(region=="Isle of Man")
data <- world.cities %>% filter(country.etc=="UK")

# 4 ECN sites
ECN_site_info <- data.frame(
  "sitecode" = 1:12, 
  "name" = c("Drayton","Glensaugh","Hillsborough","Moor House",
             "North Wyke","Rothamsted","Sourhope","Wytham",
             "Alice Holt","Porton Down","Snowdon","Cairngorms"),
  "lat" = c(52.193875,56.9092667,54.4534,54.6950416666,
            50.781933335,51.803425,55.4898527778,51.781349999999996,
            51.15457222,51.127175,53.07455,57.11634444), 
  "long" = c(-1.764430555,-2.55337222,-6.0781277778,-2.38785,
             -3.91780555553,-0.372683333,-2.2120333,-1.3360583333333333,
             -0.86321666,-1.63985,-4.03351111,-3.82971667))

ECN_site_info <-ECN_site_info[input$site,]
ggplot() +
  geom_polygon(data = UK, aes(x=long, y = lat, group = group), fill="darkgrey", alpha=0.7) +
  geom_polygon(data = Ireland, aes(x=long, y = lat, group = group), fill="grey", alpha=0.5) +
  geom_polygon(data = IoM, aes(x=long, y = lat, group = group), fill="grey", alpha=0.3) +
  #geom_point( data=data, aes(x=long, y=lat, alpha=pop)) +
  geom_text_repel( data=ECN_site_info, aes(x=long, y=lat, label=name), size=5) +
  geom_point( data=ECN_site_info, aes(x=long, y=lat), color="red", size=3) +
  theme_void() + ylim(50,61) + coord_map() +
  theme(legend.position="none")
})
```
## Figure 2: Lamb cyclones/westerlies counts (1871-2020)
This plots the yearly Lamb cyclones/westerlies counts (1871-2020). The top graph fits a changepoint alogorithm to it, while the seconds shows the second derivative of its deviation from mean, as well as overlaying the Atlantic multidecadal oscillation (AMO).

```{r fig2, warning=FALSE, exercise=TRUE, exercise.lines=30,fig.width=9, fig.height=5,fig.fullwidth=TRUE}

source(here('R','cpt_helpers.R'))
source(here('R','Deriv.R'))
library(mgcv);

lamb_type_lookup <-  read_csv(here("data","lamb-type-lookup.csv")) ;

lamb_data = read_csv(here('data','20CR_1871-1947_ncep_1948-2020_12hrs_UK.csv')) %>% 
    dplyr::mutate(DATE = make_date(year,month,day)) %>% 
    dplyr::select(DATE,LWT) %>% 
    dplyr::mutate(LWT = as.character(LWT)) %>% 
    dplyr::mutate(LWT = setNames(lamb_type_lookup$Type,lamb_type_lookup$Num)[LWT]);

currData = lamb_data %>% mutate(wk_year=(yday(DATE)-1)%/%7 + 1, yday = yday(DATE),year=year(DATE), 
        lamb_grp2= case_when(
          LWT %in% c('C','W','SW','CW','CSW') ~ "(C,W,SW,CW,CSW)",
          TRUE ~ "others")) %>% 
  group_by(year)%>%  
  filter(lamb_grp2 == "(C,W,SW,CW,CSW)") %>% 
  summarise(N = n());

cpt.mean(currData$N,penalty="Manual",pen.value=0.8,method="AMOC",test.stat="CUSUM");

out=cpt.meanvar(currData$N,pen.value=c(2*log(length(currData$N)),100*log(length(currData$N))),penalty="CROPS",method="PELT",test.stat = "Poisson");

currData = currData %>% rename(value=N) %>% mutate(segment = 1);
for (ii in 1:length(!is.na(out@cpts.full[1,]))) {
  if (ii == length(!is.na(out@cpts.full[1,]))) {
     currData$segment[out@cpts.full[1,ii]: length(out@data.set)] = ii+1
  }else {
      currData$segment[out@cpts.full[1,ii]:out@cpts.full[1,ii+1]] = ii + 1
  }
  
};

currData = currData %>% group_by(segment) %>% mutate(mean=mean(value),var=var(value),meanGroup=segment,varGroup=segment);

p1 = createChangePointPlot('cpt.meanvar', currData %>% mutate(temporalField = make_date(year,1,1))) + 
   labs(title = paste0('Number of Lamb westerlies/cyclones days per year (1871-', max(currData$year) ,')'),
              subtitle = "Changepoints using cpt.meanvar (pentaly = 'CROPS',method='PELT',test.stat='Poission')" #,
        #      caption = "Data source: Climate Research Unit"
        )  + ylab('count') + xlab('Year') + 
  theme_half_open(12) +
  background_grid(major = 'y', minor = "none") +
  panel_border() + background_grid();

AMO = read_csv(here('data','AMO_smoothed_from_the_Kaplan_SST_V2.csv')) %>% 
  gather(key="month",value="value",-Year) %>% 
  mutate(month= match(month,toupper(month.abb)), Date = make_date(Year,month,1)) %>% 
  select(-Year,-month) %>% 
  filter(!value < -99) %>% 
  mutate(year=year(Date)+yday(Date)/366);

lamb_type_lookup <-  read_csv(here('data',"lamb-type-lookup.csv"))  ;

lamb_data = read_csv(here('data','20CR_1871-1947_ncep_1948-2020_12hrs_UK.csv')) %>% 
  dplyr::mutate(DATE = make_date(year,month,day)) %>% 
  dplyr::select(DATE,LWT) %>% 
  dplyr::mutate(LWT = as.character(LWT)) %>% 
  dplyr::mutate(LWT = setNames(lamb_type_lookup$Type,lamb_type_lookup$Num)[LWT], 
                season = case_when(
                  month(DATE) %in% 3:5 ~ 'Spring (MAM)',
                  month(DATE) %in% 6:8 ~ 'Summer (JJA)',
                  month(DATE) %in% 9:11 ~ 'Autumn (SON)',
                  month(DATE) %in% c(12,1,2) ~ 'Winter (DJF)',
                  TRUE                      ~ NA_character_));
currData = lamb_data %>% mutate(wk_year=(yday(DATE)-1)%/%7 + 1, yday = yday(DATE),year=year(DATE), 
                                lamb_grp2= case_when(
                                  LWT %in% c('C','W','SW','CW','CSW') ~ "(C,W,SW,CW,CSW)",
                                  TRUE ~ "others")) %>% 
  group_by(year)%>%  
  filter(lamb_grp2 == "(C,W,SW,CW,CSW)") %>% 
  summarise(N = n());

#### FIT ADDITIVE MODEL
ctrl <- list(niterEM = 0, msVerbose = TRUE, optimMethod="L-BFGS-B");

m2 <- gamm(N ~ s(year, k =9),
           data = currData, 
           control = ctrl);

want <- seq(1, nrow(currData), length.out = 200);
pdat <- with(currData, data.frame(year = year[want]) );
p2 <- predict(m2$gam, newdata = pdat, type = "terms", se.fit = TRUE);
pdat <- transform(pdat, p2 = p2$fit[,1], se2 = p2$se.fit[,1]);
df.res <- df.residual(m2$gam);
crit.t <- qt(0.025, df.res, lower.tail = FALSE);
pdat <- transform(pdat,
                  upper = p2 + (crit.t * se2),
                  lower = p2 - (crit.t * se2));

Term <- "year";
m2.d <- Deriv(m2);
m2.dci <- confint(m2.d, term = Term);
m2.dsig <- signifD(pdat$p2, d = m2.d[[Term]]$deriv,
                   m2.dci[[Term]]$upper, m2.dci[[Term]]$lower);


p2=ggplot(pdat) + 
  geom_line(data=AMO,aes(x=year,y=value*100),col='firebrick4')+
  geom_area(data=AMO,aes(x=year,y=value*100),fill='firebrick4',alpha=0.2,col=NA)+
  geom_line(aes(year,p2))+
  geom_line(aes(year,lower),linetype='dashed')+geom_line(aes(year,upper),linetype='dashed')+
  geom_line(aes(year,unlist(m2.dsig$incr)), col='blue', size=2 )+
  geom_line(aes(year,unlist(m2.dsig$decr)), col='red', size=2 )+
  ylab('count anomaly or AMO (x100)')+ xlab('Year') + 
  theme_half_open(12) +
  background_grid(major = 'y', minor = "none") +
  panel_border() + background_grid()+
  ggtitle('Lamb cyclones/westerlies days count fitted anomaly (1871-2020)\nand Atlantic multi-decadal oscillation (AMO)');


pALL = plot_grid(p1,p2,labels = "AUTO", align = 'h', axis = "tb",nrow = 2);
pALL;

#save_plot(paste0("lamb_1871_2020.png"), pALL, base_height =NULL, base_width = 7.5, dpi=128) # 714 x 660
  
  
```

## Figure 3: Lamb vs chemistry boxplots
This code plots concentration over time for selected chemical species across ECN sites grouped by number of Lamb westerlies and cyclones per week. Chemical samples are taken weekly.

```{r fig3, warning=FALSE, exercise=TRUE, exercise.lines=30,fig.width=9, fig.height=5,fig.fullwidth=TRUE}

plot_chembox = function(date_chem, FIELDNAME,grp_name,flux=FALSE,skip_legend=1){
  
  if(flux==FALSE){
    y = date_chem[[FIELDNAME]]
  } else {
    y = date_chem[[FIELDNAME]]*date_chem[['VOLUME']]/1000/(pi*(0.152/2)^2)*(365.25/7) # per area per year
  }
  
  out = ggplot(date_chem, #%>% filter(!is.na(get(FIELDNAME))),
               #aes(x=yr_grp, y=get(FIELDNAME)*VOLUME, fill=get(grp_name))) +
               aes(x=yr_grp, y= !!y, fill=get(grp_name))) +
    geom_boxplot()+
    facet_grid(SITECODE ~ .) +
    theme_half_open(12) +
    background_grid(major = 'y', minor = "none") +
    panel_border() + background_grid()+
    scale_fill_brewer(palette = 'Spectral')+
    theme(axis.text.x=element_text(angle=30, hjust=1))+
    scale_y_continuous(limits = quantile(y, c(0., 0.95),na.rm=TRUE))

  if(skip_legend == 1){
    out = out + theme(legend.position = "none")
  }else{
    out = out +  theme(legend.position = "bottom",legend.justification = 'center',
                     legend.key.size=unit(36,units = "points") ,
                     legend.text = element_text(size = 18))
  }
  return(out)
}

grp_name = 'lamb_grp2'

p1 = plot_chembox(date_chem, "CONDY","lamb_grp2") + xlab('Year')+ylab('Electrical Conductivity (\u03BCS/cm)')
p4 = plot_chembox(date_chem, "NO3N","lamb_grp2") + xlab('Year')+ylab( expression("NO"[3]^-{}*" (\u03BCeq/L)"))
p5 = plot_chembox(date_chem, "NH4N","lamb_grp2") + xlab('Year')+ylab( expression("NH"[4]^+{}*" (\u03BCeq/L)"))+
  theme(legend.position = "bottom",legend.justification = 'center',
        legend.key.size=unit(36,units = "points") ,
        legend.text = element_text(size = 18)) + 
  guides(fill=guide_legend("Number of Westerly days per week: " )) 
p2 = plot_chembox(date_chem, "CHLORIDE","lamb_grp2") + xlab('Year')+ylab( expression("Cl"[]^-{}*" (\u03BCeq/L)"))+
  theme(legend.position = "bottom",legend.justification = 'center',
        legend.key.size=unit(36,units = "points") ,
        legend.text = element_text(size = 18)) + 
  guides(fill=guide_legend("Number of Westerly days per week: " ))
p3 = plot_chembox(date_chem, "SODIUM","lamb_grp2") + xlab('Year')+ylab( expression("Na"[]^+{}*" (\u03BCeq/L)")) 
p6 = plot_chembox(date_chem, "non-marine sulphate","lamb_grp2") + xlab('Year')+ylab( expression("xSO"[4]^2^-{}*" (\u03BCeq/L)"))

pALL = plot_grid(p1,p2,p3,labels = "AUTO", align = 'h', axis = "tb",nrow = 1) 
pALL;

pALL = plot_grid(p4,p5,p6,labels = c("D","E","F"), align = 'h', axis = "tb",nrow = 1) 
pALL
```

    