---
title: "AccSamplingDesign - Benchmarking Beta-based Sampling Plan Optimization"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE, warning = FALSE)
library(kableExtra)
library(dplyr)
library(AccSamplingDesign)
library(bench)
```

This benchmark reproduces the results reported in **Section 7: Computational Performance and Benchmarking**. We compare the grid search method `beta_plan_grid_search()` (adapted from Govindaraju and Kissling, 2015) with the `optPlan()` function from the `AccSamplingDesign` package. 

**Please allow over 1 minute for the grid search to complete.**

```{r grid-search-func, echo=FALSE, message=FALSE}
#-------------------------------------------------------------------------------
# This function describes the grid search method for the Beta optimal plan 
# Code from Govindaraju and Kissling (2015)
#-------------------------------------------------------------------------------
beta_plan_grid_search <- function(LSL, theta, p1, p2, alpha, beta, nsim = 3000) {
  mulimit <- function(p, LSL, theta) {
    f <- function(mu) pbeta(LSL, theta * mu, theta * (1 - mu)) - p
    uniroot(f, interval = c(0, 1), tol = 1e-9)$root
  }
  Pa_est <- function(nsim, mu, theta, k, m, LSL) {
    shape1 <- m * mu * theta
    shape2 <- m * (1 - mu) * theta
    ym <- rbeta(nsim, shape1, shape2)
    acc_crit <- ym - k * sqrt(ym * (1 - ym) / theta)
    mean(acc_crit > LSL)
  }

  if (!requireNamespace("AcceptanceSampling", quietly = TRUE)) {
    stop("Package 'AcceptanceSampling' is required but not installed.")
  }

  muU <- mulimit(p1, LSL, theta)
  muL <- mulimit(p2, LSL, theta)

  vplan <- AcceptanceSampling::find.plan(
    c(p1, 1 - alpha), c(p2, beta),
    type = "normal", s.type = "known"
  )

  kvalues <- seq(0.5 * vplan$k, 2 * vplan$k, by = 0.01)
  mvalues <- seq(0.5 * vplan$n, 2 * vplan$n, by = 0.15)
  grid <- expand.grid(k = kvalues, m = mvalues)

  CR <- mapply(Pa_est, nsim = nsim, mu = muL, theta = theta,
               k = grid$k, m = grid$m, LSL = LSL)
  PR <- 1 - mapply(Pa_est, nsim = nsim, mu = muU, theta = theta,
                   k = grid$k, m = grid$m, LSL = LSL)

  feasible <- subset(cbind(grid, CR, PR), CR < beta & PR < alpha)
  feasible <- feasible[order(feasible$m, feasible$k), ]

  if (nrow(feasible) == 0) stop("No feasible plan found.")

  opt <- feasible[1, ]
  opt$m <- round(opt$m, 4) 
  opt$k <- round(opt$k, 4)

  return(list(n = opt$m, k = opt$k))
}
```

### Define Parameters (from Example 6.3 of 'Vitamin-A sampling')

```{r define-params}
LSL    <- 5.65e-6
theta  <- 6.6e+08
p1     <- 0.025
p2     <- 0.10
alpha  <- 0.05
beta   <- 0.10
```

```{r run-methods, echo=FALSE, message=FALSE}
#-------------------------------------------------------------------------------
# Benchmark performance (please allow about 1+ minute for grid_search to run)
#-------------------------------------------------------------------------------

set.seed(123)

cat("Running Grid Search (may take 1+ minute)...\n")
res_grid <- beta_plan_grid_search(LSL, theta, p1, p2, alpha, beta)

cat("Running AccSamplingDesign NLP method...\n")
res_opt <- AccSamplingDesign::optPlan(
  PRQ = p1, CRQ = p2, alpha = alpha, beta = beta,
  distribution = "beta", theta_type = "known",
  theta = theta, LSL = LSL
)
```

### Benchmark Runtime and Memory Usage

```{r benchmark, echo=FALSE, message=FALSE}
bm <- bench::mark(
  Govindaraju_and_Kissling_2015_GridSearch = beta_plan_grid_search(LSL, theta, p1, p2, alpha, beta),
  AccSamplingDesign_NLP = AccSamplingDesign::optPlan(
    PRQ = p1, CRQ = p2, alpha = alpha, beta = beta,
    distribution = "beta", theta_type = "known",
    theta = theta, LSL = LSL
  ),
  iterations = 1,
  check = FALSE
)
```

```{r summary-table, echo=FALSE, message=FALSE, results='asis'}
#-------------------------------------------------------------------------------
# Display final table of benchmark results
#-------------------------------------------------------------------------------

summary_tbl <- tibble(
  Method = c("`beta_plan_grid_search()`", "`AccSamplingDesign::optPlan()`"),
  n = c(res_grid$n, res_opt$n),
  k = c(res_grid$k, res_opt$k),
  `Median Time (s)` = round(as.numeric(bm$median), 3),
  `Memory Used` = format(bm$mem_alloc, units = "auto")
)

kable(summary_tbl, caption = "Benchmarking AccSamplingDesign performance: Grid Search vs. NLP for Beta-based sampling plan optimization.") %>%
  kable_styling(full_width = FALSE)
```

```{r speedup, echo=FALSE, message=FALSE, results='asis'}
#-------------------------------------------------------------------------------
# Print speed-up factor
#-------------------------------------------------------------------------------

time_grid <- as.numeric(bm$median[1])
time_nlp  <- as.numeric(bm$median[2])
speedup   <- round(time_grid / time_nlp, 1)

cat(sprintf("\n>> `AccSamplingDesign::optPlan()` (NLP method) is approximately **%.1f× faster** than the Grid Search method for finding optimal Beta-based sampling plans.\n", speedup))
```
